---
title: "Ignition Efficiency Model"
author: "Charlotte Uden"
date: "2026-06-03"
output: html_document
---

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(patchwork)
library(ggplot2)
library(gridExtra)
library(terra)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(glmnet)
library(pROC)
library(PRROC)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
```


Read in drivers
```{r}
df <- read.csv(file.path(drivers_root, "lightning_fires_burned_area_jja_2002_2011.csv"))
df <- df[,2:29]

#Random forest results indicate that U (% open land), C (% conifer cover), and sd (snow depth) should be excluded
df <- df[,c("lat","lon","year","month","rh","tair","precip","wind","precip_1m","precip_2m","precip_3m","precip_5y","tair_1m","tair_2m","tair_3m","tair_5y","B","strike_count","fire_count")]
head(df)
```



Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]
```

```{r}
# ── Save scaling parameters ───────────────────────────────────────────────────
predictors_ignition <- c("rh", "tair", "precip", "wind",
                          "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                          "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")

mu_rh        <- mean(train$rh);        sigma_rh        <- sd(train$rh)
mu_tair      <- mean(train$tair);      sigma_tair      <- sd(train$tair)
mu_precip    <- mean(train$precip);    sigma_precip    <- sd(train$precip)
mu_wind      <- mean(train$wind);      sigma_wind      <- sd(train$wind)
mu_precip_1m <- mean(train$precip_1m); sigma_precip_1m <- sd(train$precip_1m)
mu_precip_2m <- mean(train$precip_2m); sigma_precip_2m <- sd(train$precip_2m)
mu_precip_3m <- mean(train$precip_3m); sigma_precip_3m <- sd(train$precip_3m)
mu_precip_5y <- mean(train$precip_5y); sigma_precip_5y <- sd(train$precip_5y)
mu_tair_1m   <- mean(train$tair_1m);   sigma_tair_1m   <- sd(train$tair_1m)
mu_tair_2m   <- mean(train$tair_2m);   sigma_tair_2m   <- sd(train$tair_2m)
mu_tair_3m   <- mean(train$tair_3m);   sigma_tair_3m   <- sd(train$tair_3m)
mu_tair_5y   <- mean(train$tair_5y);   sigma_tair_5y   <- sd(train$tair_5y)
mu_B         <- mean(train$B);         sigma_B         <- sd(train$B)

# Store and save
mean_sd_ignition <- data.frame(
  var   = predictors_ignition,
  mu    = c(mu_rh, mu_tair, mu_precip, mu_wind,
            mu_precip_1m, mu_precip_2m, mu_precip_3m, mu_precip_5y,
            mu_tair_1m, mu_tair_2m, mu_tair_3m, mu_tair_5y, mu_B),
  sigma = c(sigma_rh, sigma_tair, sigma_precip, sigma_wind,
            sigma_precip_1m, sigma_precip_2m, sigma_precip_3m, sigma_precip_5y,
            sigma_tair_1m, sigma_tair_2m, sigma_tair_3m, sigma_tair_5y, sigma_B)
)

write.csv(mean_sd_ignition,
          file.path(mu_sigma_root, "mu_sigma_ignition_predictors.csv"),
          row.names = FALSE)

# ── Standardize train ─────────────────────────────────────────────────────────
train <- train %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B
  )

train <- train[, c("lat", "lon", "year", "month",
                   "strike_count", "fire_count",
                   "rh", "tair", "precip", "wind",
                   "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                   "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")]

# ── Standardize test ──────────────────────────────────────────────────────────
# Use training means and sds 
test <- test %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B
  )

test <- test[, c("lat", "lon", "year", "month",
                 "strike_count", "fire_count",
                 "rh", "tair", "precip", "wind",
                 "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                 "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")]

head(train)

```

```{r}

# ── Predictors ────────────────────────────────────────────────────────────────
predictors_ignition <- c("rh", "tair", "precip", "wind",
                          "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                          "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")

# ── Filter to gridcell-months with at least one strike ────────────────────────
# Rows with strike_count == 0 have no denominator for the binomial
train_strikes <- train |> filter(strike_count > 0)
test_strikes  <- test  |> filter(strike_count > 0)

cat("Training rows with strikes:", nrow(train_strikes), "\n")
cat("Test rows with strikes:    ", nrow(test_strikes),  "\n")
cat("Training fire prevalence:  ",
    round(sum(train_strikes$fire_count) / sum(train_strikes$strike_count), 6), "\n")
```

# ── Model 1: Standard binomial GLM ────────────────────────────────────────────
```{r}
formula_ignition <- as.formula(
  paste("cbind(fire_count, strike_count - fire_count) ~",
        paste(predictors_ignition, collapse = " + "))
)

I1_glm <- glm(formula_ignition,
               data   = train_strikes,
               family = binomial(link = "logit"))

summary(I1_glm)
```

# ── Model 2: Ridge-penalized logistic regression (glmnet) ─────────────────────
# Following Hessilt et al. (2022) and Uden et al. (2026)
```{r}

# Build model matrices
X_train <- model.matrix(
  as.formula(paste("~", paste(predictors_ignition, collapse = " + "), "- 1")),
  data = train_strikes
)

X_test <- model.matrix(
  as.formula(paste("~", paste(predictors_ignition, collapse = " + "), "- 1")),
  data = test_strikes
)

# Two-column response: cbind(successes, failures)
Y_train <- cbind(train_strikes$fire_count,
                 train_strikes$strike_count - train_strikes$fire_count)

# 5-fold cross-validation to select lambda (ridge: alpha = 0)
set.seed(123)
cv_ridge <- cv.glmnet(
  x        = X_train,
  y        = Y_train,
  family   = "binomial",
  alpha    = 0,        # ridge penalty
  nfolds   = 5,
  type.measure = "deviance"
)

plot(cv_ridge)
cat("Optimal lambda (min deviance):", cv_ridge$lambda.min, "\n")
cat("Optimal lambda (1se rule):    ", cv_ridge$lambda.1se, "\n")

# Fit final ridge model with optimal lambda
I1_ridge <- glmnet(
  x      = X_train,
  y      = Y_train,
  family = "binomial",
  alpha  = 0,
  lambda = cv_ridge$lambda.min
)

# Coefficients
coef(I1_ridge)
```

# ── Predictions on test set ───────────────────────────────────────────────────
```{r}
# GLM predictions
test_strikes$pignite_glm <- predict(I1_glm,
                                     newdata = test_strikes,
                                     type    = "response")

# Ridge predictions
test_strikes$pignite_ridge <- as.numeric(
  predict(I1_ridge,
          newx = X_test,
          type = "response")
)
```

# ── Evaluation ────────────────────────────────────────────────────────────────
```{r}
# Create binary outcome for ROC: 1 if any fire occurred in gridcell-month
test_strikes <- test_strikes |>
  mutate(fire_binary = as.integer(fire_count > 0))

# ROC AUC
roc_glm   <- roc(test_strikes$fire_binary, test_strikes$pignite_glm,   quiet = TRUE)
roc_ridge <- roc(test_strikes$fire_binary, test_strikes$pignite_ridge,  quiet = TRUE)

cat("\nROC AUC — GLM:  ", round(auc(roc_glm),   4), "\n")
cat("ROC AUC — Ridge:", round(auc(roc_ridge),  4), "\n")

# Brier score (mean squared error of probability predictions)
brier_glm   <- mean((test_strikes$pignite_glm   - test_strikes$fire_binary)^2)
brier_ridge <- mean((test_strikes$pignite_ridge  - test_strikes$fire_binary)^2)

cat("Brier score — GLM:  ", round(brier_glm,   4), "\n")
cat("Brier score — Ridge:", round(brier_ridge,  4), "\n")

# Calibration: plot predicted probability vs observed frequency
# Bin predictions into deciles
calibration_plot <- function(preds, obs, model_name) {
  df_cal <- data.frame(pred = preds, obs = obs) |>
    mutate(bin = ntile(pred, 10)) |>
    group_by(bin) |>
    summarise(mean_pred = mean(pred),
              mean_obs  = mean(obs),
              .groups   = "drop")

  ggplot(df_cal, aes(x = mean_pred, y = mean_obs)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", color = "grey50") +
    geom_point(size = 3) +
    labs(title = paste("Calibration —", model_name),
         x = "Mean predicted probability",
         y = "Observed frequency") +
    theme_bw()
}

calibration_plot(test_strikes$pignite_glm,   test_strikes$fire_binary, "GLM")
calibration_plot(test_strikes$pignite_ridge,  test_strikes$fire_binary, "Ridge")

# ── PR AUC ────────────────────────────────────────────────────────────────────

pr_glm <- pr.curve(
  scores.class0 = test_strikes$pignite_glm[test_strikes$fire_binary == 1],
  scores.class1 = test_strikes$pignite_glm[test_strikes$fire_binary == 0],
  curve         = TRUE
)

pr_ridge <- pr.curve(
  scores.class0 = test_strikes$pignite_ridge[test_strikes$fire_binary == 1],
  scores.class1 = test_strikes$pignite_ridge[test_strikes$fire_binary == 0],
  curve         = TRUE
)

cat("PR AUC — GLM:  ", round(pr_glm$auc.integral,   4), "\n")
cat("PR AUC — Ridge:", round(pr_ridge$auc.integral,  4), "\n")

# Baseline PR AUC = event prevalence (no-skill classifier)
baseline_pr <- mean(test_strikes$fire_binary)
cat("Baseline PR AUC (event prevalence):", round(baseline_pr, 6), "\n")
cat("Baseline:", mean(test_strikes$fire_binary), "\n")
# ── Calibration slope and intercept ──────────────────────────────────────────
# Following your published paper: logit(obs) ~ a + b * logit(pred)
# Intercept (a): 0 = no bias, positive = underprediction, negative = overprediction
# Slope (b):     1 = perfect, <1 = overconfident, >1 = underconfident

calibration_stats <- function(preds, obs, model_name) {

  # Clip predictions away from 0 and 1 to avoid infinite logits
  preds_clipped <- pmin(pmax(preds, 1e-6), 1 - 1e-6)

  logit_pred <- log(preds_clipped / (1 - preds_clipped))

  cal_model <- glm(obs ~ logit_pred,
                   family = binomial(link = "logit"))

  intercept <- coef(cal_model)[1]
  slope     <- coef(cal_model)[2]
  ci        <- confint(cal_model)

  cat(sprintf("\nCalibration — %s:\n", model_name))
  cat(sprintf("  Intercept: %6.3f  (95%% CI: %6.3f, %6.3f)\n",
              intercept, ci[1, 1], ci[1, 2]))
  cat(sprintf("  Slope:     %6.3f  (95%% CI: %6.3f, %6.3f)\n",
              slope, ci[2, 1], ci[2, 2]))

  data.frame(
    model     = model_name,
    intercept = intercept,
    int_low   = ci[1, 1],
    int_high  = ci[1, 2],
    slope     = slope,
    slope_low = ci[2, 1],
    slope_high= ci[2, 2]
  )
}

cal_glm   <- calibration_stats(test_strikes$pignite_glm,
                                test_strikes$fire_binary,
                                "GLM")

cal_ridge <- calibration_stats(test_strikes$pignite_ridge,
                                test_strikes$fire_binary,
                                "Ridge")

cal_results <- bind_rows(cal_glm, cal_ridge)
print(cal_results)

# ── Summary table of all metrics ──────────────────────────────────────────────
metrics_ignition <- data.frame(
  model      = c("GLM", "Ridge"),
  roc_auc    = c(round(auc(roc_glm),  4), round(auc(roc_ridge),  4)),
  pr_auc     = c(round(pr_glm$auc.integral,  4), round(pr_ridge$auc.integral, 4)),
  brier      = c(round(brier_glm,  4), round(brier_ridge,  4)),
  int        = round(cal_results$intercept, 3),
  int_low    = round(cal_results$int_low,   3),
  int_high   = round(cal_results$int_high,  3),
  slope      = round(cal_results$slope,     3),
  slope_low  = round(cal_results$slope_low, 3),
  slope_high = round(cal_results$slope_high,3)
)

print(metrics_ignition)

```

# ── Bayesian Binomial Logistic Regression ──────────────────────────────────────────────────────────

```{r}
# Use parallel chains and pre-compiled Stan model
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

# ── Build model matrices ───────────────────────────────────────────────────────
predictors_ignition <- c("rh", "tair", "precip", "wind",
                          "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                          "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")

X_train_bayes <- as.matrix(train_strikes[, predictors_ignition])
X_test_bayes  <- as.matrix(test_strikes[,  predictors_ignition])

# ── Stan data list ─────────────────────────────────────────────────────────────
stan_data <- list(
  N         = nrow(train_strikes),
  K         = length(predictors_ignition),
  X         = X_train_bayes,
  n_trials  = train_strikes$strike_count,
  n_success = train_strikes$fire_count,
  sigma_beta0 = 1.0,   # prior sd for intercept — matches published paper
  sigma_beta  = 1.0    # prior sd for coefficients
)
```

# ── Compile and fit ────────────────────────────────────────────────────────────
```{r}
cat("Compiling Stan model...\n")
stan_model_compiled <- stan_model("/raid/cuden/Paleo_Lightning_Fire/Models/Ignition_Efficiency/binomial_logistic.stan")

cat("Fitting Bayesian binomial logistic regression...\n")
I1_bayes <- sampling(
  object  = stan_model_compiled,
  data    = stan_data,
  chains  = 4,
  iter    = 2000,      # 1000 warmup + 1000 sampling per chain = 4000 draws
  warmup  = 1000,
  seed    = 123,
  control = list(adapt_delta = 0.95)  # increase if divergent transitions occur
)
```

# ── Check convergence ──────────────────────────────────────────────────────────
```{r}
print(I1_bayes, pars = c("beta0", "beta"), probs = c(0.025, 0.5, 0.975))

# R-hat values — all should be < 1.01
rhat_vals <- summary(I1_bayes)$summary[, "Rhat"]
cat("Max R-hat:", round(max(rhat_vals, na.rm = TRUE), 4), "\n")
cat("Any R-hat > 1.01:", any(rhat_vals > 1.01, na.rm = TRUE), "\n")

# Trace plots for coefficients
traceplot(I1_bayes, pars = c("beta0", "beta"))
```

# ── Posterior predictions on test set ─────────────────────────────────────────
```{r}
# Extract posterior draws of coefficients
posterior <- extract(I1_bayes)

# Compute predicted probability for each test observation
# using posterior mean of beta0 and beta
beta0_mean <- mean(posterior$beta0)
beta_mean  <- colMeans(posterior$beta)

logit_pred_test <- beta0_mean + X_test_bayes %*% beta_mean
test_strikes$pignite_bayes <- as.numeric(1 / (1 + exp(-logit_pred_test)))

# Also compute posterior uncertainty (sd across draws)
# This gives a credible interval width per gridcell-month
n_draws <- length(posterior$beta0)
pred_draws <- matrix(NA, nrow = nrow(test_strikes), ncol = n_draws)

for (d in seq_len(n_draws)) {
  eta <- posterior$beta0[d] + X_test_bayes %*% posterior$beta[d, ]
  pred_draws[, d] <- 1 / (1 + exp(-eta))
}

test_strikes$pignite_bayes_sd  <- apply(pred_draws, 1, sd)
test_strikes$pignite_bayes_low <- apply(pred_draws, 1, quantile, 0.025)
test_strikes$pignite_bayes_high<- apply(pred_draws, 1, quantile, 0.975)

cat("Posterior mean predictions summary:\n")
print(summary(test_strikes$pignite_bayes))
```

# ── Evaluation metrics ─────────────────────────────────────────────────────────
```{r}
# ROC AUC
roc_bayes <- roc(test_strikes$fire_binary,
                  test_strikes$pignite_bayes,
                  quiet = TRUE)
cat("ROC AUC — Bayes:", round(auc(roc_bayes), 4), "\n")

# PR AUC
pr_bayes <- pr.curve(
  scores.class0 = test_strikes$pignite_bayes[test_strikes$fire_binary == 1],
  scores.class1 = test_strikes$pignite_bayes[test_strikes$fire_binary == 0],
  curve         = TRUE
)
cat("PR AUC — Bayes:", round(pr_bayes$auc.integral, 4), "\n")

# Brier score
brier_bayes <- mean((test_strikes$pignite_bayes - test_strikes$fire_binary)^2)
cat("Brier score — Bayes:", round(brier_bayes, 4), "\n")

# Calibration slope and intercept
cal_bayes <- calibration_stats(test_strikes$pignite_bayes,
                                test_strikes$fire_binary,
                                "Bayes")
```

# ── Add Bayes row to metrics_ignition ─────────────────────────────────────────
```{r}
metrics_ignition_updated <- bind_rows(
  metrics_ignition,
  data.frame(
    model      = "Bayes",
    roc_auc    = round(auc(roc_bayes),            4),
    pr_auc     = round(pr_bayes$auc.integral,      4),
    brier      = round(brier_bayes,                4),
    int        = round(cal_bayes$intercept,        3),
    int_low    = round(cal_bayes$int_low,          3),
    int_high   = round(cal_bayes$int_high,         3),
    slope      = round(cal_bayes$slope,            3),
    slope_low  = round(cal_bayes$slope_low,        3),
    slope_high = round(cal_bayes$slope_high,       3)
  )
)

print(metrics_ignition_updated)

# Save Stan model fit object for later use
saveRDS(I1_bayes,
        file.path(rds_root, "binomial_logisitc_regression_stanfit.rds"))

# ── Coefficient plot ───────────────────────────────────────────────────────────
# Extract posterior summaries for each coefficient
coef_summary <- as.data.frame(
  summary(I1_bayes,
          pars  = "beta",
          probs = c(0.025, 0.25, 0.75, 0.975))$summary
)
coef_summary$variable <- predictors_ignition

ggplot(coef_summary,
       aes(x = mean, y = reorder(variable, mean))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = `2.5%`, xmax = `97.5%`),
                 height = 0.2, alpha = 0.5) +
  geom_errorbarh(aes(xmin = `25%`,  xmax = `75%`),
                 height = 0, linewidth = 1.2) +
  geom_point(size = 3, shape = 21, fill = "white") +
  labs(title = "Posterior coefficients — Bayesian ignition model",
       x     = "Coefficient (posterior mean ± 95% CI)",
       y     = NULL) +
  theme_bw() +
  theme(panel.grid = element_blank())


```

Next, we'll try the Bayesian Logistic Regression:


```{r}

fit_bayes_logit_ridge <- function(data,
                                  predictors,
                                  response = "caused_fire",
                                  train_frac = 0.8,
                                  seed = 123,
                                  iter = 2000,
                                  chains = 4,
                                  sigma_beta = 1,
                                  sigma_alpha = 5,
                                  compiled_model = NULL,
                                  boot = TRUE,
                                  n_boot = 1000) {
  #-------------------------------
  # 0. Stan model code (inline)
  #-------------------------------
  stan_code <- "
  data {
    int<lower=1> N;
    int<lower=1> K;
    matrix[N, K] X;
    int<lower=0, upper=1> y[N];
    real<lower=0> sigma_beta;
    real<lower=0> sigma_alpha;
  }
  parameters {
    real alpha;
    vector[K] beta;
  }
  model {
    alpha ~ normal(0, sigma_alpha);
    beta  ~ normal(0, sigma_beta);
    y ~ bernoulli_logit(alpha + X * beta);
  }
  generated quantities {
    vector[N] log_lik;
    for (n in 1:N) {
      log_lik[n] = bernoulli_logit_lpmf(y[n] | alpha + X[n] * beta);
    }
  }"
  
  # ---- helper: PR-AUC (average precision style) ---------------------------
  pr_auc_ap <- function(y_true, p_hat) {
    o <- order(p_hat, decreasing = TRUE)
    y <- y_true[o]
    tp <- cumsum(y == 1)
    fp <- cumsum(y == 0)
    total_pos <- sum(y == 1)
    if (total_pos == 0) return(NA_real_)
    recall <- tp / total_pos
    precision <- tp / (tp + fp)
    recall_prev <- c(0, recall[-length(recall)])
    sum((recall - recall_prev) * precision)
  }
  
  # ---- helper: calibration intercept & slope ------------------------------
  calibration_stats <- function(y_true, p_hat) {
    eps <- 1e-6
    p_clipped <- pmin(pmax(p_hat, eps), 1 - eps)
    logit_p <- qlogis(p_clipped)
    cal_fit <- glm(y_true ~ logit_p, family = binomial())
    coefs <- coef(cal_fit)
    list(
      intercept = unname(coefs[1]),
      slope     = unname(coefs["logit_p"])
    )
  }
  
  #-------------------------------
  # 1. Train/test split
  #-------------------------------
  set.seed(seed)
  n <- nrow(data)
  train_size <- floor(train_frac * n)
  train_index <- sample(seq_len(n), size = train_size)
  
  train <- data[train_index, , drop = FALSE]
  test  <- data[-train_index, , drop = FALSE]
  
  #-------------------------------
  # 2. Standardization (training-based)
  #-------------------------------
  scale_list <- lapply(train[, predictors, drop = FALSE],
                       function(x) c(mean = mean(x), sd = sd(x)))
  scale_values <- do.call(rbind, scale_list)
  
  standardize_df <- function(df) {
    out <- df
    for (p in predictors) {
      m <- scale_values[p, "mean"]
      s <- scale_values[p, "sd"]
      out[[p]] <- (out[[p]] - m) / s
    }
    out
  }
  
  train_std <- standardize_df(train)
  test_std  <- standardize_df(test)
  
  #-------------------------------
  # 3. Model matrices
  #-------------------------------
  formula_str <- paste(response, "~", paste(predictors, collapse = " + "))
  formula_obj <- as.formula(formula_str)
  
  X_train <- model.matrix(formula_obj, data = train_std)[, -1, drop = FALSE]
  y_train <- train_std[[response]]
  
  X_test  <- model.matrix(formula_obj, data = test_std)[, -1, drop = FALSE]
  y_test  <- test_std[[response]]
  
  N <- nrow(X_train)
  K <- ncol(X_train)
  
  stan_data <- list(
    N = N,
    K = K,
    X = X_train,
    y = y_train,
    sigma_beta = sigma_beta,
    sigma_alpha = sigma_alpha
  )
  
  #-------------------------------
  # 4. Compile Stan model (once per session)
  #-------------------------------
  if (is.null(compiled_model)) {
    sm <- rstan::stan_model(model_code = stan_code)
  } else {
    sm <- compiled_model
  }
  
  #-------------------------------
  # 5. Sample from posterior
  #-------------------------------
  fit <- rstan::sampling(
    sm,
    data = stan_data,
    iter = iter,
    chains = chains,
    seed = seed,
    refresh = 0
  )
  
  #-------------------------------
  # 6. Posterior summaries (alpha, beta)
  #-------------------------------
  summ <- rstan::summary(fit, pars = c("alpha", "beta"))$summary
  alpha_mean <- summ["alpha", "mean"]
  beta_means <- summ[grep("^beta\\[", rownames(summ)), "mean"]
  names(beta_means) <- colnames(X_train)
  
  #-------------------------------
  # 7. Test-set predictions (using posterior means)
  #-------------------------------
  linpred_test <- as.vector(alpha_mean + X_test %*% beta_means)
  prob_test <- plogis(linpred_test)
  class_test <- ifelse(prob_test >= 0.5, 1L, 0L)
  
  #-------------------------------
  # 8. Point-estimate performance metrics
  #-------------------------------
  roc_obj <- roc(response = y_test, predictor = prob_test)
  auc_val <- as.numeric(auc(roc_obj))
  
  accuracy <- mean(class_test == y_test)
  brier <- mean((prob_test - y_test)^2)
  pr_auc_val <- pr_auc_ap(y_test, prob_test)
  cal <- calibration_stats(y_test, prob_test)
  
  #-------------------------------
  # 9. Bootstrap CIs (optional)
  #-------------------------------
  AUC_CI <- PR_AUC_CI <- ACC_CI <- Brier_CI <- 
    calib_int_CI <- calib_slope_CI <- c(NA_real_, NA_real_)
  
  if (boot) {
    n_test <- length(y_test)
    AUC_boot   <- numeric(n_boot)
    PR_boot    <- numeric(n_boot)
    ACC_boot   <- numeric(n_boot)
    Brier_boot <- numeric(n_boot)
    INT_boot   <- numeric(n_boot)
    SLOPE_boot <- numeric(n_boot)
    
    for (b in seq_len(n_boot)) {
      idx <- sample.int(n_test, n_test, replace = TRUE)
      y_b <- y_test[idx]
      p_b <- prob_test[idx]
      c_b <- ifelse(p_b >= 0.5, 1L, 0L)
      
      # AUC
      if (length(unique(y_b)) == 2) {
        AUC_boot[b] <- tryCatch(
          as.numeric(auc(roc(response = y_b, predictor = p_b))),
          error = function(e) NA_real_
        )
      } else {
        AUC_boot[b] <- NA_real_
      }
      
      # PR-AUC
      PR_boot[b] <- pr_auc_ap(y_b, p_b)
      
      # Accuracy
      ACC_boot[b] <- mean(c_b == y_b)
      
      # Brier
      Brier_boot[b] <- mean((p_b - y_b)^2)
      
      # Calibration
      cal_b <- tryCatch(
        calibration_stats(y_b, p_b),
        error = function(e) list(intercept = NA_real_, slope = NA_real_)
      )
      INT_boot[b]   <- cal_b$intercept
      SLOPE_boot[b] <- cal_b$slope
    }
    
    qfun <- function(x) quantile(x[is.finite(x) & !is.na(x)], probs = c(0.025, 0.975))
    
    if (any(is.finite(AUC_boot)))       AUC_CI         <- qfun(AUC_boot)
    if (any(is.finite(PR_boot)))        PR_AUC_CI      <- qfun(PR_boot)
    if (any(is.finite(ACC_boot)))       ACC_CI         <- qfun(ACC_boot)
    if (any(is.finite(Brier_boot)))     Brier_CI       <- qfun(Brier_boot)
    if (any(is.finite(INT_boot)))       calib_int_CI   <- qfun(INT_boot)
    if (any(is.finite(SLOPE_boot)))     calib_slope_CI <- qfun(SLOPE_boot)
  }
  
  #-------------------------------
  # 10. Return everything
  #-------------------------------
  list(
    call = match.call(),
    formula = formula_obj,
    predictors = predictors,
    stan_fit = fit,
    scale_values = scale_values,
    posterior_summary = summ,
    coef_means = list(
      alpha = alpha_mean,
      beta = beta_means
    ),
    predictions = data.frame(
      row_id = rownames(test),
      observed = y_test,
      predicted_prob = prob_test,
      predicted_class = class_test
    ),
    performance = list(
      AUC = auc_val,
      AUC_CI = AUC_CI,
      PR_AUC = pr_auc_val,
      PR_AUC_CI = PR_AUC_CI,
      accuracy = accuracy,
      accuracy_CI = ACC_CI,
      Brier = brier,
      Brier_CI = Brier_CI,
      calib_intercept = cal$intercept,
      calib_intercept_CI = calib_int_CI,
      calib_slope = cal$slope,
      calib_slope_CI = calib_slope_CI
    ),
    train_index = train_index,
    compiled_model = sm
  )
}

```

Function to check if model is saved
```{r}
run_or_load_model <- function(path, fit_expr, verbose = TRUE) {
  # path: where to save the fitted model (.rds)
  # fit_expr: expression that returns the fitted object (e.g., your fit_bayes_logit_ridge call)
  
  if (file.exists(path)) {
    if (verbose) message("Loading saved model from: ", path)
    fit <- readRDS(path)
  } else {
    if (verbose) message("Fitting model (cache not found): ", path)
    fit <- eval.parent(substitute(fit_expr))
    
    # Ensure directory exists
    dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
    
    if (verbose) message("Saving fitted model to: ", path)
    saveRDS(fit, path)
  }
  
  fit
}
```


Fit model
```{r}
predictors_BLR.1  <- c("temperature", "precipitation", "wind", "relative_humidity")#, 
                   #"FFMC", "DMC", "DC", 
                   #"C", "B", "U")
model_path <- file.path(rds_root, "logistic_regression.rds")

BLR.1 <- run_or_load_model(
  path = model_path,
  fit_bayes_logit_ridge(
  data       = df,              
  predictors = predictors_m1.1,
  response   = "caused_fire",
  sigma_beta = 1,               # prior SD for coefficients (ridge strength)
  sigma_alpha = 5,              # vague prior for intercept
  iter       = 2000,
  chains     = 4
))

```





# ── Save predictions ──────────────────────────────────────────────────────────
```{r}
# write.csv(metrics_ignition,
#           file.path(predictions_root, "ignition_model_metrics.csv"),
#           row.names = FALSE)
# # ── Save updated metrics and predictions ───────────────────────────────────────
# write.csv(metrics_ignition_updated,
#           file.path(predictions_root, "ignition_model_metrics.csv"),
#           row.names = FALSE)
# 
# write.csv(test_strikes,
#           file.path(predictions_root, "ignition_predictions.csv"),
#           row.names = FALSE)
```

```{r}


```