---
title: "Ignition Efficiency Random Forest"
output: html_document
---

```{r}
#library(randomForest)
library(ranger)
library(dplyr)
library(ggplot2)
library(tidyr)
```


Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
```


```{r}
df <- read.csv(file.path(drivers_root, "lightning_fires_burned_area_jja_2002_2011.csv"))
df <- df[,2:29]
head(df)
```

----------------------------------------------
Run random forest for Strike Count
----------------------------------------------

```{r}

# ── Define predictor set ───────────────────────────────────────────────────────
# Adjust column names to match your actual df
predictors <- c("precip", "tair", "wind", "rh",
                "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                "tair_1m", "tair_2m", "tair_3m", "tair_5y",
                "C", "B", "U", "sd")

# ── Random forest 1: Lightning strike count (rstrike) ─────────────────────────
df_strikes <- df[, c("strike_count", predictors)]
df_strikes <- na.omit(df_strikes)

# Correlation check on a sample
#df_sample <- df_strikes %>% sample_frac(0.1)
#plot(df_sample)

#remove a percent of 0s
zero_indices <- which(df_strikes$strike_count == 0)
num_to_remove <- round(length(zero_indices) * 0.75)
set.seed(42)
rows_to_remove <- sample(zero_indices, num_to_remove)
df_strikes <- df_strikes[-rows_to_remove, ]

# ── Fit ranger twice to get both importance metrics ───────────────────────────

rf_permutation <- ranger(
  strike_count ~ .,
  data        = df_strikes,
  num.trees   = 500,
  importance  = "permutation",
  num.threads = 4
)

rf_impurity <- ranger(
  strike_count ~ .,
  data        = df_strikes,
  num.trees   = 500,
  importance  = "impurity",
  num.threads = 4
)

# ── Extract and combine importance scores ─────────────────────────────────────

imp_mse <- data.frame(
  variable   = names(importance(rf_permutation)),
  importance = importance(rf_permutation),
  metric     = "%IncMSE"
)

imp_purity <- data.frame(
  variable   = names(importance(rf_impurity)),
  importance = importance(rf_impurity),
  metric     = "IncNodePurity"
)

imp_all <- bind_rows(imp_mse, imp_purity)

# Order variables by %IncMSE value (so both panels share the same y-axis order)
var_order <- imp_mse |>
  arrange(importance) |>
  pull(variable)

imp_all <- imp_all |>
  mutate(variable = factor(variable, levels = var_order))

# ── Plot ──────────────────────────────────────────────────────────────────────

# Label formatting: IncNodePurity values can be large — use scientific notation
imp_all <- imp_all |>
  mutate(metric = factor(metric, levels = c("%IncMSE", "IncNodePurity")))

p <- ggplot(imp_all, aes(x = importance, y = variable)) +
  geom_hline(
    aes(yintercept = as.numeric(variable)),
    linetype  = "dashed",
    color     = "grey75",
    linewidth = 0.4
  ) +
  geom_point(
    shape  = 21,
    size   = 3,
    color  = "black",
    fill   = "white",
    stroke = 0.7
  ) +
  facet_wrap(
    ~ metric,
    scales = "free_x",   # each panel gets its own x-axis scale
    ncol   = 2
  ) +
  labs(
    title = "Variable Importance — Strike Count",
    x     = NULL,
    y     = NULL
  ) +
  theme_bw() +
  theme(
    panel.grid        = element_blank(),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.8),
    strip.background  = element_blank(),
    strip.text        = element_text(size = 10),
    axis.text.y       = element_text(size = 9),
    axis.text.x       = element_text(size = 8),
    plot.title        = element_text(size = 11, hjust = 0.5)
  )

print(p)

```

Save
```{r}
#ggsave("varimp_burned_area.png", p, width = 7, height = 4.5, dpi = 300)

write.csv(imp_all, (file.path(table_root, "strike_count_random_forest_075_zeros.csv")), row.names = FALSE)
```

----------------------------------------------
Run random forest for Lightning-caused Fire Count
----------------------------------------------
```{r}
#peak at data
summary(df$fire_count)
sum(df$fire_count == 0, na.rm = TRUE)
nrow(df)
sum(df$fire_count == 0, na.rm = TRUE)/nrow(df)
hist(df$fire_count)
ggplot(data=filter(df, year==2002), aes(x=lon, y=lat, color=fire_count)) + geom_point()
ggplot(data=filter(df, fire_count>0), aes(x=lon, y=lat, color=fire_count)) + geom_point()
```
Over 99% of observations have burned_area=0. Let's remove 100% and then 75% of these rows before running the random forest. 

```{r}

# ── Define predictor set ───────────────────────────────────────────────────────
# Adjust column names to match your actual df
predictors <- c("precip", "tair", "wind", "rh",
                "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                "tair_1m", "tair_2m", "tair_3m", "tair_5y",
                "C", "B", "U", "sd")

# ── Random forest 1: Lightning strike count (rstrike) ─────────────────────────
df_fire <- df[, c("fire_count", predictors)]
df_fire <- na.omit(df_fire)

# Correlation check on a sample
#df_sample <- df_fire %>% sample_frac(0.1)
#plot(df_sample)

#remove a percent of 0s
zero_indices <- which(df_fire$fire_count == 0)
num_to_remove <- round(length(zero_indices) * 0.5) # % of fire_count==0 rows to remove
set.seed(42)
rows_to_remove <- sample(zero_indices, num_to_remove)
df_fire <- df_fire[-rows_to_remove, ]

# ── Fit ranger twice to get both importance metrics ───────────────────────────

rf_permutation <- ranger(
  fire_count ~ .,
  data        = df_fire,
  num.trees   = 500,
  importance  = "permutation",
  num.threads = 4
)

rf_impurity <- ranger(
  fire_count ~ .,
  data        = df_fire,
  num.trees   = 500,
  importance  = "impurity",
  num.threads = 4
)

# ── Extract and combine importance scores ─────────────────────────────────────

imp_mse <- data.frame(
  variable   = names(importance(rf_permutation)),
  importance = importance(rf_permutation),
  metric     = "%IncMSE"
)

imp_purity <- data.frame(
  variable   = names(importance(rf_impurity)),
  importance = importance(rf_impurity),
  metric     = "IncNodePurity"
)

imp_all <- bind_rows(imp_mse, imp_purity)

# Order variables by %IncMSE value (so both panels share the same y-axis order)
var_order <- imp_mse |>
  arrange(importance) |>
  pull(variable)

imp_all <- imp_all |>
  mutate(variable = factor(variable, levels = var_order))

# ── Plot ──────────────────────────────────────────────────────────────────────

# Label formatting: IncNodePurity values can be large — use scientific notation
imp_all <- imp_all |>
  mutate(metric = factor(metric, levels = c("%IncMSE", "IncNodePurity")))

p <- ggplot(imp_all, aes(x = importance, y = variable)) +
  geom_hline(
    aes(yintercept = as.numeric(variable)),
    linetype  = "dashed",
    color     = "grey75",
    linewidth = 0.4
  ) +
  geom_point(
    shape  = 21,
    size   = 3,
    color  = "black",
    fill   = "white",
    stroke = 0.7
  ) +
  facet_wrap(
    ~ metric,
    scales = "free_x",   # each panel gets its own x-axis scale
    ncol   = 2
  ) +
  labs(
    title = "Variable Importance — Fire Count",
    x     = NULL,
    y     = NULL
  ) +
  theme_bw() +
  theme(
    panel.grid        = element_blank(),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.8),
    strip.background  = element_blank(),
    strip.text        = element_text(size = 10),
    axis.text.y       = element_text(size = 9),
    axis.text.x       = element_text(size = 8),
    plot.title        = element_text(size = 11, hjust = 0.5)
  )

print(p)
```

Save
```{r}
#ggsave("varimp_burned_area.png", p, width = 7, height = 4.5, dpi = 300)

write.csv(imp_all, (file.path(table_root, "fire_count_random_forest_050_zeros.csv")), row.names = FALSE)
```

```{r}

# ── Compare importance across both models ─────────────────────────────────────
imp_strikes <- as.data.frame(importance(rf_strikes)) %>%
  tibble::rownames_to_column("variable") %>%
  mutate(model = "Strike count")

imp_fire <- as.data.frame(importance(rf_fire)) %>%
  tibble::rownames_to_column("variable") %>%
  mutate(model = "Fire count")

imp_combined <- bind_rows(imp_strikes, imp_fire)

# Plot %IncMSE side by side for both models
ggplot(imp_combined, aes(x = reorder(variable, `%IncMSE`),
                          y = `%IncMSE`,
                          fill = model)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Variable Importance: Strike Count vs Fire Count",
    x     = "Predictor",
    y     = "%IncMSE",
    fill  = "Model"
  ) +
  theme_minimal())


```


# Mean Decrease Accuracy (%IncMSE):
- remove variables from the model and asses which have the greatest impact on model accuracy.
- Higher the value of mean decrease accuracy or mean decrease gini score,
       higher the importance of the variable in the model.
- tells you effect of randomly shuffling values of a variable on increasing mse of model predictions
- a higher number indicates that the variable is more important

# Mean Decrease Gini or Increase in Node Purity (IncNodePurity):
- expresses the change in the homogeneity of the of the groups created by the trees
     (using the Gini coefficient as a measure).
- What is expressed is the decrease in said purity if a particular variable
      has no information. If a variable has no information to begin with,
      the decrase would be zero. (from: https://www.reagro.org/methods/statistical/randomforest.html)

Save random forest results for appendix figure. 
```{r}
rf_results <- importance(rf_burned_area)
rf_results <- data.frame(Variable = rownames(rf_results),
                     IncMSE = rf_results[, "%IncMSE"],
                     IncNodePurity = rf_results[, "IncNodePurity"])
# Export
write.csv(rf_results, (file.path(table_root, "burned_area_random_forest.csv")), row.names = FALSE)
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```