---
title: "Rate_Calibration"
output: html_document
---

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(patchwork)
library(ggplot2)
library(gridExtra)
library(terra)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(glmnet)
library(pROC)
library(PRROC)
```
Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]

```


Read in drivers
```{r}
df <- read.csv(file.path(drivers_root, "lightning_fires_burned_area_jja_2002_2011.csv"))
df <- df[,2:29]

#Random forest results indicate that U (% open land), C (% conifer cover), and sd (snow depth) should be excluded
df <- df[,c("lat","lon","year","month","rh","tair","precip","wind","precip_1m","precip_2m","precip_3m","precip_5y","tair_1m","tair_2m","tair_3m","tair_5y","B","strike_count","fire_count")]
head(df)
```

Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]
```

```{r}
# ── Save scaling parameters ───────────────────────────────────────────────────
predictors_ignition <- c("rh", "tair", "precip", "wind",
                          "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                          "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")

mu_rh        <- mean(train$rh);        sigma_rh        <- sd(train$rh)
mu_tair      <- mean(train$tair);      sigma_tair      <- sd(train$tair)
mu_precip    <- mean(train$precip);    sigma_precip    <- sd(train$precip)
mu_wind      <- mean(train$wind);      sigma_wind      <- sd(train$wind)
mu_precip_1m <- mean(train$precip_1m); sigma_precip_1m <- sd(train$precip_1m)
mu_precip_2m <- mean(train$precip_2m); sigma_precip_2m <- sd(train$precip_2m)
mu_precip_3m <- mean(train$precip_3m); sigma_precip_3m <- sd(train$precip_3m)
mu_precip_5y <- mean(train$precip_5y); sigma_precip_5y <- sd(train$precip_5y)
mu_tair_1m   <- mean(train$tair_1m);   sigma_tair_1m   <- sd(train$tair_1m)
mu_tair_2m   <- mean(train$tair_2m);   sigma_tair_2m   <- sd(train$tair_2m)
mu_tair_3m   <- mean(train$tair_3m);   sigma_tair_3m   <- sd(train$tair_3m)
mu_tair_5y   <- mean(train$tair_5y);   sigma_tair_5y   <- sd(train$tair_5y)
mu_B         <- mean(train$B);         sigma_B         <- sd(train$B)

# Store and save
mean_sd_ignition <- data.frame(
  var   = predictors_ignition,
  mu    = c(mu_rh, mu_tair, mu_precip, mu_wind,
            mu_precip_1m, mu_precip_2m, mu_precip_3m, mu_precip_5y,
            mu_tair_1m, mu_tair_2m, mu_tair_3m, mu_tair_5y, mu_B),
  sigma = c(sigma_rh, sigma_tair, sigma_precip, sigma_wind,
            sigma_precip_1m, sigma_precip_2m, sigma_precip_3m, sigma_precip_5y,
            sigma_tair_1m, sigma_tair_2m, sigma_tair_3m, sigma_tair_5y, sigma_B)
)

write.csv(mean_sd_ignition,
          file.path(mu_sigma_root, "mu_sigma_ignition_predictors.csv"),
          row.names = FALSE)

# ── Standardize train ─────────────────────────────────────────────────────────
train <- train %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B
  )

train <- train[, c("lat", "lon", "year", "month",
                   "strike_count", "fire_count",
                   "rh", "tair", "precip", "wind",
                   "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                   "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")]

# ── Standardize test ──────────────────────────────────────────────────────────
# Use training means and sds 
test <- test %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B
  )

test <- test[, c("lat", "lon", "year", "month",
                 "strike_count", "fire_count",
                 "rh", "tair", "precip", "wind",
                 "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                 "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B")]

head(train)

```

##########################################
##### IGNITION EFFICINCY (p_ignite) #####
##########################################

Y ~ Binomial(n, p_ignite) --> technically, this is applied in the next step, to estimate fire count.

logit(p_ignite) = alpha + X * beta

alpha ~ normal(mu_alpha, sigma_alpha)
beta  ~ normal(0, sigma_beta)

p_ignite_corrected = plogis(qlogis(pie) + cal_shift)

p_ignite: probability of 1 strike resulting in an ignition in a given gridcell and month. 
X: average precipitation rate, temperature, relative humidity, cumulative precipitation for the previous 1, 2, 3 months, and 5 years, mean temperature for the previous 1, 2, 3 months, and 5 years, and % broadleaf cover. 


```{r}

#Bayes recalibrated
#Load calibrated Bayesian model
ignition_bayes_obj <- readRDS(file.path(rds_root, 
                                         "Ignition_Efficiency/m2_bayes_calibrated.rds"))

ignition_bayes_fit <- ignition_bayes_obj$fit        # stanfit object
cal_shift_bayes    <- ignition_bayes_obj$cal_shift  # intercept shift
coef_means         <- ignition_bayes_obj$coef_means # posterior mean alpha and beta

cat("Bayesian calibration shift:", round(cal_shift_bayes, 3), "\n")

# ── Build predictor matrix from standardised data
predictors_m2 <- c("rh", "tair", "precip",
                    "precip_1m", "precip_2m", "precip_3m",
                    "precip_5y", "tair_1m", "tair_2m",
                    "tair_3m", "tair_5y", "B")

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m2, collapse = " + "), "- 1")),
  data = train
)

#generate predictions using posterior mean
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

train <- train |>
  mutate(
    #inverse_logit of linear predictor, linpred
    p_ignite_bayes     = plogis(linpred),
    #add calibration shift 
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,     #pmin and pmax make sure arguments in plogis from reaching infinity
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

#verify
cat("Raw Bayes p_ignite summary:\n")
print(summary(train$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(train$p_ignite_bayes_cal))

```

##### RATE CALIBRATION SHIFT FOR P_IGNITE ##### 

Ignition efficiency model not capture the observed fire/strike rate. 


This is due to a combination of model type and data structure.

What the model was trained to predict
The binomial logistic regression was trained on gridcell-month outcomes — cbind(fire_count, strike_count - fire_count) — where the outcome is whether each strike results in a fire. The model maximises the binomial likelihood across all gridcell-months simultaneously.
However, the vast majority of gridcell-months have fire_count = 0 even when strike_count > 0. The model sees thousands of zero-fire months and adjusts its intercept downward to fit them. But "zero fires in a gridcell-month with 50 strikes" doesn't mean each individual strike had zero probability of igniting — it means the probability was low enough that none of 50 independent trials succeeded. The model conflates these two things.

Rare event bias in logistic regression
When events are rare (ie. lightning-caused fires are extremely rare at the gridcell-month level (prevalence ~0.004)) maximum likelihood estimation of the intercept in logistic regression is biased downward. The intercept gets pulled toward negative infinity to fit the overwhelming proportion of zeros.

The binary calibration only partially fixes it
The binary calibration corrects the model to match the observed gridcell-month fire rate — the fraction of gridcell-months with at least one fire. But this is still not the same as the per-strike ignition rate.
Consider a gridcell-month with 100 strikes and 1 fire:
* Binary outcome: 1 (fire occurred) — contributes to binary calibration target
* Per-strike rate: 1/100 = 0.01 — this is what you need for p_fire = 1-(1-p)^n
The binary calibration makes the model predict the right fraction of fire-months, but p_ignite ends up being the gridcell-month fire probability, not the per-strike probability. When you then raise it to the power of hundreds of strikes, you get enormous fire probabilities.

The rate calibration fixes the scale
The rate calibration specifically targets sum(fire_count) / sum(strike_count) — the true per-strike ignition rate — and estimates the additional shift needed to align p_ignite with this quantity. This is the number that belongs in 1 - (1-p)^n_strikes.

In summary, as a flow:

Model training:
  Binomial GLM learns P(fire | gridcell-month climate)
  → Rare event bias pulls intercept too low
  → Binary calibration corrects to P(fire-month | climate)
  → But P(fire-month) ≠ P(fire | single strike)
  → Rate calibration corrects to P(fire | single strike)
  → Now correctly used in 1-(1-p)^n_strikes
The need for two calibration steps reflects the fact that the model was specified at the gridcell-month level for practical reasons (that's the resolution of your climate data) but needs to produce predictions at the per-strike level for the fire probability formula. These are fundamentally different quantities that happen to be related through the binomial formula, and the two calibration steps bridge that gap.

so the model is trained to predict the probability of a fire given x strikes. but not the probability of a fire for each strike? the rate caliibration 


```{r}
#Estimate rate calibration shift from train data
#wses gridcell-months with strikes > 0
#weights by strike_count so high-strike months have more influence

train <- train |>
  filter(strike_count > 0) |>   #the lightning model takes care of 0 stirke counts, so dont need the P_ignite model to worry about zeros.
  mutate(
    obs_rate    = fire_count / strike_count,
    eps         = 1e-8,
    logit_p_cal = qlogis(pmin(pmax(p_ignite_bayes_cal, eps), 1 - eps))
  )

rate_cal_model <- glm(
  obs_rate ~ offset(logit_p_cal),
  data    = train,
  family  = binomial(),
  weights = strike_count
)

rate_cal_shift <- coef(rate_cal_model)[1]
cat("Rate calibration shift:", round(rate_cal_shift, 3), "\n")
# Expect a negative value — shifts predictions down toward true per-strike rate

#apply combined shift to all predictions
eps <- 1e-8

train <- train |>
  mutate(
    p_ignite_rate_cal = plogis(
      qlogis(pmin(pmax(p_ignite_bayes_cal, eps), 1 - eps)) + rate_cal_shift
    )
  )

# verify annual rates now match observed 
train |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate           = sum(fire_count) / sum(strike_count),
    pred_binary_cal    = mean(p_ignite_bayes_cal),
    pred_rate_cal      = mean(p_ignite_rate_cal),
    .groups            = "drop"
  )

#check expected fire count
cat("Expected fires (binary cal):  ",
    sum(test$strike_count * test$p_ignite_bayes_cal, na.rm = TRUE), "\n")
cat("Expected fires (rate cal):    ",
    sum(test$strike_count * test$p_ignite_rate_cal,  na.rm = TRUE), "\n")
cat("Observed fires:               ",
    sum(test$lightning_ignition), "\n")

#save shift for paleo application
saveRDS(list(
  cal_shift_bayes = cal_shift_bayes,   # binary calibration shift
  rate_cal_shift  = rate_cal_shift,    # rate calibration shift
  combined_shift  = cal_shift_bayes + rate_cal_shift
), file.path(rds_root, "Ignition_Efficiency/calibration_shifts.rds"))

```


### Apply full ignition model to test set


```{r}

############################
# Intercept calibration
############################

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m2, collapse = " + "), "- 1")),
  data = test
)

#generate predictions using posterior mean
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

test <- test |>
  mutate(
    #inverse_logit of linear predictor, linpred
    p_ignite_bayes     = plogis(linpred),
    #add calibration shift 
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,     #pmin and pmax make sure arguments in plogis from reaching infinity
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

#verify
cat("Raw Bayes p_ignite summary:\n")
print(summary(test$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(test$p_ignite_bayes_cal))

############################
# Rate Calibration Shift
############################


test <- test |>
  mutate(
    p_ignite_rate_cal = plogis(
      qlogis(pmin(pmax(p_ignite_bayes_cal, eps), 1 - eps)) + rate_cal_shift
    )
  )

# verify annual rates now match observed 
test |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate           = sum(fire_count) / sum(strike_count),
    pred_binary_cal    = mean(p_ignite_bayes_cal),
    pred_rate_cal      = mean(p_ignite_rate_cal),
    .groups            = "drop"
  )





```

Performance


```{r}
# Annual aggregation
test |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate  = sum(fire_count) / sum(strike_count),
    pred_rate = mean(p_ignite_rate_cal),
    .groups   = "drop"
  ) |>
  summarise(
    mean_obs_rate = mean(obs_rate),
    mean_pred_rate = mean(pred_rate), 
    spearman = cor(obs_rate, pred_rate, method = "spearman"),
    ratio    = sum(pred_rate) / sum(obs_rate)
  )

# Annual gridcell aggregation
test |>
  filter(strike_count > 0) |>
  group_by(year, lat, lon) |>
  summarise(
    obs_rate  = sum(fire_count) / sum(strike_count),
    pred_rate = mean(p_ignite_rate_cal),
    .groups   = "drop"
  ) |>
  summarise(
    mean_obs_rate = mean(obs_rate),
    mean_pred_rate = mean(pred_rate), 
    spearman = cor(obs_rate, pred_rate, method = "spearman"),
    ratio    = sum(pred_rate) / sum(obs_rate)
  )
# Monthly gridcell aggregation

test |>
  filter(strike_count > 0) |>
  group_by(year, lat, lon, month) |>
  summarise(
    obs_rate  = sum(fire_count) / sum(strike_count),
    pred_rate = mean(p_ignite_rate_cal),
    .groups   = "drop"
  ) |>
  summarise(
    mean_obs_rate = mean(obs_rate),
    mean_pred_rate = mean(pred_rate), 
    spearman = cor(obs_rate, pred_rate, method = "spearman"),
    ratio    = sum(pred_rate) / sum(obs_rate)
  )
```


Ignition efficiency: rate calibrated (per-strike ignition probability)

Observed vs predicted per-strike ignition rate: sum(fire_count) / sum(strike_count) vs mean(p_ignite_rate_cal) aggregated by year, directly measures whether the calibrated probability matches the true per-strike ignition rate that enters the fire probability formula

- Expected vs observed fire count ratio — sum(n_strikes × p_ignite_rate_cal) vs observed fire count, evaluates whether the calibrated probability produces the right aggregate number of fires when multiplied by strike counts across all gridcell-months. 1 is perfect. 
- Spearman correlation between observed and predicted ignition rate — computed at the annual or spatial level (not gridcell-month, where noise is too high), assesses whether the model correctly ranks years or regions by ignition efficiency. 1 = perfect, 0.7 - 0.9, strong correlation

These differ from the binary fire count metrics because the rate-calibrated model is no longer evaluated as a 0/1 classifier — it's evaluated as a rate predictor. ROC AUC and PR AUC are not appropriate here since you're no longer predicting a binary outcome but a continuous per-strike probability that feeds into the count simulation.

The aggregation level matters — gridcell-month observed rates (fire_count / strike_count) are too noisy for meaningful correlation since most are exactly 0 or 1. Annual or multi-year spatial aggregation gives stable estimates

```{r}
plot_df <- test |>
  filter(strike_count > 0) |>
  group_by(year, lat, lon, month) |>
  summarise(
    obs_rate  = sum(fire_count) / sum(strike_count),
    pred_rate = mean(p_ignite_rate_cal),
    .groups   = "drop"
  ) 

ggplot(plot_df, aes(x=pred_rate, y=obs_rate)) + geom_point()

plot_df <- test |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate  = sum(fire_count) / sum(strike_count),
    pred_rate = mean(p_ignite_rate_cal),
    .groups   = "drop"
  ) 

df_long <- plot_df %>% 
  pivot_longer(
    cols = c(obs_rate, pred_rate),  # Columns to transform (can also use: c(Year2024, Year2025) or 2:3)
    names_to = "outcome",           # Name of the new category column
    values_to = "rate"          # Name of the new metrics column
  )

ggplot(df_long, aes(x=year, y=rate, color=outcome)) + geom_point()

```
