---
title: "For Model Pipeline"
author: "Charlotte Uden"
date: "2026-06-14"
output: html_document
---


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(pROC)
library(loo)
library(cowplot)
library(stats)
library(caret)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

##########################################
#####       LIGHTNING    (r_strike)  #####
##########################################

r_strike ~ Gamma(alpha, beta)

alpha = exp(a_alpha + b_alpha * X)

beta = exp(a_beta + b_beta * X)

a_alpha, b_alpha, a_beta, b_beta ~ Normal(0, 1)

r_strike: lightning strikes / km2 / month
X: temperature, precipitation, relative humidity, shortwave radiation, surface pressure

exp(): ensures alpha and beta stay positive, which the gamma distribution requires. 

```{r}
lightning <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))
head(lightning)
hist(lightning$strikes)
```

Predict lightning strike rate for all gridcells and years
```{r}
#standardize:
lightning_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) lightning_mu_sigma$mu[lightning_mu_sigma$var == var]
get_sigma <- function(var) lightning_mu_sigma$sigma[lightning_mu_sigma$var == var]

lightning_scaled <- lightning |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    swr    = (swr    - get_mu("swr"))    / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh")
  )

#read in model
#lightning_model <- readRDS(file.path(rds_root, "G13_gamma_glm_lightning.rds"))

#make predictions
# lightning_scaled <- lightning_scaled |>
#   mutate(
#     strikes_pred = predict(lightning_model,
#                            newdata = lightning_scaled,
#                            type    = "response")
#   )

# ── Load B12 Bayesian lightning model ─────────────────────────────────────────
B12 <- readRDS(file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))

# ── Extract posterior samples ─────────────────────────────────────────────────
posterior_samples <- as.data.frame(B12)

# Generate predictions using posterior mean coefficients
# B12 excludes wind — predictors: swr, tair, rh, precip, sp

a_alpha_mean <- mean(posterior_samples$a_alpha)
b_alpha_mean <- mean(posterior_samples$b_alpha)
c_alpha_mean <- mean(posterior_samples$c_alpha)
d_alpha_mean <- mean(posterior_samples$d_alpha)
e_alpha_mean <- mean(posterior_samples$e_alpha)
f_alpha_mean <- mean(posterior_samples$f_alpha)

a_beta_mean  <- mean(posterior_samples$a_beta)
b_beta_mean  <- mean(posterior_samples$b_beta)
c_beta_mean  <- mean(posterior_samples$c_beta)
d_beta_mean  <- mean(posterior_samples$d_beta)
e_beta_mean  <- mean(posterior_samples$e_beta)
f_beta_mean  <- mean(posterior_samples$f_beta)

# Compute alpha and beta per observation using posterior means
# Using vectorised operations rather than a loop for speed

lp_alpha <- a_alpha_mean +
            b_alpha_mean * lightning_scaled$swr   +
            c_alpha_mean * lightning_scaled$tair  +
            d_alpha_mean * lightning_scaled$rh    +
            e_alpha_mean * lightning_scaled$precip +
            f_alpha_mean * lightning_scaled$sp

lp_beta  <- a_beta_mean  +
            b_beta_mean  * lightning_scaled$swr   +
            c_beta_mean  * lightning_scaled$tair  +
            d_beta_mean  * lightning_scaled$rh    +
            e_beta_mean  * lightning_scaled$precip +
            f_beta_mean  * lightning_scaled$sp

alpha_pred <- exp(lp_alpha)
beta_pred  <- exp(lp_beta)

set.seed(123)

# Gamma alpha and beta
# if you want the mean: alpha / beta
lightning_scaled <- lightning_scaled |>
  mutate(
    gamma_alpha   = alpha_pred,
    gamma_beta    = beta_pred,
    #r_strike_mean  = alpha_pred / beta_pred   # posterior mean prediction
    r_strike = rgamma(n(), shape = gamma_alpha, rate = gamma_beta)
  )

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Predicted strike rate summary:\n")
print(summary(lightning_scaled$r_strike))
cat("Observed strike rate summary:\n")
print(summary(lightning_scaled$strikes))

lightning_scaled <- lightning_scaled[, c("lat", "lon", "year", "strikes", "r_strike")]
```


##########################################
##### STRIKE COUNT       (n_strikes) #####
##########################################

n_strikes ~ NegBinomial(mu, theta)

mu = r_strike * area

theta = exp(gamma_0 + gamma * X)

Gamma_0 ~ Normal(2,2)
Gamma ~ Normal(0, sigma2)
Sigma = 1 

r_strike: strikes/km2/month
area: gridcell area, km2
n_strikes: number of strikes in a gridcell-month
X: monthly climate predictors, temperature, precipitation, wind, relative humidity, shortwave radiation, surface pressure
theta: Precision/dispersion parameter of the Negative Binomial distribution predicted from climate. Theta > 0.
mu: mean lightning strike rate, strikes/gridcell/month. This is predicted from the Gamma model above and is the same for all summer months (June, July, August) in a year. The summer mean was used to ensure that mu > 0, which is required by the negative binomial distribution. 


```{r}
# ── Apply negative binomial model to fire dataframe ────────────────────────────────────────────────────

#read in drivers
df <- read.csv(file.path(drivers_root, "lightning_strike_count_climate_jja_2002_2011.csv"))

#merge summer strike rate (strikes/km2/month) predictions, r_strike - months will be duplicated
df <- df %>%
  left_join(
    lightning_scaled %>% dplyr::select(lat, lon, year, strikes, r_strike),
    by = c("lat", "lon", "year")
  )

df <- na.omit(df)

names(df)[names(df) == "strikes"] <- "r_strike_obs"
names(df)[names(df) == "strike_count"] <- "n_strikes_obs"

#Scale strike rate from strikes/km2/month to strikes/gridcell/month
area <- read.csv("~/raid/cuden/data/paleo_fire/era5_grid_area.csv")

df <- df %>%
  left_join(
    area %>% dplyr::select(lat, lon, area),
    by = c("lat", "lon")
  )

#standardize:
n_strike_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_count_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) n_strike_mu_sigma$mu[n_strike_mu_sigma$var == var]
get_sigma <- function(var) n_strike_mu_sigma$sigma[n_strike_mu_sigma$var == var]

n_strike_scaled <- df |>
  dplyr::select(lat, lon, year, month, r_strike_obs, n_strikes_obs, r_strike, area, precip, tair, wind, rh, swr, sp) |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh"),
    swr     = (swr     - get_mu("swr"))     / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp")
  )

#caluclate mu for the negative binomial distribution: r_strike * area = strikes/month/gridcell
n_strike_scaled <- n_strike_scaled |>
  mutate(mu_r_strike_gridcell = r_strike * area)

#Predictors for theta (negative binomial parameter we need to sample strike count)
theta_predictors <- c("swr", "tair", "rh", "precip", "sp", "wind")

#read in model
nb_model <- readRDS(file.path(rds_root, "Lightning/neg_bin_strike_count_small.rds"))

#Posterior mean gamma coefficients
gamma_0_mean <- nb_model$coef_means$gamma_0
gamma_means  <- nb_model$coef_means$gamma

#Build predictor matrix for full fire dataframe
Z_fire <- model.matrix(
  as.formula(paste("~", paste(theta_predictors, collapse = " + "), "- 1")),
  data = n_strike_scaled   # must have same standardised predictors
)

#Compute posterior mean theta per gridcell-month
log_theta_pred <- as.numeric(gamma_0_mean + Z_fire %*% gamma_means)
theta_pred     <- exp(log_theta_pred)

#sample strike count using mu=r_strike*area and theta
n_strike_scaled <- n_strike_scaled |>
  mutate(
    nb_theta         = theta_pred,
    log_mu_pred      = log(pmax(mu_r_strike_gridcell, 1e-8)),
    n_strikes = rnbinom(n    = n(),
                                 mu   = mu_r_strike_gridcell,
                                 size = nb_theta)
  )

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Observed zeros:    ", mean(n_strike_scaled$n_strikes_obs == 0), "\n")
cat("Simulated zeros:   ", mean(n_strike_scaled$n_strikes == 0), "\n")
cat("Observed mean:     ", mean(n_strike_scaled$n_strikes_obs), "\n")
cat("Simulated mean:    ", mean(n_strike_scaled$n_strikes), "\n")
cat("Observed count:     ", sum(n_strike_scaled$n_strikes_obs), "\n")
cat("Simulated count:    ", sum(n_strike_scaled$n_strikes), "\n")
cat("Observed variance: ", var(n_strike_scaled$n_strikes_obs), "\n")
cat("Simulated variance:", var(n_strike_scaled$n_strikes), "\n")
summary(n_strike_scaled$n_strikes)
summary(n_strike_scaled$n_strikes_obs)
hist(n_strike_scaled$n_strikes)
hist(n_strike_scaled$n_strikes_obs)


```
Simulated counts come close to observations, capturing the zero outcomes well (zero strike make up 55% of observed outcomes and 64% of simulated outcomes). This issue comes with the model's ability to capture variance in strike counts (simulated var = 453, observed var = 22580). This is driving both a reduction in the simulated mean (8.99, while simulated is 4.94) and total count (1,192,293, while simulated is 654,492). 

It this point, I feel pretty good with the lightning models. 


Read in ingition efficiency and burned area predictors
```{r}
fire <- read.csv(file.path(drivers_root, "lightning_fires_burned_area_jja_2002_2011.csv"))

fire <- fire |>
  mutate(lat = round(lat, 2), lon = round(lon, 2)) |>
  distinct(lat, lon, year, month, .keep_all = TRUE)
```

##########################################
##### IGNITION EFFICINCY (p_ignite) #####
##########################################

Y ~ Binomial(n, p_ignite) --> technically, this is applied in the next step, to estimate fire count.

logit(p_ignite) = alpha + X * beta

alpha ~ normal(mu_alpha, sigma_alpha)
beta  ~ normal(0, sigma_beta)

p_ignite_corrected = plogis(qlogis(pie) + cal_shift)

p_ignite: probability of 1 strike resulting in an ignition in a given gridcell and month. 
X: average precipitation rate, temperature, relative humidity, cumulative precipitation for the previous 1, 2, 3 months, and 5 years, mean temperature for the previous 1, 2, 3 months, and 5 years, and % broadleaf cover. 


```{r}
#standardize:
ignition_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_ignition_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) ignition_mu_sigma$mu[ignition_mu_sigma$var == var]
get_sigma <- function(var) ignition_mu_sigma$sigma[ignition_mu_sigma$var == var]

ignition_scaled <- fire |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_2m   = (precip_2m   - get_mu("precip_2m"))   / get_sigma("precip_2m"),
    precip_3m   = (precip_3m   - get_mu("precip_3m"))   / get_sigma("precip_3m"),
    precip_5y     = (precip_5y     - get_mu("precip_5y"))     / get_sigma("precip_5y"),
    tair_1m = (tair_1m - get_mu("tair_1m")) / get_sigma("tair_1m"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_3m   = (tair_3m   - get_mu("tair_3m"))   / get_sigma("tair_3m"),
    tair_5y     = (tair_5y     - get_mu("tair_5y"))     / get_sigma("tair_5y"),
    B     = (B     - get_mu("B"))     / get_sigma("B")
  )

#Bayes recalibrated
# ── Load calibrated Bayesian model ────────────────────────────────────────────
ignition_bayes_obj <- readRDS(file.path(rds_root, 
                                         "Ignition_Efficiency/m2_bayes_calibrated.rds"))
ignition_bayes_fit <- ignition_bayes_obj$fit        # stanfit object
cal_shift_bayes    <- ignition_bayes_obj$cal_shift  # intercept shift
coef_means         <- ignition_bayes_obj$coef_means # posterior mean alpha and beta

cat("Bayesian calibration shift:", round(cal_shift_bayes, 3), "\n")

# ── Build predictor matrix from standardised data ─────────────────────────────
predictors_m2 <- c("rh", "tair", "precip",
                    "precip_1m", "precip_2m", "precip_3m",
                    "precip_5y", "tair_1m", "tair_2m",
                    "tair_3m", "tair_5y", "B")

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m2, collapse = " + "), "- 1")),
  data = ignition_scaled
)

# ── Generate predictions using posterior mean ─────────────────────────────────
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

ignition_scaled <- ignition_scaled |>
  mutate(
    #inverse_logit of linear predictor, linpred
    p_ignite_bayes     = plogis(linpred),
    #add calibration shift 
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,     #pmin and pmax make sure arguments in plogis from reaching infinity
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Raw Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes_cal))

ignition_scaled <- ignition_scaled[, c("lat", "lon", "year", "month",
                          "fire_count", "strike_count",
                          "p_ignite_bayes_cal")]

ignition_scaled <- ignition_scaled %>% dplyr::rename(p_ignite = p_ignite_bayes_cal)

#add monthly ignition probability, p_ignite, to fire df
fire <- fire %>%
  left_join(
    ignition_scaled %>% dplyr::select(lat, lon, year, month, p_ignite),
    by = c("lat", "lon", "year", "month"))

```


##### RATE CALIBRATION SHIFT FOR P_IGNITE ##### 

```{r}

rate_cal_shift <- readRDS(file.path(rds_root, "Ignition_Efficiency/calibration_shifts.rds"))$rate_cal_shift

eps <- 1e-8

fire <- fire |>
  mutate(
    p_ignite_rate_cal = plogis(
      qlogis(pmin(pmax(p_ignite, eps), 1 - eps)) + rate_cal_shift
    )
  )

# Dannual rates now match observed
fire |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate           = sum(fire_count) / sum(strike_count),
    pred_binary_cal    = mean(p_ignite),
    pred_rate_cal      = mean(p_ignite_rate_cal),
    .groups            = "drop"
  )

#loook at expected fire count
cat("Expected fires (binary cal):  ",
    sum(fire$strike_count * fire$p_ignite, na.rm = TRUE), "\n")
cat("Expected fires (rate cal):    ",
    sum(fire$strike_count * fire$p_ignite_rate_cal,  na.rm = TRUE), "\n")
cat("Observed fires:               ",
    sum(fire$lightning_ignition), "\n")
```


##########################################
#####     BURNED AREA | 1 ignition  #####
##########################################

Burned_area ~ Gamma(theta, theta/mu)

log(mu) = alpha + alpha_year + X * beta

alpha ~ Normal(0, sigma^2_alpha)
beta ~ Normal(0, sigma^2_beta)
theta ~ Exponential(1)
alpha_year ~ Normal(0, sigma^2_year)
sigma^2_year ~ Exponential(1)


Burned_area: area of 1 lightning-caused ignition (km2)
X: Relative humidity, temperature, precipitation, wind, cumulative precipitation for the previous 1, 3 months, and 5 years, mean temperature for the previous 2 months and 5 years, proximity to urban area (binary; 1 = within 40 km of an urban area), proximity to road (binary; 1 = within 40 km of an major road)
alpha_year: year random affect

Apply model
```{r}
#standardize:
burned_area_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_burned_area_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) burned_area_mu_sigma$mu[burned_area_mu_sigma$var == var]
get_sigma <- function(var) burned_area_mu_sigma$sigma[burned_area_mu_sigma$var == var]

burned_area_scaled <- fire |>
  mutate(
    rh        = (rh        - get_mu("rh"))        / get_sigma("rh"),
    tair      = (tair      - get_mu("tair"))      / get_sigma("tair"),
    precip    = (precip    - get_mu("precip"))    / get_sigma("precip"),
    wind      = (wind      - get_mu("wind"))      / get_sigma("wind"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_3m = (precip_3m - get_mu("precip_3m")) / get_sigma("precip_3m"),
    precip_5y = (precip_5y - get_mu("precip_5y")) / get_sigma("precip_5y"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_5y   = (tair_5y   - get_mu("tair_5y"))   / get_sigma("tair_5y")
  ) #no scaling for binary urban and road data.

#read in binary distance data:
proximity <- read.csv(file.path(proximity_data_root, "era5_proximity_road_river_urban.csv"))
proximity <- dplyr::select(proximity, lon, lat, road_proximity_binary, urban_proximity_binary)
#check:
# ggplot() +
#   geom_point(data = proximity |> distinct(lat, lon),
#              aes(x = lon, y = lat),
#              color = "grey70", size = 0.8, alpha = 0.5) 

# Round to 2 decimal places in both dataframes
burned_area_scaled <- burned_area_scaled |>
  mutate(lat = round(lat, 2), lon = round(lon, 2))

proximity <- proximity |>
  mutate(lat = round(lat, 2), lon = round(lon, 2))

#merge by lat lon
burned_area_scaled <- burned_area_scaled |>
  left_join(
    proximity |> dplyr::select(lat, lon, 
                                road_proximity_binary, 
                                urban_proximity_binary),
    by = c("lat", "lon")
  )


preds_final <- c("rh", "tair", "precip", "wind",
                 "precip_1m", "precip_3m", "precip_5y",
                 "tair_2m", "tair_5y",
                 "road_proximity_binary",
                 "urban_proximity_binary")



# ── Load model and extract posterior parameters ───────────────────────────────
m6_obj     <- readRDS(file.path(rds_root, "Burned_Area/m6_bayes_gamma_re.rds"))
m6_fit     <- m6_obj$fit           # stanfit object
coef_means <- m6_obj$coef_means    # list: alpha, beta, alpha_year

alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

cat("Global intercept (alpha):", round(alpha_mean, 3), "\n")
cat("Year random effects:\n")
print(round(coef_means$alpha_year, 3))

# ── Build predictor matrix ────────────────────────────────────────────────────
X_new <- model.matrix(
  as.formula(paste("~", paste(preds_final, collapse = " + "), "- 1")),
  data = burned_area_scaled
)

# ── Generate predictions ──────────────────────────────────────────────────────
# Year random effects set to ZERO for new data (paleo or ERA5 application)
# This represents "climatologically average year" — variation comes from
# climate predictors only, not year-specific synoptic noise

# ── Option A: posterior mean prediction (fast, point estimate) ────────────────
linpred <- as.numeric(alpha_mean + X_new %*% beta_means)
# alpha_year term is omitted — equivalent to setting it to 0

burned_area_scaled <- burned_area_scaled |>
  mutate(burned_area_pred = exp(linpred))

# Extract posterior draws
post <- rstan::extract(m6_fit)

# ── Extract shape parameter (phi) from posterior ───────────────────────────────
phi_mean <- mean(post$phi)
phi_sd   <- sd(post$phi)

cat("Gamma shape (phi) posterior mean:", round(phi_mean, 3), "\n")
cat("Gamma shape (phi) posterior sd:  ", round(phi_sd, 3), "\n")

# ── Add gamma_mean and gamma_shape columns to burned_area_scaled ──────────────
burned_area_scaled <- burned_area_scaled |>
  mutate(
    gamma_mean  = burned_area_pred,   # mean (mu) per gridcell-month
    gamma_shape = phi_mean             # shape (phi) — constant across rows
  )

# rgamma requires rate, not the mean. calculate this as:
# rgamma(n, shape, rate) where rate = shape / mean
burned_area_scaled$gamma_rate <- burned_area_scaled$gamma_shape / burned_area_scaled$gamma_mean

```

##########################################
#####         STRIKE COUNT         #######
##########################################

strike_count = Poisson(r_strike * area)

```{r}
# #merge summer strike rate (strikes/km2/month) predictions - months will be duplicated
# fire <- fire %>%
#   left_join(
#     lightning_scaled %>% dplyr::select(lat, lon, year, strikes, r_strike),
#     by = c("lat", "lon", "year")
#   )
# 
# #Scale strike rate from strikes/km2/month to strikes/gridcell/month
# area <- read.csv("~/raid/cuden/data/paleo_fire/era5_grid_area.csv")
# 
# fire <- fire %>%
#   left_join(
#     area %>% dplyr::select(lat, lon, area),
#     by = c("lat", "lon")
#   )
# 
# #calculate grudcell-level striike rate (strikes/gridcell/month) and use as lambda in poisson distrubution to get strike count
# set.seed(123) 
# #simulated count from predicted rate
# fire$strike_count_pred <- sapply(fire$r_strike * fire$area, function(lam) rpois(1, lambda = lam))
# #simulated count from observed rate
# fire$strike_count_test <- sapply(fire$strikes * fire$area, function(lam) rpois(1, lambda = lam))

```

##########################################
#####         FIRE COUNT         #########
##########################################

fire_count | lightning = Binomial(n_strikes, p_ignite)

```{r}

#add monthly strike count, n_strikes, to fire df
fire <- fire %>%
  left_join(
    n_strike_scaled %>% dplyr::select(lat, lon, year, month, r_strike_obs, n_strikes_obs, r_strike, n_strikes),
    by = c("lat", "lon", "year", "month"))

#adjust p_ignite
#fire <- fire %>% mutate(p_ignite = p_ignite/10)

#Use strike_count and p_ingite to sample fire count from a binomial distribution
fire$fire_count_pred <- mapply(
  function(n, p) rbinom(1, size = n, prob = p),
  fire$n_strikes,
  fire$p_ignite_rate_cal
)   

#Use OBSERVED strike_count and ignition rate (fires/strikes);  to sample fire count from a binomial distribution

#caluclate ignintion frequency from obs fires/strikes, if strieks == 0, p_ignite_test==0. 
fire$p_ignite_test <- mapply(
  function(n_fires, n_strikes) {
    if (is.na(n_fires) || is.na(n_strikes) || n_strikes == 0) return(0L)
    n_fires/n_strikes
  },
  fire$fire_count,
  fire$n_strikes_obs
)
  
#simulate fire count
fire$fire_count_test <- mapply(
  function(n, p) {
    if (is.na(n) || is.na(p) || n == 0) return(0L)
    p_safe <- pmin(pmax(p, 0), 1)   # clip to [0,1] as safety guard
    rbinom(1, size = n, prob = p_safe)
  },
  fire$n_strikes_obs,
  fire$p_ignite_test
)
  
```

#####################################
######### TOTAL BURNED AREA #########
#####################################

total_burned = rgamma(fire_count, burned_area_shape, burned_area_rate)

```{r}
#add monthly burned area gamma parameters to fire df
fire <- fire %>%
  left_join(
    burned_area_scaled %>% dplyr::select(lat, lon, year, month, gamma_shape, gamma_rate),
    by = c("lat", "lon", "year", "month")
  )

#Use fire_count and burned area shape and rate estimates to simulate total burned area from the gamma distribution
fire$total_burned_area <- mapply(
  function(fire_count, BA_shape, BA_rate) {
    if (fire_count == 0 || is.na(fire_count)) return(0)
    sum(rgamma(n = fire_count, shape = BA_shape, rate = BA_rate))
  },
  fire$fire_count_pred,
  fire$gamma_shape,
  fire$gamma_rate
)

#Use observed fire_count and predicted burned area shape and rate estimates to simulate total burned area from the gamma distribution
fire$total_burned_area_test <- mapply(
  function(fire_count, BA_shape, BA_rate) {
    if (fire_count == 0 || is.na(fire_count)) return(0)
    sum(rgamma(n = fire_count, shape = BA_shape, rate = BA_rate))
  },
  fire$fire_count_test,
  fire$gamma_shape,
  fire$gamma_rate
)

```

Quick comparison between observed and predicted 10 year summaries

```{r}
pred_r_strike <- mean(fire$r_strike)
obs_r_strike <- mean(fire$r_strike_obs)

pred_strike_count <- sum(fire$n_strikes)
obs_strike_count <- sum(fire$n_strikes_obs)

pred_p_ignite <- mean(fire$p_ignite_rate_cal)
sim_obs_p_ignite <- mean(fire$p_ignite_test)

pred_fire_count <- sum(fire$fire_count_pred)
obs_fire_count <- sum(fire$fire_count)

pred_burned_area <- sum(fire$total_burned_area)
sim_obs_burned_area <- sum(fire$total_burned_area_test)
obs_burned_area <- sum(fire$burned_area_km2)

summary_table <- as.data.frame(matrix(c(pred_r_strike, obs_r_strike, NA, pred_strike_count, NA, obs_strike_count, pred_p_ignite, sim_obs_p_ignite, NA, pred_fire_count, NA, obs_fire_count, pred_burned_area, sim_obs_burned_area, obs_burned_area), ncol=3, byrow=TRUE))

colnames(summary_table) <- c("predicted", "sim_from_obs", "observed")

summary_table <- cbind(model_outpit = c("strike_rate", "strike_count", "p_ignite", "fire_count", "total_burned_area"), summary_table)

summary_table_fmt <- summary_table |>
  mutate(across(where(is.numeric), ~ formatC(., format = "f", digits = 4, big.mark = ",")))

print(summary_table_fmt)

```

Export Drivers and Predictions
```{r}
# df_out <- fire |> dplyr::select(lat,lon,year,month,
#                                 rh,tair,precip,wind,precip_1m,precip_2m,precip_3m,precip_5y,tair_1m,tair_2m,tair_3m,tair_5y,C,B,U, fire_count, burned_area_km2, lightning_ignition, dist_road, dist_river, dist_urban, 
#                                 total_ignition_count, single_lightning_ignition, burned_area_km2_total, p_ignite, p_ignite_rate_cal, r_strike_obs, n_strikes_obs, r_strike, n_strikes, fire_count_pred, p_ignite_test, fire_count_test, gamma_shape, gamma_rate, total_burned_area, total_burned_area_test)

df_out <- fire |> dplyr::select(lat,lon,year,month,
                                r_strike_obs, n_strikes_obs, burned_area_km2, lightning_ignition, total_ignition_count, fire_count, single_lightning_ignition, burned_area_km2_total, 
                                r_strike, n_strikes, p_ignite, p_ignite_rate_cal, fire_count_pred, gamma_shape, gamma_rate, total_burned_area)


write.csv(df_out, file.path(predictions_root, "predictions_2002_2011.csv"))


```








July 3rd
- check gamma sampling for burned area. - is it using fire count as size correctly?
- same for fire count sampling using p_ignite and strike_count. 
- Why does sampling from poisson not work?? what does this plot tell us??
```{r}
plot(fire$strike_count_pred, fire$r_strike)
plot(fire$strike_count, fire$strikes)
plot(fire$strike_count_test, fire$strikes)
plot(fire$strike_count, fire$strike_count_test)

hist(fire$r_strike)
hist(fire$strikes)

hist(fire$strike_count_pred)
hist(fire$strike_count_test)
hist(fire$strike_count)

```



JULY 1

LIGHTNING 

mean strike rates are similar between observed and predicted. Issue arises when counts are sampled from the poisson distribution. Could it be an issue with tails - moments when strike rates are exceptionally high are not being captured by the gamma GLM - could the gamma Bayesian approach handle this better?

IGNITION
Predictions are too high, by an order of magnitude! 

FIRE
Best fire count: simulated from fire count (strikes * p_ignite) and conditioned on p_fire (rbinom(p_fire)==1)
However, multiplying the mean observed strike count (8.989618 strikes per gridcell) by mean observed ignition rate (0.00023 ignitions/strikes) gets very close to the observed fire count (273, whereas actual fire count is 265)

BURNED AREA
Best burned area: simulated from fire count (strikes * P_ignite)

Keep:
strike count in a gridcell * p_ignite 

- or try sampling form binomial distribution: 
n_fires = rbinom(strike_count, ignition_efficiency)
then:
total burned area = sum(rgamma(n_fires, shape = gamma_shape, rate  = gamma_shape / gamma_mean))

gamma_shape and gamma_mean are estimated by the burned area model. 

What about strike rate? Is the model picking up tails

Since the lightning model doesn't capture the tails, strike counts sampled from poisson have very small tail. 
The model captures the mean well, but not gridcell/months when climate results in extreme lightning events. Max predicted fire count is > 60, while observed max count is > 20,000.




