---
title: "Fire Model Pipeline"
author: "Charlotte Uden"
date: "2026-06-14"
output: html_document
---


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(pROC)
library(loo)
library(cowplot)
library(stats)
library(caret)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")
config <- read_properties(config_file_path)

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

##########################################
#####       LIGHTNING    (r_strike)  #####
##########################################

r_strike ~ Gamma(alpha, beta)

alpha = exp(a_alpha + b_alpha * X)

beta = exp(a_beta + b_beta * X)

a_alpha, b_alpha, a_beta, b_beta ~ Normal(0, 1)

r_strike: lightning strikes / km2 / month
X: temperature, precipitation, relative humidity, shortwave radiation, surface pressure

exp(): ensures alpha and beta stay positive, which the gamma distribution requires. 

```{r}
lightning <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))
head(lightning)
hist(lightning$strikes)
```


Predict lightning strike rate for all gridcells and years
```{r}
#standardize:
lightning_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) lightning_mu_sigma$mu[lightning_mu_sigma$var == var]
get_sigma <- function(var) lightning_mu_sigma$sigma[lightning_mu_sigma$var == var]

lightning_scaled <- lightning |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    swr    = (swr    - get_mu("swr"))    / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh")
  )

#read in model
#lightning_model <- readRDS(file.path(rds_root, "G13_gamma_glm_lightning.rds"))

#make predictions
# lightning_scaled <- lightning_scaled |>
#   mutate(
#     strikes_pred = predict(lightning_model,
#                            newdata = lightning_scaled,
#                            type    = "response")
#   )

# ── Load B12 Bayesian lightning model ─────────────────────────────────────────
B12 <- readRDS(file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))

# ── Extract posterior samples ─────────────────────────────────────────────────
posterior_samples <- as.data.frame(B12)

# Generate predictions using posterior mean coefficients
# B12 excludes wind — predictors: swr, tair, rh, precip, sp

a_alpha_mean <- mean(posterior_samples$a_alpha)
b_alpha_mean <- mean(posterior_samples$b_alpha)
c_alpha_mean <- mean(posterior_samples$c_alpha)
d_alpha_mean <- mean(posterior_samples$d_alpha)
e_alpha_mean <- mean(posterior_samples$e_alpha)
f_alpha_mean <- mean(posterior_samples$f_alpha)

a_beta_mean  <- mean(posterior_samples$a_beta)
b_beta_mean  <- mean(posterior_samples$b_beta)
c_beta_mean  <- mean(posterior_samples$c_beta)
d_beta_mean  <- mean(posterior_samples$d_beta)
e_beta_mean  <- mean(posterior_samples$e_beta)
f_beta_mean  <- mean(posterior_samples$f_beta)

# Compute alpha and beta per observation using posterior means
# Using vectorised operations rather than a loop for speed

lp_alpha <- a_alpha_mean +
            b_alpha_mean * lightning_scaled$swr   +
            c_alpha_mean * lightning_scaled$tair  +
            d_alpha_mean * lightning_scaled$rh    +
            e_alpha_mean * lightning_scaled$precip +
            f_alpha_mean * lightning_scaled$sp

lp_beta  <- a_beta_mean  +
            b_beta_mean  * lightning_scaled$swr   +
            c_beta_mean  * lightning_scaled$tair  +
            d_beta_mean  * lightning_scaled$rh    +
            e_beta_mean  * lightning_scaled$precip +
            f_beta_mean  * lightning_scaled$sp

alpha_pred <- exp(lp_alpha)
beta_pred  <- exp(lp_beta)

set.seed(123)

# Gamma alpha and beta
# if you want the mean: alpha / beta
lightning_scaled <- lightning_scaled |>
  mutate(
    gamma_alpha   = alpha_pred,
    gamma_beta    = beta_pred,
    r_strike_mean  = alpha_pred / beta_pred,   # posterior mean prediction
    r_strike = rgamma(n(), shape = gamma_alpha, rate = gamma_beta)
  )

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Predicted strike rate summary:\n")
print(summary(lightning_scaled$r_strike))
cat("Observed strike rate summary:\n")
print(summary(lightning_scaled$strikes))

lightning_scaled <- lightning_scaled[, c("lat", "lon", "year", "strikes", "r_strike", "r_strike_mean")]
```


##########################################
##### STRIKE COUNT       (n_strikes) #####
##########################################

n_strikes ~ NegBinomial(mu, theta)

mu = r_strike * area

theta = exp(gamma_0 + gamma * X)

Gamma_0 ~ Normal(2,2)
Gamma ~ Normal(0, sigma2)
Sigma = 1 

r_strike: strikes/km2/month
area: gridcell area, km2
n_strikes: number of strikes in a gridcell-month
X: monthly climate predictors, temperature, precipitation, wind, relative humidity, shortwave radiation, surface pressure
theta: Precision/dispersion parameter of the Negative Binomial distribution predicted from climate. Theta > 0.
mu: mean lightning strike rate, strikes/gridcell/month. This is predicted from the Gamma model above and is the same for all summer months (June, July, August) in a year. The summer mean was used to ensure that mu > 0, which is required by the negative binomial distribution. 


```{r}
# ── Apply negative binomial model to fire dataframe ────────────────────────────────────────────────────

#read in drivers
df <- read.csv(file.path(drivers_root, "lightning_strike_count_climate_jja_2002_2011.csv"))

#merge summer strike rate (strikes/km2/month) predictions, r_strike - months will be duplicated
df <- df %>%
  left_join(
    lightning_scaled %>% dplyr::select(lat, lon, year, strikes, r_strike, r_strike_mean),
    by = c("lat", "lon", "year")
  )

df <- na.omit(df)

names(df)[names(df) == "strikes"] <- "r_strike_obs"
names(df)[names(df) == "strike_count"] <- "n_strikes_obs"


#Scale strike rate from strikes/km2/month to strikes/gridcell/month
area <- read.csv("~/raid/cuden/data/paleo_fire/era5_grid_area.csv")

df <- df %>%
  left_join(
    area %>% dplyr::select(lat, lon, area),
    by = c("lat", "lon")
  )

#standardize:
n_strike_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_count_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) n_strike_mu_sigma$mu[n_strike_mu_sigma$var == var]
get_sigma <- function(var) n_strike_mu_sigma$sigma[n_strike_mu_sigma$var == var]

n_strike_scaled <- df |>
  dplyr::select(lat, lon, year, month, r_strike_obs, n_strikes_obs, r_strike, area, precip, tair, wind, rh, swr, sp) |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh"),
    swr     = (swr     - get_mu("swr"))     / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp")
  )

#caluclate mu for the negative binomial distribution: r_strike * area = strikes/month/gridcell
n_strike_scaled <- n_strike_scaled |>
  mutate(mu_r_strike_gridcell = r_strike * area)

# n_strike_scaled <- n_strike_scaled |>
#   mutate(mu_r_strike_gridcell = r_strike_mean * area)

#Predictors for theta (negative binomial parameter we need to sample strike count)
theta_predictors <- c("swr", "tair", "rh", "precip", "sp", "wind")

#read in model
nb_model <- readRDS(file.path(rds_root, "Lightning/neg_bin_strike_count_small.rds"))

#Posterior mean gamma coefficients
gamma_0_mean <- nb_model$coef_means$gamma_0
gamma_means  <- nb_model$coef_means$gamma

#Build predictor matrix for full fire dataframe
Z_fire <- model.matrix(
  as.formula(paste("~", paste(theta_predictors, collapse = " + "), "- 1")),
  data = n_strike_scaled   # must have same standardised predictors
)

#Compute posterior mean theta per gridcell-month
log_theta_pred <- as.numeric(gamma_0_mean + Z_fire %*% gamma_means)
theta_pred     <- exp(log_theta_pred)

#sample strike count using mu=r_strike*area and theta
n_strike_scaled <- n_strike_scaled |>
  mutate(
    nb_theta         = theta_pred,
    log_mu_pred      = log(pmax(mu_r_strike_gridcell, 1e-8)),
    n_strikes = rnbinom(n    = n(),
                                 mu   = mu_r_strike_gridcell,
                                 size = nb_theta)
  )

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Observed zeros:    ", mean(n_strike_scaled$n_strikes_obs == 0), "\n")
cat("Simulated zeros:   ", mean(n_strike_scaled$n_strikes == 0), "\n")
cat("Observed mean:     ", mean(n_strike_scaled$n_strikes_obs), "\n")
cat("Simulated mean:    ", mean(n_strike_scaled$n_strikes), "\n")
cat("Observed count:     ", sum(n_strike_scaled$n_strikes_obs), "\n")
cat("Simulated count:    ", sum(n_strike_scaled$n_strikes), "\n")
cat("Observed variance: ", var(n_strike_scaled$n_strikes_obs), "\n")
cat("Simulated variance:", var(n_strike_scaled$n_strikes), "\n")
summary(n_strike_scaled$n_strikes)
summary(n_strike_scaled$n_strikes_obs)
hist(n_strike_scaled$n_strikes)
hist(n_strike_scaled$n_strikes_obs)


```
Simulated counts come close to observations, capturing the zero outcomes well (zero strike make up 55% of observed outcomes and 64% of simulated outcomes). This issue comes with the model's ability to capture variance in strike counts (simulated var = 453, observed var = 22580). This is driving both a reduction in the simulated mean (8.99, while simulated is 4.94) and total count (1,192,293, while simulated is 654,492). 

It this point, I feel pretty good with the lightning models. 


Read in ingition efficiency and burned area predictors
```{r}
fire <- read.csv(file.path(drivers_root, "lightning_fires_burned_area_jja_2002_2011.csv"))

#new training data from NIFC. Need to replace observed fire count and burned area
ignition_new <- read.csv(file.path(drivers_root, "lightning_ignition_jja_2002_2011_new.csv"))[,2:26]
ignition_new <- ignition_new[,c("lat","lon","year","month", "fire_count", "strike_count")]


 #during training, rows where fire count exceeded strike count were removed - reoeat here:
ignition_new  <- ignition_new %>% filter(!(strike_count > 0 & fire_count > strike_count))

#replace old fire count data with new
fire <- fire %>%
  dplyr::select(-fire_count) %>%
  left_join(ignition_new %>% dplyr::select(lat, lon, year, month, fire_count),
            by = c("lat", "lon", "year", "month")) %>%
  mutate(fire_count = ifelse(is.na(fire_count), 0, fire_count)) #some fire counts removed above - replace with 0

#replace old burned area with new:
ba_new <- read.csv(file.path(drivers_root, "burned_area_jja_2002_2011_new.csv"))

fire <- fire %>%
  dplyr::select(-burned_area_km2, -lightning_ignition, -grid_id, 
                -single_lightning_ignition, -total_ignition_count, -burned_area_km2_total) %>%
  left_join(ba_new %>% dplyr::select(lat, lon, year, month, burned_area_km2_total),
            by = c("lat", "lon", "year", "month")) %>%
  mutate(burned_area_km2_total = ifelse(is.na(burned_area_km2_total), 0, burned_area_km2_total))

rm(ignition_new)
rm(ba_new)

fire <- fire |>
  mutate(lat = round(lat, 2), lon = round(lon, 2)) |>
  distinct(lat, lon, year, month, .keep_all = TRUE)
```

##########################################
##### IGNITION EFFICINCY (p_ignite) #####
##########################################

Y ~ Binomial(n, p_ignite) --> technically, this is applied in the next step, to estimate fire count.

logit(p_ignite) = alpha + X * beta

alpha ~ normal(mu_alpha, sigma_alpha)
beta  ~ normal(0, sigma_beta)

p_ignite_corrected = plogis(qlogis(pie) + cal_shift)

p_ignite: probability of 1 strike resulting in an ignition in a given gridcell and month. 
X: average precipitation rate, temperature, relative humidity, cumulative precipitation for the previous 1, 2, 3 months, and 5 years, mean temperature for the previous 1, 2, 3 months, and 5 years, and % broadleaf cover. 


```{r}
#standardize:
ignition_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_ignition_predictors_redo.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) ignition_mu_sigma$mu[ignition_mu_sigma$var == var]
get_sigma <- function(var) ignition_mu_sigma$sigma[ignition_mu_sigma$var == var]

ignition_scaled <- fire |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_2m   = (precip_2m   - get_mu("precip_2m"))   / get_sigma("precip_2m"),
    precip_3m   = (precip_3m   - get_mu("precip_3m"))   / get_sigma("precip_3m"),
    precip_5y     = (precip_5y     - get_mu("precip_5y"))     / get_sigma("precip_5y"),
    tair_1m = (tair_1m - get_mu("tair_1m")) / get_sigma("tair_1m"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_3m   = (tair_3m   - get_mu("tair_3m"))   / get_sigma("tair_3m"),
    tair_5y     = (tair_5y     - get_mu("tair_5y"))     / get_sigma("tair_5y"),
    B     = (B     - get_mu("B"))     / get_sigma("B"),
    C     = (C     - get_mu("C"))     / get_sigma("C"),
    U     = (U     - get_mu("U"))     / get_sigma("U")
  )
```

```{r}
#Bayes recalibrated
# ── Load calibrated Bayesian model ────────────────────────────────────────────
# ignition_bayes_obj <- readRDS(file.path(rds_root, 
#                                          "Ignition_Efficiency/m2_bayes_calibrated.rds"))

ignition_bayes_obj <- readRDS(file.path(rds_root, 
                                         "Ignition_Efficiency/redo/m9_bayes_calibrated.rds"))

ignition_bayes_fit <- ignition_bayes_obj$fit        # stanfit object
cal_shift_bayes    <- ignition_bayes_obj$cal_shift  # intercept shift
coef_means         <- ignition_bayes_obj$coef_means # posterior mean alpha and beta

cat("Bayesian calibration shift:", round(cal_shift_bayes, 3), "\n")

# ── Build predictor matrix from standardised data ─────────────────────────────
predictors_m9 <- c("rh", "tair", "precip", "precip_1m", "precip_2m", "precip_3m", "tair_2m", "tair_3m", "tair_5y", "C", "U")

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m9, collapse = " + "), "- 1")),
  data = ignition_scaled
)

# ── Generate predictions using posterior mean ─────────────────────────────────
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

ignition_scaled <- ignition_scaled |>
  mutate(
    #inverse_logit of linear predictor, linpred
    p_ignite_bayes     = plogis(linpred),
    #add calibration shift 
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,     #pmin and pmax make sure arguments in plogis from reaching infinity
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Raw Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes_cal))

ignition_scaled <- ignition_scaled[, c("lat", "lon", "year", "month",
                          "fire_count", "strike_count",
                          "p_ignite_bayes_cal")]

ignition_scaled <- ignition_scaled %>% dplyr::rename(p_ignite = p_ignite_bayes_cal)

#add monthly ignition probability, p_ignite, to fire df
fire <- fire %>%
  left_join(
    ignition_scaled %>% dplyr::select(lat, lon, year, month, p_ignite),
    by = c("lat", "lon", "year", "month"))

```


##### RATE CALIBRATION SHIFT FOR P_IGNITE ##### 

```{r}

rate_cal_shift <- readRDS(file.path(rds_root, "Ignition_Efficiency/redo/calibration_shifts.rds"))$rate_cal_shift

eps <- 1e-8

fire <- fire |>
  mutate(
    p_ignite_rate_cal = plogis(
      qlogis(pmin(pmax(p_ignite, eps), 1 - eps)) + rate_cal_shift
    )
  )

# Do annual rates now match observed?
annual_ignition_rates <- fire |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate           = sum(fire_count) / sum(strike_count),
    #pred_binary_cal    = mean(p_ignite),
    pred_rate_cal      = mean(p_ignite_rate_cal),
    .groups            = "drop"
  )
long_df <- annual_ignition_rates %>%
  pivot_longer(
    cols = c("obs_rate","pred_rate_cal"), # Specify columns to reshape
    names_to = "dataset", # New column for original column names
    values_to = "ignition_rates" # New column for values
)
mean_obs_rate <- mean(annual_ignition_rates$obs_rate)
ggplot(long_df, aes(x=year, y=ignition_rates, color=dataset)) + geom_point() + geom_hline(yintercept=mean_obs_rate)

#loook at expected fire count
cat("Expected fires (binary cal):  ",
    sum(fire$strike_count * fire$p_ignite, na.rm = TRUE), "\n")
cat("Expected fires (rate cal):    ",
    sum(fire$strike_count * fire$p_ignite_rate_cal,  na.rm = TRUE), "\n")
cat("Observed fires:               ",
    sum(fire$fire_count), "\n")
```


##########################################
#####     BURNED AREA | 1 ignition  #####
##########################################



MIXTURE MODEL
```{r}
#standardize:
burned_area_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_burned_area_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) burned_area_mu_sigma$mu[burned_area_mu_sigma$var == var]
get_sigma <- function(var) burned_area_mu_sigma$sigma[burned_area_mu_sigma$var == var]

burned_area_scaled <- fire |>
  mutate(
    rh        = (rh        - get_mu("rh"))        / get_sigma("rh"),
    tair      = (tair      - get_mu("tair"))      / get_sigma("tair"),
    precip    = (precip    - get_mu("precip"))    / get_sigma("precip"),
    wind      = (wind      - get_mu("wind"))      / get_sigma("wind"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_3m = (precip_3m - get_mu("precip_3m")) / get_sigma("precip_3m"),
    precip_5y = (precip_5y - get_mu("precip_5y")) / get_sigma("precip_5y"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_5y   = (tair_5y   - get_mu("tair_5y"))   / get_sigma("tair_5y"),
    C     = (C     - get_mu("C"))     / get_sigma("C"),
    U     = (U     - get_mu("U"))     / get_sigma("U")
  ) #no scaling for binary urban and road data.

#read in binary distance data:
proximity <- read.csv(file.path(proximity_data_root, "era5_proximity_road_river_urban.csv"))
proximity <- dplyr::select(proximity, lon, lat, road_proximity_binary, urban_proximity_binary)
#check:
# ggplot() +
#   geom_point(data = proximity |> distinct(lat, lon),
#              aes(x = lon, y = lat),
#              color = "grey70", size = 0.8, alpha = 0.5) 

# Round to 2 decimal places in both dataframes
burned_area_scaled <- burned_area_scaled |>
  mutate(lat = round(lat, 2), lon = round(lon, 2))

proximity <- proximity |>
  mutate(lat = round(lat, 2), lon = round(lon, 2))

#merge by lat lon
burned_area_scaled <- burned_area_scaled |>
  left_join(
    proximity |> dplyr::select(lat, lon, 
                                road_proximity_binary, 
                                urban_proximity_binary),
    by = c("lat", "lon")
  )


# ── Predictor sets, one per sub-model (from mixture_gamma_custom fit) ──

preds_classifier <- c("urban_proximity_binary", "rh", "tair", "precip", "tair_5y")
preds_extreme     <- c("U", "tair")
preds_typical     <- c("urban_proximity_binary", "tair_2m", "tair")

# ── Load the fitted mixture model ────────────────────────────────────────────

mixture_obj <- readRDS(file.path(rds_root, "Burned_Area/redo/m3_bayes_mixture_custom.rds"))

fit_logit   <- mixture_obj$fit_logit
fit_extreme <- mixture_obj$fit_extreme
fit_normal  <- mixture_obj$fit_normal

# ── Posterior means for each component (analogous to old alpha_mean/beta_means) ──

summ_logit   <- rstan::summary(fit_logit,   pars = c("alpha", "beta"))$summary
summ_extreme <- rstan::summary(fit_extreme, pars = c("alpha", "beta", "phi"))$summary
summ_normal  <- rstan::summary(fit_normal,  pars = c("alpha", "beta", "phi"))$summary

alpha_logit_mean <- summ_logit["alpha", "mean"]
beta_logit_mean  <- summ_logit[grep("^beta\\[", rownames(summ_logit)), "mean"]

alpha_extreme_mean <- summ_extreme["alpha", "mean"]
beta_extreme_mean  <- summ_extreme[grep("^beta\\[", rownames(summ_extreme)), "mean"]
phi_extreme_mean   <- summ_extreme["phi", "mean"]

alpha_typical_mean <- summ_normal["alpha", "mean"]
beta_typical_mean  <- summ_normal[grep("^beta\\[", rownames(summ_normal)), "mean"]
phi_typical_mean   <- summ_normal["phi", "mean"]

cat("Classifier intercept (alpha):", round(alpha_logit_mean, 3), "\n")
cat("Extreme-Gamma shape (phi):   ", round(phi_extreme_mean, 3), "\n")
cat("Typical-Gamma shape (phi):   ", round(phi_typical_mean, 3), "\n")

# ── Build predictor matrices — three separate ones, one per sub-model ───────

X_classifier <- model.matrix(
  as.formula(paste("~", paste(preds_classifier, collapse = " + "), "- 1")),
  data = burned_area_scaled
)
X_extreme <- model.matrix(
  as.formula(paste("~", paste(preds_extreme, collapse = " + "), "- 1")),
  data = burned_area_scaled
)
X_typical <- model.matrix(
  as.formula(paste("~", paste(preds_typical, collapse = " + "), "- 1")),
  data = burned_area_scaled
)

# ── Generate predictions (posterior mean, point-estimate — same approach
# as "Option A" in your original code) ───────────────────────────────────────

p_extreme_linpred <- as.numeric(alpha_logit_mean + X_classifier %*% beta_logit_mean)
p_extreme         <- plogis(p_extreme_linpred)

mu_extreme <- as.numeric(exp(alpha_extreme_mean + X_extreme %*% beta_extreme_mean))
mu_typical <- as.numeric(exp(alpha_typical_mean + X_typical %*% beta_typical_mean))

# The blended MEAN (for reporting / a point prediction) is a valid weighted average -
# this is fine to report as "expected burned area," but is NOT itself a Gamma mean,
# so it should NOT be paired with a single shape parameter for sampling purposes.
burned_area_pred <- p_extreme * mu_extreme + (1 - p_extreme) * mu_typical

burned_area_scaled <- burned_area_scaled |>
  mutate(
    p_extreme        = p_extreme,
    mu_extreme       = mu_extreme,
    mu_typical       = mu_typical,
    phi_extreme      = phi_extreme_mean,   # constant across rows, like old phi_mean
    phi_typical      = phi_typical_mean,   # constant across rows
    burned_area_pred = burned_area_pred    # blended point estimate, for reference/plots only
  )

# rate = shape / mean, computed PER COMPONENT (not a single blended rate)
burned_area_scaled <- burned_area_scaled |>
  mutate(
    rate_extreme = phi_extreme / mu_extreme,
    rate_typical = phi_typical / mu_typical
  )
```

Abandoned year random effect model:

Burned_area ~ Gamma(theta, theta/mu)

log(mu) = alpha + alpha_year + X * beta

alpha ~ Normal(0, sigma^2_alpha)
beta ~ Normal(0, sigma^2_beta)
theta ~ Exponential(1)
alpha_year ~ Normal(0, sigma^2_year)
sigma^2_year ~ Exponential(1)


Burned_area: area of 1 lightning-caused ignition (km2)
X: Relative humidity, temperature, precipitation, wind, cumulative precipitation for the previous 1, 3 months, and 5 years, mean temperature for the previous 2 months and 5 years, proximity to urban area (binary; 1 = within 40 km of an urban area), proximity to road (binary; 1 = within 40 km of an major road)
alpha_year: year random affect
```{r}
# preds_final <- c("rh", "tair", "precip", "wind",
#                   "precip_1m", "precip_3m", "precip_5y",
#                   "tair_2m", "tair_5y", 
#                   "C", "U", 
#                   "urban_proximity_binary")
# 
# 
# 
# # ── Load model and extract posterior parameters ───────────────────────────────
# m1_obj     <- readRDS(file.path(rds_root, "Burned_Area/redo/m1_bayes_gamma_re.rds"))
# m1_fit     <- m1_obj$fit           # stanfit object
# coef_means <- m1_obj$coef_means    # list: alpha, beta, alpha_year
# 
# alpha_mean <- coef_means$alpha
# beta_means <- coef_means$beta
# 
# cat("Global intercept (alpha):", round(alpha_mean, 3), "\n")
# cat("Year random effects:\n")
# print(round(coef_means$alpha_year, 3))
# 
# # ── Build predictor matrix ────────────────────────────────────────────────────
# X_new <- model.matrix(
#   as.formula(paste("~", paste(preds_final, collapse = " + "), "- 1")),
#   data = burned_area_scaled
# )
# 
# # ── Generate predictions ──────────────────────────────────────────────────────
# # Year random effects set to ZERO for new data (paleo or ERA5 application)
# # This represents "climatologically average year" — variation comes from
# # climate predictors only, not year-specific synoptic noise
# 
# # ── Option A: posterior mean prediction (fast, point estimate) ────────────────
# linpred <- as.numeric(alpha_mean + X_new %*% beta_means)
# # alpha_year term is omitted — equivalent to setting it to 0
# 
# burned_area_scaled <- burned_area_scaled |>
#   mutate(burned_area_pred = exp(linpred))
# 
# # Extract posterior draws
# post <- rstan::extract(m1_fit)
# 
# # ── Extract shape parameter (phi) from posterior ───────────────────────────────
# phi_mean <- mean(post$phi)
# phi_sd   <- sd(post$phi)
# 
# cat("Gamma shape (phi) posterior mean:", round(phi_mean, 3), "\n")
# cat("Gamma shape (phi) posterior sd:  ", round(phi_sd, 3), "\n")
# 
# # ── Add gamma_mean and gamma_shape columns to burned_area_scaled ──────────────
# burned_area_scaled <- burned_area_scaled |>
#   mutate(
#     gamma_mean  = burned_area_pred,   # mean (mu) per gridcell-month
#     gamma_shape = phi_mean             # shape (phi) — constant across rows
#   )
# 
# # rgamma requires rate, not the mean. calculate this as:
# # rgamma(n, shape, rate) where rate = shape / mean
# burned_area_scaled$gamma_rate <- burned_area_scaled$gamma_shape / burned_area_scaled$gamma_mean

```

##########################################
#####         STRIKE COUNT         #######
##########################################

strike_count = Poisson(r_strike * area)

```{r}
# #merge summer strike rate (strikes/km2/month) predictions - months will be duplicated
# fire <- fire %>%
#   left_join(
#     lightning_scaled %>% dplyr::select(lat, lon, year, strikes, r_strike),
#     by = c("lat", "lon", "year")
#   )
# 
# #Scale strike rate from strikes/km2/month to strikes/gridcell/month
# area <- read.csv("~/raid/cuden/data/paleo_fire/era5_grid_area.csv")
# 
# fire <- fire %>%
#   left_join(
#     area %>% dplyr::select(lat, lon, area),
#     by = c("lat", "lon")
#   )
# 
# #calculate grudcell-level striike rate (strikes/gridcell/month) and use as lambda in poisson distrubution to get strike count
# set.seed(123)
# #simulated count from predicted rate
# fire$strike_count_pred <- sapply(fire$r_strike * fire$area, function(lam) rpois(1, lambda = lam))
# #simulated count from observed rate
# fire$strike_count_test <- sapply(fire$strikes * fire$area, function(lam) rpois(1, lambda = lam))

```

##########################################
#####         FIRE COUNT         #########
##########################################

fire_count | lightning = Binomial(n_strikes, p_ignite)

```{r}
# 
#add monthly strike count, n_strikes, to fire df
fire <- fire %>%
  left_join(
    n_strike_scaled %>% dplyr::select(lat, lon, year, month, r_strike_obs, n_strikes_obs, r_strike, n_strikes, mu_r_strike_gridcell),
    by = c("lat", "lon", "year", "month"))


#Use strike_count and p_ingite to sample fire count from a binomial distribution
fire$fire_count_pred <- mapply(
  function(n, p) rbinom(1, size = n, prob = p),
  fire$n_strikes,
  fire$p_ignite_rate_cal
)

#OR (test) multiply dridcell strike rate by p_ignite
fire <- fire  %>% mutate(fire_count_gridcell_rstrike = mu_r_strike_gridcell*p_ignite_rate_cal) 

# #Use OBSERVED strike_count and ignition rate (fires/strikes);  to sample fire count from a binomial distribution
# 
# #caluclate ignintion frequency from obs fires/strikes, if strieks == 0, p_ignite_test==0. 
# fire$p_ignite_test <- mapply(
#   function(n_fires, n_strikes) {
#     if (is.na(n_fires) || is.na(n_strikes) || n_strikes == 0) return(0L)
#     n_fires/n_strikes
#   },
#   fire$fire_count,
#   fire$n_strikes_obs
# )
#   
# #simulate fire count
# fire$fire_count_test <- mapply(
#   function(n, p) {
#     if (is.na(n) || is.na(p) || n == 0) return(0L)
#     p_safe <- pmin(pmax(p, 0), 1)   # clip to [0,1] as safety guard
#     rbinom(1, size = n, prob = p_safe)
#   },
#   fire$n_strikes_obs,
#   fire$p_ignite_test
# )
#   
```

#####################################
######### TOTAL BURNED AREA #########
#####################################

total_burned = rgamma(fire_count, burned_area_shape, burned_area_rate)

```{r}
# ── Derive a calibration shift for p_extreme, analogous to rate_cal_shift ────
# Goal: align p_extreme's distribution among SIMULATED fire-months with what
# it should be, using the same logit-shift approach already used for p_ignite

burned_area_scaled <- burned_area_scaled %>%
  left_join(
    fire %>% dplyr::select(lat, lon, year, month, fire_count_pred),
    by = c("lat", "lon", "year", "month")
  )

# Does p_ignite correlate more strongly with p_extreme-favoring conditions
# in SIMULATED fires than it should, based on real fires?
#cor(fire$p_ignite_rate_cal[fire$fire_count_pred > 0], fire$p_extreme[fire$fire_count_pred > 0])
#cor(fire$p_ignite_rate_cal[fire$burned_area_km2_total > 0], fire$p_extreme[fire$burned_area_km2_total > 0])

# Reference: p_extreme among REAL historical fires (what the classifier was trained on)
p_extreme_real_fires <- burned_area_scaled$p_extreme[burned_area_scaled$burned_area_km2_total > 0]

# Current: p_extreme among SIMULATED fires (what the pipeline actually samples)
p_extreme_simulated <- burned_area_scaled$p_extreme[burned_area_scaled$fire_count_pred > 0]

cat("Real-fire p_extreme mean:", mean(p_extreme_real_fires, na.rm = TRUE), "\n")
cat("Simulated-fire p_extreme mean:", mean(p_extreme_simulated, na.rm = TRUE), "\n")

# Fit the shift as the difference in mean LOGIT-space p_extreme between the two populations
eps <- 1e-8
mean_logit_real      <- mean(qlogis(pmin(pmax(p_extreme_real_fires, eps), 1 - eps)), na.rm = TRUE)
mean_logit_simulated <- mean(qlogis(pmin(pmax(p_extreme_simulated, eps), 1 - eps)), na.rm = TRUE)

extreme_cal_shift <- mean_logit_real - mean_logit_simulated
cat("Extreme calibration shift:", round(extreme_cal_shift, 3), "\n")

# ── Apply the shift to p_extreme wherever it's used for sampling ────────────
burned_area_scaled <- burned_area_scaled %>%
  mutate(
    p_extreme_cal = plogis(pmin(pmax(qlogis(pmin(pmax(p_extreme, eps), 1 - eps)) + extreme_cal_shift,
                                       qlogis(eps)), qlogis(1 - eps)))
  )
```
Save calibration parameters for p_extreme
```{r}
# ── Export the p_extreme calibration shift for reuse on calibration/paleo data ──

extreme_cal_params <- list(
  extreme_cal_shift    = extreme_cal_shift,
  mean_logit_real      = mean_logit_real,
  mean_logit_simulated = mean_logit_simulated,
  n_real_fires         = length(p_extreme_real_fires),
  n_simulated_fires    = sum(!is.na(p_extreme_simulated)),
  eps                  = eps,
  fitted_on            = "ERA5 training period, 2002-2011",
  fitted_date          = Sys.Date()
)

cat("Exporting extreme calibration parameters:\n")
print(extreme_cal_params)

saveRDS(
  extreme_cal_params,
  file.path(rds_root, "Burned_Area/extreme_cal_shift.rds")
)
```

```{r}
# Sample burned area

#add monthly burned area gamma parameters to fire df
fire <- fire %>%
  left_join(
    burned_area_scaled %>% dplyr::select(lat, lon, year, month, p_extreme, p_extreme_cal,
                                          mu_extreme, mu_typical, rate_extreme, rate_typical,
                                          phi_extreme, phi_typical),
    by = c("lat", "lon", "year", "month")
  )

n_rows <- nrow(fire)
fc <- fire$fire_count_pred

# ── Sampling ──────────────────────────────────────────────────────────────────
# Uses p_extreme_cal for classification, but the ORIGINAL (uncalibrated)
# rate_extreme/rate_typical for magnitude - no post-hoc magnitude shift

valid_idx <- which(!is.na(fc) & fc > 0)
rep_idx   <- rep(valid_idx, times = fc[valid_idx])

rep_p_extreme <- rep(fire$p_extreme_cal[valid_idx], times = fc[valid_idx])
rep_rate_ext  <- rep(fire$rate_extreme[valid_idx], times = fc[valid_idx])
rep_rate_typ  <- rep(fire$rate_typical[valid_idx], times = fc[valid_idx])

is_extreme_draw <- rbinom(length(rep_idx), size = 1, prob = rep_p_extreme)

samples <- ifelse(
  is_extreme_draw == 1,
  rgamma(length(rep_idx), shape = fire$phi_extreme[1], rate = rep_rate_ext),
  rgamma(length(rep_idx), shape = fire$phi_typical[1], rate = rep_rate_typ)
)

row_sums <- tapply(samples, rep_idx, sum)
total_burned_area <- rep(0, n_rows)
total_burned_area[as.integer(names(row_sums))] <- row_sums
fire$total_burned_area <- total_burned_area
```


```{r}
# # ── Derive a calibration shift for p_extreme, analogous to rate_cal_shift ────
# # Goal: align p_extreme's distribution among SIMULATED fire-months with what
# # it should be, using the same logit-shift approach already used for p_ignite
# 
# burned_area_scaled <- burned_area_scaled %>%
#   left_join(
#     fire %>% dplyr::select(lat, lon, year, month, fire_count_pred),
#     by = c("lat", "lon", "year", "month")
#   )
# 
# # Reference: p_extreme among REAL historical fires (what the classifier was trained on)
# p_extreme_real_fires <- burned_area_scaled$p_extreme[burned_area_scaled$burned_area_km2_total > 0]
# 
# # Current: p_extreme among SIMULATED fires (what the pipeline actually samples)
# p_extreme_simulated <- burned_area_scaled$p_extreme[burned_area_scaled$fire_count_pred > 0]
# 
# cat("Real-fire p_extreme mean:", mean(p_extreme_real_fires, na.rm = TRUE), "\n")
# cat("Simulated-fire p_extreme mean:", mean(p_extreme_simulated, na.rm = TRUE), "\n")
# 
# # Fit the shift as the difference in mean LOGIT-space p_extreme between the two populations
# eps <- 1e-8
# mean_logit_real      <- mean(qlogis(pmin(pmax(p_extreme_real_fires, eps), 1 - eps)), na.rm = TRUE)
# mean_logit_simulated <- mean(qlogis(pmin(pmax(p_extreme_simulated, eps), 1 - eps)), na.rm = TRUE)
# 
# extreme_cal_shift <- mean_logit_real - mean_logit_simulated
# cat("Extreme calibration shift:", round(extreme_cal_shift, 3), "\n")
# 
# # ── Apply the shift to p_extreme wherever it's used for sampling ────────────
# burned_area_scaled <- burned_area_scaled %>%
#   mutate(
#     p_extreme_cal = plogis(pmin(pmax(qlogis(pmin(pmax(p_extreme, eps), 1 - eps)) + extreme_cal_shift,
#                                        qlogis(eps)), qlogis(1 - eps)))
#   )

```

Sample burned area
```{r}
# #add monthly burned area gamma parameters to fire df
# fire <- fire %>%
#   left_join(
#     burned_area_scaled %>% dplyr::select(lat, lon, year, month, p_extreme, p_extreme_cal,
#                                           mu_extreme, mu_typical, rate_extreme, rate_typical,
#                                           phi_extreme, phi_typical),
#     by = c("lat", "lon", "year", "month")
#   )
# 
# # ── Define fc/n_rows FIRST, right after the join, since everything below needs them ──
# n_rows <- nrow(fire)
# fc <- fire$fire_count_pred
# 
# # ── Magnitude calibration ────────────────────────────────────────────────────
# mu_blend_simulated <- fire$p_extreme_cal[fc > 0] * fire$mu_extreme[fc > 0] +
#                        (1 - fire$p_extreme_cal[fc > 0]) * fire$mu_typical[fc > 0]
# 
# current_mean_per_fire  <- mean(mu_blend_simulated, na.rm = TRUE)
# observed_mean_per_fire <- sum(fire$burned_area_km2_total) / sum(fire$fire_count)
# 
# cat("Current predicted mean per fire:", round(current_mean_per_fire, 1), "\n")
# cat("Observed mean per fire:", round(observed_mean_per_fire, 1), "\n")
# 
# magnitude_cal_shift <- log(observed_mean_per_fire) - log(current_mean_per_fire)
# cat("Magnitude calibration shift (log-scale):", round(magnitude_cal_shift, 3), "\n")
# 
# fire <- fire %>%
#   mutate(
#     mu_extreme_cal   = mu_extreme * exp(magnitude_cal_shift),
#     mu_typical_cal   = mu_typical * exp(magnitude_cal_shift),
#     rate_extreme_cal = phi_extreme / mu_extreme_cal,
#     rate_typical_cal = phi_typical / mu_typical_cal
#   )
# 
# # ── Sampling ──────────────────────────────────────────────────────────────────
# valid_idx <- which(!is.na(fc) & fc > 0)
# rep_idx   <- rep(valid_idx, times = fc[valid_idx])
# 
# rep_p_extreme <- rep(fire$p_extreme_cal[valid_idx], times = fc[valid_idx])
# rep_rate_ext  <- rep(fire$rate_extreme_cal[valid_idx], times = fc[valid_idx])
# rep_rate_typ  <- rep(fire$rate_typical_cal[valid_idx], times = fc[valid_idx])
# 
# is_extreme_draw <- rbinom(length(rep_idx), size = 1, prob = rep_p_extreme)
# 
# samples <- ifelse(
#   is_extreme_draw == 1,
#   rgamma(length(rep_idx), shape = fire$phi_extreme[1], rate = rep_rate_ext),
#   rgamma(length(rep_idx), shape = fire$phi_typical[1], rate = rep_rate_typ)
# )
# 
# row_sums <- tapply(samples, rep_idx, sum)
# total_burned_area <- rep(0, n_rows)
# total_burned_area[as.integer(names(row_sums))] <- row_sums
fire$total_burned_area <- total_burned_area
# #Use fire_count and burned area shape and rate estimates to simulate total burned area from the gamma distribution
# fire$total_burned_area <- mapply(
#   function(fire_count, BA_shape, BA_rate) {
#     if (fire_count == 0 || is.na(fire_count)) return(0)
#     sum(rgamma(n = fire_count, shape = BA_shape, rate = BA_rate))
#   },
#   fire$fire_count_pred,
#   fire$gamma_shape,
#   fire$gamma_rate
# )
# 
# #Use observed fire_count and predicted burned area shape and rate estimates to simulate total burned area from the gamma distribution
# fire$total_burned_area_test <- mapply(
#   function(fire_count, BA_shape, BA_rate) {
#     if (fire_count == 0 || is.na(fire_count)) return(0)
#     sum(rgamma(n = fire_count, shape = BA_shape, rate = BA_rate))
#   },
#   fire$fire_count_test,
#   fire$gamma_shape,
#   fire$gamma_rate
# )

```




Quick comparison between observed and predicted 10 year summaries

```{r}
pred_r_strike <- mean(fire$r_strike)
obs_r_strike <- mean(fire$r_strike_obs)

pred_strike_count <- sum(fire$n_strikes)
obs_strike_count <- sum(fire$n_strikes_obs)

pred_p_ignite <- mean(fire$p_ignite_rate_cal)
sim_obs_p_ignite <- sum(fire$fire_count)/sum(fire$n_strikes_obs)

pred_fire_count <- sum(fire$fire_count_pred)
obs_fire_count <- sum(fire$fire_count)

pred_burned_area <- sum(fire$total_burned_area)
obs_burned_area <- sum(fire$burned_area_km2_total)

summary_table <- as.data.frame(matrix(c(pred_r_strike, obs_r_strike, pred_strike_count,  obs_strike_count, pred_p_ignite, sim_obs_p_ignite, pred_fire_count,obs_fire_count, pred_burned_area, obs_burned_area), ncol=2, byrow=TRUE))

colnames(summary_table) <- c("predicted", "observed")

summary_table <- cbind(model_outpit = c("strike_rate", "strike_count", "p_ignite", "fire_count", "total_burned_area"), summary_table)

summary_table_fmt <- summary_table |>
  mutate(across(where(is.numeric), ~ formatC(., format = "f", digits = 4, big.mark = ",")))

print(summary_table_fmt)

```


```{r}
mean(area$area)
summary(fire$total_burned_area)
filter(fire, total_burned_area > area)

# fire <- fire %>%
#   left_join(
#     area %>% dplyr::select(lat, lon, area),
#     by = c("lat", "lon")
#   )

sum(filter(fire, total_burned_area > area)$total_burned_area)

fire <- fire |> mutate(total_burned_area_capped = ifelse(total_burned_area > area, area, total_burned_area))
#totally burned cells without being capped to gridcell size
sum(filter(fire, total_burned_area > area)$total_burned_area)
#totalburned cells capped
sum(filter(fire, total_burned_area > area)$total_burned_area_capped)

#total burned area with cap:
sum(fire$total_burned_area_capped)
```


Calculate cumulateve burned area with cap:

```{r}
# ── Generic sequential cap-and-recovery function (unchanged core logic) ─────
apply_area_cap <- function(df, raw_col = "total_burned_area", recovery_years = Inf) {
  # df must be pre-sorted chronologically (year, month) for ONE gridcell,
  # and must already have a single, validated `area` column

  n <- nrow(df)
  area <- df$area[1]

  capped_burned_area <- numeric(n)
  was_capped         <- logical(n)

  cumulative <- 0
  recovery_until_year <- -Inf

  for (i in seq_len(n)) {
    this_year <- df$year[i]
    raw_val   <- df[[raw_col]][i]

    if (this_year < recovery_until_year) {
      capped_burned_area[i] <- 0
      was_capped[i] <- (raw_val > 0)
      next
    }

    remaining_capacity <- area - cumulative

    if (remaining_capacity <= 0) {
      capped_burned_area[i] <- 0
      was_capped[i] <- (raw_val > 0)
      next
    }

    applied_val <- min(raw_val, remaining_capacity)
    capped_burned_area[i] <- applied_val
    was_capped[i] <- (applied_val < raw_val)

    cumulative <- cumulative + applied_val

    if (cumulative >= area) {
      cumulative <- 0
      recovery_until_year <- this_year + recovery_years
    }
  }

  df$total_burned_area_capped <- capped_burned_area
  df$burned_area_was_capped   <- was_capped
  df
}

# ── Wrapper: explicitly joins the CORRECT area table by lat/lon before capping ──

apply_cap_to_dataset <- function(data, area_lookup, recovery_years = Inf,
                                  lat_round = 4, lon_round = 4) {
  # area_lookup must have columns: lat, lon, area
  # (e.g. era5_grid_area.csv for the training period,
  #  TraCE_21ka_gridcell_area.csv / area_clipped_km2 for calibration and paleo)

  data <- data %>% mutate(lat = round(lat, lat_round), lon = round(lon, lon_round))
  area_lookup <- area_lookup %>%
    mutate(lat = round(lat, lat_round), lon = round(lon, lon_round)) %>%
    dplyr::select(lat, lon, area) %>%
    distinct(lat, lon, .keep_all = TRUE)   # guard against duplicate area rows per gridcell

  data <- data %>%
    left_join(area_lookup, by = c("lat", "lon"))

  # ── Validate before doing any of the sequential capping work ────────────────
  n_missing <- sum(is.na(data$area))
  if (n_missing > 0) {
    warning(sprintf(
      "%d rows (%.2f%%) had no matching area after the join - these gridcells will be excluded from capping. Check lat/lon rounding or coverage of area_lookup.",
      n_missing, 100 * n_missing / nrow(data)
    ))
  }

  # Confirm area really is constant within each gridcell (it should be, by construction,
  # but worth checking given how many silent join/duplication bugs this pipeline has had)
  area_check <- data %>%
    filter(!is.na(area)) %>%
    group_by(lat, lon) %>%
    summarise(n_distinct_area = n_distinct(area), .groups = "drop") %>%
    filter(n_distinct_area > 1)

  if (nrow(area_check) > 0) {
    stop(sprintf(
      "%d gridcells have MORE THAN ONE distinct area value - area_lookup likely has duplicate/conflicting entries for these lat/lon. Fix area_lookup before proceeding.",
      nrow(area_check)
    ))
  }

  data %>%
    filter(!is.na(area)) %>%
    arrange(lat, lon, year, month) %>%
    group_by(lat, lon) %>%
    group_modify(~ apply_area_cap(.x, recovery_years = recovery_years)) %>%
    ungroup()
}
```

Apply funciton
```{r}
era5_area <- read.csv(file.path(paleoclimate_data_root, "era5_grid_area.csv")) |> dplyr::select(lat, lon, area)

era5_area_clip <- era5_area %>%
  mutate(lat = round(lat, 4), lon = round(lon, 4)) %>%
  semi_join(
    fire %>% mutate(lat = round(lat, 4), lon = round(lon, 4)) %>% distinct(lat, lon),
    by = c("lat", "lon")
  )

plot(era5_area$lon, era5_area$lat)

fire <- fire %>% dplyr::select(-area)

fire_training_capped <- apply_cap_to_dataset(
  fire, area_lookup = era5_area, recovery_years = Inf
)
```

```{r}
# Sanity check the join/capping worked as expected
sum(is.na(fire_training_capped$area))   # should be 0 or very low
summary(fire_training_capped$total_burned_area_capped)
sum(fire_training_capped$burned_area_was_capped)

# Compare the corrected total against your earlier manual-cap result (67,837) and observed (62,826)
sum(fire_training_capped$total_burned_area_capped, na.rm = TRUE)
```
plot over time
```{r}

# ── Annual totals: observed, uncapped predicted, capped predicted ───────────

annual_comparison <- fire_training_capped %>%
  group_by(year) %>%
  summarise(
    Observed              = sum(burned_area_km2_total, na.rm = TRUE),
    `Predicted (uncapped)` = sum(total_burned_area, na.rm = TRUE),
    # `Predicted (capped)`   = sum(total_burned_area_capped, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Observed, `Predicted (uncapped)`, `Predicted (capped)`),
    names_to = "series",
    values_to = "burned_area"
  )

# ── Plot ───────────────────────────────────────────────────────────────────

ggplot(annual_comparison, aes(x = year, y = burned_area, color = series)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "Observed"              = "black",
    "Predicted (uncapped)"  = "firebrick",
    "Predicted (capped)"    = "steelblue"
  )) +
  labs(
    x = "Year", y = "Total burned area (km²)", color = NULL,
    title = "Annual total burned area: observed vs. predicted (with and without gridcell area cap)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}

# ── Annual totals: observed, predicted ───────────

annual_comparison <- fire_training_capped %>%
  group_by(year) %>%
  summarise(
    Observed              = sum(fire_count, na.rm = TRUE),
    `Predicted` = sum(fire_count_pred, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Observed, `Predicted`),
    names_to = "series",
    values_to = "fire_count"
  )

# ── Plot ───────────────────────────────────────────────────────────────────

ggplot(annual_comparison, aes(x = year, y = fire_count, color = series)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "Observed"              = "black",
    "Predicted"  = "firebrick"
  )) +
  labs(
    x = "Year", y = "Total fire count", color = NULL,
    title = "Annual total fire count: observed vs. predicted"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}

# ── Annual totals: observed, predicted ───────────

annual_comparison <- fire_training_capped %>%
  # group_by(year) %>%
  summarise(
    Observed              = sum(n_strikes_obs, na.rm = TRUE),
    `Predicted` = sum(n_strikes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Observed, `Predicted`),
    names_to = "series",
    values_to = "strike_count"
  )

# ── Plot ───────────────────────────────────────────────────────────────────

ggplot(annual_comparison, aes(x = year, y = strike_count, color = series)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "Observed"              = "black",
    "Predicted"  = "firebrick"
  )) +
  labs(
    x = "Year", y = "Total strike count", color = NULL,
    title = "Annual total strike count: observed vs. predicted"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

FOr other period datasets (calibration and paelo):
```{r}
# fire_calibration_capped <- apply_cap_to_dataset(
#   fire_calibration, area_lookup = trace_area, recovery_years = 12
# )
# 
# # Paleo period (TraCE grid, full Holocene) - same TraCE area table
# fire_paleo_capped <- apply_cap_to_dataset(
#   fire_paleo, area_lookup = trace_area, recovery_years = 12
# )
```


Export Drivers and Predictions
```{r}
df_out <- fire %>%
  left_join(
    area %>% dplyr::select(lat, lon, area),
    by = c("lat", "lon")
  )

write.csv(df_out, file.path(predictions_root, "predictions_2002_2011_mixture.csv"))

```

```{r}
df_old <- read.csv(file.path(predictions_root, "predictions_2002_2011.csv"))
```

```{r}
pred_r_strike <- mean(df_old$r_strike)
obs_r_strike <- mean(fire$r_strike_obs)

pred_strike_count <- sum(df_old$n_strikes)
obs_strike_count <- sum(df_old$n_strikes_obs)

pred_p_ignite <- mean(df_old$p_ignite_rate_cal)
sim_obs_p_ignite <- sum(df_old$fire_count)/sum(df_old$n_strikes_obs)

pred_fire_count <- sum(df_old$fire_count_pred)
obs_fire_count <- sum(fire$fire_count)

pred_burned_area <- sum(df_old$total_burned_area)
obs_burned_area <- sum(df_old$burned_area_km2_total)

summary_table <- as.data.frame(matrix(c(pred_r_strike, obs_r_strike, pred_strike_count,  obs_strike_count, pred_p_ignite, sim_obs_p_ignite, pred_fire_count,obs_fire_count, pred_burned_area, obs_burned_area), ncol=2, byrow=TRUE))

colnames(summary_table) <- c("predicted", "observed")

summary_table <- cbind(model_outpit = c("strike_rate", "strike_count", "p_ignite", "fire_count", "total_burned_area"), summary_table)

summary_table_fmt <- summary_table |>
  mutate(across(where(is.numeric), ~ formatC(., format = "f", digits = 4, big.mark = ",")))

print(summary_table_fmt)

```

```{r}
mean(area$area)
summary(df_old$total_burned_area)
filter(df_old, total_burned_area > mean(area$area))

sum(filter(df_old, total_burned_area > mean(area$area))$total_burned_area)

df_old <- df_old |> mutate(df_old, total_burned_area_capped = ifelse(total_burned_area > mean(area$area), mean(area$area), total_burned_area))
#totally burned cells without being capped to gridcell size
sum(filter(df_old, total_burned_area > mean(area$area))$total_burned_area)
#totalburned cells capped
sum(filter(df_old, total_burned_area > mean(area$area))$total_burned_area_capped)

#total burned area with cap:
sum(df_old$total_burned_area_capped)
```



```{r}
ba_extreme_obs <- fire$p_extreme[fire$burned_area_km2_total > 0] 
ggplot(obs_fire_happened, aes(x=burned_area_km2_total, y=p_extreme)) + geom_point()

#predicted fire count and ignition efficiency compounding to generate larger than expected fire sizes
obs_fire_happened <- filter(fire, burned_area_km2_total > 0)
obs_fire_happened  <- obs_fire_happened  %>% filter(!(fire_count > strike_count)) %>% mutate(fire_rate = fire_count/n_strikes_obs)
ggplot(obs_fire_happened, aes(x=fire_rate, y=burned_area_km2_total, size=p_extreme, alpha=0.1)) + geom_point() + ggtitle("observed fire counts/strikes vs. fire size")

#How does this behave in the observed data?
pred_fire_happened <- fire  %>% filter(total_burned_area > 0)
ggplot(pred_fire_happened, aes(x=p_ignite_rate_cal, y=p_extreme, size=total_burned_area, alpha=0.1)) + geom_point()+ ggtitle("predicted ignition efficiency vs. fire size")
```







July 3rd
- check gamma sampling for burned area. - is it using fire count as size correctly?
- same for fire count sampling using p_ignite and strike_count. 
- Why does sampling from poisson not work?? what does this plot tell us??
```{r}
plot(fire$strike_count_pred, fire$r_strike)
plot(fire$strike_count, fire$strikes)
plot(fire$strike_count_test, fire$strikes)
plot(fire$strike_count, fire$strike_count_test)

hist(fire$r_strike)
hist(fire$strikes)

hist(fire$strike_count_pred)
hist(fire$strike_count_test)
hist(fire$strike_count)

```



JULY 1

LIGHTNING 

mean strike rates are similar between observed and predicted. Issue arises when counts are sampled from the poisson distribution. Could it be an issue with tails - moments when strike rates are exceptionally high are not being captured by the gamma GLM - could the gamma Bayesian approach handle this better?

IGNITION
Predictions are too high, by an order of magnitude! 

FIRE
Best fire count: simulated from fire count (strikes * p_ignite) and conditioned on p_fire (rbinom(p_fire)==1)
However, multiplying the mean observed strike count (8.989618 strikes per gridcell) by mean observed ignition rate (0.00023 ignitions/strikes) gets very close to the observed fire count (273, whereas actual fire count is 265)

BURNED AREA
Best burned area: simulated from fire count (strikes * P_ignite)

Keep:
strike count in a gridcell * p_ignite 

- or try sampling form binomial distribution: 
n_fires = rbinom(strike_count, ignition_efficiency)
then:
total burned area = sum(rgamma(n_fires, shape = gamma_shape, rate  = gamma_shape / gamma_mean))

gamma_shape and gamma_mean are estimated by the burned area model. 

What about strike rate? Is the model picking up tails

Since the lightning model doesn't capture the tails, strike counts sampled from poisson have very small tail. 
The model captures the mean well, but not gridcell/months when climate results in extreme lightning events. Max predicted fire count is > 60, while observed max count is > 20,000.




