---
title: "Untitled"
output: html_document
---

```{r}
library(dplyr)
library(tidyr)
library(gt)
```



# =============================================================================
# Reads the same model objects the unified pipeline loads, extracts posterior
# mean + 95% CI for every regression coefficient in every sub-model, and
# assembles one wide table: rows = predictors, columns = outcome/parameter.
#
# NAMING NOTES (see chat for full explanation):
# - Lightning Gamma model: `alpha` is the SHAPE parameter, `beta` is the RATE
#   parameter (rgamma(shape = alpha, rate = beta) in predict_lightning_rate()) --
#   labeled here by true statistical role, not the a/b naming in the pipeline.
# - Typical/extreme fire "Gamma" models are mean-dispersion parameterized:
#   alpha/beta are coefficients on log(mean); phi is a separate, non-regression
#   dispersion parameter (not in the user's original list, added here since it
#   is a real fitted quantity with no predictor to report against).
# =============================================================================
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]
  key_vals <- strsplit(lines, "=")
  setNames(trimws(sapply(key_vals, `[`, 2)), trimws(sapply(key_vals, `[`, 1)))
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
modern_fire_stats_root <- config[["modern_fire_stats_root"]]
calibration_thunderstorm <- config[["calibration_thunderstorm"]]
```

```{r}
# ---- Helper: mean + 95% CI as a formatted string, or "--" if absent --------
fmt_param <- function(draws) {
  if (is.null(draws)) return("--")
  sprintf("%.3f [%.3f, %.3f]", mean(draws), quantile(draws, 0.025), quantile(draws, 0.975))
}
```

```{r}
# =============================================================================
# 1. LIGHTNING STRIKE RATE (Gamma): alpha = shape, beta = rate
#    Posterior columns: a_alpha..f_alpha (intercept, swr, tair, rh, precip, sp)
#                        a_beta..f_beta  (same order)
# =============================================================================
B12 <- readRDS(file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))
lightning_posterior <- as.data.frame(B12)

lightning_predictors <- c("swr", "tair", "rh", "precip", "sp")
alpha_cols <- c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "f_alpha")
beta_cols  <- c("a_beta",  "b_beta",  "c_beta",  "d_beta",  "e_beta",  "f_beta")

strike_rate_shape <- setNames(
  c(fmt_param(lightning_posterior[[alpha_cols[1]]]),
    sapply(2:6, function(i) fmt_param(lightning_posterior[[alpha_cols[i]]]))),
  c("Intercept", lightning_predictors)
)
strike_rate_rate <- setNames(
  c(fmt_param(lightning_posterior[[beta_cols[1]]]),
    sapply(2:6, function(i) fmt_param(lightning_posterior[[beta_cols[i]]]))),
  c("Intercept", lightning_predictors)
)

# =============================================================================
# 2. STRIKE COUNT DISPERSION (theta): NegBinomial, linear predictor on
#    log(theta). Posterior: gamma_0 (intercept) + gamma[,1:6] matching
#    theta_predictors <- c("swr","tair","rh","precip","sp","wind")
# =============================================================================
nb_model_obj  <- readRDS(file.path(rds_root, "Lightning/neg_bin_strike_count_no_outliers.rds"))
theta_predictors <- c("swr", "tair", "rh", "precip", "sp", "wind")
nb_posterior  <- rstan::extract(nb_model_obj$fit, pars = c("gamma_0", "gamma"))

strike_count_theta <- setNames(
  c(fmt_param(nb_posterior$gamma_0),
    sapply(seq_along(theta_predictors), function(i) fmt_param(nb_posterior$gamma[, i]))),
  c("Intercept", theta_predictors)
)

# =============================================================================
# 3. IGNITION EFFICIENCY (pi | 1 strike): Beta-Binomial, linear predictor on
#    logit(pi). Posterior: alpha (intercept) + beta[,1:k] matching
#    ignition_bayes_obj$predictors (order as saved with the model)
# =============================================================================
ignition_bayes_obj <- readRDS(file.path(rds_root, "Ignition_Efficiency/redo/beta_binomial_ignition_bayes.rds"))
ignition_posterior <- rstan::extract(ignition_bayes_obj$fit, pars = c("alpha", "beta"))
predictors_ignition <- ignition_bayes_obj$predictors

ignition_pi <- setNames(
  c(fmt_param(ignition_posterior$alpha),
    sapply(seq_along(predictors_ignition), function(i) fmt_param(ignition_posterior$beta[, i]))),
  c("Intercept", predictors_ignition)
)

# =============================================================================
# 4-7. BURNED AREA MIXTURE MODEL: classifier (P) + typical Gamma + extreme
#       Gamma, each with its own predictor set
# =============================================================================
mixture_obj <- readRDS(file.path(rds_root, "Burned_Area/redo/m3_bayes_mixture_custom.rds"))

logit_posterior   <- rstan::extract(mixture_obj$fit_logit,   pars = c("alpha", "beta"))
extreme_posterior <- rstan::extract(mixture_obj$fit_extreme, pars = c("alpha", "beta", "phi"))
normal_posterior  <- rstan::extract(mixture_obj$fit_normal,  pars = c("alpha", "beta", "phi"))

preds_classifier <- c("urban_proximity_binary", "rh", "tair", "precip", "tair_5y")
preds_extreme    <- c("U", "tair")
preds_typical    <- c("urban_proximity_binary", "tair_2m", "tair")

extreme_classifier_P <- setNames(
  c(fmt_param(logit_posterior$alpha),
    sapply(seq_along(preds_classifier), function(i) fmt_param(logit_posterior$beta[, i]))),
  c("Intercept", preds_classifier)
)

typical_fire_logmu <- setNames(
  c(fmt_param(normal_posterior$alpha),
    sapply(seq_along(preds_typical), function(i) fmt_param(normal_posterior$beta[, i]))),
  c("Intercept", preds_typical)
)
typical_fire_phi <- fmt_param(normal_posterior$phi)   # scalar, no predictor

extreme_fire_logmu <- setNames(
  c(fmt_param(extreme_posterior$alpha),
    sapply(seq_along(preds_extreme), function(i) fmt_param(extreme_posterior$beta[, i]))),
  c("Intercept", preds_extreme)
)
extreme_fire_phi <- fmt_param(extreme_posterior$phi)   # scalar, no predictor
```


# =============================================================================
# ASSEMBLE: union of every predictor name across all models, in a sensible
# fixed order (Intercept first, then alphabetical), one row per predictor,
# one column per model/parameter, "--" where a predictor isn't in that model
# =============================================================================
```{r}
all_predictors <- union(
  names(strike_rate_shape),
  union(names(strike_count_theta),
        union(names(ignition_pi),
              union(names(extreme_classifier_P),
                    union(names(typical_fire_logmu), names(extreme_fire_logmu)))))
)
all_predictors <- c("Intercept", sort(setdiff(all_predictors, "Intercept")))

lookup_or_dash <- function(named_vec, predictor) {
  if (predictor %in% names(named_vec)) named_vec[[predictor]] else "--"
}

param_table <- tibble(
  Predictor = all_predictors,
  `r_strike: shape` = sapply(all_predictors, lookup_or_dash, named_vec = strike_rate_shape),
  `r_strike: rate`  = sapply(all_predictors, lookup_or_dash, named_vec = strike_rate_rate),
  `n_strikes: dispersion` = sapply(all_predictors, lookup_or_dash, named_vec = strike_count_theta),
  `P_ignite` = sapply(all_predictors, lookup_or_dash, named_vec = ignition_pi),
  `P_extreme` = sapply(all_predictors, lookup_or_dash, named_vec = extreme_classifier_P),
  `B_typical: log(mean)` = sapply(all_predictors, lookup_or_dash, named_vec = typical_fire_logmu),
  `B_extreme: log(mean)` = sapply(all_predictors, lookup_or_dash, named_vec = extreme_fire_logmu)
)

# Add the two dispersion (phi) rows separately -- scalars, no predictor row
# to attach to, so they get their own labeled rows at the bottom
phi_row <- tibble(
  Predictor = "Dispersion (\u03c6)",
  `r_strike: shape` = "--", `r_strike: rate` = "--",
  `n_strikes: dispersion` = "--", `P_ignite` = "--",
  `P_extreme` = "--",
  `B_typical: log(mean)` = typical_fire_phi,
  `B_extreme: log(mean)` = extreme_fire_phi
)
param_table <- bind_rows(param_table, phi_row)

param_table
```



# =============================================================================
# EXPORT: gt table -> LaTeX body only (booktabs, no float wrapper), matching
# the export convention already used for the metrics table
# =============================================================================
```{r}
export_table_body_only <- function(gt_tbl, path) {
  full_tex <- as_latex(gt_tbl) %>% as.character()
  body <- sub(".*(\\\\begin\\{tabular\\*\\}.*\\\\end\\{tabular\\*\\})[\\s\\S]*", "\\1", full_tex, perl = TRUE)
  writeLines(body, path)
}

gt_param_tbl <- param_table %>%
  gt(rowname_col = "Predictor") %>%
  tab_header(title = "Posterior parameter estimates (mean [95% CI]) by sub-model")

export_table_body_only(gt_param_tbl, "parameter_table.tex")

print(param_table, n = Inf)
```

