---
title: "Outliers"
output: html_document
---


```{r, echo=FALSE, include=FALSE}
library(rstan)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(patchwork)
library(ggplot2)
library(gridExtra)
library(terra)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 

```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]
```


```{r}
df <- read.csv(file.path(drivers_root, "lightning_strike_count_climate_jja_2002_2011.csv"))
head(df)
```

What's expected?
Number reported from Bieniek: "The gridded monthly averaged observed lightning over 1986–2015 is shown for May, June, July, and August in [Fig. 3](https://journals.ametsoc.org/view/journals/apme/59/6/JAMC-D-19-0209.1.xml#fig3). Most lightning activity occurs in June and July, with some 20-km grid cells in the interior of Alaska receiving more than 30 strikes per month on average." 

Bienek's gridcells are a little larger (20x20=400 km2)
```{r}
area <- read.csv(file.path(paleoclimate_data_root, "era5_grid_area.csv"))
alaska_max_rate_per_km2 <- 30 / 400   # Bieniek's peak interior rate, per km²
physical_cap <- alaska_max_rate_per_km2 * max(area$area)
physical_cap
```



Location
```{r}
filter(df, strike_count > 1000) %>%
  count(lat, lon) %>%
  arrange(desc(n))

filter(df, strike_count > 100) %>%
  count(lat, lon) %>%
  arrange(desc(n))
```

Years
```{r}
nrow(filter(df, strike_count > 1000))
filter(df, strike_count > 1000) %>%
  count(year) %>%
  arrange(desc(n))

nrow(filter(df, strike_count > 100))
filter(df, strike_count > 100) %>%
  count(year) %>%
  arrange(desc(n))
```


Are extreme clusters non random?
```{r}
library(data.table)
setDT(df)

# How many unique gridcells exist in the dataset overall?
n_total_gridcells <- df[, uniqueN(paste(lat, lon))]

# Observed: 7 gridcells account for how many of the >1000 rows?
top7 <- df[strike_count > 1000, .N, by = .(lat, lon)][order(-N)][1:7]
pct_from_top7 <- sum(top7$N) / nrow(df[strike_count > 1000])

# Compare to what you'd expect if >1000 events were spread randomly across
# all gridcells with equal probability
expected_pct_if_random <- 7 / n_total_gridcells
cat(sprintf("Observed: %.1f%% of extreme rows from top 7 gridcells\n", pct_from_top7 * 100))
cat(sprintf("Expected if random: %.4f%%\n", expected_pct_if_random * 100))
```
observed concentration is (orders of magnitude) above the random-expectation baseline, likely given 7 out of ~88 rows above 1000 all falling into just 7 of presumably thousands of gridcells. this isn't random weather signal, worth reporting as a stated justification rather than an eyeball judgment.

Where should the threshold be set?

point where the plot's slope changes noticeably or the confidence band widens erratically are where the EVT-suggested threshold above which the data stops following the bulk distribution's tail behavior.
```{r}
library(ismev)  # or POT package — either works
strike_vals <- df[strike_count > 0, strike_count]

mrl.plot(strike_vals, umin = 50, umax = 5000)

mrl.plot(strike_vals, umin = 50, umax = 1100)
```


The physically expected maximum based on Bieniek et al.'s peak interior lightning rate scaled to ERA5 grid area (~35 strikes/month/gridcell), the extreme value threshold identified via mean residual life analysis (~1,000), and the strong spatial non-randomness of exceedances (100% concentrated in 7 of 15,041 gridcells, versus 0.0465% expected under random occurrence) jointly motivated a cap of 1,000 strikes per gridcell-month. Strike count observations exceeding this cap were excluded from the training data (rather than truncated). 
