---
title: "paleo_predicitons"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(pROC)
library(loo)
library(cowplot)
library(stats)
library(caret)
library(arrow)
library(patchwork)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

##########################################
#####       LIGHTNING    (r_strike)  #####
##########################################

r_strike ~ Gamma(alpha, beta)

alpha = exp(a_alpha + b_alpha * X)

beta = exp(a_beta + b_beta * X)

a_alpha, b_alpha, a_beta, b_beta ~ Normal(0, 1)

r_strike: lightning strikes / km2 / month
X: temperature, precipitation, relative humidity, shortwave radiation, surface pressure

exp(): ensures alpha and beta stay positive, which the gamma distribution requires. 

```{r}
#corrected trace data:
df_summer <- read_parquet(file.path(paleoclimate_data_root, "trace21ka_processed/trace_lightning_seasonal_corrected.parquet"))
head(df_summer)
df_month <- read_parquet(file.path(paleoclimate_data_root, "trace21ka_processed/trace_monthly_with_lags_and_landcover.parquet"))

#only want holocene:
df_summer <- filter(df_summer, year > -9750)
df_month <- filter(df_month, year > -9750)

#to compare with training period observations and predictions:
df_modern <- read.csv(file.path(predictions_root, "predictions_2002_2011.csv"))

#climate drivers for lightning strike rate modern trainin gperiod:
lightning <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))

#raw trace data:
#df_summer <- read_parquet(file.path(paleoclimate_data_root, "trace21ka_processed/trace_lightning_model.parquet"))
#df_month <- read_parquet(file.path(paleoclimate_data_root, "trace21ka_processed/trace_ignition_burnedarea_model.parquet"))

```

```{r}
# #make year match conventional dating
# df_summer <- df_summer %>%
#   mutate(year = ce_year + 1900)
# 
# df_month <- df_month %>%
#   mutate(year = ce_year + 1900)
# 
# # Sanity check
# range(df_summer$year)
# range(df_month$year)
# 
# #clip to Alaska
# trace_cells <- readRDS(file.path(rds_root, "trace_cells.rds"))
# 
# df_summer <- df_summer %>%
#   mutate(lon = ifelse(lon > 180, lon - 360, lon)) %>%   # match trace_cells' -180/180 convention
#   semi_join(trace_cells, by = c("lat", "lon")) %>%
#   left_join(trace_elev_lookup, by = c("lat", "lon"))
# 
# df_month <- df_month %>%
#   mutate(lon = ifelse(lon > 180, lon - 360, lon)) %>%   # match trace_cells' -180/180 convention
#   semi_join(trace_cells, by = c("lat", "lon")) %>%
#   left_join(trace_elev_lookup, by = c("lat", "lon"))
# 
# # Sanity checks
# nrow(df_summer)
# n_distinct(df_summer$lat, df_summer$lon)   # should be 23
# names(df_summer)
# # Sanity checks
# nrow(df_month)
# n_distinct(df_month$lat, df_month$lon)   # should be 23
# names(df_month)
```

Predict lightning strike rate for all gridcells and years
```{r}
#standardize:
lightning_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) lightning_mu_sigma$mu[lightning_mu_sigma$var == var]
get_sigma <- function(var) lightning_mu_sigma$sigma[lightning_mu_sigma$var == var]

lightning_scaled <- df_summer |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    #wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    swr    = (swr    - get_mu("swr"))    / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh")
  )

# ── Load B12 Bayesian lightning model ─────────────────────────────────────────
B12 <- readRDS(file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))

# ── Extract posterior samples ─────────────────────────────────────────────────
posterior_samples <- as.data.frame(B12)

# Generate predictions using posterior mean coefficients
# B12 excludes wind — predictors: swr, tair, rh, precip, sp

a_alpha_mean <- mean(posterior_samples$a_alpha)
b_alpha_mean <- mean(posterior_samples$b_alpha)
c_alpha_mean <- mean(posterior_samples$c_alpha)
d_alpha_mean <- mean(posterior_samples$d_alpha)
e_alpha_mean <- mean(posterior_samples$e_alpha)
f_alpha_mean <- mean(posterior_samples$f_alpha)

a_beta_mean  <- mean(posterior_samples$a_beta)
b_beta_mean  <- mean(posterior_samples$b_beta)
c_beta_mean  <- mean(posterior_samples$c_beta)
d_beta_mean  <- mean(posterior_samples$d_beta)
e_beta_mean  <- mean(posterior_samples$e_beta)
f_beta_mean  <- mean(posterior_samples$f_beta)

# Compute alpha and beta per observation using posterior means
# Using vectorised operations rather than a loop for speed

lp_alpha <- a_alpha_mean +
            b_alpha_mean * lightning_scaled$swr   +
            c_alpha_mean * lightning_scaled$tair  +
            d_alpha_mean * lightning_scaled$rh    +
            e_alpha_mean * lightning_scaled$precip +
            f_alpha_mean * lightning_scaled$sp

lp_beta  <- a_beta_mean  +
            b_beta_mean  * lightning_scaled$swr   +
            c_beta_mean  * lightning_scaled$tair  +
            d_beta_mean  * lightning_scaled$rh    +
            e_beta_mean  * lightning_scaled$precip +
            f_beta_mean  * lightning_scaled$sp

alpha_pred <- exp(lp_alpha)
beta_pred  <- exp(lp_beta)

set.seed(123)

# Gamma alpha and beta
# if you want the mean: alpha / beta
lightning_scaled <- lightning_scaled |>
  mutate(
    gamma_alpha   = alpha_pred,
    gamma_beta    = beta_pred,
    #r_strike_mean  = alpha_pred / beta_pred   # posterior mean prediction
    r_strike = rgamma(n(), shape = gamma_alpha, rate = gamma_beta)
  )
```

```{r}
hist(filter(df_summer, year>1950)$tair, main="tair")
hist(lightning$tair, main="rh")

hist(filter(df_summer, year>1950)$rh, main="rh")
hist(lightning$rh, main="rh")

hist(filter(df_summer, year>1950)$precip, main="precip")
hist(lightning$precip, main="precip")

hist(filter(df_summer, year>1950)$sp, main="sp")
hist(lightning$sp, main="sp")

hist(filter(df_summer, year>1950)$swr, main="swr")
hist(lightning$swr, main="swr")

```


```{r}
# ── plot ────────────────────────────────────────────────────────────────────
cat("Observed training period strike rate summary:\n")
print(summary(df_modern$r_strike_obs))
hist(df_modern$r_strike_obs, main="Observed training period strike rate")

cat("Predicted training period strike rate summary:\n")
print(summary(df_modern$r_strike))
hist(df_modern$r_strike, main="Predicted training period strike rate")

lightning_last_century <- filter(lightning_scaled, year > 1900)
cat("Predicted strike during the last century:\n")
print(summary(lightning_last_century$r_strike))
hist(lightning_last_century$r_strike, main="Last century simulated strike rate")

cat("Predicted strike rate summary:\n")
print(summary(lightning_scaled$r_strike))
hist(lightning_scaled$r_strike, main="Paleo simulated strike rate")

lightning_post_dryas <- filter(lightning_scaled, year > -10000)
cat("Predicted strike rate post dryas:\n")
print(summary(lightning_post_dryas$r_strike))
hist(lightning_post_dryas$r_strike, main="Holocene simulated strike rate")

df_summary <- lightning_scaled %>%
  dplyr::select(year, r_strike) %>%
  #filter(year > -10600) %>%
  mutate(century_bin = floor(year / 100) * 100) %>%   # bin into 100-year windows
  group_by(century_bin) %>%
  summarise(
    mean_r_strike = mean(r_strike, na.rm = TRUE),
    ci_low   = quantile(r_strike, 0.025, na.rm = TRUE),
    ci_high  = quantile(r_strike, 0.975, na.rm = TRUE),
    n_obs    = n(),
    .groups  = "drop"
  )

ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_r_strike)) +
  geom_point(size = 3, alpha=0.2) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.1, alpha=0.2) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin), 
                                              max(df_summary$century_bin), by = 500))) +
  theme_minimal() + 
  labs(x = "Year (100-yr bin)", y = "100 year mean strike rate",
       title = "") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme(legend.title = element_blank())

```


##########################################
##### STRIKE COUNT       (n_strikes) #####
##########################################

n_strikes ~ NegBinomial(mu, theta)

mu = r_strike * area

theta = exp(gamma_0 + gamma * X)

Gamma_0 ~ Normal(2,2)
Gamma ~ Normal(0, sigma2)
Sigma = 1 

r_strike: strikes/km2/month
area: gridcell area, km2
n_strikes: number of strikes in a gridcell-month
X: monthly climate predictors, temperature, precipitation, wind, relative humidity, shortwave radiation, surface pressure
theta: Precision/dispersion parameter of the Negative Binomial distribution predicted from climate. Theta > 0.
mu: mean lightning strike rate, strikes/gridcell/month. This is predicted from the Gamma model above and is the same for all summer months (June, July, August) in a year. The summer mean was used to ensure that mu > 0, which is required by the negative binomial distribution. 




```{r}
# ── Apply negative binomial model to fire dataframe ────────────────────────────────────────────────────

#merge summer strike rate (strikes/km2/month) predictions, r_strike - months will be duplicated
df <- df_month %>%
  left_join(
    lightning_scaled %>% dplyr::select(lat, lon, year, r_strike),
    by = c("lat", "lon", "year")
  )


#Scale strike rate from strikes/km2/month to strikes/gridcell/month
area <- read.csv(file.path(paleoclimate_data_root, "TraCE_21ka_gridcell_area.csv")) 

colnames(area)[colnames(area) == "area_clipped_km2"] <- "area"
head(area)

area <- area %>%
  mutate(lat = round(lat, 4), lon = round(lon, 4))

df <- df %>%
  mutate(lat = round(lat, 4), lon = round(lon, 4)) %>%
  left_join(
    area %>% dplyr::select(lat, lon, area),
    by = c("lat", "lon")
  )

#standardize:
n_strike_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_count_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) n_strike_mu_sigma$mu[n_strike_mu_sigma$var == var]
get_sigma <- function(var) n_strike_mu_sigma$sigma[n_strike_mu_sigma$var == var]

n_strike_scaled <- df |>
  dplyr::select(lat, lon, year, month, r_strike, area, precip, tair, wind, rh, swr, sp) |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh"),
    swr     = (swr     - get_mu("swr"))     / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp")
  )

#caluclate mu for the negative binomial distribution: r_strike * area = strikes/month/gridcell
n_strike_scaled <- n_strike_scaled |>
  mutate(mu_r_strike_gridcell = r_strike * area)

#Predictors for theta (negative binomial parameter we need to sample strike count)
theta_predictors <- c("swr", "tair", "rh", "precip", "sp", "wind")

#read in strike count model
nb_model <- readRDS(file.path(rds_root, "Lightning/neg_bin_strike_count_small.rds"))

#Posterior mean gamma coefficients
gamma_0_mean <- nb_model$coef_means$gamma_0
gamma_means  <- nb_model$coef_means$gamma

#Build predictor matrix for full fire dataframe
Z_fire <- model.matrix(
  as.formula(paste("~", paste(theta_predictors, collapse = " + "), "- 1")),
  data = n_strike_scaled   # must have same standardised predictors
)

#Compute posterior mean theta per gridcell-month
log_theta_pred <- as.numeric(gamma_0_mean + Z_fire %*% gamma_means)
theta_pred     <- exp(log_theta_pred)

#sample strike count using mu=r_strike*area and theta
n_strike_scaled <- n_strike_scaled |>
  mutate(
    nb_theta         = theta_pred,
    log_mu_pred      = log(pmax(mu_r_strike_gridcell, 1e-8)),
    n_strikes = rnbinom(n    = n(),
                                 mu   = mu_r_strike_gridcell,
                                 size = nb_theta)
  )
```

Look at simulated strike count:
```{r}
# ── Verify ────────────────────────────────────────────────────────────────────
cat("Simulated zeros:   ", sum(n_strike_scaled$n_strikes == 0), "\n")
cat("Simulated mean:    ", mean(n_strike_scaled$n_strikes), "\n")
cat("Simulated count:    ", sum(n_strike_scaled$n_strikes), "\n")
cat("Simulated variance:", var(n_strike_scaled$n_strikes), "\n")
cat("Simulated strike distribution:   ", print(summary(n_strike_scaled$n_strikes)), "\n")

hist(df_modern$n_strikes_obs, main="training period observed strike count")
hist(df_modern$n_strikes, main="training period predicted strike count")
hist(n_strike_scaled$n_strikes, main="trace strike count")

lightning_post_dryas <- filter(n_strike_scaled, year > -10000)

cat("Simulated strike distribution:   ", print(summary(lightning_post_dryas$n_strikes)), "\n")
hist(lightning_post_dryas$n_strikes, main="Holocene strike count")

df_summary <- lightning_scaled %>%
  dplyr::select(year, r_strike) %>%
  #filter(year > -10600) %>%
  mutate(century_bin = floor(year / 100) * 100) %>%   # bin into 100-year windows
  group_by(century_bin) %>%
  summarise(
    mean_r_strike = mean(r_strike, na.rm = TRUE),
    ci_low   = quantile(r_strike, 0.025, na.rm = TRUE),
    ci_high  = quantile(r_strike, 0.975, na.rm = TRUE),
    n_obs    = n(),
    .groups  = "drop"
  )

ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_r_strike)) +
  geom_point(size = 3, alpha=0.2) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.1, alpha=0.2) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin), 
                                              max(df_summary$century_bin), by = 500))) +
  theme_minimal() + 
  labs(x = "Year (100-yr bin)", y = "100 year mean strike count",
       title = "") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme(legend.title = element_blank())
```


#add urban and road proximity - all 0 for paleo period
```{r}
# Add suppression predictors — all zero for paleoclimate (no roads/urban areas)
df$road_proximity_binary  <- 0L
df$urban_proximity_binary <- 0L
```



##########################################
##### IGNITION EFFICINCY (p_ignite) #####
##########################################

Y ~ Binomial(n, p_ignite) --> technically, this is applied in the next step, to estimate fire count.

logit(p_ignite) = alpha + X * beta

alpha ~ normal(mu_alpha, sigma_alpha)
beta  ~ normal(0, sigma_beta)

p_ignite_corrected = plogis(qlogis(pie) + cal_shift)

p_ignite: probability of 1 strike resulting in an ignition in a given gridcell and month. 
X: average precipitation rate, temperature, relative humidity, cumulative precipitation for the previous 1, 2, 3 months, and 5 years, mean temperature for the previous 1, 2, 3 months, and 5 years, and % broadleaf cover. 

Land cover predictors are not available pre -9749 and post 1950. assign 1950 values to years after then Filter out these NA values for years prior to min(df$year[!is.na(df$B)]) = -9749
```{r}
max(df$year[!is.na(df$B)])
min(df$year[!is.na(df$B)])

#make landcover for 1950 = landcover till 1989. 
#get each gridcell's landcover values AT year 1950
landcover_1950 <- df %>%
  filter(year == 1950) %>%
  dplyr::select(lat, lon, C_1950 = C, B_1950 = B, U_1950 = U) %>%
  distinct(lat, lon, .keep_all = TRUE)   # in case of multiple months per year — see note below

# Sanity check: should be one row per gridcell
nrow(landcover_1950)
n_distinct(df$lat, df$lon)

#join those 1950 values onto the full dataframe, and overwrite C/B/U for any row where year > 1950
df <- df %>%
  left_join(landcover_1950, by = c("lat", "lon")) %>%
  mutate(
    C = if_else(year > 1950, C_1950, C),
    B = if_else(year > 1950, B_1950, B),
    U = if_else(year > 1950, U_1950, U)
  ) %>%
  dplyr::select(-C_1950, -B_1950, -U_1950)

#remove rows prior to -9749:
df <- filter(df, year > -9750)
```

```{r}

#standardize:
ignition_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_ignition_predictors_redo.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) ignition_mu_sigma$mu[ignition_mu_sigma$var == var]
get_sigma <- function(var) ignition_mu_sigma$sigma[ignition_mu_sigma$var == var]

ignition_scaled <- df |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_2m   = (precip_2m   - get_mu("precip_2m"))   / get_sigma("precip_2m"),
    precip_3m   = (precip_3m   - get_mu("precip_3m"))   / get_sigma("precip_3m"),
    precip_5y     = (precip_5y     - get_mu("precip_5y"))     / get_sigma("precip_5y"),
    tair_1m = (tair_1m - get_mu("tair_1m")) / get_sigma("tair_1m"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_3m   = (tair_3m   - get_mu("tair_3m"))   / get_sigma("tair_3m"),
    tair_5y     = (tair_5y     - get_mu("tair_5y"))     / get_sigma("tair_5y"),
    B     = (B     - get_mu("B"))     / get_sigma("B"),
    C     = (C     - get_mu("C"))     / get_sigma("C"),
    U     = (U     - get_mu("U"))     / get_sigma("U")
  )
```


```{r}
#Bayes recalibrated
# ── Load calibrated Bayesian model ────────────────────────────────────────────
# ignition_bayes_obj <- readRDS(file.path(rds_root, 
#                                       "Ignition_Efficiency/m2_bayes_calibrated.rds")) #comment out when rerunning - takes a while to load

ignition_bayes_obj <- readRDS(file.path(rds_root, 
                                         "Ignition_Efficiency/redo/m9_bayes_calibrated.rds"))

ignition_bayes_fit <- ignition_bayes_obj$fit        # stanfit object
cal_shift_bayes    <- ignition_bayes_obj$cal_shift  # intercept shift
coef_means         <- ignition_bayes_obj$coef_means # posterior mean alpha and beta

cat("Bayesian calibration shift:", round(cal_shift_bayes, 3), "\n")

# ── Build predictor matrix from standardised data ─────────────────────────────
# predictors_m2 <- c("rh", "tair", "precip",
#                     "precip_1m", "precip_2m", "precip_3m",
#                     "precip_5y", "tair_1m", "tair_2m",
#                     "tair_3m", "tair_5y", "B")

predictors_m9 <- c("rh", "tair", "precip", "precip_1m", "precip_2m", "precip_3m", "tair_2m", "tair_3m", "tair_5y", "C", "U")

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m9, collapse = " + "), "- 1")),
  data = ignition_scaled
)

# ── Generate predictions using posterior mean ─────────────────────────────────
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

ignition_scaled <- ignition_scaled |>
  mutate(
    #inverse_logit of linear predictor, linpred
    p_ignite_bayes     = plogis(linpred),
    #add calibration shift 
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,     #pmin and pmax make sure arguments in plogis from reaching infinity
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Raw Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes_cal))

ignition_scaled <- ignition_scaled[, c("lat", "lon", "year", "month", "p_ignite_bayes_cal")]

ignition_scaled <- ignition_scaled %>% dplyr::rename(p_ignite = p_ignite_bayes_cal)

#add monthly ignition probability, p_ignite, to fire df
df <- df %>%
  left_join(
    ignition_scaled %>% dplyr::select(lat, lon, year, month, p_ignite),
    by = c("lat", "lon", "year", "month"))

```


##### RATE CALIBRATION SHIFT FOR P_IGNITE ##### 

```{r}

rate_cal_shift <- readRDS(file.path(rds_root, "Ignition_Efficiency/redo/calibration_shifts.rds"))$rate_cal_shift

eps <- 1e-8

df <- df |>
  mutate(
    p_ignite_rate_cal = plogis(
      qlogis(pmin(pmax(p_ignite, eps), 1 - eps)) + rate_cal_shift
    )
  )
```

Ignition efficiency over time
```{r}
hist(df$p_ignite_rate_cal, main="Holocene ignition efficiency")

df_summary <- df %>%
  dplyr::select(year, p_ignite_rate_cal) %>%
  #filter(year > -10600) %>%
  mutate(century_bin = floor(year / 100) * 100) %>%   # bin into 100-year windows
  group_by(century_bin) %>%
  summarise(
    mean_p_ignite = mean(p_ignite_rate_cal, na.rm = TRUE),
    ci_low   = quantile(p_ignite_rate_cal, 0.025, na.rm = TRUE),
    ci_high  = quantile(p_ignite_rate_cal, 0.975, na.rm = TRUE),
    n_obs    = n(),
    .groups  = "drop"
  )

ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_p_ignite)) +
  geom_point(size = 3, alpha=0.2) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.1, alpha=0.2) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin), 
                                              max(df_summary$century_bin), by = 500))) +
  theme_minimal() + 
  labs(x = "Year (100-yr bin)", y = "100 year mean ignition efficiency",
       title = "") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme(legend.title = element_blank())
```

##########################################
#####     BURNED AREA | 1 ignition  #####
##########################################

Burned_area ~ Gamma(theta, theta/mu)

log(mu) = alpha + alpha_year + X * beta

alpha ~ Normal(0, sigma^2_alpha)
beta ~ Normal(0, sigma^2_beta)
theta ~ Exponential(1)
alpha_year ~ Normal(0, sigma^2_year)
sigma^2_year ~ Exponential(1)


Burned_area: area of 1 lightning-caused ignition (km2)
X: Relative humidity, temperature, precipitation, wind, cumulative precipitation for the previous 1, 3 months, and 5 years, mean temperature for the previous 2 months and 5 years, proximity to urban area (binary; 1 = within 40 km of an urban area), proximity to road (binary; 1 = within 40 km of an major road)
alpha_year: year random affect

Apply model
```{r}
#standardize:
burned_area_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_burned_area_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) burned_area_mu_sigma$mu[burned_area_mu_sigma$var == var]
get_sigma <- function(var) burned_area_mu_sigma$sigma[burned_area_mu_sigma$var == var]

burned_area_scaled <- df |>
  mutate(
    rh        = (rh        - get_mu("rh"))        / get_sigma("rh"),
    tair      = (tair      - get_mu("tair"))      / get_sigma("tair"),
    precip    = (precip    - get_mu("precip"))    / get_sigma("precip"),
    wind      = (wind      - get_mu("wind"))      / get_sigma("wind"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_3m = (precip_3m - get_mu("precip_3m")) / get_sigma("precip_3m"),
    precip_5y = (precip_5y - get_mu("precip_5y")) / get_sigma("precip_5y"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_5y   = (tair_5y   - get_mu("tair_5y"))   / get_sigma("tair_5y"),
    C     = (C     - get_mu("C"))     / get_sigma("C"),
    U     = (U     - get_mu("U"))     / get_sigma("U")
  ) #no scaling for binary urban and road data.

# ── Predictor sets, one per sub-model (from mixture_gamma_custom fit) ──

preds_classifier <- c("urban_proximity_binary", "rh", "tair", "precip", "tair_5y")
preds_extreme     <- c("U", "tair")
preds_typical     <- c("urban_proximity_binary", "tair_2m", "tair")

# ── Load the fitted mixture model ────────────────────────────────────────────

mixture_obj <- readRDS(file.path(rds_root, "Burned_Area/redo/m3_bayes_mixture_custom.rds"))

fit_logit   <- mixture_obj$fit_logit
fit_extreme <- mixture_obj$fit_extreme
fit_normal  <- mixture_obj$fit_normal

# ── Posterior means for each component (analogous to old alpha_mean/beta_means) ──

summ_logit   <- rstan::summary(fit_logit,   pars = c("alpha", "beta"))$summary
summ_extreme <- rstan::summary(fit_extreme, pars = c("alpha", "beta", "phi"))$summary
summ_normal  <- rstan::summary(fit_normal,  pars = c("alpha", "beta", "phi"))$summary

alpha_logit_mean <- summ_logit["alpha", "mean"]
beta_logit_mean  <- summ_logit[grep("^beta\\[", rownames(summ_logit)), "mean"]

alpha_extreme_mean <- summ_extreme["alpha", "mean"]
beta_extreme_mean  <- summ_extreme[grep("^beta\\[", rownames(summ_extreme)), "mean"]
phi_extreme_mean   <- summ_extreme["phi", "mean"]

alpha_typical_mean <- summ_normal["alpha", "mean"]
beta_typical_mean  <- summ_normal[grep("^beta\\[", rownames(summ_normal)), "mean"]
phi_typical_mean   <- summ_normal["phi", "mean"]

cat("Classifier intercept (alpha):", round(alpha_logit_mean, 3), "\n")
cat("Extreme-Gamma shape (phi):   ", round(phi_extreme_mean, 3), "\n")
cat("Typical-Gamma shape (phi):   ", round(phi_typical_mean, 3), "\n")

# ── Build predictor matrices — three separate ones, one per sub-model ───────

X_classifier <- model.matrix(
  as.formula(paste("~", paste(preds_classifier, collapse = " + "), "- 1")),
  data = burned_area_scaled
)
X_extreme <- model.matrix(
  as.formula(paste("~", paste(preds_extreme, collapse = " + "), "- 1")),
  data = burned_area_scaled
)
X_typical <- model.matrix(
  as.formula(paste("~", paste(preds_typical, collapse = " + "), "- 1")),
  data = burned_area_scaled
)

# ── Generate predictions (posterior mean, point-estimate — same approach
# as "Option A" in your original code) ───────────────────────────────────────

p_extreme_linpred <- as.numeric(alpha_logit_mean + X_classifier %*% beta_logit_mean)
p_extreme         <- plogis(p_extreme_linpred)

mu_extreme <- as.numeric(exp(alpha_extreme_mean + X_extreme %*% beta_extreme_mean))
mu_typical <- as.numeric(exp(alpha_typical_mean + X_typical %*% beta_typical_mean))

# The blended MEAN (for reporting / a point prediction) is a valid weighted average -
# this is fine to report as expected burned area, but is NOT itself a Gamma mean,
# so it should NOT be paired with a single shape parameter for sampling purposes.
# ie. this is expected mean fire size in a gridcell-month given | 1 fire
burned_area_pred <- p_extreme * mu_extreme + (1 - p_extreme) * mu_typical

burned_area_scaled <- burned_area_scaled |>
  mutate(
    p_extreme        = p_extreme,
    mu_extreme       = mu_extreme,
    mu_typical       = mu_typical,
    phi_extreme      = phi_extreme_mean,   # constant across rows, like old phi_mean
    phi_typical      = phi_typical_mean,   # constant across rows
    burned_area_pred = burned_area_pred    # blended point estimate, for reference/plots only
  )

# rate = shape / mean, computed PER COMPONENT (not a single blended rate)
burned_area_scaled <- burned_area_scaled |>
  mutate(
    rate_extreme = phi_extreme / mu_extreme,
    rate_typical = phi_typical / mu_typical
  )

################################################
## Logit-space calibration shift on p_extreme 
################################################
#Apply the pre-fit p_extreme calibration shift to a new period's data 

extreme_cal_params <- readRDS(file.path(rds_root, "Burned_Area/extreme_cal_shift.rds"))

extreme_cal_shift <- extreme_cal_params$extreme_cal_shift
eps               <- extreme_cal_params$eps

cat("Applying extreme calibration shift:", round(extreme_cal_shift, 3),
    "(fitted on", extreme_cal_params$fitted_on, ")\n")

burned_area_scaled <- burned_area_scaled %>%
  mutate(
    p_extreme_cal = plogis(pmin(pmax(qlogis(pmin(pmax(p_extreme, eps), 1 - eps)) + extreme_cal_shift,
                                       qlogis(eps)), qlogis(1 - eps)))
  )
```


Mean Burned area | 1 ignition over time
```{r}
hist(burned_area_scaled$burned_area_pred, main="Holocene ignition efficiency")

df_summary <- burned_area_scaled %>%
  dplyr::select(year, burned_area_pred) %>%
  #filter(year > -10600) %>%
  mutate(century_bin = floor(year / 100) * 100) %>%   # bin into 100-year windows
  group_by(century_bin) %>%
  summarise(
    mean_ba = mean(burned_area_pred, na.rm = TRUE),
    ci_low   = quantile(burned_area_pred, 0.025, na.rm = TRUE),
    ci_high  = quantile(burned_area_pred, 0.975, na.rm = TRUE),
    n_obs    = n(),
    .groups  = "drop"
  )

ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_ba)) +
  geom_point(size = 3, alpha=0.2) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.1, alpha=0.2) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin), 
                                              max(df_summary$century_bin), by = 500))) +
  theme_minimal() + 
  labs(x = "Year (100-yr bin)", y = "100 year mean BA | 1 ignition",
       title = "") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme(legend.title = element_blank())
```

##########################################
#####         FIRE COUNT         #########
##########################################

fire_count | lightning = Binomial(n_strikes, p_ignite)

```{r}

#add monthly strike count, n_strikes, to fire df
df <- df %>%
  left_join(
    n_strike_scaled %>% dplyr::select(lat, lon, year, month, n_strikes),
    by = c("lat", "lon", "year", "month"))

#Use strike_count and p_ingite to sample fire count from a binomial distribution
df$fire_count_pred <- mapply(
  function(n, p) rbinom(1, size = n, prob = p),
  df$n_strikes,
  df$p_ignite_rate_cal
)   
  
```

Mean fire count over time
```{r}
hist(filter(df, fire_count_pred>0)$fire_count_pred, main="Holocene fire count")

summary(filter(df, fire_count_pred>0)$fire_count_pred)
     
df_summary <- df %>%
  dplyr::select(year, fire_count_pred) %>%
  #filter(year > -10600) %>%
  mutate(century_bin = floor(year / 100) * 100) %>%   # bin into 100-year windows
  group_by(century_bin) %>%
  summarise(
    mean_fire_count = mean(fire_count_pred, na.rm = TRUE),
    ci_low   = quantile(fire_count_pred, 0.025, na.rm = TRUE),
    ci_high  = quantile(fire_count_pred, 0.975, na.rm = TRUE),
    n_obs    = n(),
    .groups  = "drop"
  )

ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_fire_count)) +
  geom_point(size = 3, alpha=0.2) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.1, alpha=0.2) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin), 
                                              max(df_summary$century_bin), by = 500))) +
  theme_minimal() + 
  labs(x = "Year (100-yr bin)", y = "100 year mean fire count",
       title = "") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme(legend.title = element_blank())
```

#####################################
######### TOTAL BURNED AREA #########
#####################################

total_burned = rgamma(fire_count, burned_area_shape, burned_area_rate)

```{r}
df <- df %>%
  left_join(
    burned_area_scaled %>% dplyr::select(lat, lon, year, month, p_extreme, p_extreme_cal,
                                          mu_extreme, mu_typical, rate_extreme, rate_typical,
                                          phi_extreme, phi_typical),
    by = c("lat", "lon", "year", "month")
  )

n_rows <- nrow(df)
fc <- df$fire_count_pred

# ── Sampling ──────────────────────────────────────────────────────────────────
# Uses p_extreme_cal for classification, but the ORIGINAL (uncalibrated)
# rate_extreme/rate_typical for magnitude - no post-hoc magnitude shift

valid_idx <- which(!is.na(fc) & fc > 0)
rep_idx   <- rep(valid_idx, times = fc[valid_idx])

rep_p_extreme <- rep(df$p_extreme_cal[valid_idx], times = fc[valid_idx])
rep_rate_ext  <- rep(df$rate_extreme[valid_idx], times = fc[valid_idx])
rep_rate_typ  <- rep(df$rate_typical[valid_idx], times = fc[valid_idx])

is_extreme_draw <- rbinom(length(rep_idx), size = 1, prob = rep_p_extreme)

samples <- ifelse(
  is_extreme_draw == 1,
  rgamma(length(rep_idx), shape = df$phi_extreme[1], rate = rep_rate_ext),
  rgamma(length(rep_idx), shape = df$phi_typical[1], rate = rep_rate_typ)
)

row_sums <- tapply(samples, rep_idx, sum)
total_burned_area <- rep(0, n_rows)
total_burned_area[as.integer(names(row_sums))] <- row_sums
df$total_burned_area <- total_burned_area

```

Mean total burned area over time
```{r}
hist(df$total_burned_area, main="Holocene total burned area")

df_summary <- df %>%
  dplyr::select(year, total_burned_area) %>%
  #filter(year > -10600) %>%
  mutate(century_bin = floor(year / 100) * 100) %>%   # bin into 100-year windows
  group_by(century_bin) %>%
  summarise(
    mean_total_ba = mean(total_burned_area, na.rm = TRUE),
    ci_low   = quantile(total_burned_area, 0.025, na.rm = TRUE),
    ci_high  = quantile(total_burned_area, 0.975, na.rm = TRUE),
    n_obs    = n(),
    .groups  = "drop"
  )

ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_total_ba)) +
  geom_point(size = 3, alpha=0.2) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.1, alpha=0.2) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin), 
                                              max(df_summary$century_bin), by = 500))) +
  theme_minimal() + 
  labs(x = "Year (100-yr bin)", y = "100 year mean total gridcell burned area",
       title = "") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme(legend.title = element_blank())
```

Apply burn area cap (limits burn size to gridcell size):
```{r}

# ── Per-gridcell engine: processes ONE gridcell's chronologically-sorted
# time series at a time. Never call this directly on a multi-gridcell
# dataframe - use apply_cap_to_dataset() below instead. ─────────────────────

apply_area_cap <- function(df, raw_col = "total_burned_area", recovery_years = Inf) {
  n <- nrow(df)
  area <- df$area[1]

  capped_burned_area <- numeric(n)
  was_capped         <- logical(n)
  reset_year         <- rep(NA_real_, n)

  cumulative <- 0
  recovery_until_year <- -Inf

  for (i in seq_len(n)) {
    this_year <- df$year[i]
    raw_val   <- df[[raw_col]][i]

    if (this_year < recovery_until_year) {
      capped_burned_area[i] <- 0
      was_capped[i] <- (raw_val > 0)
      next
    }

    remaining_capacity <- area - cumulative
    if (remaining_capacity <= 0) {
      capped_burned_area[i] <- 0
      was_capped[i] <- (raw_val > 0)
      next
    }

    applied_val <- min(raw_val, remaining_capacity)
    capped_burned_area[i] <- applied_val
    was_capped[i] <- (applied_val < raw_val)
    cumulative <- cumulative + applied_val

    if (cumulative >= area) {
      cumulative <- 0
      recovery_until_year <- this_year + recovery_years
      reset_year[i] <- this_year
    }
  }

  df$total_burned_area_capped <- capped_burned_area
  df$burned_area_was_capped   <- was_capped
  df$reset_year               <- reset_year
  df
}

# ── Wrapper: joins the correct per-gridcell area, validates it, sorts
# chronologically WITHIN each gridcell, then applies apply_area_cap()
# separately to each gridcell's own rows. THIS is the function to call. ─────

apply_cap_to_dataset <- function(data, area_lookup, recovery_years = Inf,
                                  lat_round = 4, lon_round = 4) {

  # Drop any pre-existing area column from `data` BEFORE joining, so there's
  # no ambiguity/suffix collision with area_lookup's own `area` column
  if ("area" %in% names(data)) {
    data <- data %>% dplyr::select(-area)
  }

  data <- data %>% mutate(lat = round(lat, lat_round), lon = round(lon, lon_round))
  area_lookup <- area_lookup %>%
    mutate(lat = round(lat, lat_round), lon = round(lon, lon_round)) %>%
    dplyr::select(lat, lon, area) %>%
    distinct(lat, lon, .keep_all = TRUE)

  data <- data %>%
    left_join(area_lookup, by = c("lat", "lon"))

  n_missing <- sum(is.na(data$area))
  if (n_missing > 0) {
    warning(sprintf(
      "%d rows (%.2f%%) had no matching area after the join - these will be excluded from capping.",
      n_missing, 100 * n_missing / nrow(data)
    ))
  }

  area_check <- data %>%
    filter(!is.na(area)) %>%
    group_by(lat, lon) %>%
    summarise(n_distinct_area = n_distinct(area), .groups = "drop") %>%
    filter(n_distinct_area > 1)

  if (nrow(area_check) > 0) {
    stop(sprintf(
      "%d gridcells have more than one distinct area value - fix area_lookup before proceeding.",
      nrow(area_check)
    ))
  }

  data %>%
    filter(!is.na(area)) %>%
    arrange(lat, lon, year, month) %>%
    group_by(lat, lon) %>%
    group_modify(~ apply_area_cap(.x, recovery_years = recovery_years)) %>%
    ungroup()
}
```

```{r}
area <- read.csv(file.path(paleoclimate_data_root, "TraCE_21ka_gridcell_area.csv")) %>% dplyr::select(lat, lon, area_clipped_km2)  %>% rename(area = area_clipped_km2)

df_capped <- apply_cap_to_dataset(
  df, area_lookup = area, recovery_years = 20
)

```


```{r}
emergent_fri <- df_capped %>%
  filter(!is.na(reset_year)) %>%
  distinct(lat, lon, reset_year) %>%
  arrange(lat, lon, reset_year) %>%
  group_by(lat, lon) %>%
  mutate(interval = reset_year - lag(reset_year)) %>%
  ungroup() %>%
  filter(!is.na(interval))

summary(emergent_fri$interval)

hist(emergent_fri$interval, breaks = 30,
     main = "Emergent simulated fire return interval",
     xlab = "Years between full-burn resets")

nrow(filter(emergent_fri, interval<50))
```

```{r}

# ── Per-gridcell: for each row, find the YEAR and BURNED AREA of the next
# fire STRICTLY AFTER that row (not including the row itself). A single
# backward pass, so it's fast even across the full paleo record. ────────────

add_next_fire_info <- function(df) {
  n <- nrow(df)
  next_fire_year <- rep(NA_real_, n)
  next_fire_area <- rep(NA_real_, n)

  carry_year <- NA_real_
  carry_area <- NA_real_

  for (i in seq(n, 1, by = -1)) {
    # Assign BEFORE updating carry, so a fire row never "sees" itself
    next_fire_year[i] <- carry_year
    next_fire_area[i] <- carry_area

    if (df$total_burned_area_capped[i] > 0) {
      carry_year <- df$year[i]
      carry_area <- df$total_burned_area_capped[i]
    }
  }

  df$next_fire_year          <- next_fire_year
  df$next_fire_burned_area   <- next_fire_area
  df
}

# ── Apply per gridcell, same chronological-sort pattern as before ───────────

df_capped_with_next_fire <- df_capped %>%
  arrange(lat, lon, year, month) %>%
  group_by(lat, lon) %>%
  group_modify(~ add_next_fire_info(.x)) %>%
  ungroup()

# ── Build emergent_fri, now including burned area of the first fire post-reset ──

emergent_fri <- df_capped_with_next_fire %>%
  filter(!is.na(reset_year)) %>%
  distinct(lat, lon, reset_year, .keep_all = TRUE) %>%   # keep_all so next_fire_* columns survive
  arrange(lat, lon, reset_year) %>%
  group_by(lat, lon) %>%
  mutate(interval = reset_year - lag(reset_year)) %>%
  ungroup() %>%
  filter(!is.na(interval)) %>%
  dplyr::select(lat, lon, reset_year, interval,
                first_fire_after_reset_year = next_fire_year,
                first_fire_after_reset_burned_area = next_fire_burned_area)%>%
  mutate(lat = round(lat, 4), lon = round(lon, 4)) %>%
  left_join(
    area %>% mutate(lat = round(lat, 4), lon = round(lon, 4)),
    by = c("lat", "lon")) %>% mutate(pct_first_fire_of_gridcell = (first_fire_after_reset_burned_area/area)*100)

filter(emergent_fri, interval<50)

filter(emergent_fri, interval<100 & first_fire_after_reset_burned_area>100)

ggplot(filter(emergent_fri, interval < 500),
       aes(x = interval, y = first_fire_after_reset_burned_area,
           size = pct_first_fire_of_gridcell, color = pct_first_fire_of_gridcell)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "#556B2F", high = "firebrick") +
  guides(
    color = guide_legend(title = "% of gridcell burned\nby first fire"),
    size  = guide_legend(title = "% of gridcell burned\nby first fire")
  ) +
  theme_minimal() +
  ggtitle("Simulated fire return time") +
  ylab("First fire burned area (km²)") +
  xlab("Fire return interval (years)")

summary(filter(emergent_fri, interval < 500)$pct_first_fire_of_gridcell)
```

Convert km2 to acres:
```{r}
df <- df %>%
  mutate(total_burned_area_acres = total_burned_area * 247.105,
         total_burned_area_acres_capped = total_burned_area_capped * 247.105)
```

```{r}
write_parquet(df, file.path(predictions_root, "paleo_fire_lightning_predictions.parquet"))
```


##########################################################################################
#################################### VEGETATION FEEDBACK ################################
##########################################################################################

There are 425.8 million acres in Alaska. Do burned area predictions make sense - ie. in the paleo period does more area burn in a time interval than is plausible given the size of a gridcell. 

```{r}
#within a month, no fire size exceeds a gridcell size - good. 
filter(df, total_burned_area>area) 
#within a year, does total burned area exceed gridcell size? - no, good!
df %>%
  dplyr::select(year, lat, lon, area, total_burned_area) %>%
  group_by(year, lat, lon) %>%
  summarise(
    annual_total = sum(total_burned_area, na.rm = TRUE),
    area = first(area),
    .groups = "drop"
  ) %>%
  filter(annual_total > area)

#within a century? yes, just one gridcell-year: -7400
century_bins <- df %>%
  dplyr::select(year, lat, lon, area, total_burned_area, B, C, U) %>%
  mutate(century_bin = floor(year / 100) * 100) %>%
  group_by(century_bin, lat, lon) %>%
  summarise(
    century_total = sum(total_burned_area, na.rm = TRUE),
    area   = first(area),
    mean_B = mean(B, na.rm = TRUE),
    mean_C = mean(C, na.rm = TRUE),
    mean_U = mean(U, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(percent_burned = century_total / area)

# Alaska-wide mean vegetation cover per century, averaging the per-gridcell century means so every gridcell counts equally
alaska_veg_by_century <- century_bins %>%
  group_by(century_bin) %>%
  summarise(
    alaska_mean_B = mean(mean_B, na.rm = TRUE),
    alaska_mean_C = mean(mean_C, na.rm = TRUE),
    alaska_mean_U = mean(mean_U, na.rm = TRUE),
    .groups = "drop"
  )

#join the Alaska-wide values back onto the gridcell-level table
century_bins <- century_bins %>%
  left_join(alaska_veg_by_century, by = "century_bin")

century_anomaly <- century_bins %>% filter(century_total > area)




```

Look at percent of gridcell burned at varying century bins

```{r}
plot_percent_burned <- function(df,
                                 year_interval = 200,
                                 main = paste0("Percent of gridcell burned over ", year_interval, " year intervals"),
                                 x_label = paste0("Year (", year_interval, "-yr bin)"),
                                 y_label = "Percent",
                                 x_breaks_every = 500) {

  century_bins <- df %>%
    dplyr::select(year, lat, lon, area, total_burned_area, B, C, U) %>%
    mutate(century_bin = floor(year / year_interval) * year_interval) %>%
    group_by(century_bin, lat, lon) %>%
    summarise(
      century_total = sum(total_burned_area, na.rm = TRUE),
      area   = first(area),
      mean_B = mean(B, na.rm = TRUE),
      mean_C = mean(C, na.rm = TRUE),
      mean_U = mean(U, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(percent_burned = century_total / area)

  alaska_veg_by_century <- century_bins %>%
    group_by(century_bin) %>%
    summarise(
      alaska_mean_B = mean(mean_B, na.rm = TRUE),
      alaska_mean_C = mean(mean_C, na.rm = TRUE),
      alaska_mean_U = mean(mean_U, na.rm = TRUE),
      .groups = "drop"
    )

  century_bins <- century_bins %>%
    left_join(alaska_veg_by_century, by = "century_bin")
  
  df_long <- century_bins %>%
  dplyr::select(century_bin, alaska_mean_B, alaska_mean_C, alaska_mean_U, percent_burned) %>%
  pivot_longer(
    cols = c(alaska_mean_B, alaska_mean_C, alaska_mean_U, percent_burned),
    names_to = "metric",
    values_to = "percent"
  )

  p <- ggplot(df_long, aes(x = as.factor(century_bin), y = percent, color=metric)) +
    geom_point(size = 3, alpha = 0.6) +
    scale_x_discrete(breaks = as.character(seq(min(century_bins$century_bin),
                                                max(century_bins$century_bin),
                                                by = x_breaks_every))) +
    theme_minimal() +
    labs(x = x_label, y = y_label, title = main, color = NULL) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

  return(p)
}

p1 <- plot_percent_burned(df, year_interval = 100)
p2 <- plot_percent_burned(df, year_interval = 200)
p3 <- plot_percent_burned(df, year_interval = 300)
p4 <- plot_percent_burned(df, year_interval = 400)
p5 <- plot_percent_burned(df, year_interval = 500)

combined_plot <- p1/p2/p3/p4/p5
combined_plot
# ggsave(
#   filename = file.path(figure_root, "Figure_time_percent_burned_and_cover.png"),
#   plot = combined_plot,
#   width = 8,
#   height = 20,     
#   units = "in",
#   dpi = 600
# )
```


Need to find out how soon a burn area size exceeds the total area of a gridcell to get an idea of whether my fire return time predictions are appropriate. For each gridcell, find the number of years it takes for the sum of each year month to surpass the total area of the gridcell:

```{r}

# ── Step 1: annual burned area per gridcell (sum across months within each year) ──
gridcell_annual <- df %>%
  dplyr::select(lat, lon, year, area, total_burned_area) %>%
  group_by(lat, lon, year) %>%
  summarise(
    annual_burned = sum(total_burned_area, na.rm = TRUE),
    area = first(area),
    .groups = "drop"
  )

# ── Step 2: cumulative burned area over time, per gridcell ────────────────────
gridcell_cumulative <- gridcell_annual %>%
  arrange(lat, lon, year) %>%
  group_by(lat, lon) %>%
  mutate(
    cum_burned   = cumsum(annual_burned),
    exceeded     = cum_burned > area,        # TRUE once cumulative burn > gridcell area
    years_elapsed = row_number()             # 1st year in record = 1, 2nd = 2, etc.
  ) %>%
  ungroup()

# ── Step 3: for each gridcell, find the FIRST year the threshold was exceeded ──
first_exceedance <- gridcell_cumulative %>%
  filter(exceeded) %>%
  group_by(lat, lon) %>%
  slice_min(year, n = 1) %>%   # earliest exceedance year per gridcell
  ungroup() %>%
  dplyr::select(lat, lon, year, years_elapsed, cum_burned, area)

#the calendar year its cumulative burned area first surpassed its own total area, years_elapsed (how many years into that gridcell's record this took), cum_burned at that point, and gridcell area:
first_exceedance

sum_next_100yr <- function(lat_val, lon_val, exceed_year) {
  gridcell_annual %>%
    filter(lat == lat_val, lon == lon_val,
           year > exceed_year, year <= exceed_year + 100) %>%
    summarise(total = sum(annual_burned, na.rm = TRUE)) %>%
    pull(total)
}

first_exceedance <- first_exceedance %>%
  mutate(
    burned_next_100yr = pmap_dbl(
      list(lat, lon, year),
      sum_next_100yr
    )
  )

first_exceedance
```


if burned area > gridcell area over a 100 year period, nothing can burn... what model component should this suppress? I think burned area... 

Factors to consider: 
- time since last fire
- time to regrowth 
- vegetation type/cover. I have C/B/U cover. 
- fire severity
- climate, region, elevation

Ideas:
regrowth rate for boreal forest: 



```{r}
summary(df$r_strike)
summary(df$n_strikes)
summary(df$fire_count_pred)
summary(df$total_burned_area)
```

```{r}
var <-  "n_strikes"
y_lable <- "Strike count" # (gridcell-month predictions)" #, 100-yr bins, log scale)"
main <- "(a)"

#training period summaries:
train_n_strikes_annual_totals <- df_modern %>%
  group_by(year) %>%
  summarise(annual_total = sum(n_strikes_obs, na.rm = TRUE), .groups = "drop")

max_train_n_strikes <- max(train_n_strikes_annual_totals$annual_total)
mean_train_n_strikes <- mean(train_n_strikes_annual_totals$annual_total)
median_train_n_strikes <- median(train_n_strikes_annual_totals$annual_total)

hline_df <- tibble::tibble(
  value = c(max_train_n_strikes, mean_train_n_strikes, median_train_n_strikes),
  label = c("2002-2011 max", "2002-2011 mean", "2002-2011 median")
)

#collapse to one Alaska-wide total per summer (year), sum across all gridcells and all three summer months.
df_year <- df %>%
  dplyr::select(year, n_strikes) %>%
  rename(prediction = n_strikes) %>%
  group_by(year) %>%
  summarise(annual_total = sum(prediction, na.rm = TRUE), .groups = "drop")

df_summary <- df_year %>%
  mutate(century_bin = floor(year / 100) * 100) %>%
  group_by(century_bin) %>%
  summarise(
    mean      = mean(annual_total, na.rm = TRUE),
    median    = median(annual_total, na.rm = TRUE),
    ci_low    = quantile(annual_total, 0.025, na.rm = TRUE),
    ci_high   = quantile(annual_total, 0.975, na.rm = TRUE),
    max       = max(annual_total, na.rm = TRUE),
    pct_zero  = mean(annual_total == 0, na.rm = TRUE),
    n_obs     = n(),   # now = number of YEARS in the bin (~100), not gridcell-month rows
    .groups   = "drop"
  )


df_long <- df_summary %>%
  dplyr::select(century_bin, mean, max, median) %>%
  pivot_longer(
    cols = c(mean, max, median),
    names_to = "metric",
    values_to = "value"
  )

p1 <- ggplot() +
  # Mean with CI error bars (kept separate since only mean has ci_low/ci_high)
  geom_point(data = df_summary, aes(x = as.factor(century_bin), y = mean, color = "mean"),
             size = 3, alpha = 0.6) +
  geom_errorbar(data = df_summary, aes(x = as.factor(century_bin), ymin = ci_low, ymax = ci_high),
                width = 0.1, alpha = 0.2, color = "black") +
  # Max points
  geom_point(data = df_long %>% filter(metric != "mean"),
             aes(x = as.factor(century_bin), y = value, color = metric),
             size = 3, alpha = 0.6) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin),
                                              max(df_summary$century_bin), by = 500))) +
  scale_y_log10(breaks = scales::breaks_log(n = 5), labels = scales::label_number()) +
  scale_color_manual(values = c(
    "mean"      = "black",
    "median" = "blue",
    "max" = "firebrick"
  )) +
  geom_hline(data = hline_df, aes(yintercept = value, color = label),
             linetype = "dotted", linewidth = 0.5) +
  theme_minimal() +
  labs(x = "Year (100-yr bin)", y = y_lable,
       title = main, color = NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p1
```

```{r}
var <-  "fire_count_pred"
y_lable <- "Fire count" # (Alaska-wide annual predictions)" #, 100-yr bins, log scale)"
main <- "(b)"

#training period summaries:
train_fire_count_annual_totals <- df_modern %>%
  group_by(year) %>%
  summarise(annual_total = sum(fire_count, na.rm = TRUE), .groups = "drop")

max_train_fire_count <- max(train_fire_count_annual_totals$annual_total)
mean_train_fire_count <- mean(train_fire_count_annual_totals$annual_total)
median_train_fire_count <- median(train_fire_count_annual_totals$annual_total)

hline_df <- tibble::tibble(
  value = c(max_train_fire_count, mean_train_fire_count, median_train_fire_count),
  label = c("2002-2011 max", "2002-2011 mean", "2002-2011 median")
)

#collapse to one Alaska-wide total per summer (year), sum across all gridcells and all three summer months.
df_year <- df %>%
  dplyr::select(year, fire_count_pred) %>%
  rename(prediction = fire_count_pred) %>%
  group_by(year) %>%
  summarise(annual_total = sum(prediction, na.rm = TRUE), .groups = "drop")

df_summary <- df_year %>%
  mutate(century_bin = floor(year / 100) * 100) %>%
  group_by(century_bin) %>%
  summarise(
    mean      = mean(annual_total, na.rm = TRUE),
    median    = median(annual_total, na.rm = TRUE),
    ci_low    = quantile(annual_total, 0.025, na.rm = TRUE),
    ci_high   = quantile(annual_total, 0.975, na.rm = TRUE),
    max       = max(annual_total, na.rm = TRUE),
    pct_zero  = mean(annual_total == 0, na.rm = TRUE),
    n_obs     = n(),   # now = number of YEARS in the bin (~100), not gridcell-month rows
    .groups   = "drop"
  )


df_long <- df_summary %>%
  dplyr::select(century_bin, mean, max, median) %>%
  pivot_longer(
    cols = c(mean, max, median),
    names_to = "metric",
    values_to = "value"
  )

p2 <- ggplot() +
  # Mean with CI error bars (kept separate since only mean has ci_low/ci_high)
  geom_point(data = df_summary, aes(x = as.factor(century_bin), y = mean, color = "mean"),
             size = 3, alpha = 0.6) +
  geom_errorbar(data = df_summary, aes(x = as.factor(century_bin), ymin = ci_low, ymax = ci_high),
                width = 0.1, alpha = 0.2, color = "black") +
  # Max points
  geom_point(data = df_long %>% filter(metric != "mean"),
             aes(x = as.factor(century_bin), y = value, color = metric),
             size = 3, alpha = 0.6) +
  geom_hline(data = hline_df, aes(yintercept = value, color = label),
             linetype = "dotted", linewidth = 0.5) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin),
                                              max(df_summary$century_bin), by = 500))) +
  scale_y_log10(breaks = scales::breaks_log(n = 6), labels = scales::label_number()) +
  scale_color_manual(values = c(
    "mean"      = "black",
    "median" = "blue",
    "max" = "firebrick"
  )) +
  theme_minimal() +
  labs(x = "Year (100-yr bin)", y = y_lable,
       title = main, color = NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p2
```

```{r}

#training period summaries:
#convert km to acres
df_modern <- df_modern %>%
  mutate(obs_total_burned_area_acres = burned_area_km2_total * 247.105)

train_ba_annual_totals <- df_modern %>%
  group_by(year) %>%
  summarise(annual_total = sum(obs_total_burned_area_acres, na.rm = TRUE), .groups = "drop")

max_train_ba <- max(train_ba_annual_totals$annual_total)
mean_train_ba <- mean(train_ba_annual_totals$annual_total)
median_train_ba <- median(train_ba_annual_totals$annual_total)

hline_df <- tibble::tibble(
  value = c(max_train_ba, mean_train_ba, median_train_ba),
  label = c("2002-2011 max", "2002-2011 mean", "2002-2011 median")
)

#convert trace burned area km to acres
df_capped <- df_capped %>%
  mutate(total_burned_area_acres_capped = total_burned_area_capped * 247.105)

var <-  "total_burned_area_acres"
y_lable <- "Burned area (acres)" # (Alaska-wide annual total)" # (100-yr summary, log scale)"
main <- "(c)"

#collapse to one Alaska-wide total per summer (year), sum across all gridcells and all three summer months.
df_year <- df_capped %>%
  dplyr::select(year, total_burned_area_acres_capped) %>%
  rename(prediction = total_burned_area_acres_capped) %>%
  group_by(year) %>%
  summarise(annual_total = sum(prediction, na.rm = TRUE), .groups = "drop")

df_summary <- df_year %>%
  mutate(century_bin = floor(year / 100) * 100) %>%
  group_by(century_bin) %>%
  summarise(
    mean      = mean(annual_total, na.rm = TRUE),
    median    = median(annual_total, na.rm = TRUE),
    ci_low    = quantile(annual_total, 0.025, na.rm = TRUE),
    ci_high   = quantile(annual_total, 0.975, na.rm = TRUE),
    max       = max(annual_total, na.rm = TRUE),
    pct_zero  = mean(annual_total == 0, na.rm = TRUE),
    n_obs     = n(),   # now = number of YEARS in the bin (~100), not gridcell-month rows
    .groups   = "drop"
  )

df_long <- df_summary %>%
  dplyr::select(century_bin, mean, max, median) %>%
  pivot_longer(
    cols = c(mean, max, median),
    names_to = "metric",
    values_to = "value"
  )

p3 <- ggplot() +
  geom_point(data = df_summary, aes(x = as.factor(century_bin), y = mean, color = "mean"),
             size = 3, alpha = 0.6) +
  geom_errorbar(data = df_summary, aes(x = as.factor(century_bin), ymin = ci_low, ymax = ci_high),
                width = 0.1, alpha = 0.2, color = "black") +
  geom_point(data = df_long %>% filter(metric != "mean"),
             aes(x = as.factor(century_bin), y = value, color = metric),
             size = 3, alpha = 0.6) +
  geom_hline(data = hline_df, aes(yintercept = value, color = label),
             linetype = "dotted", linewidth = 0.5) +
  scale_x_discrete(breaks = as.character(seq(min(df_summary$century_bin),
                                              max(df_summary$century_bin), by = 500))) +
  scale_y_log10(breaks = scales::breaks_log(n = 6), labels = scales::label_number()) +
  scale_color_manual(values = c(
    "mean"      = "black",
    "median" = "blue",
    "max" = "firebrick"
  )) +
  theme_minimal() +
  labs(x = "Year (100-yr bin)", y = y_lable,
       title = main, color = NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p3
```

```{r}
p1/p2/p3
combined_plot <- p1/p2/p3
ggsave(
  filename = file.path(figure_root, "Figure_time_paleo_predictions.png"),
  plot = combined_plot,
  width = 8,
  height = 12,     
  units = "in",
  dpi = 600
)
```



