---
title: "Trace_gridcell_area"
author: "Charlotte Uden"
date: "2026-08-10"
output: html_document
---


```{r}
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggplot2)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

```{r}
trace_cells <- readRDS(file.path(rds_root, "trace_cells.rds"))
```

# ── 1. Get Alaska's coastline/boundary polygon ───────────────────────────────
# rnaturalearth's state-level data includes Alaska as a US state polygon,
# already handling the coastline (not just a bounding box).
```{r}

usa_states <- ne_states(country = "United States of America", returnclass = "sf")
alaska_boundary <- usa_states %>% filter(name == "Alaska")

if (nrow(alaska_boundary) == 0) {
  stop("Alaska boundary not found — check the 'name' field in usa_states (try View(usa_states) to inspect).")
}

```

# ── 2. Reproject to an equal-area CRS suitable for Alaska ───────────────────
# EPSG:3338 (NAD83 / Alaska Albers) is the standard equal-area projection for
# Alaska and handles the antimeridian (180° longitude) crossing cleanly —
# critical since Alaska's Aleutian chain crosses the date line, and plain
# lat/lon-based area math breaks down there.
```{r}

ALASKA_CRS <- 3338

alaska_proj <- st_transform(alaska_boundary, ALASKA_CRS)
```


# ── 3. Rebuild grid cell polygons from your TraCE lat/lon centers ───────────
# Reuses dlat/dlon computed in the previous script (from trace_lats/trace_lons).
# If not already in your environment, recompute:
```{r}
trace_lats <- sort(unique(trace_cells$lat))
trace_lons <- sort(unique(trace_cells$lon))
dlat <- median(diff(trace_lats))
dlon <- median(diff(trace_lons))

# Unique gridcell centers actually present in the data
grid_centers <- n_strike_scaled %>%
  dplyr::select(lat, lon) %>%
  distinct()

cat(sprintf("Building %d gridcell polygons...\n", nrow(grid_centers)))

# Build one rectangular polygon per gridcell center, in lon/lat space (EPSG:4326)
make_cell_polygon <- function(lon, lat, dlon, dlat) {
  lon_min <- lon - dlon / 2
  lon_max <- lon + dlon / 2
  lat_min <- max(lat - dlat / 2, -90)
  lat_max <- min(lat + dlat / 2,  90)

  st_polygon(list(rbind(
    c(lon_min, lat_min),
    c(lon_max, lat_min),
    c(lon_max, lat_max),
    c(lon_min, lat_max),
    c(lon_min, lat_min)
  )))
}

grid_polygons <- lapply(seq_len(nrow(grid_centers)), function(i) {
  make_cell_polygon(grid_centers$lon[i], grid_centers$lat[i], dlon, dlat)
})

grid_sf <- st_sf(
  lat = grid_centers$lat,
  lon = grid_centers$lon,
  geometry = st_sfc(grid_polygons, crs = 4326)
)
```
# NOTE: if your TraCE lon values are in 0-360 convention rather than -180/180,
# reproject/wrap them before this step, e.g.:
#   grid_centers$lon <- ifelse(grid_centers$lon > 180, grid_centers$lon - 360, grid_centers$lon)
# Do this consistently for both grid_centers and n_strike_scaled$lon before joining later.

# ── 4. Reproject grid polygons to the same equal-area CRS ───────────────────
```{r}
grid_proj <- st_transform(grid_sf, ALASKA_CRS)

# Fix any invalid geometries before intersecting (common with lon/lat rectangles
# reprojected near the antimeridian or poles)
grid_proj <- st_make_valid(grid_proj)
alaska_proj <- st_make_valid(alaska_proj)
```
# ── 5. Clip each gridcell polygon to Alaska's actual land boundary ──────────
# st_intersection cuts each rectangle down to only the portion overlapping land.
# Cells entirely outside Alaska will simply not appear in the output (0 area).
```{r}
alaska_union <- st_union(alaska_proj)   # dissolve multi-part Alaska geometry into one

clipped <- st_intersection(grid_proj, st_sf(geometry = alaska_union))
```
# ── 6. Compute clipped area in km² ───────────────────────────────────────────
```{r}
clipped <- clipped %>%
  mutate(area_clipped_km2 = as.numeric(st_area(geometry)) / 1e6)   # m² -> km²

clipped_areas <- clipped %>%
  st_drop_geometry() %>%
  dplyr::select(lat, lon, area_clipped_km2)

cat(sprintf("Gridcells with land overlap: %d / %d\n",
            nrow(clipped_areas), nrow(grid_centers)))
```
# ── 7. Join clipped area back to full grid (cells with 0 land overlap get 0) ─
```{r}
grid_with_area <- grid_centers %>%
  left_join(clipped_areas, by = c("lat", "lon")) %>%
  mutate(area_clipped_km2 = ifelse(is.na(area_clipped_km2), 0, area_clipped_km2))

cat("\nClipped area summary (km²):\n")
print(summary(grid_with_area$area_clipped_km2))
```

# ═══════════════════════════════════════════════════════════════════════════
# PLOT: Alaska outline, clipped grid, and gridcell center points
# ═══════════════════════════════════════════════════════════════════════════
```{r}
# Reproject the full (unclipped) grid polygons too, for outline reference
grid_full_proj <- grid_proj   # already reprojected above

# Center points, as an sf object in the same projected CRS
centers_sf <- st_as_sf(grid_centers, coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(ALASKA_CRS)

p <- ggplot() +
  # Alaska outline
  geom_sf(data = alaska_proj, fill = NA, color = "black", linewidth = 0.6) +
  # Full (unclipped) grid cell borders, light, for reference
  geom_sf(data = grid_full_proj, fill = NA, color = "grey70", linewidth = 0.2) +
  # Clipped gridcells (actual land-overlap polygons), shaded
  geom_sf(data = clipped, aes(fill = area_clipped_km2), color = "steelblue", linewidth = 0.3, alpha = 0.5) +
  # Gridcell center points
  geom_sf(data = centers_sf, color = "red", size = 0.8) +
  scale_fill_viridis_c(name = "Clipped\narea (km²)", option = "viridis") +
  labs(
    title = "TraCE-21ka gridcells clipped to Alaska coastline",
    subtitle = "Grey = full gridcell extent, blue = land-clipped area, red = gridcell centers"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

print(p)

# ggsave("trace_alaska_clipped_grid.png", p, width = 9, height = 7, dpi = 300)
```

```{r}
write.csv(grid_with_area, file.path(paleoclimate_data_root, "TraCE_21ka_gridcell_area.csv"))
```

