---
title: "Lightning Prediction using a Gamma GLM"
author: "Charlotte Uden"
date: "2026-06-03"
output: html_document
---

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(patchwork)
library(ggplot2)
library(gridExtra)
library(terra)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
```

Read in drivers
```{r}
df <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))
head(df)
```

look at data
```{r}
hist(df$strikes)
summary(df$strikes)
sum(df$strikes == 0, na.rm = TRUE)
nrow(df)
ggplot(data=filter(df, year==2002), aes(x=lon, y=lat, color=strikes)) + geom_point()
```

The random forest indicated that every variable contributes something to lightning strike rate, nothing is 'hurting' the model.
However, we'll try testing the Gamma GLM with and without wind (lowest on both %IncMSE and IncNodePurity) to see if excluding it increases model performance. 

Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]
```


#Standardize the data
```{r}
#-----
mu_precip <- mean(train$precip)
sigma_precip <- sd(train$precip)
mu_tair <- mean(train$tair)
sigma_tair <- sd(train$tair)
mu_wind <- mean(train$wind)
sigma_wind <- sd(train$wind)
mu_swr <- mean(train$swr)
sigma_swr <- sd(train$swr)
mu_sp <- mean(train$sp)
sigma_sp <- sd(train$sp)
mu_rh <- mean(train$rh)
sigma_rh <- sd(train$rh)

#store means and sds
mu <- c(mu_precip, mu_tair, mu_wind, mu_swr, mu_sp, mu_rh)
sigma <- c(sigma_precip, sigma_tair, sigma_wind, sigma_swr, sigma_sp, sigma_rh)
var <- c("precip", "tair", "wind", "swr", "sp", "rh")
mean_sd <- data.frame(cbind(var, mu, sigma))
write.csv(mean_sd, file.path(mu_sigma_root, "mu_sigma_lightning_predictors.csv"))
#-----

train <- train %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

train <- train[,c("strikes","precip","tair", "wind", "swr", "sp", "rh")]
head(train)

test <- test %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) %>%
  mutate(lon = lon) %>%
  mutate(lat = lat)

test <- test[,c("lon", "lat", "year", "strikes", "precip", "tair", "wind", "swr", "sp", "rh")]

plot(test)

hist(test$strikes)
```

The gamma distribution cannot handle 0 values. the monthly summaries do have 0's, so we need to add a very small value so that the gamma distribution can handle the data:
```{r}
epsilon <- 1e-4  # small value
train$strikes<- train$strikes + epsilon
```

Functions for calculating performance metrics
```{r}
# ============================================================
# Bootstrap CI versions of each performance metric
# Each function returns: estimate, ci_low, ci_high
# Percentile CI: 2.5% and 97.5% from n_boot bootstrap samples
# Resampling is done on the TEST SET rows / paired vectors
# ============================================================

# ---- helpers (small + boring, used by all) ----
.rmse <- function(obs, pred) {
  ok <- is.finite(obs) & is.finite(pred)
  obs <- obs[ok]; pred <- pred[ok]
  if (length(obs) == 0) return(NA_real_)
  sqrt(mean((obs - pred)^2))
}

.safe_cor <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]; y <- y[ok]
  if (length(x) < 2) return(NA_real_)
  if (sd(x) == 0 || sd(y) == 0) return(NA_real_)
  cor(x, y, use = "complete.obs")
}

.ci_from_boot <- function(boot_vals, conf = 0.95) {
  alpha <- (1 - conf) / 2
  boot_vals <- boot_vals[is.finite(boot_vals)]
  if (length(boot_vals) == 0) {
    return(list(ci_low = NA_real_, ci_high = NA_real_))
  }
  list(
    ci_low  = as.numeric(stats::quantile(boot_vals, probs = alpha, na.rm = TRUE)),
    ci_high = as.numeric(stats::quantile(boot_vals, probs = 1 - alpha, na.rm = TRUE))
  )
}

# ============================================================
# 1) Normalized RMSE (normalized by SD of observed)
# ============================================================
metric_nrmse_ci <- function(obs, pred, n_boot = 1000, conf = 0.95, seed = 123) {
  ok <- is.finite(obs) & is.finite(pred)
  obs <- obs[ok]; pred <- pred[ok]
  N <- length(obs)

  if (N == 0) return(list(estimate = NA_real_, ci_low = NA_real_, ci_high = NA_real_))

  denom <- sd(obs)
  est <- if (!is.finite(denom) || denom == 0) NA_real_ else .rmse(obs, pred) / denom

  set.seed(seed)
  boot <- replicate(n_boot, {
    idx <- sample.int(N, size = N, replace = TRUE)
    o <- obs[idx]; p <- pred[idx]
    d <- sd(o)
    if (!is.finite(d) || d == 0) return(NA_real_)
    .rmse(o, p) / d
  })

  ci <- .ci_from_boot(boot, conf = conf)
  list(estimate = est, ci_low = ci$ci_low, ci_high = ci$ci_high)
}

# ============================================================
# 2) Correlation coefficient (Pearson)
# ============================================================
metric_cor_ci <- function(obs, pred, n_boot = 1000, conf = 0.95, seed = 123) {
  ok <- is.finite(obs) & is.finite(pred)
  obs <- obs[ok]; pred <- pred[ok]
  N <- length(obs)

  if (N == 0) return(list(estimate = NA_real_, ci_low = NA_real_, ci_high = NA_real_))

  est <- .safe_cor(obs, pred)

  set.seed(seed)
  boot <- replicate(n_boot, {
    idx <- sample.int(N, size = N, replace = TRUE)
    .safe_cor(obs[idx], pred[idx])
  })

  ci <- .ci_from_boot(boot, conf = conf)
  list(estimate = est, ci_low = ci$ci_low, ci_high = ci$ci_high)
}

# ============================================================
# 3) Skill score (Perkins S-score)
# ============================================================
metric_sscore_ci <- function(obs, pred, nbins = 15,
                             n_boot = 1000, conf = 0.95, seed = 123) {
  ok <- is.finite(obs) & is.finite(pred)
  obs <- obs[ok]; pred <- pred[ok]
  N <- length(obs)

  if (N == 0) return(list(estimate = NA_real_, ci_low = NA_real_, ci_high = NA_real_))

  calculate_sscore <- function(obs, pred, nbins = 15) {
    min_val <- min(c(obs, pred))
    max_val <- max(c(obs, pred))
    if (!is.finite(min_val) || !is.finite(max_val) || min_val == max_val) return(NA_real_)

    breaks <- seq(min_val, max_val, length.out = nbins + 1)
    hist_obs  <- hist(obs,  breaks = breaks, plot = FALSE)
    hist_pred <- hist(pred, breaks = breaks, plot = FALSE)

    prob_obs  <- hist_obs$counts  / sum(hist_obs$counts)
    prob_pred <- hist_pred$counts / sum(hist_pred$counts)

    sum(pmin(prob_obs, prob_pred))
  }

  est <- calculate_sscore(obs, pred, nbins = nbins)

  set.seed(seed)
  boot <- replicate(n_boot, {
    idx <- sample.int(N, size = N, replace = TRUE)
    calculate_sscore(obs[idx], pred[idx], nbins = nbins)
  })

  ci <- .ci_from_boot(boot, conf = conf)
  list(estimate = est, ci_low = ci$ci_low, ci_high = ci$ci_high)
}

# ============================================================
# 4) Spatial correlation
#    (correlation of gridcell means across space)
#    test must contain lon, lat, and strikes (obs)
# ============================================================
metric_spatial_cor_ci <- function(test, y_pred,
                                  n_boot = 1000, conf = 0.95, seed = 123) {
  test <- dplyr::as_tibble(test)
  test$pred <- y_pred

  test <- test %>% dplyr::filter(is.finite(strikes), is.finite(pred))
  N <- nrow(test)
  if (N == 0) return(list(estimate = NA_real_, ci_low = NA_real_, ci_high = NA_real_))

  # point estimate on full test set
  spatial_means <- test %>%
    dplyr::group_by(lon, lat) %>%
    dplyr::summarise(
      obs_mean  = mean(strikes, na.rm = TRUE),
      pred_mean = mean(pred,    na.rm = TRUE),
      .groups = "drop"
    )

  est <- .safe_cor(spatial_means$obs_mean, spatial_means$pred_mean)

  # bootstrap resampling rows of the test set
  set.seed(seed)
  boot <- replicate(n_boot, {
    idx <- sample.int(N, size = N, replace = TRUE)
    tb <- test[idx, , drop = FALSE]

    sm <- tb %>%
      dplyr::group_by(lon, lat) %>%
      dplyr::summarise(
        obs_mean  = mean(strikes, na.rm = TRUE),
        pred_mean = mean(pred,    na.rm = TRUE),
        .groups = "drop"
      )

    .safe_cor(sm$obs_mean, sm$pred_mean)
  })

  ci <- .ci_from_boot(boot, conf = conf)
  list(estimate = est, ci_low = ci$ci_low, ci_high = ci$ci_high)
}

# ============================================================
# 5) Anomaly Correlation Coefficient (ACC)
#    computed from annual means, then anomalies, then correlation
# ============================================================
metric_acc_ci <- function(year, obs, pred, n_boot = 1000, conf = 0.95, seed = 123) {
  df <- dplyr::tibble(year = year, obs = obs, pred = pred) %>%
    dplyr::filter(is.finite(year), is.finite(obs), is.finite(pred))

  N <- nrow(df)
  if (N == 0) return(list(estimate = NA_real_, ci_low = NA_real_, ci_high = NA_real_))

  calc_acc_core <- function(df_in) {
    annual <- df_in %>%
      dplyr::group_by(year) %>%
      dplyr::summarise(
        obs  = mean(obs,  na.rm = TRUE),
        pred = mean(pred, na.rm = TRUE),
        .groups = "drop"
      )
    obs_anom  <- annual$obs  - mean(annual$obs,  na.rm = TRUE)
    pred_anom <- annual$pred - mean(annual$pred, na.rm = TRUE)
    .safe_cor(obs_anom, pred_anom)
  }

  est <- calc_acc_core(df)

  set.seed(seed)
  boot <- replicate(n_boot, {
    idx <- sample.int(N, size = N, replace = TRUE)
    calc_acc_core(df[idx, , drop = FALSE])
  })

  ci <- .ci_from_boot(boot, conf = conf)
  list(estimate = est, ci_low = ci$ci_low, ci_high = ci$ci_high)
}

# ============================================================
# 6) Normalized RMSE of anomalies (nRMSE_anom)
#    = RMSE(annual anomalies) / SD(observed annual anomalies)
# ============================================================
metric_nrmse_anom_ci <- function(year, obs, pred, n_boot = 1000, conf = 0.95, seed = 123) {
  df <- dplyr::tibble(year = year, obs = obs, pred = pred) %>%
    dplyr::filter(is.finite(year), is.finite(obs), is.finite(pred))

  N <- nrow(df)
  if (N == 0) return(list(estimate = NA_real_, ci_low = NA_real_, ci_high = NA_real_))

  calc_nrmse_anom_core <- function(df_in) {
    annual <- df_in %>%
      dplyr::group_by(year) %>%
      dplyr::summarise(
        obs  = mean(obs,  na.rm = TRUE),
        pred = mean(pred, na.rm = TRUE),
        .groups = "drop"
      )
    obs_anom  <- annual$obs  - mean(annual$obs,  na.rm = TRUE)
    pred_anom <- annual$pred - mean(annual$pred, na.rm = TRUE)

    denom <- sd(obs_anom, na.rm = TRUE)
    if (!is.finite(denom) || denom == 0) return(NA_real_)
    sqrt(mean((obs_anom - pred_anom)^2, na.rm = TRUE)) / denom
  }

  est <- calc_nrmse_anom_core(df)

  set.seed(seed)
  boot <- replicate(n_boot, {
    idx <- sample.int(N, size = N, replace = TRUE)
    calc_nrmse_anom_core(df[idx, , drop = FALSE])
  })

  ci <- .ci_from_boot(boot, conf = conf)
  list(estimate = est, ci_low = ci$ci_low, ci_high = ci$ci_high)
}

# ============================================================
# Example calls
# ============================================================
# nrmse_res  <- metric_nrmse_ci(test$strikes, y_pred)
# cor_res    <- metric_cor_ci(test$strikes, y_pred)
# sscore_res <- metric_sscore_ci(test$strikes, y_pred, nbins = 15)
# spcor_res  <- metric_spatial_cor_ci(test, y_pred)
# acc_res    <- metric_acc_ci(test$year, test$strikes, y_pred)
# nrmseA_res <- metric_nrmse_anom_ci(test$year, test$strikes, y_pred)

```

Train Gamma GLMs
```{r}
#swr + tair + rh + wind + precip +sp
G13 <- glm(strikes ~ swr + tair + rh + precip + sp + wind,
                 data = train,
                 family = Gamma(link =log))

# Save the model
saveRDS(G13, file.path(rds_root, "G13_gamma_glm_lightning.rds"))

# Verify it saved correctly by reloading and checking
G13_check <- readRDS(file.path(rds_root, "G13_gamma_glm_lightning.rds"))
cat("Model class:", class(G13_check), "\n")
cat("Formula:    ", deparse(formula(G13_check)), "\n")
cat("Predictors: ", paste(names(coef(G13_check))[-1], collapse = ", "), "\n")

a <- G13$coefficients[1]
b <- G13$coefficients[2]
c <- G13$coefficients[3]
d <- G13$coefficients[4]
e <- G13$coefficients[5]
f <- G13$coefficients[6]
g <- G13$coefficients[7]

G13_coefficients <- c(a, b, c, d, e, f, g)

y_pred_G13 <- exp(a + b*test$swr + c*test$tair + d*test$rh + e*test$precip + f*test$sp + g*test$wind)

nrmse_res_G13  <- metric_nrmse_ci(test$strikes, y_pred_G13)
cor_res_G13    <- metric_cor_ci(test$strikes, y_pred_G13)
sscore_res_G13 <- metric_sscore_ci(test$strikes, y_pred_G13, nbins = 15)
spcor_res_G13  <- metric_spatial_cor_ci(test, y_pred_G13)
acc_res_G13  <- metric_acc_ci(test$year, test$strikes, y_pred_G13)
nrmseA_res_G13 <- metric_nrmse_anom_ci(test$year, test$strikes, y_pred_G13)

#EXCLUDE WIND

#swr + tair + rh + precip +sp
G12 <- glm(strikes ~ swr + tair + rh + precip + sp,
                 data = train,
                 family = Gamma(link =log))

a <- G12$coefficients[1]
b <- G12$coefficients[2]
c <- G12$coefficients[3]
d <- G12$coefficients[4]
e <- G12$coefficients[5]
f <- G12$coefficients[6]

G12_coefficients <- c(a, b, c, d, e, f)

y_pred_G12 <- exp(a + b*test$swr + c*test$tair + d*test$rh + e*test$precip + f*test$sp)

nrmse_res_G12  <- metric_nrmse_ci(test$strikes, y_pred_G12)
cor_res_G12    <- metric_cor_ci(test$strikes, y_pred_G12)
sscore_res_G12 <- metric_sscore_ci(test$strikes, y_pred_G12, nbins = 15)
spcor_res_G12  <- metric_spatial_cor_ci(test, y_pred_G12)
acc_res_G12  <- metric_acc_ci(test$year, test$strikes, y_pred_G12)
nrmseA_res_G12 <- metric_nrmse_anom_ci(test$year, test$strikes, y_pred_G12)
```

# B13: Short Wave Radiation + Temperature + Precipitation + Wind + Precipitation + Surface Pressure

Lightning strike rate modeled from a gamma distribution where mu and sigma are dependent on climate: shortwave radiation,  temperature, relative humidity, wind, precipitation, and surface pressure
. 
    S ~ dgamma( alpha , beta ),
    alpha <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W + f_alpha * P + g_alpha * SP,
    beta <- a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W + f_beta * P + g_beta * SP,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  W = train$wind,
  P = train$precip,
  SP = train$sp
)


# Compile and fit the Stan model
init_list <- function() {
  list(a_alpha = 1, b_alpha = 0.1, c_alpha = 0.1, d_alpha = 0.1, e_alpha = 0.1, f_alpha = 0.1, g_alpha = 0.1,
       a_beta = 1, b_beta = 0.1, c_beta = 0.1, d_beta = 0.1, e_beta = 0.1, f_beta = 0.1, g_beta = 0.1)
}


if (!file.exists(file.path(rds_root, "Lightning/B13_gamma_bayes.rds"))) {
  B13 <- stan(file = "01_gamma_swr_tair_rh_wind_precip_sp.stan", 
            data = dat,
            iter = 2000, 
            chains = 4, 
            init = init_list,
            cores = min(parallel::detectCores(), 4))
    saveRDS(B13, file = file.path(rds_root, "Lightning/B13_gamma_bayes.rds"))
    } else {
      B13 <- readRDS(file.path(rds_root, "Lightning/B13_gamma_bayes.rds"))
    }


```


```{r}

#print(B13, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "f_alpha", "g_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta", "f_beta", "g_beta"))

traceplot(B13)

parameter_stats <- summary(B13)$summary[, c("mean", "sd")][1:14, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

# ggplot(summary_table, aes(x = parameter, y = mean)) +
#   geom_point(color = "blue", size = 3) +  # Point for the mean
#   geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
#   labs(y = "Parameter Estimate", x = "Parameter") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   ggtitle("Parameter Means and SDs from Stan Model")

#store mean posterior sample for coefficient table
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
f_alpha <- summary_table["f_alpha", "mean"]
g_alpha <- summary_table["g_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]
f_beta <- summary_table["f_beta", "mean"]
g_beta <- summary_table["g_beta", "mean"]

B13_alpha_coefficients <- c(a_alpha, b_alpha, c_alpha, d_alpha, e_alpha, f_alpha, g_alpha)

B13_beta_coefficients <- c(a_beta, b_beta, c_beta, d_beta, e_beta, f_beta, g_beta)
```

Make predictions
```{r}
#Extract posterior samples
posterior_samples <- as.data.frame(B13)

#Extract individual parameters
a_alpha_samples <- posterior_samples$a_alpha
b_alpha_samples <- posterior_samples$b_alpha
c_alpha_samples <- posterior_samples$c_alpha
d_alpha_samples <- posterior_samples$d_alpha
e_alpha_samples <- posterior_samples$e_alpha
f_alpha_samples <- posterior_samples$f_alpha
g_alpha_samples <- posterior_samples$g_alpha
a_beta_samples <- posterior_samples$a_beta
b_beta_samples <- posterior_samples$b_beta
c_beta_samples <- posterior_samples$c_beta
d_beta_samples <- posterior_samples$d_beta
e_beta_samples <- posterior_samples$e_beta
f_beta_samples <- posterior_samples$f_beta
g_beta_samples <- posterior_samples$g_beta

#Number of posterior samples
num_samples <- length(a_alpha_samples)

#Number of test observations
num_test <- length(test$swr)

#Create vectors to store sampled alpha, beta, and predictions
sampled_alpha <- numeric(num_test)
sampled_beta <- numeric(num_test)
sampled_predictions <- numeric(num_test)

#Randomly select a posterior sample for each test observation
posterior_index <- sample(1:num_samples, num_test, replace = TRUE)

for (j in 1:num_test) {
  #Pick a random posterior sample index
  i <- posterior_index[j]  

  #Compute alpha and beta for the test observation
  sampled_alpha[j] <- exp(a_alpha_samples[i] + b_alpha_samples[i] * test$swr[j] + c_alpha_samples[i] * test$tair[j] + d_alpha_samples[i] * test$rh[j] + e_alpha_samples[i] * test$wind[j] + f_alpha_samples[i] * test$precip[j] + g_alpha_samples[i] * test$sp[j])
  sampled_beta[j] <- exp(a_beta_samples[i] + b_beta_samples[i] * test$swr[j] + c_beta_samples[i] * test$tair[j] + d_beta_samples[i] * test$rh[j] + e_beta_samples[i] * test$wind[j] + f_beta_samples[i] * test$precip[j] + g_beta_samples[i] * test$sp[j])
  
  #Generate a single prediction from the Gamma distribution
  sampled_predictions[j] <- rgamma(1, shape = sampled_alpha[j], rate = sampled_beta[j])
}

#Create a dataframe with observed and predicted values
predictions_df <- data.frame(
  actual = test$strikes,      #Observed lightning strikes
  swr = test$swr,            #Climate predictor
  tair = test$tair,            #Climate predictor
  rh = test$rh,            #Climate predictor
  wind = test$wind,            #Climate predictor
  precip = test$precip,            #Climate predictor
  sp = test$sp,            #Climate predictor
  alpha = sampled_alpha,      #Sampled alpha for each observation
  beta = sampled_beta,        #Sampled beta for each observation
  sampled_pred = sampled_predictions  # Sampled prediction
)

#add small value back 
predictions_df <- mutate(predictions_df, sampled_pred = sampled_pred + epsilon)

#View predictions
print(head(predictions_df))

hist(predictions_df$sampled_pred, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$swr, predictions_df$sampled_pred, color=test$sp, main="Lightning Predictions")
plot(test$swr, test$strikes, color=test$sp, main="Lightning Observations")
plot(predictions_df$sampled_pred, test$strikes, main="Observed vs. Predicted")
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

#store predictions for figures 
y_pred_B13 <- predictions_df$sampled_pred
```

```{r}
nrmse_res_B13  <- metric_nrmse_ci(test$strikes, y_pred_B13)
cor_res_B13    <- metric_cor_ci(test$strikes, y_pred_B13)
sscore_res_B13 <- metric_sscore_ci(test$strikes, y_pred_B13, nbins = 15)
spcor_res_B13  <- metric_spatial_cor_ci(test, y_pred_B13)
acc_res_B13  <- metric_acc_ci(test$year, test$strikes, y_pred_B13)
nrmseA_res_B13 <- metric_nrmse_anom_ci(test$year, test$strikes, y_pred_B13)
```

```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  P = train$precip,
  SP = train$sp
)


# Compile and fit the Stan model
init_list <- function() {
  list(a_alpha = 1, b_alpha = 0.1, c_alpha = 0.1, d_alpha = 0.1, e_alpha = 0.1, f_alpha = 0.1, 
       a_beta = 1, b_beta = 0.1, c_beta = 0.1, d_beta = 0.1, e_beta = 0.1, f_beta = 0.1)
}

if (!file.exists(file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))) {
  B12 <- stan(file = "01_gamma_swr_tair_rh_precip_sp.stan", 
            data = dat,
            iter = 2000, 
            chains = 4, 
            init = init_list,
            cores = min(parallel::detectCores(), 4))
    saveRDS(B12, file = file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))
    } else {
      B12 <- readRDS(file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))
    }

```


```{r}

#print(B12, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "f_alpha", "g_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta", "f_beta", "g_beta"))

traceplot(B12)

parameter_stats <- summary(B12)$summary[, c("mean", "sd")][1:14, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#store mean posterior sample for coefficient table
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
f_alpha <- summary_table["f_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]
f_beta <- summary_table["f_beta", "mean"]


B12_alpha_coefficients <- c(a_alpha, b_alpha, c_alpha, d_alpha, e_alpha, f_alpha)

B12_beta_coefficients <- c(a_beta, b_beta, c_beta, d_beta, e_beta, f_beta)
```

Make predictions
```{r}
#Extract posterior samples
posterior_samples <- as.data.frame(B12)

#Extract individual parameters
a_alpha_samples <- posterior_samples$a_alpha
b_alpha_samples <- posterior_samples$b_alpha
c_alpha_samples <- posterior_samples$c_alpha
d_alpha_samples <- posterior_samples$d_alpha
e_alpha_samples <- posterior_samples$e_alpha
f_alpha_samples <- posterior_samples$f_alpha
a_beta_samples <- posterior_samples$a_beta
b_beta_samples <- posterior_samples$b_beta
c_beta_samples <- posterior_samples$c_beta
d_beta_samples <- posterior_samples$d_beta
e_beta_samples <- posterior_samples$e_beta
f_beta_samples <- posterior_samples$f_beta


#Number of posterior samples
num_samples <- length(a_alpha_samples)

#Number of test observations
num_test <- length(test$swr)

#Create vectors to store sampled alpha, beta, and predictions
sampled_alpha <- numeric(num_test)
sampled_beta <- numeric(num_test)
sampled_predictions <- numeric(num_test)

#Randomly select a posterior sample for each test observation
posterior_index <- sample(1:num_samples, num_test, replace = TRUE)

for (j in 1:num_test) {
  #Pick a random posterior sample index
  i <- posterior_index[j]  

  #Compute alpha and beta for the test observation
  sampled_alpha[j] <- exp(a_alpha_samples[i] + b_alpha_samples[i] * test$swr[j] + c_alpha_samples[i] * test$tair[j] + d_alpha_samples[i] * test$rh[j] + e_alpha_samples[i] * test$precip[j] + f_alpha_samples[i] * test$sp[j])
  sampled_beta[j] <- exp(a_beta_samples[i] + b_beta_samples[i] * test$swr[j] + c_beta_samples[i] * test$tair[j] + d_beta_samples[i] * test$rh[j] + e_beta_samples[i] * test$precip[j] + f_beta_samples[i] * test$sp[j])
  
  #Generate a single prediction from the Gamma distribution
  sampled_predictions[j] <- rgamma(1, shape = sampled_alpha[j], rate = sampled_beta[j])
}

#Create a dataframe with observed and predicted values
predictions_df <- data.frame(
  actual = test$strikes,      #Observed lightning strikes
  swr = test$swr,            #Climate predictor
  tair = test$tair,            #Climate predictor
  rh = test$rh,            #Climate predictor
  precip = test$precip,            #Climate predictor
  sp = test$sp,            #Climate predictor
  alpha = sampled_alpha,      #Sampled alpha for each observation
  beta = sampled_beta,        #Sampled beta for each observation
  sampled_pred = sampled_predictions  # Sampled prediction
)

#add small value back 
predictions_df <- mutate(predictions_df, sampled_pred = sampled_pred + epsilon)

#View predictions
print(head(predictions_df))

hist(predictions_df$sampled_pred, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$swr, predictions_df$sampled_pred, color=test$sp, main="Lightning Predictions")
plot(test$swr, test$strikes, color=test$sp, main="Lightning Observations")
plot(predictions_df$sampled_pred, test$strikes, main="Observed vs. Predicted")
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

#store predictions for figures 
y_pred_B12 <- predictions_df$sampled_pred
```


```{r}
nrmse_res_B12  <- metric_nrmse_ci(test$strikes, y_pred_B12)
cor_res_B12    <- metric_cor_ci(test$strikes, y_pred_B12)
sscore_res_B12 <- metric_sscore_ci(test$strikes, y_pred_B12, nbins = 15)
spcor_res_B12  <- metric_spatial_cor_ci(test, y_pred_B12)
acc_res_B12  <- metric_acc_ci(test$year, test$strikes, y_pred_B12)
nrmseA_res_B12 <- metric_nrmse_anom_ci(test$year, test$strikes, y_pred_B12)
```


Put performace metrics together and plot
```{r}

nrmse <- c(nrmse_res_G13$estimate, nrmse_res_G12$estimate, nrmse_res_B13$estimate, nrmse_res_B12$estimate)
cor <- c(cor_res_G13$estimate, cor_res_G12$estimate, cor_res_B13$estimate, cor_res_B12$estimate)
sscore <- c(sscore_res_G13$estimate, sscore_res_G12$estimate, sscore_res_B13$estimate, sscore_res_B12$estimate)
spcor <- c(spcor_res_G13$estimate, spcor_res_G12$estimate, spcor_res_B13$estimate, spcor_res_B12$estimate)
acc <- c(acc_res_G13$estimate, acc_res_G12$estimate, acc_res_B13$estimate, acc_res_B12$estimate)
nrmseA <- c(nrmseA_res_G13$estimate, nrmseA_res_G12$estimate, nrmseA_res_B13$estimate, nrmseA_res_B12$estimate)

nrmse_ci_low <- c(nrmse_res_G13$ci_low, nrmse_res_G12$ci_low, nrmse_res_B13$ci_low, nrmse_res_B12$ci_low)
cor_ci_low <- c(cor_res_G13$ci_low, cor_res_G12$ci_low, cor_res_B13$ci_low, cor_res_B12$ci_low)
sscore_ci_low <- c(sscore_res_G13$ci_low, sscore_res_G12$ci_low, sscore_res_B13$ci_low, sscore_res_B12$ci_low)
spcor_ci_low <- c(spcor_res_G13$ci_low, spcor_res_G12$ci_low, spcor_res_B13$ci_low, spcor_res_B12$ci_low)
acc_ci_low <- c(acc_res_G13$ci_low, acc_res_G12$ci_low, acc_res_B13$ci_low, acc_res_B12$ci_low)
nrmseA_ci_low <- c(nrmseA_res_G13$ci_low, nrmseA_res_G12$ci_low, nrmseA_res_B13$ci_low, nrmseA_res_B12$ci_low)

nrmse_ci_high <- c(nrmse_res_G13$ci_high, nrmse_res_G12$ci_high, nrmse_res_B13$ci_high, nrmse_res_B12$ci_high)
cor_ci_high <- c(cor_res_G13$ci_high, cor_res_G12$ci_high, cor_res_B13$ci_high, cor_res_B12$ci_high)
sscore_ci_high <- c(sscore_res_G13$ci_high, sscore_res_G12$ci_high, sscore_res_B13$ci_high, sscore_res_B12$ci_high)
spcor_ci_high <- c(spcor_res_G13$ci_high, spcor_res_G12$ci_high, spcor_res_B13$ci_high, spcor_res_B12$ci_high)
acc_ci_high <- c(acc_res_G13$ci_high, acc_res_G12$ci_high, acc_res_B13$ci_high, acc_res_B12$ci_high)
nrmseA_ci_high <- c(nrmseA_res_G13$ci_high, nrmseA_res_G12$ci_high, nrmseA_res_B13$ci_high, nrmseA_res_B12$ci_high)

name <- c("G13", "G12", "B13", "B12")

method <- c("Gamma GLM", "Gamma GLM", "Gamma Bayes", "Gamma Bayes")

predictors <- c("RH + Rsds + T + Ps + Pr + U10", "RH + Rsds + T + Ps + Pr", "RH + Rsds + T + Ps + Pr + U10", "RH + Rsds + T + Ps + Pr")

metrics <- data.frame(cbind(name, method, predictors, nrmse, nrmse_ci_low, nrmse_ci_high, cor, cor_ci_low, cor_ci_high, sscore, sscore_ci_low, sscore_ci_high, spcor, spcor_ci_low, spcor_ci_high, acc, acc_ci_low, acc_ci_high, nrmseA, nrmseA_ci_low, nrmseA_ci_high))

metrics <- metrics |>
  mutate(across(-c(name, method, predictors), as.numeric))

#export
write.csv(metrics, file.path(performance_root, "lightning_performance.csv"))
```



######
Try other Priors for the bayes models
```{r}
# ── Step 1: Add prior parameters to data list ─────────────────────────────────
# These will be passed as data so the Stan model can use them
# without recompilation when changed

make_dat <- function(prior_mu_a_beta, prior_sigma_intercepts, prior_sigma_slopes) {
  list(
    N                   = nrow(train),
    S                   = train$strikes,
    SWR                 = train$swr,
    T                   = train$tair,
    RH                  = train$rh,
    P                   = train$precip,
    SP                  = train$sp,
    mu_a_beta           = prior_mu_a_beta,
    sigma_intercepts    = prior_sigma_intercepts,
    sigma_slopes        = prior_sigma_slopes
  )
}


# ── Step 3: Compile ONCE ──────────────────────────────────────────────────────
cat("Compiling Stan model...\n")
sm_compiled <- stan_model(file = "gamma_swr_tair_rh_precip_sp.stan")
saveRDS(sm_compiled, file.path(rds_root, "Lightning/compiled_gamma_bayes.rds"))
# Next session: sm_compiled <- readRDS(file.path(rds_root, "Lightning/compiled_gamma_bayes.rds"))

# ── Step 4: Define prior configurations ───────────────────────────────────────
configs <- list(
  tight  = list(mu_a_beta = 0, sigma_intercepts = 1, sigma_slopes = 1),  # original
  medium = list(mu_a_beta = 5, sigma_intercepts = 3, sigma_slopes = 2),
  wide   = list(mu_a_beta = 5, sigma_intercepts = 5, sigma_slopes = 3)
)

init_list <- function() {
  list(
    a_alpha = 1,   b_alpha = 0.1, c_alpha = 0.1,
    d_alpha = 0.1, e_alpha = 0.1, f_alpha = 0.1,
    a_beta  = 4,   b_beta  = 0.1, c_beta  = 0.1,   # initialise a_beta near expected value
    d_beta  = 0.1, e_beta  = 0.1, f_beta  = 0.1
  )
}

# ── Step 5: Fit all configurations ────────────────────────────────────────────
fits_B12 <- lapply(names(configs), function(cfg_name) {

  rds_path <- file.path(rds_root,
                         sprintf("Lightning/B12_gamma_bayes_%s.rds", cfg_name))

  if (file.exists(rds_path)) {
    message("Loading cached model: ", cfg_name)
    return(readRDS(rds_path))
  }

  message("Fitting model: ", cfg_name)
  cfg <- configs[[cfg_name]]

  fit <- sampling(
    sm_compiled,
    data    = make_dat(cfg$mu_a_beta,
                       cfg$sigma_intercepts,
                       cfg$sigma_slopes),
    iter    = 2000,
    chains  = 4,
    init    = init_list,
    cores   = min(parallel::detectCores(), 4),
    seed    = 123,
    refresh = 200
  )

  saveRDS(fit, rds_path)
  fit
})

names(fits_B12) <- names(configs)

# ── Step 6: Quick coefficient comparison across configurations ────────────────
lapply(names(fits_B12), function(cfg_name) {
  fit <- fits_B12[[cfg_name]]
  summ <- summary(fit, pars = c("a_alpha", "a_beta"))$summary
  cat(sprintf("\n[%s] a_alpha = %.3f | a_beta = %.3f\n",
              cfg_name,
              summ["a_alpha", "mean"],
              summ["a_beta",  "mean"]))
})

```

```{r}
# ── Helper: generate predictions from a fitted Stan model ─────────────────────
predict_bayes_lightning <- function(fit, test_df, has_wind = FALSE,
                                    epsilon = 1e-10, seed = 123) {
  set.seed(seed)
  posterior_samples <- as.data.frame(fit)

  # Parameter names depend on whether wind (e_alpha/e_beta) is included
  param_names_alpha <- c("a_alpha", "b_alpha", "c_alpha",
                          "d_alpha", "e_alpha", "f_alpha")
  param_names_beta  <- c("a_beta",  "b_beta",  "c_beta",
                          "d_beta",  "e_beta",  "f_beta")
  if (has_wind) {
    param_names_alpha <- c(param_names_alpha, "g_alpha")
    param_names_beta  <- c(param_names_beta,  "g_beta")
  }

  num_samples  <- nrow(posterior_samples)
  num_test     <- nrow(test_df)
  post_idx     <- sample(seq_len(num_samples), num_test, replace = TRUE)

  sampled_pred <- numeric(num_test)

  for (j in seq_len(num_test)) {
    i <- post_idx[j]

    # Build linear predictor for alpha
    lp_alpha <- posterior_samples[i, "a_alpha"] +
                posterior_samples[i, "b_alpha"] * test_df$swr[j]   +
                posterior_samples[i, "c_alpha"] * test_df$tair[j]  +
                posterior_samples[i, "d_alpha"] * test_df$rh[j]    +
                posterior_samples[i, "e_alpha"] * test_df$precip[j] +
                posterior_samples[i, "f_alpha"] * test_df$sp[j]

    # Build linear predictor for beta
    lp_beta  <- posterior_samples[i, "a_beta"]  +
                posterior_samples[i, "b_beta"]  * test_df$swr[j]   +
                posterior_samples[i, "c_beta"]  * test_df$tair[j]  +
                posterior_samples[i, "d_beta"]  * test_df$rh[j]    +
                posterior_samples[i, "e_beta"]  * test_df$precip[j] +
                posterior_samples[i, "f_beta"]  * test_df$sp[j]

    if (has_wind) {
      lp_alpha <- lp_alpha + posterior_samples[i, "g_alpha"] * test_df$wind[j]
      lp_beta  <- lp_beta  + posterior_samples[i, "g_beta"]  * test_df$wind[j]
    }

    alpha_j <- exp(lp_alpha)
    beta_j  <- exp(lp_beta)
    sampled_pred[j] <- rgamma(1, shape = alpha_j, rate = beta_j) + epsilon
  }

  sampled_pred
}

# ── Compute metrics for each configuration in fits_B12 ────────────────────────
# Assumes fits_B12 is a named list: list(tight = ..., medium = ..., wide = ...)
# Adjust has_wind = FALSE for B12 (no wind), TRUE for B13

new_rows <- lapply(names(fits_B12), function(cfg_name) {

  cat("Computing metrics for B12 —", cfg_name, "...\n")
  fit <- fits_B12[[cfg_name]]

  # Generate predictions
  y_pred <- predict_bayes_lightning(fit, test,
                                     has_wind = FALSE,  # B12 excludes wind
                                     seed     = 123)

  # Compute metrics using your existing functions
  nrmse_res  <- metric_nrmse_ci(test$strikes,  y_pred)
  cor_res    <- metric_cor_ci(test$strikes,    y_pred)
  sscore_res <- metric_sscore_ci(test$strikes, y_pred, nbins = 15)
  spcor_res  <- metric_spatial_cor_ci(test,    y_pred)
  acc_res    <- metric_acc_ci(test$year, test$strikes, y_pred)
  nrmseA_res <- metric_nrmse_anom_ci(test$year, test$strikes, y_pred)

  data.frame(
    name           = paste0("B12_", cfg_name),
    method         = "Gamma Bayes",
    predictors     = "RH + Rsds + T + Ps + Pr",
    nrmse          = nrmse_res$estimate,
    nrmse_ci_low   = nrmse_res$ci_low[1],
    nrmse_ci_high  = nrmse_res$ci_high[2],
    cor            = cor_res$estimate,
    cor_ci_low     = cor_res$ci_low[1],
    cor_ci_high    = cor_res$ci_high[2],
    sscore         = sscore_res$estimate,
    sscore_ci_low  = sscore_res$ci_low[1],
    sscore_ci_high = sscore_res$ci_high[2],
    spcor          = spcor_res$estimate,
    spcor_ci_low   = spcor_res$ci_low[1],
    spcor_ci_high  = spcor_res$ci_high[2],
    acc            = acc_res$estimate,
    acc_ci_low     = acc_res$ci_low[1],
    acc_ci_high    = acc_res$ci_high[2],
    nrmseA         = nrmseA_res$estimate,
    nrmseA_ci_low  = nrmseA_res$ci_low[1],
    nrmseA_ci_high = nrmseA_res$ci_high[2],
    row.names      = NULL
  )
})

# ── Append to existing metrics dataframe ──────────────────────────────────────
metrics <- bind_rows(metrics, new_rows)
rownames(metrics) <- NULL
print(metrics)


write.csv(metrics, file.path(performance_root, "lightning_performance.csv"))
```



```{r}
p_nrmse <- ggplot(metrics,
                aes(x = nrmse, y = predictors, color=method, shape=name)) +
  geom_point(size = 2.8, alpha = 0.9,
                 position = position_dodge(width = 0.6)) +
  # --- confidence intervals ---
  geom_errorbarh(aes(xmin = nrmse_ci_low,
                     xmax = nrmse_ci_high),
                 height = 0.25,
                 alpha = 0.6,
                 linewidth = 0.5,
                 position = position_dodge(width = 0.6)) +
  scale_x_reverse() +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  labs(
    x = "nRMSE",
    y = NULL
  ) +
  ggtitle("(a)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.margin=ggplot2::margin(t = 10, r =10, b = 10, l = 30)) +
  theme(legend.position = "none")

p_cor <- ggplot(metrics,
                aes(x = cor, y = predictors, color=method, shape=name)) +
  geom_point(size = 2.8, alpha = 0.9,
                 position = position_dodge(width = 0.6)) +
  # --- confidence intervals ---
  geom_errorbarh(aes(xmin = cor_ci_low,
                     xmax = cor_ci_high),
                 height = 0.25,
                 alpha = 0.6,
                 linewidth = 0.5,
                 position = position_dodge(width = 0.6)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  labs(
    x = "Correlation between Obs and Pred",
    y = NULL
  ) +
  ggtitle("(b)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.margin=ggplot2::margin(t = 10, r =10, b = 10, l = 30)) +
  theme(legend.position = "none")

p_sscore <- ggplot(metrics,
                aes(x = sscore, y = predictors, color=method, shape=name)) +
  geom_point(size = 2.8, alpha = 0.9,
                 position = position_dodge(width = 0.6)) +
  # --- confidence intervals ---
  geom_errorbarh(aes(xmin = sscore_ci_low,
                     xmax = sscore_ci_high),
                 height = 0.25,
                 alpha = 0.6,
                 linewidth = 0.5,
                 position = position_dodge(width = 0.6)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  labs(
    x = "S-score",
    y = NULL
  ) +
  ggtitle("(c)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.margin=ggplot2::margin(t = 10, r =10, b = 10, l = 30)) +
  theme(legend.position = "none")

p_spcor <- ggplot(metrics,
                aes(x = spcor, y = predictors, color=method, shape=name)) +
  geom_point(size = 2.8, alpha = 0.9,
                 position = position_dodge(width = 0.6)) +
  # --- confidence intervals ---
  geom_errorbarh(aes(xmin = spcor_ci_low,
                     xmax = spcor_ci_high),
                 height = 0.25,
                 alpha = 0.6,
                 linewidth = 0.5,
                 position = position_dodge(width = 0.6)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  #xlim(0.4, 1) +
  labs(
    x = "Spatial Cor",
    y = NULL
  ) +
  ggtitle("(d)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.margin=ggplot2::margin(t = 10, r =10, b = 10, l = 30)) +
  theme(legend.position = "none")

p_acc <- ggplot(metrics,
                aes(x = acc, y = predictors, color=method, shape=name)) +
  geom_point(size = 2.8, alpha = 0.9,
                 position = position_dodge(width = 0.6)) +
  # --- confidence intervals ---
  geom_errorbarh(aes(xmin = acc_ci_low,
                     xmax = acc_ci_high),
                 height = 0.25,
                 alpha = 0.6,
                 linewidth = 0.5,
                 position = position_dodge(width = 0.6)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  labs(
    x = "Anomaly Cor Coeff",
    y = NULL
  ) +
  ggtitle("(e)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.margin=ggplot2::margin(t = 10, r =10, b = 10, l = 30)) 

p_nrmseA <- ggplot(metrics,
                aes(x = nrmseA, y = predictors, color=method, shape=name)) +
  geom_point(size = 2.8, alpha = 0.9,
                 position = position_dodge(width = 0.6)) +
  # --- confidence intervals ---
  geom_errorbarh(aes(xmin = nrmseA_ci_low,
                     xmax = nrmseA_ci_high),
                 height = 0.25,
                 alpha = 0.6,
                 linewidth = 0.5,
                 position = position_dodge(width = 0.6)) +
  scale_x_reverse() +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  labs(
    x = "nRMSE of Anomalies",
    y = NULL,
  ) +
  ggtitle("(f)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.margin=ggplot2::margin(t = 10, r =10, b = 10, l = 30)) +
  theme(legend.position = "none")

legend_plot <- ggplot(data=metrics, aes(x=nrmseA, y=predictors, color=method)) + geom_point()  +
  theme(
    legend.position   = "bottom",
    legend.box        = "horizontal",
    legend.title      = element_text(size = 10),
    legend.text       = element_text(size = 9),
    legend.key.height = unit(0.5, "cm"),
    legend.key.width  = unit(1.0, "cm"),
    legend.spacing.x  = unit(2, "cm")
  )

shared_legend <- ggpubr::get_legend(legend_plot)

blank_plot <- ggplot() + theme_void()

fig_panels <- cowplot::plot_grid(
  p_nrmse, p_cor, p_sscore, p_spcor, p_acc, p_nrmseA,
  ncol = 2,
  align = "h"
)

plot_grid(
  fig_panels,
  shared_legend,
  ncol = 1,
  rel_heights = c(1, 0.05)
)

# grid.arrange(p_nrmse, p_cor, p_sscore, p_spcor, p_acc, p_nrmseA, nrow=3, ncol=2, heights=c(1,1,1), widths=c(1,1), layout_matrix=rbind(c(1, 2), c(3, 4), c(5, 6)))
```

Map observations and predictions

Alaska polygon
```{r}
# Get USA polygon
usa <- ne_states(country = "United States of America", returnclass = "sf")
# Filter to Alaska only
alaska <- usa[usa$name == "Alaska", ]
alaska <- st_transform(alaska, crs = 4326)
```

```{r}
test <- cbind(test, strikes_pred=y_pred_G13)
hist(test$strikes)
hist(test$strikes_pred)

```

```{r}
#calculate bias for lightning prediction
plot_df <- test %>%
  group_by(lat, lon) %>%
  summarise(
    bias = mean(strikes_pred - strikes, na.rm = TRUE),
    strikes=mean(strikes),
    strikes_pred=mean(strikes_pred),
    .groups = "drop"
  )

#global min and max strike rates
min_value_strike <- min(c(plot_df$strikes, plot_df$strikes_pred), na.rm = TRUE)
max_value_strike <- max(c(plot_df$strikes, plot_df$strikes_pred), na.rm = TRUE)
```


Elevation plot
```{r}
library(geodata)
library(terra)

# Download SRTM elevation data for Alaska
# Alaska spans multiple tiles so we download by country

# 1. Download and crop
elev <- geodata::elevation_global(res = 10, path = tempdir())
alaska_ext <- ext(-170, -130, 51, 72)
elev_alaska <- crop(elev, alaska_ext)

# 2. Mask to Alaska polygon
alaska_vect <- terra::vect(alaska)
elev_masked <- terra::mask(elev_alaska, alaska_vect)

# 3. Convert to data frame — use elev_masked instead of elev_alaska
elev_df <- as.data.frame(elev_masked, xy = TRUE)
names(elev_df) <- c("lon", "lat", "elevation")
elev_df <- elev_df %>% filter(!is.na(elevation), elevation >= 0)

p_elevation <- ggplot() +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250)) +  # contour interval in metres
  geom_sf(data = alaska, fill = NA, color = "black",
          linewidth = 0.3, inherit.aes = FALSE) +
  xlim(-170, -130) +
  ylim(51, 72) +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(h)") +
  theme_minimal() +
  theme(
    axis.title.x     = element_text(size=11),
    axis.title.y     = element_text(size=11),
    axis.text.y      = element_text(size=11),
    axis.text.x      = element_text(size=11),
    panel.grid.major = element_line(color = "darkgray", size = 0.2),
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)
  )

p_elevation
```

Make maps!

```{r}
p_strike <- ggplot(plot_df, aes(x = lon, y = lat, fill = strikes_pred)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    limits = c(min_value_strike, max_value_strike), 
    name = expression(atop("Predicted lightning", "strikes/month/km"^2)),
    guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_strike

p_strike_obs <- ggplot(plot_df, aes(x = lon, y = lat, fill = strikes)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    limits = c(min_value_strike, max_value_strike), 
    name = expression(atop("Observed lightning", "strikes/month/km"^2)),
    guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(d)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE)

p_strike_obs

max_abs_bias <- max(abs(plot_df$bias), na.rm = TRUE)

p_strike_bias <- ggplot(plot_df, aes(x = lon, y = lat, fill = bias)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  # scale_fill_viridis_c(
  #   option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
  #   direction = -1,
  #   name = expression(atop("Mean Bias", "strikes/month/km"^2)),
  #   guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
   scale_fill_distiller(
  palette = "PuOr",
  limits = c(-max_abs_bias, max_abs_bias),
  name = expression(atop("Mean Bias", "strikes/month/km"^2)),
  guide = guide_colorbar(direction = "horizontal", title.position = "top")
) +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(g)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_strike_bias

p_strike + p_strike_obs + p_strike_bias
```

Export predictions
```{r}
write.csv(test, file.path(predictions_root, "lightning_predictions.csv"))
```








