---
title: "Negative_Binomial"
output: html_document
---


```{r, echo=FALSE, include=FALSE}
library(rstan)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(patchwork)
library(ggplot2)
library(gridExtra)
library(terra)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
```

# Gamma

r_strike ~ gamma(alpha, beta)
alpha = exp(a + b * X)
beta = exp(a + b * X)

r_strike: mean summer strike rate (strikes/km2/month)


# Negative Binomial

n_strike ~ negative binomial(mu, theta)
mu = r_strike * area
log(theta) = a + b * X


Read in drivers
- strike rate data is at seasonal scale (summer mean, june-august)
- strike count data is at monthly sclae (june-august)

######################
# Monthly PREDICTIONS #
######################

```{r}
df <- read.csv(file.path(drivers_root, "lightning_strike_count_climate_jja_2002_2011.csv"))
head(df)
```

Does strike count vary with month?
Yes! There are more zeros in August, and the mean strike count is higher in June and July. This tells us that climate is 
```{r}
hist(filter(df, month==8)$strike_count, main="Augst Strike Count")
hist(filter(df, month==7)$strike_count, main="July Strike Count")
hist(filter(df, month==6)$strike_count, main="June Strike Count")

as.data.frame(cbind(month = c("August", "July", "june"),
              mean_strike_count = c(
                mean(filter(df, month==8)$strike_count),
                mean(filter(df, month==7)$strike_count),
                mean(filter(df, month==6)$strike_count)),
              median_strike_count = c(
                median(filter(df, month==8)$strike_count),
                median(filter(df, month==7)$strike_count),
                median(filter(df, month==6)$strike_count)),
              sd_strike_count = c(
                sd(filter(df, month==8)$strike_count),
                sd(filter(df, month==7)$strike_count),
                sd(filter(df, month==6)$strike_count)),
              max_strike_count = c(
                max(filter(df, month==8)$strike_count),
                max(filter(df, month==7)$strike_count),
                max(filter(df, month==6)$strike_count)),
              n_zeros = c(
                nrow(filter(df, month==8 & strike_count == 0)),
                nrow(filter(df, month==7 & strike_count == 0)),
                nrow(filter(df, month==6 & strike_count == 0)))))

```

Next, lets get mu: mean summer (june-august) lightning strike rate, strikes/km2/month - this will be used as mu in the negative binomial distribution. 

Gamma model predicts:

r_strike ~ gamma(alpha, beta)
alpha = exp(a + b * X)
beta = exp(a + b * X)

r_strike: mean summer strike rate (strikes/km2/month)

n_strike ~ negative_binomial(mu, theta)
mu = r_strike * area
theta = a + b * X

Note, mu in the negative binomial distribution must be greater than 0. This meas that we cannot use monthly mean rate, as some gridcell-months expereince no lightning. To resolve this, we use monthly summer mean strike rate so that all rates are greater than 0. 

```{r}
#read in seasonal data containing observed r_strike (mean summer strikes/km2/month)  
lightning <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))

#merge summer strike rate (strikes/km2/month) observations - months will be duplicated
df <- df %>%
  left_join(
    lightning %>% dplyr::select(lat, lon, year, strikes),
    by = c("lat", "lon", "year")
  )

names(df)[names(df) == "strikes"] <- "r_strike"

#Scale strike rate from strikes/km2/month to strikes/gridcell/month
area <- read.csv("~/raid/cuden/data/paleo_fire/era5_grid_area.csv")

df <- df %>%
  left_join(
    area %>% dplyr::select(lat, lon, area),
    by = c("lat", "lon")
  )


#df grid covers a rectangle rather than the alaska border. removing NAs will deal with this
df <- na.omit(df)

ggplot(filter(df, month==8 & year==2002), aes(x=lon, y=lat, color=r_strike)) + geom_point() + ggtitle("strike rates, Summer 2002")
ggplot(filter(df, month==8 & year==2002), aes(x=lon, y=lat, color=strike_count)) + geom_point() + ggtitle("strike counts, August 2002")
ggplot(filter(df, month==8 & year==2002 & strike_count < 10), aes(x=lon, y=lat, color=strike_count)) + geom_point() + ggtitle("strike count < 10, August 2002")

cat("Number of observations:", nrow(df), "\n")
cat("Number of 0 r_strike observations:", sum(df$r_strike == 0), "\n")
cat("Frequency of 0s:", sum(df$r_strike == 0)/nrow(df), "\n")
 
```
ZEROS: Zero strike count values will need to be removed after standardization because mu parameter in the negative binomial has to be greater than 0. 


Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]
```


#Standardize the data
```{r}
#-----
mu_precip <- mean(train$precip)
sigma_precip <- sd(train$precip)
mu_tair <- mean(train$tair)
sigma_tair <- sd(train$tair)
mu_wind <- mean(train$wind)
sigma_wind <- sd(train$wind)
mu_swr <- mean(train$swr)
sigma_swr <- sd(train$swr)
mu_sp <- mean(train$sp)
sigma_sp <- sd(train$sp)
mu_rh <- mean(train$rh)
sigma_rh <- sd(train$rh)

#store means and sds
mu <- c(mu_precip, mu_tair, mu_wind, mu_swr, mu_sp, mu_rh)
sigma <- c(sigma_precip, sigma_tair, sigma_wind, sigma_swr, sigma_sp, sigma_rh)
var <- c("precip", "tair", "wind", "swr", "sp", "rh")
mean_sd <- data.frame(cbind(var, mu, sigma))
write.csv(mean_sd, file.path(mu_sigma_root, "mu_sigma_lightning_count_predictors.csv"))
#-----

train <- train %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

train <- train[,c("r_strike", "area", "strike_count", "precip","tair", "wind", "swr", "sp", "rh")]
head(train)

test <- test %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) %>%
  mutate(lon = lon) %>%
  mutate(lat = lat)

test <- test[,c("lon", "lat", "year", "r_strike", "area", "strike_count", "precip", "tair", "wind", "swr", "sp", "rh")]

hist(test$strike_count)
```


```{r}
# ── Stan model code ────────────────────────────────────────────────────────────
stan_nb_climate <- "
data {
  int<lower=0> N;              // number of observations
  int<lower=0> K;              // number of predictors for theta
  int<lower=0> y[N];           // observed strike counts
  vector[N] log_mu;            // log of predicted rate x area (offset)
  matrix[N, K] Z;              // climate predictors for theta
  real<lower=0> sigma_gamma;   // prior SD for theta coefficients
}
parameters {
  real gamma_0;                // intercept for log(theta)
  vector[K] gamma;             // coefficients for log(theta)
}
transformed parameters {
  vector[N] log_theta;
  vector[N] theta;
  log_theta = gamma_0 + Z * gamma;
  theta     = exp(log_theta);   // theta > 0 always
}
model {
  // Priors
  gamma_0 ~ normal(2, 2);      // log(theta) intercept — prior mean exp(2)~7
  gamma   ~ normal(0, sigma_gamma);
  // Likelihood
  // NB parameterised as NegBinomial2(mu, phi) where phi = theta
  // E[y] = mu, Var[y] = mu + mu^2/theta
  for (n in 1:N) {
    y[n] ~ neg_binomial_2_log(log_mu[n], theta[n]);
  }
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N) {
    log_lik[n] = neg_binomial_2_log_lpmf(y[n] | log_mu[n], theta[n]);
  }
}"

# ── Prepare data ───────────────────────────────────────────────────────────────
# Predictors for theta — use same standardised climate predictors as rate model
# Using a subset is fine — start with the most physically meaningful ones
theta_predictors <- c("swr", "tair", "rh", "precip", "sp", "wind")

# Training data must have strike_count and predicted rate
train_nb <- train |>
  filter(r_strike > 0) |> #if there are 0 strike rate summers, they will not contribute information and slow the model.
  mutate(
    log_mu = log(pmax(r_strike * area, 1e-8))   # log of rate x area
  )

Z_train <- model.matrix(
  as.formula(paste("~", paste(theta_predictors, collapse = " + "), "- 1")),
  data = train_nb
)

stan_data_nb <- list(
  N           = nrow(train_nb),
  K           = ncol(Z_train),
  y           = as.integer(train_nb$strike_count),
  log_mu      = train_nb$log_mu,
  Z           = Z_train,
  sigma_gamma = 1
)

# ── Compile and fit ────────────────────────────────────────────────────────────
sm_nb <- stan_model(
  model_code = stan_nb_climate,
  model_name = "nb_lightning_theta"
)

fit_nb <- sampling(
  sm_nb,
  data    = stan_data_nb,
  iter    = 2000,
  chains  = 4,
  cores = 4,
  seed    = 123,
  control = list(adapt_delta = 0.95),
  refresh = 200
)

# ── Check convergence ──────────────────────────────────────────────────────────
rhat_vals <- summary(fit_nb)$summary[, "Rhat"]
cat("Max R-hat:", round(max(rhat_vals, na.rm = TRUE), 4), "\n")

# ── Posterior summaries ────────────────────────────────────────────────────────
summ <- summary(fit_nb,
                 pars  = c("gamma_0", "gamma"),
                 probs = c(0.025, 0.5, 0.975))$summary

summ_df <- as.data.frame(summ)
summ_df$variable <- c("intercept", theta_predictors)
print(summ_df[, c("variable", "mean", "2.5%", "97.5%")])


# ── LOO-CV ────────────────────────────────────────────────────────────────────
library(loo)
loo_nb <- loo(fit_nb, pars = "log_lik")
print(loo_nb)

#extract posterior draws:
post_nb <- rstan::extract(fit_nb)
# Posterior mean gamma coefficients
gamma_0_mean <- mean(post_nb$gamma_0)
gamma_means  <- colMeans(post_nb$gamma)

# ── Save ──────────────────────────────────────────────────────────────────────
saveRDS(list(
  fit        = fit_nb,
  coef_means = list(gamma_0 = gamma_0_mean, gamma = gamma_means),
  predictors = theta_predictors
), file.path(rds_root, "Lightning/neg_bin_strike_count.rds"))


# Save lightweight version — coefficients only, not full stanfit. The full fit won't be needed for paleo predictions
saveRDS(list(
  #don't save fit, it's huge
  coef_means  = list(gamma_0 = gamma_0_mean, gamma = gamma_means),
  predictors  = theta_predictors,
  loo         = loo_nb,            # already computed, small
  summ        = summ_df            # posterior summaries
), file.path(rds_root, "Lightning/neg_bin_strike_count_small.rds"))

```




```{r}
#Add r_strike * area to test data:

test_nb <- test %>% 
  mutate(r_strike_gridcell = r_strike * area, 1e-8)

# ── Apply to test data ────────────────────────────────────────────────────

# Already did these steps:
# post_nb <- rstan::extract(fit_nb)
# gamma_0_mean <- mean(post_nb$gamma_0)
# gamma_means  <- colMeans(post_nb$gamma)

#Build predictor matrix for full fire dataframe
Z_test <- model.matrix(
  as.formula(paste("~", paste(theta_predictors, collapse = " + "), "- 1")),
  data = test_nb 
)

# Compute posterior mean theta per gridcell-month
log_theta_pred <- as.numeric(gamma_0_mean + Z_test %*% gamma_means)

theta_pred     <- exp(log_theta_pred)

#log_mu_pred can be used if you are usingneg_binomial_2_log Stan parameterisation, 
#but R's rnbinom uses the natural scale directly
test_nb <- test_nb |>
  mutate(
    nb_theta    = theta_pred,
    log_mu_pred = log(pmax(r_strike_gridcell, 1e-8)),
    n_strikes   = if_else(
      r_strike_gridcell > 0,
      rnbinom(n    = n(),
              mu   = r_strike_gridcell,
              size = nb_theta),
      0L
    )
  )

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Observed frequency zeros:    ", mean(test$strike_count == 0), "\n")
cat("Simulated frequency zeros:   ", mean(test$n_strikes == 0), "\n")
cat("Observed total:     ", summ_df(test$strike_count), "\n")
cat("Simulated total:    ", sum(test$n_strikes), "\n")
cat("Observed mean:     ", mean(test$strike_count), "\n")
cat("Simulated mean:    ", mean(test$n_strikes), "\n")
cat("Observed variance: ", var(test$strike_count), "\n")
cat("Simulated variance:", var(test$n_strikes), "\n")

```


Next, lets get mu from the Gamma Bayesian lightning strike rate model and see if it works when we use it to draw counts from the negative bonomial. 

r_strike ~ gamma(alpha, beta)
alpha = exp(a + b * X)
beta = exp(a + b * X)

r_strike: mean summer strike rate (strikes/km2/month) - this will be used as mu in the negative binomial distribution. 

```{r}
#read in seasonal data
lightning <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))

#standardize:
lightning_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) lightning_mu_sigma$mu[lightning_mu_sigma$var == var]
get_sigma <- function(var) lightning_mu_sigma$sigma[lightning_mu_sigma$var == var]

lightning_scaled <- lightning |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    swr    = (swr    - get_mu("swr"))    / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh")
  )


# ── Load B12 Bayesian lightning model ─────────────────────────────────────────
B12 <- readRDS(file.path(rds_root, "Lightning/B12_gamma_bayes.rds"))

# ── Extract posterior samples ─────────────────────────────────────────────────
posterior_samples <- as.data.frame(B12)

# Generate predictions using posterior mean coefficients
# B12 excludes wind — predictors: swr, tair, rh, precip, sp

a_alpha_mean <- mean(posterior_samples$a_alpha)
b_alpha_mean <- mean(posterior_samples$b_alpha)
c_alpha_mean <- mean(posterior_samples$c_alpha)
d_alpha_mean <- mean(posterior_samples$d_alpha)
e_alpha_mean <- mean(posterior_samples$e_alpha)
f_alpha_mean <- mean(posterior_samples$f_alpha)

a_beta_mean  <- mean(posterior_samples$a_beta)
b_beta_mean  <- mean(posterior_samples$b_beta)
c_beta_mean  <- mean(posterior_samples$c_beta)
d_beta_mean  <- mean(posterior_samples$d_beta)
e_beta_mean  <- mean(posterior_samples$e_beta)
f_beta_mean  <- mean(posterior_samples$f_beta)

# Compute alpha and beta per observation using posterior means
# Using vectorised operations rather than a loop for speed

lp_alpha <- a_alpha_mean +
            b_alpha_mean * lightning_scaled$swr   +
            c_alpha_mean * lightning_scaled$tair  +
            d_alpha_mean * lightning_scaled$rh    +
            e_alpha_mean * lightning_scaled$precip +
            f_alpha_mean * lightning_scaled$sp

lp_beta  <- a_beta_mean  +
            b_beta_mean  * lightning_scaled$swr   +
            c_beta_mean  * lightning_scaled$tair  +
            d_beta_mean  * lightning_scaled$rh    +
            e_beta_mean  * lightning_scaled$precip +
            f_beta_mean  * lightning_scaled$sp

alpha_pred <- exp(lp_alpha)
beta_pred  <- exp(lp_beta)

set.seed(123)

# Gamma alpha and beta
# if you want the mean: alpha / beta
lightning_scaled <- lightning_scaled |>
  mutate(
    gamma_alpha   = alpha_pred,
    gamma_beta    = beta_pred,
    #r_strike_mean  = alpha_pred / beta_pred   # posterior mean prediction
    r_strike_pred = rgamma(n(), shape = gamma_alpha, rate = gamma_beta)
  )

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Predicted strike rate summary:\n")
print(summary(lightning_scaled$r_strike_pred))
cat("Observed strike rate summary:\n")
print(summary(lightning_scaled$strikes))

lightning_scaled <- lightning_scaled[, c("lat", "lon", "year", "r_strike_pred")]

#merge summer strike rate (strikes/km2/month) predictions - months will be duplicated
df <- df %>%
  left_join(
    lightning_scaled %>% dplyr::select(lat, lon, year, r_strike_pred),
    by = c("lat", "lon", "year")
  )

#Add r_strike_gridcell = r_strike * area:
df <- df %>% 
  mutate(r_strike_pred_gridcell = r_strike_pred * area)

# ── Apply to test data ────────────────────────────────────────────────────

#already did this above.
#post_nb <- rstan::extract(fit_nb)
# Posterior mean gamma coefficients
#gamma_0_mean <- mean(post_nb$gamma_0)
#gamma_means  <- colMeans(post_nb$gamma)

#Build predictor matrix for full fire dataframe
Z_df <- model.matrix(
  as.formula(paste("~", paste(theta_predictors, collapse = " + "), "- 1")),
  data = df 
)

# Compute posterior mean theta per gridcell-month
log_theta_pred <- as.numeric(gamma_0_mean + Z_df %*% gamma_means)

theta_pred     <- exp(log_theta_pred)

df <- df |>
  mutate(
    nb_theta         = theta_pred,
    log_mu_pred      = log(pmax(r_strike_pred_gridcell, 1e-8)),
    n_strikes = rnbinom(n    = n(),
                                 mu   = r_strike_pred_gridcell,
                                 size = nb_theta)
  )

df$n_strike_poisson <- sapply(df$r_strike_pred_gridcell, function(lam) rpois(1, lambda = lam))

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Observed zeros:    ", mean(df$strike_count == 0), "\n")
cat("Simulated zeros:   ", mean(df$n_strikes == 0), "\n")
cat("Simulated zeros poisson:   ", mean(df$n_strike_poisson == 0), "\n")
cat("Observed total:    ", sum(df$strike_count), "\n")
cat("Simulated total:   ", sum(df$n_strikes), "\n")
cat("Simulated total poisson:   ", sum(df$n_strike_poisson), "\n")
cat("Observed mean:     ", mean(df$strike_count), "\n")
cat("Simulated mean:    ", mean(df$n_strikes), "\n")
cat("Simulated mean poisson:    ", mean(df$n_strike_poisson), "\n")
cat("Observed variance: ", var(df$strike_count), "\n")
cat("Simulated variance:", var(df$n_strikes), "\n")
cat("Simulated varianc poisson:", var(df$n_strike_poisson), "\n")


```


```{r}
hist(df$r_strike)
hist(df$r_strike_pred)
hist(df$strike_count)
hist(df$n_strikes)
hist(df$n_strike_poisson)

ggplot(df[sample(1:nrow(df), 0.5 * nrow(df)), ], aes(x=r_strike_pred, y=r_strike)) + geom_point()
ggplot(df[sample(1:nrow(df), 0.5 * nrow(df)), ], aes(x=n_strikes, y=strike_count)) + geom_point()
ggplot(df[sample(1:nrow(df), 0.5 * nrow(df)), ], aes(x=n_strike_poisson, y=strike_count)) + geom_point()
```


