---
title: "Random Forest for Strike Rate"
output: html_document
---

```{r}
library(randomForest)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

Read in drivers
```{r}
df <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))
head(df)
summary(df$strikes)
#remove all zeros
df <- filter(df, strikes>0)

```

```{r}
df <- df[c("strikes", "precip", "tair", "wind", "swr", "sp", "rh")]

#check for correlation between variables
#draw random sample of rows to plot 
df_sample <- df %>% sample_frac(0.1)
plot(df_sample)

#apply random forest:
rf <-randomForest(strikes~.,data=df, ntree=500) 
print(rf)

#mtry: Number of random variables used in each tree
# - reducing mtry reduces correlation and strength. 
# - the oob error rate tells you what value of mtry is optimal

mtry <- tuneRF(df[-1],df$strikes, ntreeTry=500,
               stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)
best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1]
#looking for mtry for the lowest oob error rate 
print(mtry)
print(best.m) 

#apply best mtry to model
rf <-randomForest(strikes~.,data=df, mtry=best.m, importance=TRUE,ntree=500)
print(rf)

#Evaluate variable importance
importance(rf)
varImpPlot(rf)


```


# Mean Decrease Accuracy (%IncMSE):
- remove variables from the model and asses which have the greatest impact on model accuracy.
- Higher the value of mean decrease accuracy or mean decrease gini score,
       higher the importance of the variable in the model.
- tells you effect of randomly shuffling values of a variable on increasing mse of model predictions
- a higher number indicates that the variable is more important

# Mean Decrease Gini or Increase in Node Purity (IncNodePurity):
- expresses the change in the homogeneity of the of the groups created by the trees
     (using the Gini coefficient as a measure).
- What is expressed is the decrease in said purity if a particular variable
      has no information. If a variable has no information to begin with,
      the decrase would be zero. (from: https://www.reagro.org/methods/statistical/randomforest.html)

Save random forest results for later use: 
```{r}
rf_results <- importance(rf)
rf_results <- data.frame(Variable = rownames(rf_results), 
                     IncMSE = rf_results[, "%IncMSE"],
                     IncNodePurity = rf_results[, "IncNodePurity"]) 
# Export
#write.csv(rf_results, file.path(model_performance, "strike_rate_random_forest_no_zeros.csv), row.names = FALSE)
```

```{r}
rf_results <- read.csv(file.path(performance_root, "strike_rate_random_forest_no_zeros.csv"))
```

```{r}
#-------------------------
# IncMSE plot
#-------------------------
rf_mse <- rf_results %>%
  filter(metric == "%IncMSE")

p_mse <- ggplot(rf_mse, aes(x = importance, y = reorder(variable, importance))) +
  geom_hline(
    yintercept = seq_along(unique(rf_mse$Variable)),
    linetype = "dashed",
    color = "grey75",
    linewidth = 0.4
  ) +
  geom_point(
    shape = 21,    
    size = 3,
    color = "black",
    stroke = 0.7
  ) +
  labs(
    title = "(a)",
    x = "%IncMSE",
    y = NULL,
    fill = "Training years"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) + 
  theme(legend.position = "none")


#-------------------------
# IncNodePurity plot
#------------------------

rf_node <- rf_results %>%
  filter(metric == "IncNodePurity")

p_node <- ggplot(rf_node,
                 aes(x = importance,
                     y = reorder(variable, importance))) +
  geom_hline(
    yintercept = seq_along(unique(rf_node$Variable)),
    linetype = "dashed",
    color = "grey75",
    linewidth = 0.4
  ) +
  geom_point(
    shape = 21,         # circle with border
    size = 3,
    color = "black",
    stroke = 0.7
  ) +
  labs(
    title = "(b)",
    x = "IncNodePurity",
    y = NULL,
    fill = "Training years"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    strip.background = element_blank(),
    legend.position = "bottom"
  )

(p_mse + p_node) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```
