---
title: "Fire_Probability"
author: "Charlotte Uden"
date: "2026-06-14"
output: html_document
---


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(pROC)
library(loo)
library(cowplot)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

Read in lightning data
```{r}
lightning <- read.csv(file.path(drivers_root, "lightning_strike_rate_climate_jja_2002_2011.csv"))
head(lightning)
ggplot(filter(lightning, year==2002), aes(x=lon, y=lat)) + geom_point()
```
Predict lightning strike rate for all gridcells and years
```{r}
#standardize:
lightning_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_lightning_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) lightning_mu_sigma$mu[lightning_mu_sigma$var == var]
get_sigma <- function(var) lightning_mu_sigma$sigma[lightning_mu_sigma$var == var]

lightning_scaled <- lightning |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    swr    = (swr    - get_mu("swr"))    / get_sigma("swr"),
    sp     = (sp     - get_mu("sp"))     / get_sigma("sp"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh")
  )

#read in model
lightning_model <- readRDS(file.path(rds_root, "G13_gamma_glm_lightning.rds"))

#make predictions
lightning_scaled <- lightning_scaled |>
  mutate(
    strikes_pred = predict(lightning_model,
                           newdata = lightning_scaled,
                           type    = "response")
  )
```




Read in fire data
```{r}
fire <- read.csv(file.path(drivers_root, "lightning_fires_burned_area_jja_2002_2011.csv"))
head(fire)
ggplot(filter(fire, year==2002, month==6), aes(x=lon, y=lat)) + geom_point()

fire <- fire |>
  mutate(lat = round(lat, 2), lon = round(lon, 2)) |>
  distinct(lat, lon, year, month, .keep_all = TRUE)
```

Make ignition efficiency predictions
```{r}
#standardize:
ignition_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_ignition_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) ignition_mu_sigma$mu[ignition_mu_sigma$var == var]
get_sigma <- function(var) ignition_mu_sigma$sigma[ignition_mu_sigma$var == var]

ignition_scaled <- fire |>
  mutate(
    precip = (precip - get_mu("precip")) / get_sigma("precip"),
    tair   = (tair   - get_mu("tair"))   / get_sigma("tair"),
    wind   = (wind   - get_mu("wind"))   / get_sigma("wind"),
    rh     = (rh     - get_mu("rh"))     / get_sigma("rh"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_2m   = (precip_2m   - get_mu("precip_2m"))   / get_sigma("precip_2m"),
    precip_3m   = (precip_3m   - get_mu("precip_3m"))   / get_sigma("precip_3m"),
    precip_5y     = (precip_5y     - get_mu("precip_5y"))     / get_sigma("precip_5y"),
    tair_1m = (tair_1m - get_mu("tair_1m")) / get_sigma("tair_1m"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_3m   = (tair_3m   - get_mu("tair_3m"))   / get_sigma("tair_3m"),
    tair_5y     = (tair_5y     - get_mu("tair_5y"))     / get_sigma("tair_5y"),
    B     = (B     - get_mu("B"))     / get_sigma("B")
  )

#read in model
ignition_model_obj <- readRDS(file.path(rds_root, "Ignition_Efficiency/m2_glm_calibrated.rds"))
ignition_model <- ignition_model_obj$fit
cal_shift          <- ignition_model_obj$cal_shift  # intercept shift
cat("Calibration shift:", round(cal_shift, 3), "\n")

#make predictions and apply recalibration
eps <- 1e-8

ignition_scaled <- ignition_scaled |>
  mutate(
    p_ignite     = predict(ignition_model,
                           newdata = ignition_scaled,
                           type    = "response"),
    p_ignite_cal = plogis(qlogis(pmin(pmax(p_ignite, eps), 1 - eps)) + cal_shift)
  )

cat("Raw p_ignite summary:\n");        print(summary(ignition_scaled$p_ignite))
cat("Calibrated p_ignite summary:\n"); print(summary(ignition_scaled$p_ignite_cal))

#Bayes recalibrated
# ── Load calibrated Bayesian model ────────────────────────────────────────────
ignition_bayes_obj <- readRDS(file.path(rds_root, 
                                         "Ignition_Efficiency/m2_bayes_calibrated.rds"))
ignition_bayes_fit <- ignition_bayes_obj$fit        # stanfit object
cal_shift_bayes    <- ignition_bayes_obj$cal_shift  # intercept shift
coef_means         <- ignition_bayes_obj$coef_means # posterior mean alpha and beta

cat("Bayesian calibration shift:", round(cal_shift_bayes, 3), "\n")

# ── Build predictor matrix from standardised data ─────────────────────────────
predictors_m2 <- c("rh", "tair", "precip",
                    "precip_1m", "precip_2m", "precip_3m",
                    "precip_5y", "tair_1m", "tair_2m",
                    "tair_3m", "tair_5y", "B")

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m2, collapse = " + "), "- 1")),
  data = ignition_scaled
)

# ── Generate predictions using posterior mean ─────────────────────────────────
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

ignition_scaled <- ignition_scaled |>
  mutate(
    p_ignite_bayes     = plogis(linpred),
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Raw Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(ignition_scaled$p_ignite_bayes_cal))

head(ignition_scaled[, c("lat", "lon", "year", "month",
                          "fire_count", "strike_count",
                          "p_ignite_bayes", "p_ignite_bayes_cal")])

```


Make burned area predictions
First, add proximity data
```{r}
#read in binary distance data:
proximity <- read.csv(file.path(proximity_data_root, "era5_proximity_road_river_urban.csv"))
proximity <- dplyr::select(proximity, lon, lat, road_proximity_binary, urban_proximity_binary)
#check:
ggplot() +
  geom_point(data = proximity |> distinct(lat, lon),
             aes(x = lon, y = lat),
             color = "grey70", size = 0.8, alpha = 0.5) 

# Round to 2 decimal places in both dataframes
fire <- fire |>
  mutate(lat = round(lat, 2), lon = round(lon, 2))

proximity <- proximity |>
  mutate(lat = round(lat, 2), lon = round(lon, 2))

#merge by lat lon
fire <- fire |>
  left_join(
    proximity |> dplyr::select(lat, lon, 
                                road_proximity_binary, 
                                urban_proximity_binary),
    by = c("lat", "lon")
  )
```

Apply model
```{r}
#standardize:
burned_area_mu_sigma <- read.csv(file.path(mu_sigma_root, "mu_sigma_burned_area_predictors.csv"))

# Helper function: look up mu and sigma by variable name
get_mu    <- function(var) burned_area_mu_sigma$mu[burned_area_mu_sigma$var == var]
get_sigma <- function(var) burned_area_mu_sigma$sigma[burned_area_mu_sigma$var == var]

burned_area_scaled <- fire |>
  mutate(
    rh        = (rh        - get_mu("rh"))        / get_sigma("rh"),
    tair      = (tair      - get_mu("tair"))      / get_sigma("tair"),
    precip    = (precip    - get_mu("precip"))    / get_sigma("precip"),
    wind      = (wind      - get_mu("wind"))      / get_sigma("wind"),
    precip_1m = (precip_1m - get_mu("precip_1m")) / get_sigma("precip_1m"),
    precip_3m = (precip_3m - get_mu("precip_3m")) / get_sigma("precip_3m"),
    precip_5y = (precip_5y - get_mu("precip_5y")) / get_sigma("precip_5y"),
    tair_2m   = (tair_2m   - get_mu("tair_2m"))   / get_sigma("tair_2m"),
    tair_5y   = (tair_5y   - get_mu("tair_5y"))   / get_sigma("tair_5y")
  ) #no scaling for binary urban and road data.

preds_final <- c("rh", "tair", "precip", "wind",
                 "precip_1m", "precip_3m", "precip_5y",
                 "tair_2m", "tair_5y",
                 "road_proximity_binary",
                 "urban_proximity_binary")



# ── Load model and extract posterior parameters ───────────────────────────────
m6_obj     <- readRDS(file.path(rds_root, "Burned_Area/m6_bayes_gamma_re.rds"))
m6_fit     <- m6_obj$fit           # stanfit object
coef_means <- m6_obj$coef_means    # list: alpha, beta, alpha_year

alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

cat("Global intercept (alpha):", round(alpha_mean, 3), "\n")
cat("Year random effects:\n")
print(round(coef_means$alpha_year, 3))

# ── Build predictor matrix ────────────────────────────────────────────────────
X_new <- model.matrix(
  as.formula(paste("~", paste(preds_final, collapse = " + "), "- 1")),
  data = burned_area_scaled
)

# ── Generate predictions ──────────────────────────────────────────────────────
# Year random effects set to ZERO for new data (paleo or ERA5 application)
# This represents "climatologically average year" — variation comes from
# climate predictors only, not year-specific synoptic noise

# ── Option A: posterior mean prediction (fast, point estimate) ────────────────
linpred <- as.numeric(alpha_mean + X_new %*% beta_means)
# alpha_year term is omitted — equivalent to setting it to 0

burned_area_scaled <- burned_area_scaled |>
  mutate(burned_area_pred = exp(linpred))

# ── Extract shape parameter (phi) from posterior ───────────────────────────────
phi_mean <- mean(post$phi)
phi_sd   <- sd(post$phi)

cat("Gamma shape (phi) posterior mean:", round(phi_mean, 3), "\n")
cat("Gamma shape (phi) posterior sd:  ", round(phi_sd, 3), "\n")

# ── Add gamma_mean and gamma_shape columns to burned_area_scaled ──────────────
burned_area_scaled <- burned_area_scaled |>
  mutate(
    gamma_mean  = burned_area_pred,   # mean (mu) per gridcell-month
    gamma_shape = phi_mean             # shape (phi) — constant across rows
  )

# rgamma requires rate, not the mean. calculate this as:
# rgamma(n, shape, rate) where rate = shape / mean
burned_area_scaled$gamma_rate <- burned_area_scaled$gamma_shape / burned_area_scaled$gamma_mean

# # ── Option B: full posterior uncertainty (slower, recommended) ────────────────
# post    <- rstan::extract(m6_fit)
# n_draws <- length(post$alpha)
# 
# pred_draws <- matrix(NA_real_, nrow = nrow(burned_area_scaled), ncol = n_draws)
# for (d in seq_len(n_draws)) {
#   # No alpha_year[d] term — year RE = 0 for new/paleo data
#   eta             <- post$alpha[d] + X_new %*% post$beta[d, ]
#   pred_draws[, d] <- exp(eta)
# }
# 
# burned_area_scaled <- burned_area_scaled |>
#   mutate(
#     burned_area_pred  = rowMeans(pred_draws),
#     burned_area_sd    = apply(pred_draws, 1, sd),
#     burned_area_low   = apply(pred_draws, 1, quantile, 0.025),
#     burned_area_high  = apply(pred_draws, 1, quantile, 0.975)
#   )
# 
# # ── Verify ────────────────────────────────────────────────────────────────────
# cat("Prediction summary:\n")
# print(summary(burned_area_scaled$burned_area_pred))
# cat("NAs:", sum(is.na(burned_area_scaled$burned_area_pred)), "\n")

```

Save fire df in memory
```{r}
fire_saved <- fire
```

Put data together and predict fire probability
```{r}
#summer strike rate (strikes/km2/month) predictions - months will be duplicated
fire <- fire %>%
  left_join(
    lightning_scaled %>% dplyr::select(lat, lon, year,
                                sp, swr, strikes, strikes_pred),
    by = c("lat", "lon", "year")
  )

#Scale strike rate from strikes/km2/month to strikes/gridcell/month
area <- read.csv("~/raid/cuden/data/paleo_fire/era5_grid_area.csv")

fire <- fire %>%
  left_join(
    area %>% dplyr::select(lat, lon, area),
    by = c("lat", "lon")
  )

#calculate strikes/gridcell/month
fire <- fire %>%
  mutate(
    strikes_pred_per_cell = strikes_pred * area, #strikes/gridcell/month
    strikes_obs_per_cell = strikes * area
  )

#use lambda (strikes_pred_per_cell) to sample strike counts from a poisson distrubutuion 
set.seed(123)  # for reproducibility, set once at the top

fire$strikes_pred_count <- sapply(fire$strikes_pred_per_cell, function(lam) rpois(1, lambda = lam))

#simulate strike count before scaling up to gridcell area
fire$strikes_pred_count_km2 <- sapply(fire$strikes_pred, function(lam) rpois(1, lambda = lam))
fire$strikes_obs_count_km2 <- sapply(fire$strikes, function(lam) rpois(1, lambda = lam))

fire <- fire %>%
  mutate(
    strikes_pred_per_cell_sample_first = strikes_pred_count_km2 * area, #strikes/gridcell/month
    strikes_obs_per_cell_sample_first = strikes_obs_count_km2 * area
  )

#monthly ignition probability, p_ignite 
fire <- fire %>%
  left_join(
    ignition_scaled %>% dplyr::select(lat, lon, year, month,
                                p_ignite, p_ignite_cal,
                                p_ignite_bayes, p_ignite_bayes_cal),
    by = c("lat", "lon", "year", "month")
  )

#burned area 
fire <- fire %>%
  left_join(
    burned_area_scaled |> dplyr::select(lat, lon, year, month,
                                burned_area_pred, 
                                gamma_shape,
                                gamma_rate),
    by = c("lat", "lon", "year", "month")
  )


#calculate fire probability
fire <- fire %>%
  mutate(
    # fire probability
    p_fire = 1 - (1 - p_ignite_cal)^strikes_pred_per_cell,
    p_fire_bayes = 1 - (1 - p_ignite_bayes_cal)^strikes_pred_per_cell,
    p_fire_bayes_sc = 1 - (1 - p_ignite_bayes_cal)^strikes_pred_count,
    # fire count
    fire_count_pred = p_ignite_bayes_cal*strikes_pred_count
  )

#change column structures from name to num
fire <- fire |>
  mutate(
    strikes_pred = unname(strikes_pred),
    p_ignite     = unname(p_ignite),
    p_ignite_cal     = unname(p_ignite_cal),
    p_fire       = unname(p_fire),
    p_fire_bayes       = unname(p_fire_bayes),
    p_fire_bayes_sc       = unname(p_fire_bayes_sc)
  )

#Simulate fire outcomes (burn or no burn) from p_fire
#number of burns will change each time you run this code
#set.seed(12) 
fire <- fire %>%
    rowwise() %>%
    mutate(burn_pred_binary = rbinom(n = 1, size = 1, prob = p_fire_bayes_sc)) %>%
    ungroup() %>%
    mutate(burn_area_pred_conditioned = ifelse(burn_pred_binary==1, burned_area_pred, 0))


#Simulate total burned area from fire count
sample_burned_area <- function(n, shape, rate) {
  if (n == 0) return(0)
  sum(rgamma(n, shape = shape, rate = rate))
}

fire <- fire |>
  rowwise() |>
  mutate(
    burn_area_pred_total = sample_burned_area(fire_count_pred, 
                                            gamma_shape, gamma_rate)
  ) |>
  ungroup()

```

```{r}

cat("Expected r_strike per km2:", mean(fire$strikes_pred), "\n")
cat("Observed r_strike per km2:", mean(fire$strikes), "\n\n")

cat("Simulated strike count per km2:", mean(fire$strikes_pred_count_km2), "\n")
cat("sumilated strike count from observed r_strike per km2:", mean(fire$strikes_obs_count_km2), "\n\n")

cat("Expected r_strike per gridcell:", mean(fire$strikes_pred_per_cell), "\n")
cat("Observed r_strike per gridcell:", mean(fire$strikes_obs_per_cell), "\n\n")

cat("Expected strike count per gridcell total:", sum(fire$strikes_pred_count), "\n")
cat("Observed strike count per gridcell total:", sum(fire$strike_count), "\n\n")

cat("Expected strike count per gridcell mean:", mean(fire$strikes_pred_count), "\n")
cat("Observed strike count per gridcell mean:", mean(fire$strike_count), "\n\n")

cat("Expected strike count per gridcell (poisson sample first):", sum(fire$strikes_pred_per_cell_sample_first), "\n")
cat("Observed strike count per gridcell (poisson sample first):", sum(fire$strikes_obs_per_cell_sample_first), "\n\n")

cat("Expected 10 year mean p_ignite:", mean(fire$p_ignite_bayes_cal), "\n")
cat("Observed 10 year mean ignition frequency:", mean(fire$fire_count)/mean(fire$strike_count), "\n\n")

cat("Expected 10 year mean p_ignite (simulated strikes > 0):", mean(filter(fire, strikes_pred_count > 0)$p_ignite_bayes_cal), "\n")
cat("Expected 10 year mean p_ignite (observed strikes > 0):", mean(filter(fire, strike_count > 0)$p_ignite_bayes_cal), "\n")
cat("Observed 10 year mean ignition frequency (observed strikes > 0):", sum(filter(fire, strike_count > 0)$fire_count)/sum(filter(fire, strike_count >0)$strike_count), "\n\n")

cat("simulated fire count sum(p_fire):", sum(fire$p_fire_bayes_sc), "\n")
cat("simulated fire count sum(rbinom(p_fire)):", sum(fire$burn_pred_binary), "\n")
cat("Observed fire count (n fires >= 1):", sum(fire$lightning_ignition), "\n\n")

cat("simulated fire count (strikes*p_ignite):", sum(fire$fire_count_pred), "\n")
cat("simulated fire count from fire probability and fire count:", sum(filter(fire, burn_pred_binary > 0)$fire_count_pred), "\n")
cat("Observed fire count:", sum(fire$lightning_ignition), "\n\n")

cat("simulated total burned area from p_fire:", sum(fire$burn_area_pred_conditioned), "\n")
cat("Observed 10 year total burned area (n fires = 1):", sum(filter(fire, single_lightning_ignition==1)$burned_area_km2_total), "\n\n")

cat("simulated total burned area from fire count (rbinom(p_fire)):", sum(fire$burn_area_pred_total), "\n")
cat("simulated total burned area from p_fire and fire count:", sum(filter(fire, burn_pred_binary > 0)$burn_area_pred_total), "\n")
cat("Observed 10 year total burned area (n fires >= 1):", sum(filter(fire, lightning_ignition==1)$burned_area_km2_total), "\n\n")

pred_mean_p_ig <- mean(fire$p_ignite_bayes_cal)
pred_mean_p_ig_cal <- mean(fire$p_ignite_rate_cal)
obs_mean_p_ig <- mean(fire$fire_count)/mean(fire$strike_count)
obs_mean_strike_count <- mean(fire$strike_count)
pred_mean_strike_count <- mean(fire$strikes_pred_count)

trials <- nrow(fire)

cat("fire count simulated from observed mean OBSERVED ignition frequency and strike count (132630 trials):", sum(rbinom(n=trials, size=1, obs_mean_p_ig*obs_mean_strike_count)), "\n")
cat("fire count simulated from observed mean SIMULATED ignition efficiency and strike count (132630 trials):", sum(rbinom(n=trials, size=1, pred_mean_p_ig*pred_mean_strike_count)), "\n")
cat("fire count simulated from observed mean SIMULATED ignition efficiency CALIBRATED and strike count (132630 trials):", sum(rbinom(n=trials, size=1, pred_mean_p_ig_cal*pred_mean_strike_count)), "\n")
```
LIGHTNING 
Strike rates are similar between observed and predicted. Issue arises when counts are sampled from the poisson distribution. Could it be an issue with tails - moments when strike rates are exceptionally high are not being captured by the gamma GLM - could the gamma Bayesian approach handle this better?

IGNITION
Predictions are too high, by an order of magnitude! 

FIRE
Best fire count: simulated from fire count (strikes * p_ignite) and conditioned on p_fire (rbinom(p_fire)==1)
However, multiplying the mean observed strike count (8.989618 strikes per gridcell) by mean observed ignition rate (0.00023 ignitions/strikes) gets very close to the observed fire count (273, whereas actual fire count is 265)

BURNED AREA
Best burned area: simulated from fire count (strikes * P_ignite)

Keep:
strike count in a gridcell * p_ignite 

- or try sampling form binomial distribution: 
n_fires = rbinom(strike_count, ignition_efficiency)
then:
total burned area = sum(rgamma(n_fires, shape = gamma_shape, rate  = gamma_shape / gamma_mean))

gamma_shape and gamma_mean are estimated by the burned area model. 

What is going on with ignition efficiency?
```{r}
#there are five instances in which fire count and strike count ==1. This is probably skewing the model towards higher p_ignite values. 
fire |>
  filter(strike_count > 0, fire_count / strike_count == 1)|>
  dplyr::select(fire_count, strike_count)


hist(filter(fire, strikes_pred_count > 0)$p_ignite_bayes_cal)
hist(filter(fire, strike_count > 0)$fire_count/filter(fire, strike_count > 0)$strike_count)

ggplot(data=filter(fire, strike_count > 0), aes(x=p_ignite_bayes_cal, y=fire_count/strike_count)) + geom_point()

ggplot(data=filter(fire, strike_count > 0), aes(x=p_ignite_bayes_cal, y=fire_count/strike_count)) + geom_point() + ylim(0, 0.06)

ggplot(data=filter(fire, strike_count > 0), aes(x=strike_count, y=fire_count)) + geom_point() + xlim(0, 100)

ggplot(data=filter(fire, fire_count > 0), aes(x=strike_count, y=fire_count)) + geom_point() + ylim(0, 3) + xlim(0, 100)
```
So, fires aren't necessarily happening when strike counts are high. 


What about strike rate? Is the model picking up tails?

```{r}
hist(fire$strikes)
hist(fire$strikes_pred)

hist(fire$strike_count)
hist(fire$strikes_pred_count)

ggplot(data=fire, aes(x=strikes_pred, y=strikes)) + geom_point() + xlim(0, 0.26) + ylim(0, 0.26) + ggtitle("observed and predicted strike rate (strikes/km2/month)")
ggplot(data=fire, aes(x=strikes_pred_count, y=strike_count)) + geom_point() + ggtitle("observed and predicted strike count (strikes/gridcell/month)")
```

Since the lightning model doesn't capture the tails, strike counts sampled from poisson have very small tail. 
The model captures the mean well, but not gridcell/months when climate results in extreme lightning events. Max predicted fire count is > 60, while observed max count is > 20,000.


Evaluation metrics for p_fire

```{r}
## =========================================================
## Helpers (used internally)
## =========================================================

# Drop NAs and check basic validity
.clean_xy <- function(p, y) {
  if (length(p) != length(y)) {
    stop("p and y must have the same length.")
  }
  ok <- is.finite(p) & is.finite(y)
  p <- p[ok]; y <- y[ok]
  list(p = p, y = y)
}

# Clip probabilities away from 0 and 1 to avoid log(0)
.clip_probs <- function(p, eps = 1e-15) {
  p[p < eps]      <- eps
  p[p > 1 - eps]  <- 1 - eps
  p
}

## =========================================================
## 1) Brier score
## =========================================================
# Mean squared error between probabilities and 0/1 outcomes
brier_score <- function(p, y) {
  dat <- .clean_xy(p, y); p <- dat$p; y <- dat$y
  mean((p - y)^2)
}


## =========================================================
## 2) ROC AUC (rank-based, no packages)
## =========================================================
# Area under ROC curve; threshold-free discrimination metric
roc_auc <- function(p, y) {
  dat <- .clean_xy(p, y); p <- dat$p; y <- dat$y
  
  y_vals <- unique(y)
  if (!all(y_vals %in% c(0, 1))) {
    stop("y must contain only 0 and 1.")
  }
  if (length(unique(y)) < 2) {
    warning("y has no variation (all 0s or all 1s); ROC AUC undefined.")
    return(NA_real_)
  }
  
  # Rank predictions (average ranks for ties)
  r <- rank(p, ties.method = "average")
  
  n_pos <- sum(y == 1)
  n_neg <- sum(y == 0)
  
  # Sum of ranks for positive cases
  sum_r_pos <- sum(r[y == 1])
  
  # Mann-Whitney U statistic -> AUC
  auc <- (sum_r_pos - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
  auc
}

## =========================================================
## 3) Precision–Recall AUC (PR-AUC)
## =========================================================
# PR-AUC is more informative under class imbalance (rare fires)
pr_auc <- function(p, y) {
  dat <- .clean_xy(p, y); p <- dat$p; y <- dat$y
  
  y_vals <- unique(y)
  if (!all(y_vals %in% c(0, 1))) {
    stop("y must contain only 0 and 1.")
  }
  P <- sum(y == 1)
  N <- sum(y == 0)
  if (P == 0) {
    warning("No positive events in y; PR-AUC undefined.")
    return(NA_real_)
  }
  
  # Sort by decreasing probability
  o <- order(p, decreasing = TRUE)
  y_sorted <- y[o]
  p_sorted <- p[o]
  
  # Cumulative true positives and false positives
  tp <- cumsum(y_sorted == 1)
  fp <- cumsum(y_sorted == 0)
  
  recall    <- tp / P
  precision <- tp / (tp + fp)
  
  # Add (0, baseline) point for integration
  recall    <- c(0, recall)
  precision <- c(mean(y), precision)  # baseline precision = prevalence
  
  # Trapezoidal rule in (recall, precision) space
  # Note: recall is non-decreasing by construction
  pr_auc_val <- sum(
    0.5 * (precision[-1] + precision[-length(precision)]) *
      diff(recall)
  )
  
  pr_auc_val
}

## =========================================================
## 4) Expected vs observed fire counts
## =========================================================
# Returns a named list: expected (sum of p), observed (sum of y)
expected_observed_counts <- function(p, y) {
  dat <- .clean_xy(p, y); p <- dat$p; y <- dat$y
  list(
    expected = sum(p),
    observed = sum(y)
  )
}

#note: we sum predicted probabilities because simulating 0/1 from the probability (ie sampling from a binomial distribution) adds noise - not good for model verification. 

#expected fire counts are summed rather than stochastically sampled since the expected sum of Bernoulli  trials provides a deterministic and unbiased estimate of aggregate event frequency. 

#later on, for something like forecasting or implementing into a DGVM, we would sample from a binomial distribution - variability contains information. 
```


```{r}
p <- fire$p_fire_bayes_sc
y <- fire$lightning_ignition

brier_score(p, y)
roc_auc(p, y)
pr_auc(p, y)

eo <- expected_observed_counts(p, y)
eo$expected
eo$observed
```


#ROC curve
```{r}
roc_curve_df <- function(p, y) {
  stopifnot(length(p) == length(y))
  stopifnot(all(y %in% c(0, 1)))

  # Remove NAs
  ok <- is.finite(p) & is.finite(y)
  p <- p[ok]
  y <- y[ok]

  # Sort by decreasing probability
  o <- order(p, decreasing = TRUE)
  p <- p[o]
  y <- y[o]

  P <- sum(y == 1)
  N <- sum(y == 0)

  tp <- cumsum(y == 1)
  fp <- cumsum(y == 0)

  tpr <- tp / P
  fpr <- fp / N

  # Add origin
  data.frame(
    fpr = c(0, fpr),
    tpr = c(0, tpr),
    threshold = c(Inf, p)
  )
}

roc_df <- roc_curve_df(p, y)

#add ROC AUC (area under the curve) to the plot
auc_text <- sprintf("ROC AUC = %.3f", roc_auc(p, y))

ggplot(roc_df, aes(fpr, tpr)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed", color = "grey50") +
  coord_equal() +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate",
    title = "ROC Curve"
  ) +
  annotate("text", x = 0.75, y = 0.125, label = auc_text, 
           color = "red", size = 4, fontface = "bold") +
  theme_bw()

```

ROC: If you pick a random fire day (true positive) and one non-fire day (false positive), how often does the model assign a higher probability to the fire day? ROC AUC = 0.895 means that ~89% of the time, the model ranks the fire day higher - good! this means the model separates high-risk from low-risk days. 

The steep rise near the origin means there is a high true positive rate at low false positive rates - the model can identify where fire is more likely. 

#PR Curve
```{r}
pr_curve_df <- function(p, y) {
  stopifnot(length(p) == length(y))
  stopifnot(all(y %in% c(0, 1)))

  ok <- is.finite(p) & is.finite(y)
  p <- p[ok]
  y <- y[ok]

  o <- order(p, decreasing = TRUE)
  p <- p[o]
  y <- y[o]

  P <- sum(y == 1)

  tp <- cumsum(y == 1)
  fp <- cumsum(y == 0)

  recall <- tp / P
  precision <- tp / (tp + fp)

  # Add baseline point
  data.frame(
    recall = c(0, recall),
    precision = c(mean(y), precision),
    threshold = c(Inf, p)
  )
}

pr_df <- pr_curve_df(p, y)

#annotate thresholds
annot_thresh <- c(1e-4, 5e-4, 1e-3, 2e-3)
#find closest rows to these thresholds
ann_df <- lapply(annot_thresh, function(t) {
  pr_df %>%
    mutate(diff = abs(threshold - t)) %>%
    slice_min(diff, n = 1)
}) %>%
  bind_rows() %>%
  mutate(label = paste0("t=", format(threshold, scientific = TRUE)))

auc_text <- sprintf("PR AUC = %.5f", pr_auc(p, y))

ggplot(pr_df, aes(recall, precision)) +
  geom_line(size = 1, color = "black") +
  geom_hline(yintercept = mean(y), linetype = "dashed", color = "grey50") +

  # annotation points
  geom_point(data = ann_df,
             aes(recall, precision),
             color = "red", size = 3) +

  geom_text(data = ann_df,
            aes(recall, precision, label = label),
            hjust = -0.1, vjust = 0,
            size = 3.5) +

  labs(
    x = "Recall",
    y = "Precision",
    title = "Precision–Recall Curve"
  ) +

  annotate("text", x = 0.8, y = 0.0006,
           label = auc_text,
           color = "red", size = 4, fontface = "bold") +

  theme_bw()


#add PR AUC (area under the curve) to the plot
# auc_text <- sprintf("PR AUC = %.5f", pr_auc(p, y))
# 
# ggplot(pr_df, aes(recall, precision, color=threshold)) +
#   geom_line(size = 1) +
#   geom_hline(yintercept = mean(y), linetype = "dashed", color = "grey50") +
#   labs(
#     x = "Recall",
#     y = "Precision",
#     title = "Precision–Recall Curve"
#   ) +
#   annotate("text", x = 0.8, y = 0.0006, label = auc_text, 
#            color = "red", size = 4, fontface = "bold") +
#   theme_bw()

```
Precision: high fire probabilities correspond to days with a fire
Recall: day with a fire was identified by the model as high risk 

Each row in pr_df is one decision threshold and performance you would get if you classified everything above that threshold as fire. 
For example, if threshold=0.0065, then a fire is predicted only where P_fire?=>= 0.0065. That is, when the threshold is high, only fires happen when the risk is extreme. When the threshold is low, 'warn aggressively'- more fires are triggered. In other words, start by only trusting the highest-risk predictions (threshold is high so fewer probabilites result in a fire), slowly increase the threshold (more probabilities results in fire) and watch how many fires the model catches versus how many false alarms it makes.

Looking at the plot, as thresholds decrease, more days are labeled fire, recall increases, and precision decreases. Recall increases because when more fires are predicted it catches more observed fires. Precision falls because when more fires are predicted, more of them will be on days that didn't see a fire. 

PR AUC is very small. But PR AUC should be interpreted relative to the baseline (dashed line = prevalence - fraction of fire days). 
observed fires = 272, total gridcell-days = over 4 million. So, the baseline precision (fire days / no fire days) is ~0.000067. The PR curve sits above the baseline, which means the model is better than random. 

When modeling rare events like fire, it's common for PR AUC to shrink with rarity. Here, the key takeaway is that predicted risk matches days with fire and the model is better than random across all thresholds. 

The curve concentrates fires in high-probability bins - gridcell-days with the highest predicted risk contain fires. This can be seen in the steep rise on the left hand side of the PR curve. Here, precision is high - days labeled high risk were observed to burn. Or, the model puts real fires disproportionately in the bins it labels high risk. While perfect classification is impossible, particularly for rare events, this model is able to identify high fire risk where fires cluster. 



#Reliability curve with quantile binning
(without quantile binning, all probabilities are in the first bin)
```{r}

reliability_curve_quantile <- function(p, y, nbins = 10) {
  dat <- .clean_xy(p, y)
  p <- dat$p; y <- dat$y
  
  # quantile-based breaks (to avoid empty bins) - based on the quantiles of predictions
  probs  <- seq(0, 1, length.out = nbins + 1)
  breaks <- quantile(p, probs = probs, na.rm = TRUE) 
  breaks <- unique(breaks)  # in case of ties
  
  df <- data.frame(p = p, y = y) %>%
    mutate(bin = cut(p, breaks = breaks, include.lowest = TRUE)) %>%
    group_by(bin) %>%
    summarise(
      n          = n(),
      mean_pred  = mean(p),
      event_rate = mean(y),
      .groups    = "drop"
    )
  
  # extract numeric bounds
  df$bin_lower <- as.numeric(sub("\\((.+),.*", "\\1", sub("\\[", "(", df$bin)))
  df$bin_upper <- as.numeric(sub(".*,(.+)\\]", "\\1", sub("\\)", "]", df$bin)))
  df$bin_mid   <- (df$bin_lower + df$bin_upper) / 2
  
  df %>%
    dplyr::select(bin_lower, bin_upper, bin_mid, n, mean_pred, event_rate)
}

rel_tab <- reliability_curve_quantile(p, y, nbins = 10)
rel_tab

ggplot(rel_tab, aes(x = mean_pred, y = event_rate)) +
  geom_point(shape = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlab("Mean predicted probability") +
  ylab("Observed event frequency") +
  ylim(0, 0.00065) +
  xlim(0, 0.00065) +
  theme_bw() 

rel_tab <- mutate(rel_tab, n_pred = n*mean_pred, n_event = n*event_rate)

ggplot(rel_tab, aes(x = n_pred, y = n_event)) +
  geom_point(shape = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlab("Summed predicted probability") +
  ylab("Summed observed event frequency") +
  ylim(0, 270) +
  xlim(0, 270) +
  theme_bw() 

```



Look at time
```{r}

# Position dodge setting (controls how far points are separated horizontally)
pd <- position_dodge(width = 0.4)


##########
#Lightning strike rate
##########
# Reshape to long format
  
df_long <- fire %>%
  dplyr::select(year, strikes, strikes_pred) %>%
  dplyr::filter(strikes > 0) %>%
  rename(observed=strikes, predicted=strikes_pred) %>%
  pivot_longer(cols = c("observed","predicted"),
               names_to = "lightning", values_to = "strike_rate")

df_summary <- df_long %>%
  group_by(year, lightning) %>%
  summarise(
    mean_strike_rate = mean(strike_rate, na.rm = TRUE),
    ci_low   = quantile(strike_rate, 0.025, na.rm = TRUE),
    ci_high  = quantile(strike_rate, 0.975, na.rm = TRUE),
    .groups  = "drop"
  )

# Compute 10-year observed mean 
obs_mean <- df_summary %>%
  filter(lightning == "observed") %>%
  summarise(mean_obs = mean(mean_strike_rate, na.rm = TRUE)) %>%
  pull(mean_obs)

obs_mean

# Plot
p1 <- ggplot(df_summary, aes(x = year, y = mean_strike_rate, color = lightning, shape = lightning)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.2) +
  geom_hline(yintercept = obs_mean, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(
    "predicted" = "red",
    "observed" = "gold")) +
  scale_shape_manual(values = c(
    "predicted" = 7,  # x
    "observed" = 16  # solid circle
  )) +
  labs(x = "Year", y = "Strikes/km2/month",
       title = "Lightning Strike Rate") +
  theme_minimal()

##########
#Ignition Efficiency
##########

df_long <- fire %>%
  dplyr::select(year, strike_count, fire_count, p_ignite_bayes_cal) %>%
  #filter(strike_count > 0) %>%
  dplyr::mutate(fires_per_strikes = fire_count / strike_count) %>%
  dplyr::mutate(p_ignite_bayes_cal = p_ignite_bayes_cal) %>%
  rename(observed=fires_per_strikes, predicted=p_ignite_bayes_cal) %>%
  pivot_longer(cols = c("observed","predicted"),
               names_to = "fire", values_to = "outcome")

df_summary <- df_long %>%
  group_by(year, fire) %>%
  summarise(mean_prob_freq = mean(outcome, na.rm = TRUE),
                ci_low   = quantile(outcome, 0.025, na.rm = TRUE),
                ci_high  = quantile(outcome, 0.975, na.rm = TRUE),
            .groups = "drop")

df_summary

# Compute 10-year observed mean 
obs_mean <- df_summary %>%
  filter(fire == "observed") %>%
  summarise(mean_obs = mean(mean_prob_freq, na.rm = TRUE)) %>%
  pull(mean_obs)

obs_mean

# Plot
p2 <- ggplot(df_summary, aes(x = year, y = mean_prob_freq, color = fire, shape = fire)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.2) +
  geom_hline(yintercept = obs_mean, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(
    "predicted" = "red",
    "observed" = "gold"),
    labels = c("predicted" = "monthly probability", "observed" = "fires/strikes/month")) +
  scale_shape_manual(values = c(
    "predicted" = 7,  # x
    "observed" = 16  # solid circle
  ),
    labels = c("predicted" = "monthly probability", "observed" = "fires/strikes/month")) +
  labs(x = "Year", y = "Mean probability/frequency",
       title = "Ignition Efficiency") +
  theme_minimal()

##########
#Fire Probability
##########
# Reshape to long format
df_long <- fire %>%
  dplyr::select(year, single_lightning_ignition, p_fire_bayes) %>%
  rename(observed=single_lightning_ignition, predicted=p_fire_bayes) %>%
  pivot_longer(cols = c("observed","predicted"),
               names_to = "fire", values_to = "outcome")

df_summary <- df_long %>%
  group_by(year, fire) %>%
  summarise(mean_prob_freq = mean(outcome, na.rm = TRUE),
            ci_low   = quantile(outcome, 0.025, na.rm = TRUE),
            ci_high  = quantile(outcome, 0.975, na.rm = TRUE),
            .groups = "drop")

df_summary

# Compute 10-year observed mean count
obs_mean <- df_summary %>%
  filter(fire == "observed") %>%
  summarise(mean_obs = mean(mean_prob_freq, na.rm = TRUE)) %>%
  pull(mean_obs)

obs_mean

# Plot
p3 <- ggplot(df_summary, aes(x = year, y = mean_prob_freq, color = fire, shape = fire)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.2) +
  geom_hline(yintercept = obs_mean, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(
    "predicted" = "red",
    "observed" = "gold")) +
  scale_shape_manual(values = c(
    "predicted" = 7,  # x
    "observed" = 16  # solid circle
  )) +
  labs(x = "Year", y = "probability/frequency",
       title = "Fire probability/frequency") +
  theme_minimal()

##########
#Fire count
##########
# Reshape to long format
df_long <- fire %>%
  dplyr::select(year, burn_pred_binary, single_lightning_ignition) %>%
  rename(observed=single_lightning_ignition, predicted=burn_pred_binary) %>%
  pivot_longer(cols = c("observed","predicted"),
               names_to = "fire", values_to = "count")

df_summary <- df_long %>%
  group_by(year, fire) %>%
  summarise(total_count = sum(count, na.rm = TRUE),
            .groups = "drop")

df_summary

# Compute 10-year observed mean count
obs_mean_count <- df_summary %>%
  filter(fire == "observed") %>%
  summarise(mean_obs = mean(total_count, na.rm = TRUE)) %>%
  pull(mean_obs)

obs_mean_count

# Plot
p4 <- ggplot(df_summary, aes(x = year, y = total_count, color = fire, shape = fire)) +
  geom_point(size = 3) +
  geom_hline(yintercept = obs_mean_count, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(
    "predicted" = "red",
    "observed" = "gold")) +
  scale_shape_manual(values = c(
    "predicted" = 7,  # x
    "observed" = 16  # solid circle
  )) +
  labs(x = "Year", y = "Fire count",
       title = "Fire Count") +
  theme_minimal()


##########
#Fire area
##########
# Reshape to long format
df_long <- fire %>%
  dplyr::select(year, burned_area_km2_total, burn_area_pred_conditioned) %>%
  rename(observed=burned_area_km2_total, predicted=burn_area_pred_conditioned) %>%
  pivot_longer(cols = c("observed","predicted"),
               names_to = "burned_area", values_to = "area_km")

df_long$area_km <- as.numeric(df_long$area_km)
# Summarize mean and sd for each year/model
df_summary <- df_long %>%
  group_by(year, burned_area) %>%
  summarise(mean_area = mean(area_km, na.rm = TRUE),
            ci_low   = quantile(area_km, 0.025, na.rm = TRUE),
            ci_high  = quantile(area_km, 0.975, na.rm = TRUE),
            .groups = "drop")

df_summary

# Compute 10-year observed mean
obs_mean_total <- df_summary %>%
  filter(burned_area == "observed") %>%
  summarise(mean_obs = mean(mean_area, na.rm = TRUE)) %>%
  pull(mean_obs)

obs_mean_total

#total burned area for each year
obs_mean_sum <- df_summary %>%
  filter(burned_area == "observed") %>%
  group_by(year) %>%
  summarise(obs_mean_sum = sum(mean_area, na.rm = TRUE)) %>%
  pull(obs_mean_sum)

obs_mean_sum

#10 year mean annual total
obs_mean_sum <- sum(obs_mean_sum)/10

obs_mean_sum

# Plot
p5 <- ggplot(df_summary, aes(x = year, y = mean_area, color = burned_area, shape = burned_area)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low,
                    ymax = ci_high),
                width = 0.2) +
  geom_hline(yintercept = obs_mean, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(
    "predicted" = "red",
    "observed" = "gold")) +
  scale_shape_manual(values = c(
    "predicted" = 7,  # x
    "observed" = 16  # solid circle
    #"N13"      = 17,  # solid triangle
    #"G13"      = 15,  # solid square
    #"B12"      = 18   # solid diamond
  )) +
  labs(x = "Year", y = "Burned Area (km2)",
       title = "") +
  theme_minimal()

p6 <- ggplot(df_summary, aes(x = year, y = mean_area, color = burned_area, shape = burned_area)) +
  geom_point(size = 3) +
  geom_hline(yintercept = obs_mean_sum, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(
    "predicted" = "red",
    "observed" = "gold")) +
  scale_shape_manual(values = c(
    "predicted" = 7,  # x
    "observed" = 16  # solid circle
    #"N13"      = 17,  # solid triangle
    #"G13"      = 15,  # solid square
    #"B12"      = 18   # solid diamond
  )) +
  labs(x = "Year", y = "Total burned Area (km2)",
       title = "") +
  theme_minimal()

p1 / p2 / p3 / p4 / p6

```

Alaska polygon and Elevation plot
```{r}
# Get USA polygon
usa <- ne_states(country = "United States of America", returnclass = "sf")
# Filter to Alaska only
alaska <- usa[usa$name == "Alaska", ]
alaska <- st_transform(alaska, crs = 4326)

library(geodata)
library(terra)

# Download SRTM elevation data for Alaska
# Alaska spans multiple tiles so we download by country

# 1. Download and crop
elev <- geodata::elevation_global(res = 10, path = tempdir())
alaska_ext <- ext(-170, -130, 51, 72)
elev_alaska <- crop(elev, alaska_ext)

# 2. Mask to Alaska polygon
alaska_vect <- terra::vect(alaska)
elev_masked <- terra::mask(elev_alaska, alaska_vect)

# 3. Convert to data frame — use elev_masked instead of elev_alaska
elev_df <- as.data.frame(elev_masked, xy = TRUE)
names(elev_df) <- c("lon", "lat", "elevation")
elev_df <- elev_df %>% filter(!is.na(elevation), elevation >= 0)

p_elevation <- ggplot() +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250)) +  # contour interval in metres
  geom_sf(data = alaska, fill = NA, color = "black",
          linewidth = 0.3, inherit.aes = FALSE) +
  xlim(-170, -130) +
  ylim(51, 72) +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(h)") +
  theme_minimal() +
  theme(
    axis.title.x     = element_text(size=11),
    axis.title.y     = element_text(size=11),
    axis.text.y      = element_text(size=11),
    axis.text.x      = element_text(size=11),
    panel.grid.major = element_line(color = "darkgray", size = 0.2),
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)
  )

p_elevation
```
Make map data
```{r}
#calculate bias for lightning prediction
plot_df <- fire %>%
  group_by(lat, lon) %>%
  summarise(
    fire_bias = mean(p_fire) - mean(as.numeric(single_lightning_ignition)),
    fire_obs = mean(as.numeric(single_lightning_ignition)),
    fire_pred = mean(p_fire),
    .groups = "drop"
  )

#global min and max strike rates
min_value_fire <- min(c(plot_df$fire_obs, plot_df$fire_pred), na.rm = TRUE)
max_value_fire <- max(c(plot_df$fire_obs, plot_df$fire_pred), na.rm = TRUE)
```

Make maps!

```{r}
p_fire <- ggplot(plot_df, aes(x = lon, y = lat, fill = fire_pred)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    #limits = c(min_value_fire, max_value_fire), 
    name = expression(atop("Predicted fire probability", "P(fire|p_ignite, lightning)")),
    guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_fire_obs <- ggplot(plot_df, aes(x = lon, y = lat, fill = fire_obs)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    #limits = c(min_value_fire, max_value_fire), 
    name = expression(atop("Observed fire frequency", "")),
    guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_fire_bias <- ggplot(plot_df, aes(x = lon, y = lat, fill = fire_bias)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_distiller(
    palette = "PuOr",
    #limits = c(-max_abs_bias, max_abs_bias),
    name = expression(atop("Mean Bias", "pred-obs")),
    guide = guide_colorbar(direction = "horizontal", title.position = "top")
    ) +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_fire + p_fire_obs + p_fire_bias
```



Make map data - burned area
```{r}
#calculate bias for lightning prediction
plot_df <- fire %>%
  group_by(lat, lon) %>%
  summarise(
    ba_bias = mean(burn_area_outcome) - mean(as.numeric(burned_area_km2_total)),
    ba_obs = mean(as.numeric(burned_area_km2_total)),
    ba_pred = mean(burn_area_outcome),
    .groups = "drop"
  )

#global min and max strike rates
min_value_ba <- min(c(plot_df$ba_obs, plot_df$ba_pred), na.rm = TRUE)
max_value_ba <- max(c(plot_df$ba_obs, plot_df$ba_pred), na.rm = TRUE)
```

Make maps!

```{r}
p_ba <- ggplot(plot_df, aes(x = lon, y = lat, fill = ba_pred)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    limits = c(min_value_ba, max_value_ba), 
    name = expression(atop("Predicted burned area", "km"^2)),
    guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(a)") + 
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_ba_obs <- ggplot(plot_df, aes(x = lon, y = lat, fill = ba_obs)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    limits = c(min_value_ba, max_value_ba), 
    name = expression(atop("Observed burned area", "km"^2)),
    guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(b)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_ba_bias <- ggplot(plot_df, aes(x = lon, y = lat, fill = ba_bias)) +
  geom_raster() +
  xlim(-170, -130) +
  geom_sf(data = alaska, fill = NA, color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    name = expression(atop("Burned area bias", "km"^2)),
    guide = guide_colorbar(direction = "horizontal", title.position = "top"))  +
  xlab("Lon") +
  ylab("Lat") +
  ggtitle("(c)") +
  theme_minimal() +
  theme(legend.position="bottom") +
  theme(
    axis.title.x = element_text(size=11),
    axis.title.y = element_text(size=11),
    axis.text.y = element_text(size=11),
    axis.text.x = element_text(size=11),
    legend.text = element_text(size = 8, angle = 45, hjust = 1),
    legend.title = element_text(size = 8, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.position = c(0.65, 0.16)) +
  geom_contour(data = elev_df,
               aes(x = lon, y = lat, z = elevation),
               color = "black",
               linewidth = 0.1,
               breaks = seq(0, 3000, by = 250),
               inherit.aes = FALSE) 

p_ba + p_ba_obs + p_ba_bias
```