---
title: "Monthly Surface Pressure Corrections"
author: "Charlotte Uden"
date: "2026-08-07"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(pROC)
library(loo)
library(cowplot)
library(stats)
library(caret)
library(arrow)
library(geodata)
library(terra)
library(rnaturalearth)
library(sf)
library(patchwork)
library(purrr)
library(qmap)
library(FNN)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

Read in elvation and trace points:
```{r}
trace_elev_lookup <- readRDS(file.path(rds_root, "trace_elev_lookup.rds"))
trace_cells <- readRDS(file.path(rds_root, "trace_cells.rds"))
```

From python file that converted netcdf to parquet:
Note: TraCE model years are offset from BP/CE. Year 0 in the model = 22000 BP.
    We convert to approximate CE year: ce_year = model_year - 21950
    (where 1950 CE = 0 BP = model year 21950)

```{r}
# Read the TRACE parquet file
trace_monthly_full <- read_parquet(file.path(paleocliamte_data_root,"trace21ka_processed/trace_ignition_burnedarea_model.parquet"))
names(trace_monthly_full)
range(trace_monthly_full$ce_year)

trace_monthly <- trace_monthly_full %>% filter(ce_year >= 46, ce_year <= 89)

#make year match conventional dating
trace_monthly <- trace_monthly %>%
  mutate(year = ce_year + 1900)

trace_monthly_full <- trace_monthly_full %>%
  mutate(year = ce_year + 1900)

# Sanity check
range(trace_monthly$year)
range(trace_monthly_full$year)
```

Spatial clip to Alaska
```{r}
trace_monthly_full <- trace_monthly_full %>%
  mutate(lon = ifelse(lon > 180, lon - 360, lon)) %>%   # match trace_cells' -180/180 convention
  semi_join(trace_cells, by = c("lat", "lon")) %>%
  left_join(trace_elev_lookup, by = c("lat", "lon"))

trace_monthly <- trace_monthly %>%
  mutate(lon = ifelse(lon > 180, lon - 360, lon)) %>%   # match trace_cells' -180/180 convention
  semi_join(trace_cells, by = c("lat", "lon")) %>%
  left_join(trace_elev_lookup, by = c("lat", "lon"))

# Sanity checks
nrow(trace_monthly_full)
n_distinct(trace_monthly_full$lat, trace_monthly_full$lon)   # should be 23
sum(is.na(trace_monthly_full$elevation))                      # should be 0
names(trace_monthly_full)
# Sanity checks
nrow(trace_monthly)
n_distinct(trace_monthly$lat, trace_monthly$lon)   # should be 23
sum(is.na(trace_monthly$elevation))                      # should be 0
names(trace_monthly)
```



Read in ERA5 monthly data for correction period:
```{r}
era_monthly <- read_parquet(file.path(era5_data_root,"bias_correction/processed_parquet/era5_ignition_burnedarea_model_1946_1990.parquet"))
range(era_monthly$year)
names(era_monthly)
```

A spatial mis match: Aggregate ERA5 up to TraCE's grid by veraging all ERA5 points that fall within each TraCE grid cell. This maintains TraCE's actual physical resolution and avoids inventing false precision (interpolating trace down to era5 grid). 

```{r}
#save unaggregated ERA5 data:
era_monthly_pre_aggregation <- era_monthly
# Match each ERA5 point to nearest TraCE grid cell
nn <- get.knnx(data = trace_cells[, c("lon", "lat")],
               query = era_monthly[, c("lon", "lat")], k = 1)

era_monthly$trace_lon <- trace_cells$lon[nn$nn.index]
era_monthly$trace_lat <- trace_cells$lat[nn$nn.index]

# Aggregate ERA5 to TraCE grid cells, by year, month
era_monthly<- era_monthly %>%
  group_by(trace_lat, trace_lon, year, month) %>%
  summarise(across(c(sp), mean, na.rm = TRUE),
            .groups = "drop") %>%
  rename(lat = trace_lat, lon = trace_lon)

ggplot(filter(era_monthly, year==1946 & month==6), aes(x=lon, y=lat)) + geom_point()
ggplot(filter(trace_monthly, year==1946 & month==6), aes(x=lon, y=lat)) + geom_point()
```
# ========================================================================================================================
# ==========================================    JUNE          ============================================================
# ========================================================================================================================


# ============================================================
# DEFINE VARIABLE + MONTH FOR THIS RUN
# Change these two lines and re-run top to bottom, once per month (6, 7, 8)
# ============================================================
```{r}
var_name  <- "sp"
month_val <- 6   # 6 = June, 7 = July, 8 = August

era_col   <- var_name
trace_col <- var_name

# Filter to this month only, right at the source - everything downstream inherits this scope
era_calib   <- era_monthly   %>% filter(month == month_val)
trace_calib <- trace_monthly %>% filter(month == month_val)
trace_full  <- trace_monthly_full %>% filter(month == month_val)

# Dynamically-named output columns (unchanged across months - the month scoping lives in the data, not the names)
bias_col      <- paste0(var_name, "_predicted_bias")
corrected_col <- paste0(var_name, "_corrected")
var_ratio_col <- paste0(var_name, "_var_ratio_pred")
cell_mean_col <- paste0(var_name, "_cell_mean")
rescaled_col  <- paste0(var_name, "_rescaled")
final_col     <- paste0(var_name, "_final")
```

# ============================================================
# Small reusable plotting helpers (pure boilerplate — logic stays inline below)
# ============================================================
```{r}
plot_density_compare <- function(df_list, value_cols, labels, var_label, month_label = NULL) {
  df_long <- purrr::map2_dfr(df_list, seq_along(df_list), function(df, i) {
    tibble(value = df[[value_cols[i]]], source = labels[i])
  })
  title <- if (!is.null(month_label)) paste0(var_label, " — month ", month_label) else var_label
  ggplot(df_long, aes(x = value, color = source)) +
    geom_density(linewidth = 0.9, fill = NA) +
    labs(title = title, x = var_label, y = "Density", color = NULL) +
    theme_minimal() + theme(legend.position = "bottom")
}

plot_timeseries_compare <- function(df_list, year_cols, value_cols, labels, var_label, month_label = NULL) {
  df_long <- purrr::pmap_dfr(list(df_list, year_cols, value_cols, labels),
                              function(df, yc, vc, lab) {
    df %>% group_by(year = .data[[yc]]) %>%
      summarise(mean_val = mean(.data[[vc]], na.rm = TRUE), .groups = "drop") %>%
      mutate(source = lab)
  })
  title <- if (!is.null(month_label)) paste0(var_label, " — month ", month_label) else var_label
  ggplot(df_long, aes(x = year, y = mean_val, color = source)) +
    geom_line(alpha = 0.7) +
    labs(title = title, y = paste("Mean", var_label), color = NULL) +
    theme_minimal()
}

plot_timeseries_calib <- function(era_df, era_year_col, era_value_col,
                                   trace_df, trace_year_col, trace_value_col,
                                   var_label, month_label = NULL) {
  
  era_summary <- era_df %>%
    group_by(year = .data[[era_year_col]]) %>%
    summarise(mean_val = mean(.data[[era_value_col]], na.rm = TRUE), .groups = "drop") %>%
    mutate(source = "ERA5 (calib)")
  
  trace_summary <- trace_df %>%
    group_by(year = .data[[trace_year_col]]) %>%
    summarise(mean_val = mean(.data[[trace_value_col]], na.rm = TRUE), .groups = "drop") %>%
    mutate(source = "TraCE (calib)")
  
  df_long <- bind_rows(era_summary, trace_summary)
  
  title <- if (!is.null(month_label)) paste0(var_label, " — month ", month_label) else var_label
  
  ggplot(df_long, aes(x = year, y = mean_val, color = source)) +
    geom_line(alpha = 0.7) +
    scale_color_manual(values = c("ERA5 (calib)" = "black", "TraCE (calib)" = "#2E86C1")) +
    labs(title = title, y = paste("Mean", var_label), color = NULL) +
    theme_minimal()
}

quantile_compare_table <- function(df_list, value_cols, labels,
                                    probs = c(0.01, 0.05, 0.10, 0.90, 0.95, 0.99)) {
  purrr::pmap_dfr(list(df_list, value_cols, labels), function(df, vc, lab) {
    q <- quantile(df[[vc]], probs, na.rm = TRUE)
    tibble(source = lab, prob = names(q), value = q)
  }) %>% pivot_wider(names_from = source, values_from = value)
}
```
# ============================================================
# STEP 1: Compare ERA5 calib / TraCE calib / TraCE full RAW — this month only
# Does the full paleo record fall outside the calibration TraCE range?
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)"),
  var_name, month_val
)

plot_timeseries_compare(
  list(era_calib, trace_calib, trace_full),
  c("year", "year", "year"),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = trace_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)")
)

range(trace_calib[[trace_col]], na.rm = TRUE)
range(trace_full[[trace_col]], na.rm = TRUE)
# If trace_full's range is meaningfully wider than trace_calib's -> QM will extrapolate -> flag for Step 7
```
# ============================================================
# STEP 2: Calculate bias for the calibration period — this month only
# ============================================================
```{r}
compare_df_m <- era_calib %>%
  inner_join(trace_calib, by = c("lat", "lon", "year"), suffix = c("_era", "_trace"))
  # adjust join columns if your monthly frames key differently

bias_summary_m <- tibble(
  variable = var_name, month = month_val,
  mean_bias = mean(compare_df_m[[paste0(var_name, "_trace")]] -
                    compare_df_m[[paste0(var_name, "_era")]], na.rm = TRUE),
  correlation = cor(compare_df_m[[paste0(var_name, "_trace")]],
                     compare_df_m[[paste0(var_name, "_era")]], use = "complete.obs"),
  era_sd = sd(compare_df_m[[paste0(var_name, "_era")]], na.rm = TRUE),
  trace_sd = sd(compare_df_m[[paste0(var_name, "_trace")]], na.rm = TRUE)
)
bias_summary_m

cellwise_bias_m <- compare_df_m %>%
  group_by(lat, lon, elevation) %>%
  summarise(
    bias = mean(.data[[paste0(var_name, "_trace")]] - .data[[paste0(var_name, "_era")]], na.rm = TRUE),
    .groups = "drop"
  )
cellwise_bias_m

ggplot(cellwise_bias_m, aes(x=lon, y=lat, size=abs(bias), color=bias)) + geom_point()


cor(cellwise_bias_m$bias, cellwise_bias_m$lon)
cor(cellwise_bias_m$bias, cellwise_bias_m$lat)
cor(cellwise_bias_m$bias, cellwise_bias_m$lon)
```
# ============================================================
# STEP 3: Fit candidate mean-bias models — this month only
# Fully manual, same as every variable so far — inspect, then assign the winner explicitly
# ============================================================
```{r}
m_elev   <- lm(bias ~ elevation, data = cellwise_bias_m)
m_lat    <- lm(bias ~ lat, data = cellwise_bias_m)
m_lon    <- lm(bias ~ lon, data = cellwise_bias_m)
m_latlon <- lm(bias ~ lat + lon, data = cellwise_bias_m)
m_full   <- lm(bias ~ elevation + lat + lon, data = cellwise_bias_m)
m_flat <- lm(bias ~ 1, data = cellwise_bias_m)

summary(m_elev); summary(m_lat); summary(m_lon); summary(m_latlon); summary(m_full); summary(m_flat)
anova(m_lat, m_latlon)
anova(m_latlon, m_full)
anova(m_lat, m_flat)
AIC(m_elev, m_lat, m_lon, m_latlon, m_full, m_flat)

cellwise_bias_m$resid_check <- residuals(m_latlon)  # swap in whichever you're evaluating
cor(cellwise_bias_m$resid_check, cellwise_bias_m$elevation)

cor(cellwise_bias_m$bias, cellwise_bias_m$elevation)
plot(cellwise_bias_m$elevation, cellwise_bias_m$bias)
```
Strange that elevation doesnt stant out. Maybe it will for variance ratio. Use flat bias correction. 
```{r}
# ---- Set explicitly once you've reviewed the output above ----
mean_bias_model <- m_flat

trace_calib[[bias_col]] <- predict(mean_bias_model, newdata = trace_calib)
trace_calib[[corrected_col]] <- trace_calib[[trace_col]] - trace_calib[[bias_col]]

trace_full[[bias_col]] <- predict(mean_bias_model, newdata = trace_full)
trace_full[[corrected_col]] <- trace_full[[trace_col]] - trace_full[[bias_col]]
```

# ============================================================
# STEP 4: Check corrected mean — this month only
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, corrected_col, corrected_col),
  c("ERA5 (calib)", "TraCE (calib, corrected)", "TraCE (full, corrected)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = corrected_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, corrected_col, corrected_col),
  c("ERA5 (calib)", "TraCE (calib, corrected)", "TraCE (full, corrected)")
)

cellwise_variance_m <- compare_df_m %>%
  mutate(!!corrected_col := .data[[paste0(var_name, "_trace")]] -
           predict(mean_bias_model, newdata = compare_df_m)) %>%
  group_by(lat, lon, elevation) %>%
  summarise(
    sd_era = sd(.data[[paste0(var_name, "_era")]], na.rm = TRUE),
    sd_corrected = sd(.data[[corrected_col]], na.rm = TRUE),
    var_ratio = sd_corrected / sd_era,
    .groups = "drop"
  )
cellwise_variance_m %>% arrange(desc(abs(var_ratio - 1)))
summary(cellwise_variance_m$var_ratio)

plot(cellwise_variance_m$elevation, cellwise_variance_m$var_ratio)
```
# ============================================================
# STEP 5: Decide if variance needs correcting — this month only
# ============================================================
```{r}
vr_elev   <- lm(var_ratio ~ elevation, data = cellwise_variance_m)
vr_lat    <- lm(var_ratio ~ lat, data = cellwise_variance_m)
vr_lon    <- lm(var_ratio ~ lon, data = cellwise_variance_m)
vr_latlon <- lm(var_ratio ~ lat + lon, data = cellwise_variance_m)
vr_full   <- lm(var_ratio ~ elevation + lat + lon, data = cellwise_variance_m)
vr_flat <- lm(var_ratio ~ 1, data = cellwise_variance_m)

summary(vr_elev); summary(vr_lat); summary(vr_lon); summary(vr_latlon); summary(vr_full); summary(vr_flat)

anova(vr_lat, vr_latlon)
anova(vr_full, vr_latlon)
anova(vr_full, vr_lat)

AIC(vr_elev, vr_lat, vr_lon, vr_latlon, vr_full, vr_flat)

```

June sp variance ratio shows strong latitude dependence (R²=0.81); elevation initially appeared visually associated with increased spread but did not hold up in direct model comparison, likely reflecting elevation's moderate correlation with latitude in the 23-cell domain rather than a direct elevation effect.

Go with vr_lat:
```{r}
# ---- Set explicitly, or skip Step 5 entirely if var_ratio was already close to 1 ----
var_ratio_model <- vr_lat

trace_calib[[var_ratio_col]] <- predict(var_ratio_model, newdata = trace_calib)
trace_calib <- trace_calib %>%
  group_by(lat, lon) %>%
  mutate(
    !!cell_mean_col := mean(.data[[corrected_col]], na.rm = TRUE),
    !!rescaled_col := .data[[cell_mean_col]] +
      (.data[[corrected_col]] - .data[[cell_mean_col]]) / .data[[var_ratio_col]]
  ) %>%
  ungroup()

trace_full[[var_ratio_col]] <- predict(var_ratio_model, newdata = trace_full)
trace_full <- trace_full %>%
  group_by(lat, lon) %>%
  mutate(
    !!cell_mean_col := mean(.data[[corrected_col]], na.rm = TRUE),
    !!rescaled_col := .data[[cell_mean_col]] +
      (.data[[corrected_col]] - .data[[cell_mean_col]]) / .data[[var_ratio_col]]
  ) %>%
  ungroup()
```
# ============================================================
# STEP 6: Check rescaled values — this month only
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, rescaled_col, rescaled_col),
  c("ERA5 (calib)", "TraCE (calib, rescaled)", "TraCE (full, rescaled)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = rescaled_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, rescaled_col, rescaled_col),
  c("ERA5 (calib)", "TraCE (calib, rescaled)", "TraCE (full, rescaled)")
)


cellwise_check_m <- trace_calib %>%
  left_join(
    era_calib %>% dplyr::select(lat, lon, year, !!paste0(era_col, "_era_ref") := all_of(era_col)),
    by = c("lat", "lon", "year")
  ) %>%
  group_by(lat, lon) %>%
  summarise(
    sd_era = sd(.data[[paste0(era_col, "_era_ref")]], na.rm = TRUE),
    sd_rescaled = sd(.data[[rescaled_col]], na.rm = TRUE),
    var_ratio_after = sd_rescaled / sd_era,
    .groups = "drop"
  )

cellwise_check_m %>% arrange(desc(abs(var_ratio_after - 1)))

range(cellwise_check_m$var_ratio_after)

#variance ratio still doesnt 
#how does rescaling change varaince ratio?
summary(cellwise_variance_m$var_ratio) #corrected
summary(cellwise_check_m$var_ratio_after) #rescaled

summary(era_calib$sp) 
summary(trace_calib$sp) 
summary(trace_full$sp) 
summary(trace_calib$sp_corrected) 
summary(trace_full$sp_corrected) 
summary(trace_calib$sp_rescaled)
summary(trace_full$sp_rescaled)

```

Plot century trend for June
```{r}
plot_century_trend <- function(full_data, var_col,
                                era_data, era_col,
                                calib_data, calib_col,
                                var_label = var_col,
                                bin_width = 100) {
  
  # Bin the full corrected record into century spows
  df_summary <- full_data %>%
    dplyr::select(year, all_of(var_col)) %>%
    mutate(century_bin = floor(year / bin_width) * bin_width) %>%
    group_by(century_bin) %>%
    summarise(
      mean_val = mean(.data[[var_col]], na.rm = TRUE),
      ci_low   = quantile(.data[[var_col]], 0.025, na.rm = TRUE),
      ci_high  = quantile(.data[[var_col]], 0.975, na.rm = TRUE),
      n_obs    = n(),
      .groups  = "drop"
    )
  
  # ERA5 calibration-period reference mean
  era5_mean <- era_data %>%
    group_by(year, lat, lon) %>%
    summarise(mean_obs = mean(.data[[era_col]], na.rm = TRUE), .groups = "drop") %>%
    pull(mean_obs) %>%
    mean(na.rm = TRUE)
  
  breaks_seq <- as.character(seq(min(df_summary$century_bin), max(df_summary$century_bin), by = 500))
  
  ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_val)) +
    geom_point(size = 3, alpha = 0.2) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.1, alpha = 0.2) +
    geom_hline(yintercept = era5_mean, linetype = "dotted", color = "red", linewidth = 0.5) +
    scale_x_discrete(breaks = breaks_seq) +
    theme_minimal() +
    labs(x = paste0("Year (", bin_width, "-yr bin)"), y = var_label, title = "") +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      legend.title = element_blank()
    )
}

#plot corrected full trace for all months:
plot_century_trend(
  full_data = trace_full, var_col = "sp_rescaled",
  era_data = era_monthly, era_col = "sp",
  var_label = "Corrected sp"
)
```
Accept rescaled values as final correction. No quantile mapping for sp, since calibration period sp speeds are lower than the full paleo period. Again, quantile mapping would calmp low sp values.
range(trace_calib[[trace_col]], na.rm = TRUE)
92191.31 101174.02
range(trace_full[[trace_col]], na.rm = TRUE)
89292.64 101941.20

```{r}
trace_calib[[final_col]] <- trace_calib[[rescaled_col]]
trace_full[[final_col]]  <- trace_full[[rescaled_col]]
```

# ============================================================
# STEP 7: Apply QM if Step 1 showed it's viable; otherwise finalize as-is
# ============================================================
```{r}
# # ---- OPTION A: QM viable ----
# qm_model <- fitQmap(compare_df_m[[paste0(var_name, "_era")]],
#                      trace_calib[[rescaled_col]],  # or [[corrected_col]] if Step 5 was skipped
#                      method = "QUANT", qstep = 0.01, wet.day = FALSE)
# 
# trace_calib[[final_col]] <- doQmap(trace_calib[[rescaled_col]], qm_model)
# trace_full[[final_col]]  <- doQmap(trace_full[[rescaled_col]], qm_model)
# 
# # ---- OPTION B: QM excluded ----
# # trace_calib[[final_col]] <- trace_calib[[rescaled_col]]
# # trace_full[[final_col]]  <- trace_full[[rescaled_col]]
# 
# plot_density_compare(
#   list(era_calib, trace_calib, trace_full),
#   c(era_col, final_col, final_col),
#   c("ERA5 (calib)", "TraCE (calib, final)", "TraCE (full, final)"),
#   var_name, month_val
# )
```
# ============================================================
# STORE this month's results and models before moving to the next month
# ============================================================
```{r}
if (!exists("monthly_models")) monthly_models <- list()
if (!exists("monthly_results_calib")) monthly_results_calib <- list()
if (!exists("monthly_results_full")) monthly_results_full <- list()

monthly_models[[as.character(month_val)]] <- list(
  mean_model = mean_bias_model,
  var_ratio_model = if (exists("var_ratio_model")) var_ratio_model else NULL,
  qm_model = if (exists("qm_model")) qm_model else NULL
)
monthly_results_calib[[as.character(month_val)]] <- trace_calib
monthly_results_full[[as.character(month_val)]]  <- trace_full

# Once you've done this for month_val <- 6, 7, and 8, combine:
# trace_calib_all_months <- bind_rows(monthly_results_calib)
# trace_full_all_months  <- bind_rows(monthly_results_full)
```


# ========================================================================================================================
# ==========================================    JULY          ============================================================
# ========================================================================================================================

# ============================================================
# DEFINE VARIABLE + MONTH FOR THIS RUN
# Change these two lines and re-run top to bottom, once per month (6, 7, 8)
# ============================================================

```{r}
var_name  <- "sp"
month_val <- 7   # 6 = June, 7 = July, 8 = August

era_col   <- var_name
trace_col <- var_name

# Filter to this month only, right at the source - everything downstream inherits this scope
era_calib   <- era_monthly   %>% filter(month == month_val)
trace_calib <- trace_monthly %>% filter(month == month_val)
trace_full  <- trace_monthly_full %>% filter(month == month_val)

# Dynamically-named output columns (unchanged across months - the month scoping lives in the data, not the names)
bias_col      <- paste0(var_name, "_predicted_bias")
corrected_col <- paste0(var_name, "_corrected")
var_ratio_col <- paste0(var_name, "_var_ratio_pred")
cell_mean_col <- paste0(var_name, "_cell_mean")
rescaled_col  <- paste0(var_name, "_rescaled")
final_col     <- paste0(var_name, "_final")
```


# ============================================================
# STEP 1: Compare ERA5 calib / TraCE calib / TraCE full RAW — this month only
# Does the full paleo record fall outside the calibration TraCE range?
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)"),
  var_name, month_val
)

plot_timeseries_compare(
  list(era_calib, trace_calib, trace_full),
  c("year", "year", "year"),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = trace_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)")
)

range(trace_calib[[trace_col]], na.rm = TRUE)
range(trace_full[[trace_col]], na.rm = TRUE)
# If trace_full's range is meaningfully wider than trace_calib's -> QM will extrapolate -> flag for Step 7
```
# ============================================================
# STEP 2: Calculate bias for the calibration period — this month only
# ============================================================
```{r}
compare_df_m <- era_calib %>%
  inner_join(trace_calib, by = c("lat", "lon", "year"), suffix = c("_era", "_trace"))
  # adjust join columns if your monthly frames key differently

bias_summary_m <- tibble(
  variable = var_name, month = month_val,
  mean_bias = mean(compare_df_m[[paste0(var_name, "_trace")]] -
                    compare_df_m[[paste0(var_name, "_era")]], na.rm = TRUE),
  correlation = cor(compare_df_m[[paste0(var_name, "_trace")]],
                     compare_df_m[[paste0(var_name, "_era")]], use = "complete.obs"),
  era_sd = sd(compare_df_m[[paste0(var_name, "_era")]], na.rm = TRUE),
  trace_sd = sd(compare_df_m[[paste0(var_name, "_trace")]], na.rm = TRUE)
)
bias_summary_m

cellwise_bias_m <- compare_df_m %>%
  group_by(lat, lon, elevation) %>%
  summarise(
    bias = mean(.data[[paste0(var_name, "_trace")]] - .data[[paste0(var_name, "_era")]], na.rm = TRUE),
    .groups = "drop"
  )
cellwise_bias_m

ggplot(cellwise_bias_m, aes(x=lon, y=lat, size=abs(bias), color=bias)) + geom_point()

cor(cellwise_bias_m$bias, cellwise_bias_m$lon)
cor(cellwise_bias_m$bias, cellwise_bias_m$lat)
cor(cellwise_bias_m$bias, cellwise_bias_m$elevation)
```
# ============================================================
# STEP 3: Fit candidate mean-bias models — this month only
# Fully manual, same as every variable so far — inspect, then assign the winner explicitly
# ============================================================
```{r}
m_elev   <- lm(bias ~ elevation, data = cellwise_bias_m)
m_lat    <- lm(bias ~ lat, data = cellwise_bias_m)
m_lon    <- lm(bias ~ lon, data = cellwise_bias_m)
m_latlon <- lm(bias ~ lat + lon, data = cellwise_bias_m)
m_full   <- lm(bias ~ elevation + lat + lon, data = cellwise_bias_m)
m_flat <- lm(bias ~ 1, data = cellwise_bias_m)

summary(m_elev); summary(m_lat); summary(m_lon); summary(m_latlon); summary(m_full); summary(m_flat)
anova(m_lat, m_latlon)
anova(m_latlon, m_full)
anova(m_lat, m_flat)

AIC(m_elev, m_lat, m_lon, m_latlon, m_full, m_flat)

cellwise_bias_m$resid_check <- residuals(m_flat)  # swap in whichever you're evaluating
cor(cellwise_bias_m$resid_check, cellwise_bias_m$elevation)
```

Go with flat bias correction. 
```{r}
# ---- Set explicitly once you've reviewed the output above ----
mean_bias_model <- m_flat

trace_calib[[bias_col]] <- predict(mean_bias_model, newdata = trace_calib)
trace_calib[[corrected_col]] <- trace_calib[[trace_col]] - trace_calib[[bias_col]]

trace_full[[bias_col]] <- predict(mean_bias_model, newdata = trace_full)
trace_full[[corrected_col]] <- trace_full[[trace_col]] - trace_full[[bias_col]]
```
# ============================================================
# STEP 4: Check corrected mean — this month only
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, corrected_col, corrected_col),
  c("ERA5 (calib)", "TraCE (calib, corrected)", "TraCE (full, corrected)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = corrected_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, corrected_col, corrected_col),
  c("ERA5 (calib)", "TraCE (calib, corrected)", "TraCE (full, corrected)")
)

cellwise_variance_m <- compare_df_m %>%
  mutate(!!corrected_col := .data[[paste0(var_name, "_trace")]] -
           predict(mean_bias_model, newdata = compare_df_m)) %>%
  group_by(lat, lon, elevation) %>%
  summarise(
    sd_era = sd(.data[[paste0(var_name, "_era")]], na.rm = TRUE),
    sd_corrected = sd(.data[[corrected_col]], na.rm = TRUE),
    var_ratio = sd_corrected / sd_era,
    .groups = "drop"
  )
cellwise_variance_m %>% arrange(desc(abs(var_ratio - 1)))
summary(cellwise_variance_m$var_ratio)
```
# ============================================================
# STEP 5: Decide if variance needs correcting — this month only
# ============================================================
```{r}
vr_elev   <- lm(var_ratio ~ elevation, data = cellwise_variance_m)
vr_lat    <- lm(var_ratio ~ lat, data = cellwise_variance_m)
vr_lon    <- lm(var_ratio ~ lon, data = cellwise_variance_m)
vr_latlon <- lm(var_ratio ~ lat + lon, data = cellwise_variance_m)
vr_lonelev <- lm(var_ratio ~ lon + elevation, data = cellwise_variance_m)
vr_full   <- lm(var_ratio ~ elevation + lat + lon, data = cellwise_variance_m)
vr_flat <- lm(var_ratio ~ 1, data = cellwise_variance_m)

summary(vr_elev); summary(vr_lat); summary(vr_lon); summary(vr_latlon); summary(vr_full); summary(vr_flat); summary(vr_lonelev)

anova(vr_lat, vr_latlon)
anova(vr_full, vr_latlon)
anova(vr_full, vr_lonelev)
anova(vr_lon, vr_lonelev)
anova(vr_elev, vr_lonelev)

AIC(vr_elev, vr_lat, vr_lon, vr_latlon, vr_lonelev, vr_full, vr_flat)
```
lat + lon come close to full (lat, lon, elev), but lat + lon has slightly lower AIC. 
```{r}
# ---- Set explicitly, or skip Step 5 entirely if var_ratio was already close to 1 ----
var_ratio_model <- vr_latlon

trace_calib[[var_ratio_col]] <- predict(var_ratio_model, newdata = trace_calib)
trace_calib <- trace_calib %>%
  group_by(lat, lon) %>%
  mutate(
    !!cell_mean_col := mean(.data[[corrected_col]], na.rm = TRUE),
    !!rescaled_col := .data[[cell_mean_col]] +
      (.data[[corrected_col]] - .data[[cell_mean_col]]) / .data[[var_ratio_col]]
  ) %>%
  ungroup()

trace_full[[var_ratio_col]] <- predict(var_ratio_model, newdata = trace_full)
trace_full <- trace_full %>%
  group_by(lat, lon) %>%
  mutate(
    !!cell_mean_col := mean(.data[[corrected_col]], na.rm = TRUE),
    !!rescaled_col := .data[[cell_mean_col]] +
      (.data[[corrected_col]] - .data[[cell_mean_col]]) / .data[[var_ratio_col]]
  ) %>%
  ungroup()
```
# ============================================================
# STEP 6: Check rescaled values — this month only
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, rescaled_col, rescaled_col),
  c("ERA5 (calib)", "TraCE (calib, rescaled)", "TraCE (full, rescaled)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = rescaled_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, rescaled_col, rescaled_col),
  c("ERA5 (calib)", "TraCE (calib, rescaled)", "TraCE (full, rescaled)")
)


cellwise_check_m <- trace_calib %>%
  left_join(
    era_calib %>% dplyr::select(lat, lon, year, !!paste0(era_col, "_era_ref") := all_of(era_col)),
    by = c("lat", "lon", "year")
  ) %>%
  group_by(lat, lon) %>%
  summarise(
    sd_era = sd(.data[[paste0(era_col, "_era_ref")]], na.rm = TRUE),
    sd_rescaled = sd(.data[[rescaled_col]], na.rm = TRUE),
    var_ratio_after = sd_rescaled / sd_era,
    .groups = "drop"
  )

cellwise_check_m %>% arrange(desc(abs(var_ratio_after - 1)))

summary(cellwise_variance_m$var_ratio)
summary(cellwise_check_m$var_ratio_after)

#only one gridcell has excessively high variance ratio and 2 have very low. 
cellwise_check_m$var_ratio_after

ggplot(cellwise_check_m, aes(x=lon, y=lat, color=var_ratio_after)) + geom_point()
```
varaince ratio looks great!


Plot century trend for July
```{r}
plot_century_trend <- function(full_data, var_col,
                                era_data, era_col,
                                calib_data, calib_col,
                                var_label = var_col,
                                bin_width = 100) {
  
  # Bin the full corrected record into century spows
  df_summary <- full_data %>%
    dplyr::select(year, all_of(var_col)) %>%
    mutate(century_bin = floor(year / bin_width) * bin_width) %>%
    group_by(century_bin) %>%
    summarise(
      mean_val = mean(.data[[var_col]], na.rm = TRUE),
      ci_low   = quantile(.data[[var_col]], 0.025, na.rm = TRUE),
      ci_high  = quantile(.data[[var_col]], 0.975, na.rm = TRUE),
      n_obs    = n(),
      .groups  = "drop"
    )
  
  # ERA5 calibration-period reference mean
  era5_mean <- era_data %>%
    group_by(year, lat, lon) %>%
    summarise(mean_obs = mean(.data[[era_col]], na.rm = TRUE), .groups = "drop") %>%
    pull(mean_obs) %>%
    mean(na.rm = TRUE)
  
  breaks_seq <- as.character(seq(min(df_summary$century_bin), max(df_summary$century_bin), by = 500))
  
  ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_val)) +
    geom_point(size = 3, alpha = 0.2) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.1, alpha = 0.2) +
    geom_hline(yintercept = era5_mean, linetype = "dotted", color = "red", linewidth = 0.5) +
    scale_x_discrete(breaks = breaks_seq) +
    theme_minimal() +
    labs(x = paste0("Year (", bin_width, "-yr bin)"), y = var_label, title = "") +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      legend.title = element_blank()
    )
}

#plot corrected full trace for all months:
plot_century_trend(
  full_data = trace_full, var_col = "sp_rescaled",
  era_data = era_monthly, era_col = "sp",
  var_label = "Corrected sp"
)
```

Accept rescaled values as final correction.

```{r}
trace_calib[[final_col]] <- trace_calib[[rescaled_col]]
trace_full[[final_col]]  <- trace_full[[rescaled_col]]
```

# ============================================================
# STEP 7: Apply QM if Step 1 showed it's viable; otherwise finalize as-is
# ============================================================
```{r}
# # ---- OPTION A: QM viable ----
# qm_model <- fitQmap(compare_df_m[[paste0(var_name, "_era")]],
#                      trace_calib[[rescaled_col]],  # or [[corrected_col]] if Step 5 was skipped
#                      method = "QUANT", qstep = 0.01, wet.day = FALSE)
# 
# trace_calib[[final_col]] <- doQmap(trace_calib[[rescaled_col]], qm_model)
# trace_full[[final_col]]  <- doQmap(trace_full[[rescaled_col]], qm_model)
# 
# # ---- OPTION B: QM excluded ----
# # trace_calib[[final_col]] <- trace_calib[[rescaled_col]]
# # trace_full[[final_col]]  <- trace_full[[rescaled_col]]
# 
# plot_density_compare(
#   list(era_calib, trace_calib, trace_full),
#   c(era_col, final_col, final_col),
#   c("ERA5 (calib)", "TraCE (calib, final)", "TraCE (full, final)"),
#   var_name, month_val
# )
```

# ============================================================
# STORE this month's results and models before moving to the next month
# ============================================================
```{r}
if (!exists("monthly_models")) monthly_models <- list()
if (!exists("monthly_results_calib")) monthly_results_calib <- list()
if (!exists("monthly_results_full")) monthly_results_full <- list()

monthly_models[[as.character(month_val)]] <- list(
  mean_model = mean_bias_model,
  var_ratio_model = if (exists("var_ratio_model")) var_ratio_model else NULL,
  qm_model = if (exists("qm_model")) qm_model else NULL
)
monthly_results_calib[[as.character(month_val)]] <- trace_calib
monthly_results_full[[as.character(month_val)]]  <- trace_full

# Once you've done this for month_val <- 6, 7, and 8, combine:
# trace_calib_all_months <- bind_rows(monthly_results_calib)
# trace_full_all_months  <- bind_rows(monthly_results_full)
```

# ========================================================================================================================
# ==========================================    AUGUST         ============================================================
# ========================================================================================================================

# ============================================================
# DEFINE VARIABLE + MONTH FOR THIS RUN
# Change these two lines and re-run top to bottom, once per month (6, 7, 8)
# ============================================================

```{r}
var_name  <- "sp"
month_val <- 8   # 6 = June, 7 = July, 8 = August

era_col   <- var_name
trace_col <- var_name

# Filter to this month only, right at the source - everything downstream inherits this scope
era_calib   <- era_monthly   %>% filter(month == month_val)
trace_calib <- trace_monthly %>% filter(month == month_val)
trace_full  <- trace_monthly_full %>% filter(month == month_val)

# Dynamically-named output columns (unchanged across months - the month scoping lives in the data, not the names)
bias_col      <- paste0(var_name, "_predicted_bias")
corrected_col <- paste0(var_name, "_corrected")
var_ratio_col <- paste0(var_name, "_var_ratio_pred")
cell_mean_col <- paste0(var_name, "_cell_mean")
rescaled_col  <- paste0(var_name, "_rescaled")
final_col     <- paste0(var_name, "_final")
```


# ============================================================
# STEP 1: Compare ERA5 calib / TraCE calib / TraCE full RAW — this month only
# Does the full paleo record fall outside the calibration TraCE range?
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)"),
  var_name, month_val
)

plot_timeseries_compare(
  list(era_calib, trace_calib, trace_full),
  c("year", "year", "year"),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = trace_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, trace_col, trace_col),
  c("ERA5 (calib)", "TraCE (calib)", "TraCE (full, raw)")
)

range(trace_calib[[trace_col]], na.rm = TRUE)
range(trace_full[[trace_col]], na.rm = TRUE)
# If trace_full's range is meaningfully wider than trace_calib's -> QM will extrapolate -> flag for Step 7
```
# ============================================================
# STEP 2: Calculate bias for the calibration period — this month only
# ============================================================
```{r}
compare_df_m <- era_calib %>%
  inner_join(trace_calib, by = c("lat", "lon", "year"), suffix = c("_era", "_trace"))
  # adjust join columns if your monthly frames key differently

bias_summary_m <- tibble(
  variable = var_name, month = month_val,
  mean_bias = mean(compare_df_m[[paste0(var_name, "_trace")]] -
                    compare_df_m[[paste0(var_name, "_era")]], na.rm = TRUE),
  correlation = cor(compare_df_m[[paste0(var_name, "_trace")]],
                     compare_df_m[[paste0(var_name, "_era")]], use = "complete.obs"),
  era_sd = sd(compare_df_m[[paste0(var_name, "_era")]], na.rm = TRUE),
  trace_sd = sd(compare_df_m[[paste0(var_name, "_trace")]], na.rm = TRUE)
)
bias_summary_m

cellwise_bias_m <- compare_df_m %>%
  group_by(lat, lon, elevation) %>%
  summarise(
    bias = mean(.data[[paste0(var_name, "_trace")]] - .data[[paste0(var_name, "_era")]], na.rm = TRUE),
    .groups = "drop"
  )
cellwise_bias_m

ggplot(cellwise_bias_m, aes(x=lon, y=lat, size=abs(bias), color=bias)) + geom_point()

cor(cellwise_bias_m$bias, cellwise_bias_m$lon)
cor(cellwise_bias_m$bias, cellwise_bias_m$lat)
cor(cellwise_bias_m$bias, cellwise_bias_m$lon)
```
# ============================================================
# STEP 3: Fit candidate mean-bias models — this month only
# Fully manual, same as every variable so far — inspect, then assign the winner explicitly
# ============================================================
```{r}
m_elev   <- lm(bias ~ elevation, data = cellwise_bias_m)
m_lat    <- lm(bias ~ lat, data = cellwise_bias_m)
m_lon    <- lm(bias ~ lon, data = cellwise_bias_m)
m_latlon <- lm(bias ~ lat + lon, data = cellwise_bias_m)
m_full   <- lm(bias ~ elevation + lat + lon, data = cellwise_bias_m)
m_flat <- lm(bias ~ 1, data = cellwise_bias_m)

summary(m_elev); summary(m_lat); summary(m_lon); summary(m_latlon); summary(m_full); summary(m_flat)

anova(m_lat, m_latlon)
anova(m_lon, m_latlon)
anova(m_latlon, m_full)

AIC(m_elev, m_lat, m_lon, m_latlon, m_full, m_flat)

cellwise_bias_m$resid_check <- residuals(m_latlon)  # swap in whichever you're evaluating
cor(cellwise_bias_m$resid_check, cellwise_bias_m$elevation)
```

Go with m_flat:
```{r}
# ---- Set explicitly once you've reviewed the output above ----
mean_bias_model <- m_flat

trace_calib[[bias_col]] <- predict(mean_bias_model, newdata = trace_calib)
trace_calib[[corrected_col]] <- trace_calib[[trace_col]] - trace_calib[[bias_col]]

trace_full[[bias_col]] <- predict(mean_bias_model, newdata = trace_full)
trace_full[[corrected_col]] <- trace_full[[trace_col]] - trace_full[[bias_col]]
```
# ============================================================
# STEP 4: Check corrected mean — this month only
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, corrected_col, corrected_col),
  c("ERA5 (calib)", "TraCE (calib, corrected)", "TraCE (full, corrected)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = corrected_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, corrected_col, corrected_col),
  c("ERA5 (calib)", "TraCE (calib, corrected)", "TraCE (full, corrected)")
)

cellwise_variance_m <- compare_df_m %>%
  mutate(!!corrected_col := .data[[paste0(var_name, "_trace")]] -
           predict(mean_bias_model, newdata = compare_df_m)) %>%
  group_by(lat, lon, elevation) %>%
  summarise(
    sd_era = sd(.data[[paste0(var_name, "_era")]], na.rm = TRUE),
    sd_corrected = sd(.data[[corrected_col]], na.rm = TRUE),
    var_ratio = sd_corrected / sd_era,
    .groups = "drop"
  )
cellwise_variance_m %>% arrange(desc(abs(var_ratio - 1)))
summary(cellwise_variance_m$var_ratio)
```
# ============================================================
# STEP 5: Decide if variance needs correcting — this month only
# ============================================================
```{r}
vr_elev   <- lm(var_ratio ~ elevation, data = cellwise_variance_m)
vr_lat    <- lm(var_ratio ~ lat, data = cellwise_variance_m)
vr_lon    <- lm(var_ratio ~ lon, data = cellwise_variance_m)
vr_latlon <- lm(var_ratio ~ lat + lon, data = cellwise_variance_m)
vr_full   <- lm(var_ratio ~ elevation + lat + lon, data = cellwise_variance_m)
vr_flat <- lm(var_ratio ~ 1, data = cellwise_variance_m)

summary(vr_elev); summary(vr_lat); summary(vr_lon); summary(vr_latlon); summary(vr_full); summary(vr_flat)

anova(vr_lat, vr_latlon)
anova(vr_lat, vr_flat)

AIC(vr_elev, vr_lat, vr_lon, vr_latlon, vr_full, vr_flat)
```
go with just lat. 
```{r}
# ---- Set explicitly, or skip Step 5 entirely if var_ratio was already close to 1 ----
var_ratio_model <- vr_lat

trace_calib[[var_ratio_col]] <- predict(var_ratio_model, newdata = trace_calib)
trace_calib <- trace_calib %>%
  group_by(lat, lon) %>%
  mutate(
    !!cell_mean_col := mean(.data[[corrected_col]], na.rm = TRUE),
    !!rescaled_col := .data[[cell_mean_col]] +
      (.data[[corrected_col]] - .data[[cell_mean_col]]) / .data[[var_ratio_col]]
  ) %>%
  ungroup()

trace_full[[var_ratio_col]] <- predict(var_ratio_model, newdata = trace_full)
trace_full <- trace_full %>%
  group_by(lat, lon) %>%
  mutate(
    !!cell_mean_col := mean(.data[[corrected_col]], na.rm = TRUE),
    !!rescaled_col := .data[[cell_mean_col]] +
      (.data[[corrected_col]] - .data[[cell_mean_col]]) / .data[[var_ratio_col]]
  ) %>%
  ungroup()
```
# ============================================================
# STEP 6: Check rescaled values — this month only
# ============================================================
```{r}
plot_density_compare(
  list(era_calib, trace_calib, trace_full),
  c(era_col, rescaled_col, rescaled_col),
  c("ERA5 (calib)", "TraCE (calib, rescaled)", "TraCE (full, rescaled)"),
  var_name, month_val
)

plot_timeseries_calib(
  era_df = era_calib, era_year_col = "year", era_value_col = era_col,
  trace_df = trace_calib, trace_year_col = "year", trace_value_col = rescaled_col,
  var_label = var_name, month_label = month_val
)

quantile_compare_table(
  list(era_calib, trace_calib, trace_full),
  c(era_col, rescaled_col, rescaled_col),
  c("ERA5 (calib)", "TraCE (calib, rescaled)", "TraCE (full, rescaled)")
)


cellwise_check_m <- trace_calib %>%
  left_join(
    era_calib %>% dplyr::select(lat, lon, year, !!paste0(era_col, "_era_ref") := all_of(era_col)),
    by = c("lat", "lon", "year")
  ) %>%
  group_by(lat, lon) %>%
  summarise(
    sd_era = sd(.data[[paste0(era_col, "_era_ref")]], na.rm = TRUE),
    sd_rescaled = sd(.data[[rescaled_col]], na.rm = TRUE),
    var_ratio_after = sd_rescaled / sd_era,
    .groups = "drop"
  )

cellwise_check_m %>% arrange(desc(abs(var_ratio_after - 1)))

range(cellwise_check_m$var_ratio_after)

summary(era_calib$sp)
summary(trace_calib$sp_rescaled)
summary(trace_full$sp_rescaled)
```


Plot century trend for August
```{r}
plot_century_trend <- function(full_data, var_col,
                                era_data, era_col,
                                calib_data, calib_col,
                                var_label = var_col,
                                bin_width = 100) {
  
  # Bin the full corrected record into century spows
  df_summary <- full_data %>%
    dplyr::select(year, all_of(var_col)) %>%
    mutate(century_bin = floor(year / bin_width) * bin_width) %>%
    group_by(century_bin) %>%
    summarise(
      mean_val = mean(.data[[var_col]], na.rm = TRUE),
      ci_low   = quantile(.data[[var_col]], 0.025, na.rm = TRUE),
      ci_high  = quantile(.data[[var_col]], 0.975, na.rm = TRUE),
      n_obs    = n(),
      .groups  = "drop"
    )
  
  # ERA5 calibration-period reference mean
  era5_mean <- era_data %>%
    group_by(year, lat, lon) %>%
    summarise(mean_obs = mean(.data[[era_col]], na.rm = TRUE), .groups = "drop") %>%
    pull(mean_obs) %>%
    mean(na.rm = TRUE)
  
  breaks_seq <- as.character(seq(min(df_summary$century_bin), max(df_summary$century_bin), by = 500))
  
  ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_val)) +
    geom_point(size = 3, alpha = 0.2) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.1, alpha = 0.2) +
    geom_hline(yintercept = era5_mean, linetype = "dotted", color = "red", linewidth = 0.5) +
    scale_x_discrete(breaks = breaks_seq) +
    theme_minimal() +
    labs(x = paste0("Year (", bin_width, "-yr bin)"), y = var_label, title = "") +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      legend.title = element_blank()
    )
}

#plot corrected full trace for all months:
plot_century_trend(
  full_data = trace_full, var_col = "sp_rescaled",
  era_data = era_monthly, era_col = "sp",
  var_label = "Corrected sp"
)
```

Accept rescaled values as final correction.
```{r}
trace_calib[[final_col]] <- trace_calib[[rescaled_col]]
trace_full[[final_col]]  <- trace_full[[rescaled_col]]
```

# ============================================================
# STEP 7: Apply QM if Step 1 showed it's viable; otherwise finalize as-is
# ============================================================

NO quantile mapping - full paleo range is outside the calibration range. 
```{r}
# # ---- OPTION A: QM viable ----
# qm_model <- fitQmap(compare_df_m[[paste0(var_name, "_era")]],
#                      trace_calib[[rescaled_col]],  # or [[corrected_col]] if Step 5 was skipped
#                      method = "QUANT", qstep = 0.01, wet.day = FALSE)
# 
# trace_calib[[final_col]] <- doQmap(trace_calib[[rescaled_col]], qm_model)
# trace_full[[final_col]]  <- doQmap(trace_full[[rescaled_col]], qm_model)
# 
# # ---- OPTION B: QM excluded ----
# # trace_calib[[final_col]] <- trace_calib[[rescaled_col]]
# # trace_full[[final_col]]  <- trace_full[[rescaled_col]]
# 
# plot_density_compare(
#   list(era_calib, trace_calib, trace_full),
#   c(era_col, final_col, final_col),
#   c("ERA5 (calib)", "TraCE (calib, final)", "TraCE (full, final)"),
#   var_name, month_val
# )
```

# ============================================================
# STORE this month's results and models before moving to the next month
# ============================================================
```{r}
if (!exists("monthly_models")) monthly_models <- list()
if (!exists("monthly_results_calib")) monthly_results_calib <- list()
if (!exists("monthly_results_full")) monthly_results_full <- list()

monthly_models[[as.character(month_val)]] <- list(
  mean_model = mean_bias_model,
  var_ratio_model = if (exists("var_ratio_model")) var_ratio_model else NULL,
  qm_model = if (exists("qm_model")) qm_model else NULL
)
monthly_results_calib[[as.character(month_val)]] <- trace_calib
monthly_results_full[[as.character(month_val)]]  <- trace_full

# Once you've done this for month_val <- 6, 7, and 8, combine:
trace_calib_all_months <- bind_rows(monthly_results_calib)
trace_full_all_months  <- bind_rows(monthly_results_full)
```

```{r}
# ============================================================
# Combine and save sp's monthly models
# ============================================================

# Read existing seasonal model store if it exists, so saving sp doesn't wipe out
# any other variable you may have already finished
seasonal_models_path <- file.path(rds_root, "seasonal_bias_correction_models.rds")

if (file.exists(seasonal_models_path)) {
  seasonal_bias_correction_models <- readRDS(seasonal_models_path)
} else {
  seasonal_bias_correction_models <- list()
}

seasonal_bias_correction_models[[var_name]] <- monthly_models   # monthly_models already keyed "6"/"7"/"8"

saveRDS(seasonal_bias_correction_models, seasonal_models_path)

# Confirm structure
names(seasonal_bias_correction_models)              # should include "sp"
names(seasonal_bias_correction_models[[var_name]])   # should be "6", "7", "8"
```


```{r}
# ============================================================
# Save only this variable's columns (all stages) + identifiers
# ============================================================

id_cols <- c("lat", "lon", "ce_year", "year", "month")

stage_cols <- c(
  var_name,                              # raw
  paste0(var_name, "_predicted_bias"),
  paste0(var_name, "_corrected"),
  paste0(var_name, "_var_ratio_pred"),
  paste0(var_name, "_cell_mean"),
  paste0(var_name, "_rescaled"),
  paste0(var_name, "_final")
)

# Only keep columns that actually exist (e.g. if variance rescaling was skipped for some months,
# _var_ratio_pred/_cell_mean/_rescaled might be absent for those rows after bind_rows)
export_cols <- c(id_cols, stage_cols[stage_cols %in% names(trace_full_all_months)])

trace_full_export <- trace_full_all_months %>%
  dplyr::select(all_of(export_cols))

names(trace_full_export)   # confirm this matches what you expect before saving

write_parquet(
  trace_full_export,
  file.path(paleocliamte_data_root, paste0("trace21ka_processed/trace_", var_name, "_seasonal_corrected.parquet"))
)
```


Plot century trend for all months
```{r}
plot_century_trend <- function(full_data, var_col,
                                era_data, era_col,
                                calib_data, calib_col,
                                var_label = var_col,
                                bin_width = 100) {
  
  # Bin the full corrected record into century spows
  df_summary <- full_data %>%
    dplyr::select(year, all_of(var_col)) %>%
    mutate(century_bin = floor(year / bin_width) * bin_width) %>%
    group_by(century_bin) %>%
    summarise(
      mean_val = mean(.data[[var_col]], na.rm = TRUE),
      ci_low   = quantile(.data[[var_col]], 0.025, na.rm = TRUE),
      ci_high  = quantile(.data[[var_col]], 0.975, na.rm = TRUE),
      n_obs    = n(),
      .groups  = "drop"
    )
  
  # ERA5 calibration-period reference mean
  era5_mean <- era_data %>%
    group_by(year, lat, lon) %>%
    summarise(mean_obs = mean(.data[[era_col]], na.rm = TRUE), .groups = "drop") %>%
    pull(mean_obs) %>%
    mean(na.rm = TRUE)
  
  breaks_seq <- as.character(seq(min(df_summary$century_bin), max(df_summary$century_bin), by = 500))
  
  ggplot(df_summary, aes(x = as.factor(century_bin), y = mean_val)) +
    geom_point(size = 3, alpha = 0.2) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.1, alpha = 0.2) +
    geom_hline(yintercept = era5_mean, linetype = "dotted", color = "red", linewidth = 0.5) +
    scale_x_discrete(breaks = breaks_seq) +
    theme_minimal() +
    labs(x = paste0("Year (", bin_width, "-yr bin)"), y = var_label, title = "") +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      legend.title = element_blank()
    )
}

#plot corrected full trace for all months:
plot_century_trend(
  full_data = trace_full_export, var_col = "sp_final",
  era_data = era_monthly, era_col = "sp",
  var_label = "Corrected sp"
)
```


