---
title: "Untitled"
author: "Charlotte Uden"
date: "2026-08-10"
output: html_document
---
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(pROC)
library(loo)
library(cowplot)
library(stats)
library(caret)
library(arrow)
library(geodata)
library(terra)
library(rnaturalearth)
library(sf)
library(patchwork)
library(purrr)
library(qmap)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}

#setwd("~/raid/cuden/Paleo_Lightning_Fire_Uden_2026/Bias_Correction")


read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleocliamte_data_root <- config[["paleocliamte_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]
```

Read in elvation and trace points:
```{r}
trace_elev_lookup <- readRDS(file.path(rds_root, "trace_elev_lookup.rds"))
trace_cells <- readRDS(file.path(rds_root, "trace_cells.rds"))
```

From python file that converted netcdf to parquet:
Note: TraCE model years are offset from BP/CE. Year 0 in the model = 22000 BP.
    We convert to approximate CE year: ce_year = model_year - 21950
    (where 1950 CE = 0 BP = model year 21950)

```{r}
# Read the TRACE parquet file
trace_monthly_full <- read_parquet(file.path(paleocliamte_data_root,"trace21ka_processed/trace_ignition_burnedarea_model_monthly_no_lags.parquet"))
names(trace_monthly_full)
range(trace_monthly_full$ce_year)

trace_monthly <- trace_monthly_full %>% filter(ce_year >= 46, ce_year <= 89)

#make year match conventional dating
trace_monthly <- trace_monthly %>%
  mutate(year = ce_year + 1900)

trace_monthly_full <- trace_monthly_full %>%
  mutate(year = ce_year + 1900)

# Sanity check
range(trace_monthly$year)
range(trace_monthly_full$year)
```

Spatial clip to Alaska
```{r}
trace_monthly_full <- trace_monthly_full %>%
  mutate(lon = ifelse(lon > 180, lon - 360, lon)) %>%   # match trace_cells' -180/180 convention
  semi_join(trace_cells, by = c("lat", "lon")) %>%
  left_join(trace_elev_lookup, by = c("lat", "lon"))

trace_monthly <- trace_monthly %>%
  mutate(lon = ifelse(lon > 180, lon - 360, lon)) %>%   # match trace_cells' -180/180 convention
  semi_join(trace_cells, by = c("lat", "lon")) %>%
  left_join(trace_elev_lookup, by = c("lat", "lon"))

# Sanity checks
nrow(trace_monthly_full)
n_distinct(trace_monthly_full$lat, trace_monthly_full$lon)   # should be 23
sum(is.na(trace_monthly_full$elevation))                      # should be 0
names(trace_monthly_full)
# Sanity checks
nrow(trace_monthly)
n_distinct(trace_monthly$lat, trace_monthly$lon)   # should be 23
sum(is.na(trace_monthly$elevation))                      # should be 0
names(trace_monthly)
```



Read in ERA5 monthly data for correction period:
```{r}
era_monthly <- read_parquet(file.path(era5_data_root,"bias_correction/processed_parquet/era5_ignition_burnedarea_model_MAM_JJA_1946_1990.parquet"))
range(era_monthly$year)
names(era_monthly)
```

A spatial mis match: Aggregate ERA5 up to TraCE's grid by veraging all ERA5 points that fall within each TraCE grid cell. This maintains TraCE's actual physical resolution and avoids inventing false precision (interpolating trace down to era5 grid). 

```{r}
#save unaggregated ERA5 data:
era_monthly_pre_aggregation <- era_monthly
# Match each ERA5 point to nearest TraCE grid cell
nn <- get.knnx(data = trace_cells[, c("lon", "lat")],
               query = era_monthly[, c("lon", "lat")], k = 1)

era_monthly$trace_lon <- trace_cells$lon[nn$nn.index]
era_monthly$trace_lat <- trace_cells$lat[nn$nn.index]

# Aggregate ERA5 to TraCE grid cells, by year, month
era_monthly<- era_monthly %>%
  group_by(trace_lat, trace_lon, year, month) %>%
  summarise(across(c(tair), mean, na.rm = TRUE),
            .groups = "drop") %>%
  rename(lat = trace_lat, lon = trace_lon)

ggplot(filter(era_monthly, year==1946 & month==6), aes(x=lon, y=lat)) + geom_point()
ggplot(filter(trace_monthly, year==1946 & month==6), aes(x=lon, y=lat)) + geom_point()
```

```{r}
unique(era_monthly$month)
unique(trace_monthly$month)
unique(trace_monthly_full$month)
```

