---
title: "Burned_area"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(pROC)
library(loo)
library(cowplot)
library(gt)
library(randomForest)
library(pROC)
```

Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config_file_path <- file.path(base_path, "paleo_config.properties")

config <- read_properties(config_file_path)

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]

rds_root <- file.path(rds_root, "Burned_Area/redo")
performance_root <- file.path(performance_root, "burned_area_redo")
```

```{r}
df <- read.csv(file.path(drivers_root, "burned_area_jja_2002_2011_new.csv")) %>%
  dplyr::select(-starts_with("X"))

df <- df %>% mutate(
  road_proximity_binary  = if_else(dist_road  >= 30, 0, 1),
  urban_proximity_binary = if_else(dist_urban >= 40, 0, 1),
  river_proximity_binary = if_else(dist_river >= 30, 0, 1)
)
nrow(df)
df <- na.omit(df)
nrow(df)

# read in rows with fire count exceeding strike count and compare with burned area
null_rows <- read.csv(file.path(drivers_root, "null_ignition_rows_training_data_2002-2011.csv")) %>%
  dplyr::select(-X)

null_rows <- null_rows %>%
  dplyr::select(year, lat, lon, month, fire_count) %>%
  dplyr::rename(fire_count_null = fire_count) %>%
  dplyr::mutate(null_row = 1)

df_nulls <- df %>%
  full_join(null_rows, by = c("lon", "lat", "year", "month")) %>%
  mutate(null_row = replace_na(null_row, 0))

filter(df_nulls, null_row == 1) %>%
  group_by(year, month, lat, lon) %>%
  summarise(total_burned_area = sum(burned_area_km2_total), n_fires = n(), .groups = "drop")

#these null rows need to be dropped. 

df_nulls <- filter(df_nulls, null_row == 0) 
summary(df_nulls$burned_area_km2_total)

df <- df_nulls
```

# =============================================================================
Vapor pressure deficit (VPD), computed from tair (Kelvin) and rh (%),
using the standard FAO-56 Penman-Monteith / Tetens saturation-vapor-
pressure formula. Added directly to df, BEFORE the train/test split, so
it's available to every model downstream exactly like any other predictor.
**VPD and conifer fraction was used by Coffield et al. (2019) to predict fire size
# =============================================================================
```{r}


df <- df %>%
  mutate(
    tair_C = tair - 273.15,                                  # Kelvin -> Celsius
    es     = 0.6108 * exp((17.27 * tair_C) / (tair_C + 237.3)), # saturation vapor pressure, kPa
    ea     = es * (rh / 100),                                 # actual vapor pressure, kPa
    vpd    = es - ea                                          # vapor pressure deficit, kPa
  )

summary(df$vpd)   # sanity check -- should be non-negative, roughly 0-3 kPa for these conditions
```


Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]
```


```{r}
# ── Save scaling parameters ───────────────────────────────────────────────────
predictors_ignition <- c("rh", "tair", "precip", "wind", "vpd",
                          "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                          "tair_1m", "tair_2", "tair_2m", "tair_5y", "B", "C", "U") 


mu_rh        <- mean(train$rh);        sigma_rh        <- sd(train$rh)
mu_tair      <- mean(train$tair);      sigma_tair      <- sd(train$tair)
mu_precip    <- mean(train$precip);    sigma_precip    <- sd(train$precip)
mu_wind      <- mean(train$wind);      sigma_wind      <- sd(train$wind)
mu_vpd      <- mean(train$vpd);      sigma_vpd      <- sd(train$vpd)
mu_precip_1m <- mean(train$precip_1m); sigma_precip_1m <- sd(train$precip_1m)
mu_precip_2m <- mean(train$precip_2m); sigma_precip_2m <- sd(train$precip_2m)
mu_precip_3m <- mean(train$precip_3m); sigma_precip_3m <- sd(train$precip_3m)
mu_precip_5y <- mean(train$precip_5y); sigma_precip_5y <- sd(train$precip_5y)
mu_tair_1m   <- mean(train$tair_1m);   sigma_tair_1m   <- sd(train$tair_1m)
mu_tair_2m   <- mean(train$tair_2m);   sigma_tair_2m   <- sd(train$tair_2m)
mu_tair_3m   <- mean(train$tair_3m);   sigma_tair_3m   <- sd(train$tair_3m)
mu_tair_5y   <- mean(train$tair_5y);   sigma_tair_5y   <- sd(train$tair_5y)
mu_B         <- mean(train$B);         sigma_B         <- sd(train$B)
mu_C         <- mean(train$C);         sigma_C         <- sd(train$C)
mu_U         <- mean(train$U);         sigma_U         <- sd(train$U)



# Store and save
mean_sd_ignition <- data.frame(
  var   = predictors_ignition,
  mu    = c(mu_rh, mu_tair, mu_precip, mu_wind, mu_vpd,
            mu_precip_1m, mu_precip_2m, mu_precip_3m, mu_precip_5y,
            mu_tair_1m, mu_tair_2m, mu_tair_3m, mu_tair_5y, mu_B, mu_C, mu_U),
            #mu_road_proximity_binary, mu_urban_proximity_binary, mu_river_proximity_binary),
  sigma = c(sigma_rh, sigma_tair, sigma_precip, sigma_wind, sigma_vpd,
            sigma_precip_1m, sigma_precip_2m, sigma_precip_3m, sigma_precip_5y,
            sigma_tair_1m, sigma_tair_2m, sigma_tair_3m, sigma_tair_5y, sigma_B, sigma_C, sigma_U))
            #sigma_road_proximity_binary, sigma_urban_proximity_binary, sigma_river_proximity_binary)


write.csv(mean_sd_ignition,
          file.path(mu_sigma_root, "mu_sigma_burned_area_predictors.csv"),
          row.names = FALSE)

# ── Standardize train ─────────────────────────────────────────────────────────
train <- train %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    vpd      = (vpd      - mu_vpd)      / sigma_vpd,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    B         = (B         - mu_B)         / sigma_B,
    C         = (C         - mu_C)         / sigma_C,
    U         = (U         - mu_U)         / sigma_U
  )

train <- train[, c("lat", "lon", "year", "month",
                   "burned_area_km2_total",
                   "rh", "tair", "precip", "wind", "vpd",
                   "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                   "tair_1m", "tair_2m", "tair_3m","tair_5y", "B", "C", "U",
                   "road_proximity_binary", "urban_proximity_binary", "river_proximity_binary")]

# ── Standardize test ──────────────────────────────────────────────────────────
# Use training means and sds 
test <- test %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    vpd      = (vpd      - mu_vpd)      / sigma_vpd,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    B         = (B         - mu_B)         / sigma_B,
    C         = (C         - mu_C)         / sigma_C,
    U         = (U         - mu_U)         / sigma_U
    )


test <- test[, c("lat", "lon", "year", "month",
                 "burned_area_km2_total",
                 "rh", "tair", "precip", "wind", "vpd",
                 "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                 "tair_1m", "tair_2m", "tair_3m","tair_5y", "B", "C", "U", 
                 "road_proximity_binary", "urban_proximity_binary", "river_proximity_binary")]

head(train)

```


classifier decides whether this will be an extremem fire size or typical, based on a 95% threshold (fire size above the 95th percentile is labeled extreme). Then, two separate gamma bayes models are fit for extreme versusu typical fire sizes.. First, look at predictors that will be best suited for each submodel. 

Classifier model - which predictors are good at catching whether a fire will be classified as extreme or typical?
```{r}

# ── Define "large fire" threshold ──────────────────────────────────────────
# Using top 5% of burned area as the "extreme" group; adjust as needed
threshold <- quantile(train$burned_area_km2_total, 0.95)
train <- train %>%
  mutate(is_extreme = as.integer(burned_area_km2_total >= threshold))

cat(sprintf("Threshold (95th pct): %.2f km2\n", threshold))
cat(sprintf("N extreme fires: %d / %d (%.1f%%)\n",
            sum(train$is_extreme), nrow(train), 100 * mean(train$is_extreme)))

# ── 1. Univariate comparison: does each predictor differ between groups? ──
predictors <- c("rh", "tair", "precip", "wind", "vpd",
                "precip_1m", "precip_3m", "precip_5y",
                "tair_2m", "tair_5y", "C", "U", "urban_proximity_binary")

univariate_tests <- lapply(predictors, function(v) {
  x_extreme <- train[[v]][train$is_extreme == 1]
  x_normal  <- train[[v]][train$is_extreme == 0]

  # Wilcoxon rank-sum test — robust to non-normality, appropriate given
  wt <- tryCatch(wilcox.test(x_extreme, x_normal), error = function(e) NULL)

  data.frame(
    predictor    = v,
    mean_extreme = mean(x_extreme, na.rm = TRUE),
    mean_normal  = mean(x_normal, na.rm = TRUE),
    diff         = mean(x_extreme, na.rm = TRUE) - mean(x_normal, na.rm = TRUE),
    p_value      = if (!is.null(wt)) wt$p.value else NA_real_
  )
})

univariate_results <- do.call(rbind, univariate_tests) %>%
  arrange(p_value)

print(univariate_results)

# ── 2. Visualize: boxplots of each predictor by extreme/normal ────────────
train_long <- train %>%
  dplyr::select(all_of(predictors), is_extreme) %>%
  pivot_longer(-is_extreme, names_to = "predictor", values_to = "value")

ggplot(train_long, aes(x = factor(is_extreme), y = value)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~predictor, scales = "free_y") +
  labs(x = "Extreme fire (1 = top 5% burned area)", y = "Standardised value") +
  theme_minimal()

# ── 3. Multivariate check: can predictors jointly classify extreme fires? ──
# A quick random forest AUC tells you the *ceiling* on how separable the
# groups are given these predictors, independent of the burned-area model


set.seed(123)
rf_data <- train %>% dplyr::select(all_of(predictors), is_extreme) %>%
  mutate(is_extreme = factor(is_extreme))

rf_fit <- randomForest(is_extreme ~ ., data = rf_data, ntree = 500,
                        importance = TRUE)

rf_probs <- predict(rf_fit, type = "prob")[, "1"]
rf_auc   <- auc(roc(rf_data$is_extreme, rf_probs, quiet = TRUE))
cat(sprintf("\nRandom forest AUC for separating extreme fires: %.3f\n", rf_auc))

# Variable importance — which predictors carry the most separating signal
importance_df <- as.data.frame(importance(rf_fit)) %>%
  tibble::rownames_to_column("predictor") %>%
  arrange(desc(MeanDecreaseGini))

print(importance_df)
```


worth keeping: urban_proximity_binary (p = 6.6e-13), rh (p = 3.6e-9), tair (p = 9.3e-8), precip (p = 1.75e-4), tair_5y (p = 8.95e-4)
Borderline: precip_3m, precip_1m (p ≈ 0.02–0.03)


What about fires below the 0.95 threshold? What predictors will do well at estimating smaller fire sizes?
```{r}

typical_only <- train %>%
  filter(burned_area_km2_total < threshold)

predictors_all <- c("rh", "tair", "precip", "wind", "vpd", "precip_1m", "precip_3m",
                     "precip_5y", "tair_2m", "tair_5y", "C", "U", "urban_proximity_binary")

within_typical_cor <- sapply(predictors_all, function(v) {
  cor(typical_only[[v]], log(typical_only$burned_area_km2_total), use = "complete.obs")
})

sort(within_typical_cor, decreasing = TRUE)
```


Highest correlation: "urban_proximity_binary", "tair_2m", "tair". RH might be worth including (-0.119), since its opposing negative correlation compared with lightning and ignition models will encourage contraditory outcomes between fire count and size (ie. many small fires)


And now, extremem fires - what predictors are better fit to predict extreme fires?
```{r}

extreme_only <- train %>%
  filter(burned_area_km2_total > threshold)

predictors_all <- c("rh", "tair", "precip", "wind", "vpd", "precip_1m", "precip_3m",
                     "precip_5y", "tair_2m", "tair_5y", "C", "U", "urban_proximity_binary")

within_extreme_cor <- sapply(predictors_all, function(v) {
  cor(extreme_only[[v]], log(extreme_only$burned_area_km2_total), use = "complete.obs")
})

sort(within_extreme_cor, decreasing = TRUE)
```


U is strongest (0.339), then precip_3m (-0.30384388), precip_5y (-0.28054696), C (-0.27449596), and tair (0.23708349)
Interesting that urban proximity drops!
May be worth again including RH (-0.25135813) since its influece on burned area is opposite to fire count. 

Take a look at outcome distribution
```{r}
par(mfrow = c(1, 3))
hist(train$burned_area_km2_total,
     main = "Train Burned area (raw)", xlab = "km²")
hist(log(train$burned_area_km2_total),
     main = "Train Burned area (log)", xlab = "log(km²)")
hist(sqrt(train$burned_area_km2_total),
     main = "Train Burned area (sqrt)", xlab = "sqrt(km²)")

summary(train$burned_area_km2_total)

par(mfrow = c(1, 3))
hist(df$burned_area_km2_total,
     main = "Burned area (raw)", xlab = "km²")
hist(log(df$burned_area_km2_total),
     main = "Burned area (log)", xlab = "log(km²)")
hist(sqrt(df$burned_area_km2_total),
     main = "Burned area (sqrt)", xlab = "sqrt(km²)")

summary(df$burned_area_km2_total)

#Typical fire distribution:
hist(typical_only$burned_area_km2_total, main = "Typical fire burned area (raw)", xlab = "km²")
#Extreme fire distribution:
hist(extreme_only$burned_area_km2_total, main = "Extreme fire burned area (raw)", xlab = "km²")
```


# ---- Check for colinearity between predictors --- 

Condition number (kappa(X)) is a diagnostic metric that assesses the severity of multicollinearity across the entire set of predictors. It measures how unstable or sensitive the model's coefficient estimates are to tiny changes in the data.

1 to 10 (Low Collinearity): The predictors are nearly orthogonal (independent). The model is numerically stable, and coefficients are reliable.
10 to 30 (Moderate Collinearity): Interdependence is present. The variance of the coefficients is inflating, but it may not destroy the model's predictive power.
Greater than 30 (Severe Collinearity): Strong linear dependencies exist. Minor tweaks to the data will cause massive, erratic swings in your coefficient signs and values

```{r}
preds <- c("rh", "tair", "precip", "wind", "vpd",
                          "precip_1m", "precip_2m","precip_3m","precip_5y",
                          "tair_1m", "tair_2m", "tair_3m", "tair_5y", "C", "U",
                         "road_proximity_binary", "urban_proximity_binary")

X <- model.matrix(as.formula(paste("~", paste(preds, collapse = " + "))),
                  data = train)
kappa(X)   # values > 1e6 (1,000,000) indicate severe collinearity
cor(train[, preds])  # look for correlations > 0.9

cor(train[, preds])> 0.9  # look for correlations > 0.9
```


Correlation results:
precip_2m and precip_3m: r = 0.937 — this is the problematic correlation. These two variables are carrying nearly identical information, which will destabilize the GLM fitting algorithm. remove precip_2m (between 1m and 3m). precip_1m captures very recent fuel moisture (FFMC/DMC analogue) and precip_3m captures the longer-term moisture deficit (DC analogue). The middle lag is redundant given those two endpoints are included.

tair_1m, 2m, 3m: r >0.9. remove tair_1m and tair_3m

Proximity data, r>0.9. random forest identifies road as having lower importance - remove.

B, C, U all sum to 1, also introducing autocorrelation - remove B, lowest of the three in importance according to the random forest. 

##########################
Explore interaction terms:
##########################
- humidity's suppressing effect (negative correlation) on extreme-fire probability is itself different in urban-proximate vs. remote areas (e.g., suppression resources mean humidity matters less near infrastructure, since fires get controlled early regardless), that's a physically sensible interaction and directly testable with the LRT approach

- U x rh: U (land cover fraction) is consistently one of the strongest extreme-fire-size predictors; whether more fuel amplifies or dampens the effect of dryness on fire size is a classic fire-behavior interaction (more fuel + dry conditions compounding, vs. fuel-limited growth regardless of dryness)

-----------
Classifier
-----------

```{r}
library(iml)
rf_data_features_only <- rf_data %>% dplyr::select(-is_extreme)

predictor <- Predictor$new(rf_fit, data = rf_data_features_only, y = rf_data$is_extreme)
interact <- Interaction$new(predictor, feature = "rh")
plot(interact)

```


H-statistic: urban_proximity_binary:rh (≈0.92), U:rh (≈0.90), and vpd:rh (≈0.87) are the top three

Plot: bin one continuous predictor into terciles (3 equal quantiles), then plot the outcome against the other predictor, colored/faceted by that tercile:
```{r}
train %>%
  mutate(var_tercile = ntile(vpd, 3)) %>%
  ggplot(aes(x = rh, y = log(burned_area_km2_total + 1), color = factor(var_tercile))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "vpd tercile")

train %>%
  mutate(var_tercile = ntile(urban_proximity_binary, 2)) %>%
  ggplot(aes(x = rh, y = log(burned_area_km2_total + 1), color = factor(var_tercile))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "urban_proximity_binary tercile")

train %>%
  mutate(var_tercile = ntile(U, 3)) %>%
  ggplot(aes(x = rh, y = log(burned_area_km2_total + 1), color = factor(var_tercile))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "U tercile")
```


All three plots show visibly different slopes across terciles — vpd tercile 3 (high VPD, blue) is nearly flat with a high intercept, while tercile 1 (red) drops off steeply; U shows the most dramatic split, with tercile 1 (low openland) essentially flat while terciles 2–3 fall off sharply; urban_proximity_binary shows a real but more modest difference between its two groups.

Likelihood-ratio test via a quick frequentist GLM as a quick screen. Fit the same classifier/magnitude equation as a plain glm(), once without an interaction and once with it, and compare. 

Random forest resutls without intercations: urban_proximity_binary (p = 6.6e-13), rh (p = 3.6e-9), tair (p = 9.3e-8), precip (p = 1.75e-4), tair_5y (p = 8.95e-4)
```{r}
# no interactions, using predictors identified by the non interaction random forest
glm_main <- glm(is_extreme ~ urban_proximity_binary + rh + tair + precip + tair_5y,
                 data = train, family = binomial)

# use only interaction terms identified by interaction random forest
glm_interact <- glm(is_extreme ~ rh * urban_proximity_binary + rh * U + rh * vpd,
                     data = train, family = binomial)

#a blend of both
glm_both <- glm(is_extreme ~ rh * urban_proximity_binary + rh * U + rh * vpd + tair + precip + tair_5y,
                     data = train, family = binomial)


glm_limited_2 <- glm(is_extreme ~ rh * urban_proximity_binary + rh * U + rh * vpd + tair_5y,
                     data = train, family = binomial)

glm_limited_3 <- glm(is_extreme ~ rh : urban_proximity_binary + rh : U + rh : vpd + rh + U + tair_5y,
                     data = train, family = binomial)

glm_limited_3 <- glm(is_extreme ~ rh : urban_proximity_binary + rh : U + rh + U + urban_proximity_binary + tair_5y,
                     data = train, family = binomial)

glm_limited_4 <- glm(is_extreme ~ rh : urban_proximity_binary + rh : U + rh + U + tair_5y,
                     data = train, family = binomial)

summary(glm_main); summary(glm_interact); summary(glm_both); summary(glm_limited_1); summary(glm_limited_2); summary(glm_limited_3); summary(glm_limited_4)
anova(glm_main, glm_interact, test = "LRT")
anova(glm_main, glm_both, test = "LRT")
anova(glm_limited_2, glm_limited_3, test = "LRT")
AIC(glm_main, glm_interact, glm_both, glm_limited_1, glm_limited_2, glm_limited_3, glm_limited_4)
```

Go with glm_limited_2 - include all main affects along with interactions, plus tair_5y: 
is_extreme ~ rh * urban_proximity_binary + rh * U + rh * vpd + tair_5y

-----------
Interactions for the typical fire size model: 
-----------
```{r}

predictors_all <- c("rh", "tair", "precip", "wind", "vpd",
                     "precip_1m", "precip_3m", "precip_5y",
                     "tair_2m", "tair_5y", "C", "U", "urban_proximity_binary")

typical_only <- train %>%
  filter(burned_area_km2_total < threshold)

hist(typical_only$burned_area_km2_total)
```


```{r}
rf_data_typical <- typical_only %>%
  dplyr::select(all_of(predictors_all), burned_area_km2_total) %>%
  mutate(log_burned_area = log(burned_area_km2_total + 1)) %>%
  dplyr::select(-burned_area_km2_total)

set.seed(123)
rf_fit_typical <- randomForest(log_burned_area ~ ., data = rf_data_typical,
                                ntree = 500, importance = TRUE)

# Predictor-only data frame for iml -- outcome column excluded, same fix
# applied to the classifier's interaction check earlier
rf_data_typical_features <- rf_data_typical %>% dplyr::select(-log_burned_area)

predictor_typical <- Predictor$new(rf_fit_typical,
                                    data = rf_data_typical_features,
                                    y = rf_data_typical$log_burned_area)

# Overall H-statistic: ranks EVERY pairwise interaction across all predictors
interact_typical_overall <- Interaction$new(predictor_typical)
plot(interact_typical_overall)

# rh-specific: how strongly rh interacts with each OTHER predictor
interact_typical_rh <- Interaction$new(predictor_typical, feature = "rh")
plot(interact_typical_rh)

cat("=== Typical fires: variable importance ===\n")
print(importance(rf_fit_typical))
```

Plot interactions across typical fire sizes
```{r}
typical_only %>%
  mutate(var_tercile = ntile(urban_proximity_binary, 2)) %>%
  ggplot(aes(x = rh, y = log(burned_area_km2_total + 1), color = factor(var_tercile))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "urban_proximity_binary tercile")

typical_only %>%
  mutate(var_tercile = ntile(precip_3m, 3)) %>%
  ggplot(aes(x = rh, y = log(burned_area_km2_total + 1), color = factor(var_tercile))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "precip_3m tercile")
```

Top Importance: urban_proximity_binary, rh, precip_5y, precip_3m, VPD, tair_2m
Correlation values from previous code block (Typical): urban_proximity_binary (-0.37037339), tair (0.12998461), rh (-0.11945199), VPD (0.11904348)

Top intercations: urban_proximity_binary:rh, precip_3m:rh

tair_5y correlates with 

Fit glms for quick look at candidates:
```{r}
#Check for severe collinearity among predictors
cor(typical_only[c("urban_proximity_binary", "rh", "tair", "vpd", "precip_5y", "precip_3m", "tair_2m")])


# no interactions, using predictors identified by correlation
n_preds <- 4 
glm_cor <- glm(burned_area_km2_total ~ urban_proximity_binary + rh + tair + vpd,
                 data = typical_only, family = Gamma(link = "log"),
               start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

# no interactions, using predictors identified by the non interaction random forest
n_preds <- 6
glm_rf <- glm(burned_area_km2_total ~ urban_proximity_binary + rh + precip_5y + precip_3m + vpd + tair_2m,
                 data = typical_only, family = Gamma(link = "log"),
               start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

# use only interaction terms identified by interaction random forest
n_preds <- 5
glm_interact <- glm(burned_area_km2_total ~ rh * urban_proximity_binary + rh * precip_3m,
                     data = typical_only, family = Gamma(link = "log"),
               start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 8
glm_interact_2 <- glm(burned_area_km2_total ~ rh * urban_proximity_binary + rh * precip_3m + vpd + tair_2m + tair,
                     data = typical_only, family = Gamma(link = "log"),
               start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 6
glm_interact_3 <- glm(burned_area_km2_total ~ rh * urban_proximity_binary + rh * precip_3m + tair_2m,
                     data = typical_only, family = Gamma(link = "log"),
               start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 7
glm_interact_4 <- glm(burned_area_km2_total ~ rh * urban_proximity_binary + rh * precip_3m + tair_2m + vpd,
                     data = typical_only, family = Gamma(link = "log"),
               start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 6
glm_interact_5 <- glm(burned_area_km2_total ~ rh * urban_proximity_binary + rh * precip_3m + vpd,
                          data = typical_only, family = Gamma(link = "log"),
                          start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 8
glm_interact_6 <- glm(burned_area_km2_total ~ rh * urban_proximity_binary + rh * precip_3m + rh * vpd + tair_2m,
                          data = typical_only, family = Gamma(link = "log"),
                          start = c(log(mean(typical_only$burned_area_km2_total)), rep(0, n_preds)))

summary(glm_cor); summary(glm_rf); summary(glm_interact); summary(glm_interact_2); summary(glm_interact_3); summary(glm_interact_4); summary(glm_interact_5); summary(glm_interact_6)
anova(glm_cor, glm_rf, test = "LRT")
anova(glm_rf, glm_interact, test = "LRT")
anova(glm_cor, glm_interact, test = "LRT")
anova(glm_interact, glm_interact_2, test = "LRT")
anova(glm_interact_2, glm_interact_3, test = "LRT")
anova(glm_interact_2, glm_interact_4, test = "LRT")
anova(glm_interact_4, glm_interact_5, test = "LRT")
anova(glm_interact_4, glm_interact_6, test = "LRT")

AIC(glm_cor, glm_rf, glm_interact, glm_interact_2, glm_interact_3, glm_interact_4, glm_interact_5, glm_interact_6)
```
NOTE: rh and vpd correlate (-0.935) and tair and vpd correlate (0.85). 

Use glm_interact_4 predictor set: preds_typical = c("urban_proximity_binary", "rh", "precip_3m", "tair_2m", "vpd"), with rh:urban_proximity_binary and rh:precip_3m as interaction terms
-----------
Extreme fire size model interactions
-----------
```{r}
extreme_only <- train %>%
  filter(burned_area_km2_total >= threshold)

hist(extreme_only$burned_area_km2_total, xlim = c(threshold, max(train$burned_area_km2_total)))

extreme_only <- train %>%
  filter(burned_area_km2_total >= 225)

hist(extreme_only$burned_area_km2_total, xlim = c(225, max(train$burned_area_km2_total)))

extreme_only <- train %>%
  filter(burned_area_km2_total >= 200)

hist(extreme_only$burned_area_km2_total, xlim = c(200, max(train$burned_area_km2_total)))

extreme_only <- train %>%
  filter(burned_area_km2_total >= 175)

hist(extreme_only$burned_area_km2_total, xlim = c(175, max(train$burned_area_km2_total)))

extreme_only <- train %>%
  filter(burned_area_km2_total >= 150)

hist(extreme_only$burned_area_km2_total, xlim = c(150, max(train$burned_area_km2_total)))

extreme_only <- train %>%
  filter(burned_area_km2_total >= 125)

hist(extreme_only$burned_area_km2_total, xlim = c(125, max(train$burned_area_km2_total)))

extreme_only <- train %>%
  filter(burned_area_km2_total >= 100)

hist(extreme_only$burned_area_km2_total, xlim = c(100, max(train$burned_area_km2_total)))
```



```{r}
typical_only <- train %>%
  filter(burned_area_km2_total < 175)

extreme_only <- train %>%
  filter(burned_area_km2_total >= 175)


hist(typical_only$burned_area_km2_total, xlim = c(min(train$burned_area_km2_total), 175))

hist(extreme_only$burned_area_km2_total, xlim = c(175, max(train$burned_area_km2_total)))

h_typical <- hist(typical_only$burned_area_km2_total, xlim = c(min(train$burned_area_km2_total), 175))
h_extreme <- hist(extreme_only$burned_area_km2_total, xlim = c(175, max(train$burned_area_km2_total)))
scale_value <- max(h_extreme$counts) 
h_typical$counts <- h_typical$counts / scale_value

# Plot the scaled histogram
plot(h_typical, ylab = "Scaled Frequency")
plot(log(h_extreme))

```


```{r}

rf_data_extreme <- extreme_only %>%
  dplyr::select(all_of(predictors_all), burned_area_km2_total) %>%
  mutate(log_burned_area = log(burned_area_km2_total + 1)) %>%
  dplyr::select(-burned_area_km2_total)

set.seed(123)
rf_fit_extreme <- randomForest(log_burned_area ~ ., data = rf_data_extreme,
                                ntree = 500, importance = TRUE)

rf_data_extreme_features <- rf_data_extreme %>% dplyr::select(-log_burned_area)

predictor_extreme <- Predictor$new(rf_fit_extreme,
                                    data = rf_data_extreme_features,
                                    y = rf_data_extreme$log_burned_area)

interact_extreme_overall <- Interaction$new(predictor_extreme)
plot(interact_extreme_overall)

interact_extreme_rh <- Interaction$new(predictor_extreme, feature = "rh")
plot(interact_extreme_rh)

cat("=== Extreme fires: variable importance ===\n")
print(importance(rf_fit_extreme))
```

```{r}
extreme_only %>%
  mutate(var_tercile = ntile(wind, 3)) %>%
  ggplot(aes(x = rh, y = log(burned_area_km2_total + 1), color = factor(var_tercile))) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "wind tercile")
```
Top importance: C, U, rh, tair_5y, wind
correlations (Extreme): U (0.34194954), C (-0.26921603), VPD (0.25499035), tair_2m (-0.25610951), rh (-0.25135813), tair (0.17851575)
interactions: wind:rh, with interesting outcomes. windy and humid results in smaller burned area,  while less windy, and less humid results in larger burned area. 

Fit glms for quick look at candidates:
```{r}
#Check for severe collinearity among predictors
cor(extreme_only[c("C", "U", "rh", "tair_5y", "wind", "vpd", "tair_2m", "tair")])
```
Correlation:
U and C: -0.96980175
rh and vpd: -0.9073116
tair and vpd: 0.81148662

quick glm test:
```{r}
# no interactions, using predictors identified by correlation
n_preds <- 6
glm_cor <- glm(burned_area_km2_total ~ U + C + vpd + tair_2m + rh + tair,
                 data = extreme_only, family = Gamma(link = "log"),
               start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))

# no interactions, using predictors identified by random forest
n_preds <- 5
glm_rf <- glm(burned_area_km2_total ~ C + U + rh + tair_5y + wind,
                 data = extreme_only, family = Gamma(link = "log"),
               start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))

#only interactions
n_preds <- 3
glm_3 <- glm(burned_area_km2_total ~ wind * rh,
                 data = extreme_only, family = Gamma(link = "log"),
               start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 5
glm_4 <- glm(burned_area_km2_total ~ wind * rh + C + U,
                 data = extreme_only, family = Gamma(link = "log"),
               start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 3
glm_5 <- glm(burned_area_km2_total ~ wind + C + U,
                 data = extreme_only, family = Gamma(link = "log"),
               start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))

n_preds <- 4
glm_6 <- glm(burned_area_km2_total ~ wind : rh + wind + C + U,
                 data = extreme_only, family = Gamma(link = "log"),
               start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))

# Does fuel (C or U) interact with wind, rather than rh physically. more fuel + more wind could compound fire spread more than either alone, independent of humidity
n_preds <- 4
glm_7 <- glm(burned_area_km2_total ~ wind * U + C,
             data = extreme_only, family = Gamma(link = "log"),
             start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))


# Does VPD (rather than rh) produce a cleaner wind interaction, given rh's extra noise here and VPD's stronger showing in the typical-fire model?
n_preds <- 5
glm_8 <- glm(burned_area_km2_total ~ wind * vpd + C + U,
             data = extreme_only, family = Gamma(link = "log"),
             start = c(log(mean(extreme_only$burned_area_km2_total)), rep(0, n_preds)))


anova(glm_4, glm_6, test = "LRT")

summary(glm_cor); summary(glm_rf); summary(glm_3); summary(glm_4); summary(glm_5); summary(glm_6); summary(glm_7); summary(glm_8)
AIC(glm_cor, glm_rf, glm_3, glm_4, glm_5, glm_6, glm_7, glm_8)
```
Winner is glm_4: burned_area_km2_total ~ wind * rh + C + U
The interaction itself never reaches conventional significance in any model tested. In glm_4, wind:rh has p = 0.130; in glm_6 (the marginality-violating version, which tends to inflate an interaction's apparent significance by absorbing the main effect into it), it only gets to p = 0.059

extreme_only has only 52 observations (df = 51 null), versus typical_only's 974. An interaction term asks more of the data than a main effect does. While the tercile plot shows an interesting interaction between wind and rh, its not statistically significant. Having said that, a bayes model will better handle the small sample size by allowing uncertainty in the posterior estimate. keep wind:rh interaction for now. 


predictors for simple gamma bayes model (no mixture):
```{r}
preds_m1 <- c("rh", "tair", "precip", "wind",
               "precip_1m", "precip_3m", "precip_5y",
               "tair_2m", "tair_5y", "C", "U", "urban_proximity_binary")

# Verify correlation is resolved
cor(train[, preds_m1]) > 0.9

```

What distribution should I choose for the extreme fires? At the 95% threhold, the extreme fires look gamma-distributied. But at a 175 km2 burned area cutoff, the shape of the data changes a little. 

1. Use a mean residual life (mean excess) plot, the first tool. For a true GPD tail, E[X − u | X > u] should be roughly linear in u once you're above the correct threshold. Below the correct threshold, the line will bend/curve.
2. Parameter stability plot. Fit a GPD at many different candidate thresholds and plot how the estimated shape (ξ) and modified scale parameters change. If ξ is roughly stable (flat, within CI bands of each other) above some threshold but jumps around below it, that's strong, direct evidence for exactly where the tail "truly" begins — this is the most standard, defensible way to formally answer "where does the shape change" rather than eyeballing histograms.
```{r}
library(ismev)
mrl.plot(train$burned_area_km2_total, umin = 50, umax = 500)

gpd.fitrange(train$burned_area_km2_total, umin = 50, umax = 500, nint = 20)
```


Compare a GDP distribution fit to extreme burned areas versus a gamma distribution. Lower AIC is better. 
Start with the threshold set based on the 95% threshold - fires that make up the top 5% 
```{r}
#library(MASS)
ba_threshold <- 254.2795 #basaed on 95% threshold
# Fit GPD at candidate threshold
gpd_fit <- gpd.fit(train$burned_area_km2_total, threshold = ba_threshold)
gpd.diag(gpd_fit)   # QQ-plot, return-level plot -- visual goodness-of-fit
if (xi_hat < 0) {
  implied_max <- ba_threshold - sigma_hat / xi_hat
  cat(sprintf("GPD-implied maximum burned area: %.1f km\u00b2 (observed max: %.1f km\u00b2)\n",
              implied_max, max(train$burned_area_km2_total)))
}

# Compare against a Gamma fit on the same exceedances via AIC, same
# principle as everything been doing with the GLM screens
exceedances <- train$burned_area_km2_total[train$burned_area_km2_total >= ba_threshold]
gamma_fit <- fitdistr(exceedances - ba_threshold + 0.001, "gamma")  # MASS::fitdistr
# Compare AIC: whichever is lower is the better-supported tail model AT THAT THRESHOLD
#AIC(gpd_fit) doesnt work on this class. use:
k_gpd <- length(gpd_fit$mle)   # number of estimated parameters (2: sigma, xi)
aic_gpd <- 2 * k_gpd + 2 * gpd_fit$nllh
aic_gpd 
AIC(gamma_fit)
```


Now set a threshold based on what I saw in the histogram, where a different shape appears when looking at fires greater than 175 km2. 

```{r}
library(evd)  # for dgpd()

ba_threshold <- 254.2795  
exceedances <- train$burned_area_km2_total[train$burned_area_km2_total >= ba_threshold] - ba_threshold

gpd_fit <- gpd.fit(train$burned_area_km2_total, threshold = ba_threshold)
sigma_hat <- gpd_fit$mle[1]
xi_hat    <- gpd_fit$mle[2]

cat(sprintf("Fitted GPD: sigma (scale) = %.2f, xi (shape) = %.3f\n", sigma_hat, xi_hat))

# Overlay fitted density on a histogram of the exceedances
x_seq <- seq(0, max(exceedances), length.out = 200)
gpd_density <- dgpd(x_seq, loc = 0, scale = sigma_hat, shape = xi_hat)

hist(exceedances, breaks = 15, freq = FALSE,
     main = sprintf("GPD fit at threshold = %f", ba_threshold),
     xlab = "Exceedance (burned area - threshold, km\u00b2)")
lines(x_seq, gpd_density, col = "steelblue", lwd = 2)
```


```{r}
#library(MASS)
ba_threshold <- 175 #basaed on shape change in histogram
# Fit GPD at candidate threshold
gpd_fit <- gpd.fit(train$burned_area_km2_total, threshold = ba_threshold)
gpd.diag(gpd_fit)   # QQ-plot, return-level plot -- visual goodness-of-fit

if (xi_hat < 0) {
  implied_max <- ba_threshold - sigma_hat / xi_hat
  cat(sprintf("GPD-implied maximum burned area: %.1f km\u00b2 (observed max: %.1f km\u00b2)\n",
              implied_max, max(train$burned_area_km2_total)))
}

# Compare against a Gamma fit on the same exceedances via AIC, same

exceedances <- train$burned_area_km2_total[train$burned_area_km2_total >= ba_threshold]
gamma_fit <- fitdistr(exceedances - ba_threshold + 0.001, "gamma")  # MASS::fitdistr
# Compare AIC: whichever is lower is the better-supported tail model AT THAT THRESHOLD
#AIC(gpd_fit) doesnt work on this class. use:
k_gpd <- length(gpd_fit$mle)   # number of estimated parameters (2: sigma, xi)
aic_gpd <- 2 * k_gpd + 2 * gpd_fit$nllh
aic_gpd 
AIC(gamma_fit)
```



```{r}
library(evd)  # for dgpd()

ba_threshold <- 175 
exceedances <- train$burned_area_km2_total[train$burned_area_km2_total >= ba_threshold] - ba_threshold

gpd_fit <- gpd.fit(train$burned_area_km2_total, threshold = ba_threshold)
sigma_hat <- gpd_fit$mle[1]
xi_hat    <- gpd_fit$mle[2]

cat(sprintf("Fitted GPD: sigma (scale) = %.2f, xi (shape) = %.3f\n", sigma_hat, xi_hat))

# Overlay fitted density on a histogram of the exceedances
x_seq <- seq(0, max(exceedances), length.out = 200)
gpd_density <- dgpd(x_seq, loc = 0, scale = sigma_hat, shape = xi_hat)

hist(exceedances, breaks = 15, freq = FALSE,
     main = sprintf("GPD fit at threshold = %f", ba_threshold, xi_hat),
     xlab = "Exceedance (burned area - threshold, km\u00b2)")
lines(x_seq, gpd_density, col = "steelblue", lwd = 2)
```


```{r}
your_threshold <- 275  # a defensible pick within the stable plateau of both diagnostics

gpd_fit_final <- gpd.fit(train$burned_area_km2_total, threshold = your_threshold)
gpd.diag(gpd_fit_final)

sigma_hat <- gpd_fit_final$mle[1]
xi_hat    <- gpd_fit_final$mle[2]
cat(sprintf("nexc = %d, sigma = %.2f, xi = %.3f (SE %.3f)\n",
            gpd_fit_final$nexc, sigma_hat, gpd_fit_final$se[2], xi_hat))

if (xi_hat < 0) {
  implied_max <- your_threshold - sigma_hat / xi_hat
  cat(sprintf("GPD-implied maximum burned area: %.1f km\u00b2 (observed max: %.1f km\u00b2)\n",
              implied_max, max(train$burned_area_km2_total)))
}
```


Final thoughts: the code above evaluated whether a Generalized Pareto Distribution (GPD), fit via peaks-over-threshold, would better represent the tail of the burned-area distribution than the Gamma currently used in the extreme-fire sub-model. Threshold stability diagnostics (mean residual life plot; parameter stability across thresholds 50–500 km²) indicated that the shape parameter stabilizes in the range of approximately 250–300 km², closely consistent with the 95th-percentile threshold (254.3 km²) already used to define the extreme-fire regime, supporting the current threshold choice as a reasonable, data-consistent cutoff rather than an arbitrary choice.

GDP-implied cutoff: an estimate of the theoretical upper bound on burned area implied by the fitted tail. For a Generalized Pareto Distribution with shape parameter ξ < 0 (a bounded, Weibull-type tail), the distribution has a finite right endpoint, computed directly from the fitted scale (σ) and shape parameters as threshold − σ/ξ. At a threshold set to 254.3 km² (chosen from 95% cutoff), this implied a cutoff estimate for the maximum burned area of approximately 8,583 km², four times larger than the largest fire actually observed in the training record (2,176 km²). This is not evidence of model misspecification; it reflects a 10-year observational window that contains only 52 fires classified as extreme, has almost certainly not sampled the true upper range of fire sizes the underlying physical process is capable of producing. The GPD's implied cutoff formalizes this limitation, offering a data-driven estimate of how much larger an extreme fire event could plausibly be, beyond anything captured in the data used to train the model.

When interpreting the extreme-fire Gamma component's behavior during posterior sampling: the Gamma distribution's support is unbounded, so draws from the extreme-fire model occasionally exceed the observed historical maximum , up to 6,000 km² in some posterior draws. Rather than treating these as implausible outliers or evidence of poor model behavior, the GPD-implied cutoff suggests such draws are physically reasonable: they remain comfortably below the 8,583 km² ceiling independently estimated from extreme value theory, and are consistent with the possibility of a fire larger than any yet observed occurring within the range the data support. The Gamma distribution's unbounded sampling remains within the implied threshold; artificially capping predicted burned area at the historical maximum would understate legitimate tail uncertainty that the GPD analysis independently identifies

##############################
# ----- Fit bayes model
##############################


Funciton to fit model, and calcluate performance metrics
```{r}
# =============================================================================
# fit_mixture_gamma_formula(): same three-component structure as
# fit_mixture_gamma_custom()/_truncated(), but takes FULL FORMULA OBJECTS
# for each component instead of plain predictor-name vectors -- necessary
# to support interaction terms (rh * U, wind * rh, etc.), which a simple
# "paste(preds, collapse = '+')" approach cannot express.
#
# Truncation is deliberately NOT a fitting-time argument: it only affects
# how predictions are SAMPLED from the already-fitted extreme-fire Gamma
# (see sample_mixture_predictions()'s truncate_extreme argument), not how
# the classifier/extreme/typical models are estimated. Nothing about the
# posterior itself changes based on whether you later sample truncated
# or untruncated.
# =============================================================================
fit_mixture_gamma_formula <- function(train,
                                       test,
                                       formula_classifier,   # one-sided, e.g. ~ rh * urban_proximity_binary + rh * U + rh * vpd + tair_5y
                                       formula_extreme,      # one-sided, e.g. ~ wind * rh + C + U
                                       formula_typical,      # one-sided, e.g. ~ urban_proximity_binary + rh + precip_5y + precip_3m + vpd + tair_2m
                                       outcome           = "burned_area_km2_total",
                                       extreme_quantile  = 0.95,
                                       sigma_beta        = 1,
                                       sigma_alpha       = 5,
                                       iter              = 2000,
                                       chains            = 4,
                                       seed              = 123,
                                       n_boot            = 1000,
                                       adapt_delta       = 0.95,
                                       compiled_model    = NULL) {

  if (any(train[[outcome]] <= 0)) {
    stop(sprintf("All values of '%s' must be > 0. Found %d non-positive values.",
                 outcome, sum(train[[outcome]] <= 0)))
  }

  y_train <- train[[outcome]]
  y_test  <- test[[outcome]]

  threshold <- quantile(y_train, extreme_quantile)
  cat(sprintf("[mixture_formula] Extreme threshold (%.0fth pct of train): %.2f km\u00b2\n",
              extreme_quantile * 100, threshold))

  is_extreme_train <- as.integer(y_train >= threshold)
  cat(sprintf("[mixture_formula] Extreme fires in train: %d / %d (%.1f%%)\n",
              sum(is_extreme_train), length(is_extreme_train), 100 * mean(is_extreme_train)))

  # Suppress the intercept column in each design matrix (Stan estimates
  # alpha separately) -- append "- 1" to whatever formula was supplied,
  # rather than requiring the user to remember to write it themselves
  fm_classifier <- update(formula_classifier, ~ . - 1)
  fm_extreme    <- update(formula_extreme,    ~ . - 1)
  fm_typical    <- update(formula_typical,    ~ . - 1)

  # ── Compile (or reuse) Stan models ───────────────────────────────────────
  if (is.null(compiled_model) || is.null(compiled_model$logistic)) {
    cat("[mixture_formula] Compiling classification (logistic) Stan model...\n")
    sm_logit <- stan_model(model_code = .stan_bernoulli, model_name = "mixture_formula_logit")
  } else {
    sm_logit <- compiled_model$logistic
  }
  if (is.null(compiled_model) || is.null(compiled_model$gamma)) {
    cat("[mixture_formula] Compiling magnitude (Gamma) Stan model...\n")
    sm_gamma <- stan_model(model_code = .stan_gamma, model_name = "mixture_formula_gamma")
  } else {
    sm_gamma <- compiled_model$gamma
  }

  # ── 1. Classifier ─────────────────────────────────────────────────────────
  X_train_clf <- model.matrix(fm_classifier, data = train)
  X_test_clf  <- model.matrix(fm_classifier, data = test)

  stan_data_logit <- list(
    N = nrow(train), K = ncol(X_train_clf), X = X_train_clf,
    y = is_extreme_train, sigma_beta = sigma_beta, sigma_alpha = sigma_alpha
  )

  cat("[mixture_formula] Sampling classification model...\n")
  fit_logit <- sampling(sm_logit, data = stan_data_logit,
                         iter = iter, chains = chains, seed = seed,
                         control = list(adapt_delta = adapt_delta), refresh = 200)

  div_logit  <- rstan::get_num_divergent(fit_logit)
  rhat_logit <- max(summary(fit_logit)$summary[, "Rhat"], na.rm = TRUE)
  cat(sprintf("[mixture_formula] Classifier: divergences=%d, max R-hat=%.4f %s\n",
              div_logit, rhat_logit, ifelse(rhat_logit > 1.01 || div_logit > 0, "\u26a0", "\u2713")))

  # ── 2. Extreme-fire magnitude model ──────────────────────────────────────
  train_extreme <- train[is_extreme_train == 1, , drop = FALSE]
  y_extreme     <- y_train[is_extreme_train == 1]
  X_extreme     <- model.matrix(fm_extreme, data = train_extreme)
  X_test_ext    <- model.matrix(fm_extreme, data = test)

  stan_data_extreme <- list(
    N = nrow(train_extreme), K = ncol(X_extreme), X = X_extreme, y = y_extreme,
    sigma_beta = sigma_beta, sigma_alpha = sigma_alpha
  )

  cat(sprintf("[mixture_formula] Sampling extreme-fire magnitude model (N=%d, K=%d)...\n",
              nrow(train_extreme), ncol(X_extreme)))
  fit_extreme <- sampling(sm_gamma, data = stan_data_extreme,
                           iter = iter, chains = chains, seed = seed,
                           control = list(adapt_delta = adapt_delta), refresh = 200)

  div_extreme  <- rstan::get_num_divergent(fit_extreme)
  rhat_extreme <- max(summary(fit_extreme)$summary[, "Rhat"], na.rm = TRUE)
  cat(sprintf("[mixture_formula] Extreme model: divergences=%d, max R-hat=%.4f %s\n",
              div_extreme, rhat_extreme, ifelse(rhat_extreme > 1.01 || div_extreme > 0, "\u26a0", "\u2713")))

  # ── 3. Typical-fire magnitude model ──────────────────────────────────────
  train_normal <- train[is_extreme_train == 0, , drop = FALSE]
  y_normal     <- y_train[is_extreme_train == 0]
  X_normal     <- model.matrix(fm_typical, data = train_normal)
  X_test_typ   <- model.matrix(fm_typical, data = test)

  stan_data_normal <- list(
    N = nrow(train_normal), K = ncol(X_normal), X = X_normal, y = y_normal,
    sigma_beta = sigma_beta, sigma_alpha = sigma_alpha
  )

  cat(sprintf("[mixture_formula] Sampling typical-fire magnitude model (N=%d, K=%d)...\n",
              nrow(train_normal), ncol(X_normal)))
  fit_normal <- sampling(sm_gamma, data = stan_data_normal,
                          iter = iter, chains = chains, seed = seed,
                          control = list(adapt_delta = adapt_delta), refresh = 200)

  div_normal  <- rstan::get_num_divergent(fit_normal)
  rhat_normal <- max(summary(fit_normal)$summary[, "Rhat"], na.rm = TRUE)
  cat(sprintf("[mixture_formula] Typical model: divergences=%d, max R-hat=%.4f %s\n",
              div_normal, rhat_normal, ifelse(rhat_normal > 1.01 || div_normal > 0, "\u26a0", "\u2713")))

  max_rhat  <- max(rhat_logit, rhat_extreme, rhat_normal, na.rm = TRUE)
  total_div <- div_logit + div_extreme + div_normal

  if (total_div > 0 || max_rhat > 1.05) {
    warning(sprintf(
      "One or more sub-models did not converge cleanly (total divergences: %d, max R-hat: %.4f). Inspect before trusting predictions.",
      total_div, max_rhat))
  }

  # ── Posterior summaries -- column names now naturally include the
  # interaction terms (e.g. "rh:urban_proximity_binaryX"), since they come
  # straight from model.matrix()'s own naming, not a manually-pasted list ──
  summ_logit   <- summary(fit_logit,   pars = c("alpha", "beta"))$summary
  summ_extreme <- summary(fit_extreme, pars = c("alpha", "beta", "phi"))$summary
  summ_normal  <- summary(fit_normal,  pars = c("alpha", "beta", "phi"))$summary

  coef_table <- rbind(
    transform(as.data.frame(summ_logit),
              variable  = c("alpha", paste0("beta_", colnames(X_train_clf))),
              component = "classification"),
    transform(as.data.frame(summ_extreme),
              variable  = c("alpha", paste0("beta_", colnames(X_extreme)), "phi"),
              component = "magnitude_extreme"),
    transform(as.data.frame(summ_normal),
              variable  = c("alpha", paste0("beta_", colnames(X_normal)), "phi"),
              component = "magnitude_typical")
  )
  rownames(coef_table) <- NULL

  # ── Predictions on test set: mean-blend, matching every other model in
  # this project's fitting/evaluation convention ──────────────────────────
  post_logit   <- rstan::extract(fit_logit)
  post_extreme <- rstan::extract(fit_extreme)
  post_normal  <- rstan::extract(fit_normal)

  n_draws <- min(length(post_logit$alpha), length(post_extreme$alpha), length(post_normal$alpha))

  pred_draws      <- matrix(NA_real_, nrow = nrow(test), ncol = n_draws)
  p_extreme_draws <- matrix(NA_real_, nrow = nrow(test), ncol = n_draws)

  for (d in seq_len(n_draws)) {
    p_ext_d <- plogis(post_logit$alpha[d] + X_test_clf %*% post_logit$beta[d, ])

    mu_ext_d  <- exp(post_extreme$alpha[d] + X_test_ext %*% post_extreme$beta[d, ])
    mu_norm_d <- exp(post_normal$alpha[d]  + X_test_typ %*% post_normal$beta[d, ])

    pred_draws[, d]      <- p_ext_d * mu_ext_d + (1 - p_ext_d) * mu_norm_d
    p_extreme_draws[, d] <- p_ext_d
  }

  p_hat  <- apply(pred_draws, 1, median, na.rm = TRUE)
  p_sd   <- apply(pred_draws, 1, sd,    na.rm = TRUE)
  p_low  <- apply(pred_draws, 1, quantile, 0.025, na.rm = TRUE)
  p_high <- apply(pred_draws, 1, quantile, 0.975, na.rm = TRUE)
  p_extreme_mean <- apply(p_extreme_draws, 1, median, na.rm = TRUE)

  metrics <- compute_burned_metrics(y_test, p_hat, n_boot = n_boot, seed = seed)

  cat(sprintf("[mixture_formula] nRMSE    : %.4f (%.4f, %.4f)\n",
              metrics$nrmse, metrics$nrmse_ci[1], metrics$nrmse_ci[2]))
  cat(sprintf("[mixture_formula] Pearson r: %.4f (%.4f, %.4f)\n",
              metrics$cor, metrics$cor_ci[1], metrics$cor_ci[2]))
  cat(sprintf("[mixture_formula] Spearman : %.4f (%.4f, %.4f)\n",
              metrics$spearman, metrics$spearman_ci[1], metrics$spearman_ci[2]))

  pred_df <- tibble(
    lat = test$lat, lon = test$lon, year = test$year, month = test$month,
    burned_area_km2_total = y_test,
    burned_area_pred      = p_hat,
    residual              = y_test - p_hat,
    log_residual          = log(y_test + 1) - log(p_hat + 1),
    pred_sd    = p_sd,
    pred_low   = p_low,
    pred_high  = p_high,
    p_extreme  = p_extreme_mean
  )

  list(
    method            = "mixture_gamma_formula",
    formula_classifier = formula_classifier,   # stored as FORMULAS now,
    formula_extreme     = formula_extreme,      # not predictor-name vectors --
    formula_typical      = formula_typical,      # needed to rebuild identical
                                                  # design matrices at sampling time
    outcome           = outcome,
    threshold         = threshold,
    extreme_quantile  = extreme_quantile,
    fit_logit         = fit_logit,
    fit_extreme       = fit_extreme,
    fit_normal        = fit_normal,
    coef_table        = coef_table,
    predictions       = pred_df,
    metrics           = metrics,
    max_rhat          = max_rhat,
    max_rhat_logit    = rhat_logit,
    max_rhat_extreme  = rhat_extreme,
    max_rhat_normal   = rhat_normal,
    total_divergences = total_div,
    div_logit         = div_logit,
    div_extreme       = div_extreme,
    div_normal        = div_normal,
    compiled_model    = list(logistic = sm_logit, gamma = sm_gamma)
  )
}
```

burned_area_km2_total ~ rh * urban_proximity_binary + rh * precip_3m + tair_2m + vpd
```{r}
m20_mixture_interactions <- run_or_load_model(
  file.path(rds_root, "m20_mixture_interactions.rds"),
  fit_mixture_gamma_formula(
    train, test,
    formula_classifier = ~ rh * urban_proximity_binary + rh * U + rh * vpd + tair_5y,
    formula_extreme     = ~ wind * rh + C + U,
    formula_typical      = ~ rh * urban_proximity_binary + rh * precip_3m + tair_2m + vpd , 
    chains = 4
))

m20_mixture_interactions$metrics
```

```{r}
m20_mixture_interactions$metrics$r2
predictions_m20_mixture_interactions <- m20_mixture_interactions$predictions
#m20 predicitons
summary(predictions_m20_mixture_interactions$burned_area_pred)
#observed burned area
summary(predictions_m20_mixture_interactions$burned_area_km2_total)

#10-year total pred and obs
sum(predictions_m20_mixture_interactions$burned_area_pred)
sum(predictions_m20_mixture_interactions$burned_area_km2_total)

plot(predictions_m20_mixture_interactions$burned_area_pred, predictions_m20_mixture_interactions$burned_area_km2_total)
```

A sampled-draws comparison (generating actual rgamma() samples the way the full fire pipeline does, then comparing that distribution's summary statistics — spread, tail behavior — against the real observed distribution) is the more direct visual/distributional check, complementary to the mean-based accuracy metrics rather than a replacement for them. ie s-score. 

```{r}
library(dplyr)
library(data.table)
library(ggplot2)

# =============================================================================
# 1. SAMPLING: draws n_draws random samples from a fitted mixture model
#    (custom or custom_truncated). For the NON-truncated model, the extreme
#    component is drawn from an ordinary rgamma(). For the TRUNCATED model,
#    the extreme component is drawn from a Gamma truncated below at the
#    classification threshold, via REJECTION sampling (draw ordinary
#    rgamma() values, keep only those above threshold, repeat) -- this is
#    numerically stable regardless of how far into the tail the threshold
#    sits, unlike inverse-CDF sampling (qgamma() near probability 1 is
#    unstable and can produce astronomically large values).
# =============================================================================
sample_mixture_predictions <- function(model, test, n_draws = 500, seed = 123,
                                        truncate_extreme = NULL) {
  if (is.null(truncate_extreme)) {
    truncate_extreme <- grepl("truncated", model$method)
  }

  # Prefer stored FORMULAS (fit_mixture_gamma_formula()) when present --
  # these correctly preserve interaction terms. Fall back to the older
  # predictor-vector reconstruction (fit_mixture_gamma_custom[_truncated]())
  # for models that don't have formulas stored.
  if (!is.null(model$formula_classifier)) {
    fm_classifier <- update(model$formula_classifier, ~ . - 1)
    fm_extreme    <- update(model$formula_extreme,    ~ . - 1)
    fm_typical    <- update(model$formula_typical,    ~ . - 1)
  } else {
    preds_classifier <- if (!is.null(model$preds_classifier)) model$preds_classifier else model$predictors
    preds_extreme    <- if (!is.null(model$preds_extreme))    model$preds_extreme    else model$predictors
    preds_typical     <- if (!is.null(model$preds_typical))    model$preds_typical    else model$predictors

    if (is.null(preds_classifier) || is.null(preds_extreme) || is.null(preds_typical)) {
      stop(sprintf(
        "[%s] Could not find predictor names under formulas, split fields, or shared field. Actual fields: %s",
        model$method, paste(names(model), collapse = ", ")
      ))
    }

    fm_classifier <- as.formula(paste("~", paste(preds_classifier, collapse = " + "), "- 1"))
    fm_extreme    <- as.formula(paste("~", paste(preds_extreme,    collapse = " + "), "- 1"))
    fm_typical    <- as.formula(paste("~", paste(preds_typical,    collapse = " + "), "- 1"))
  }

  threshold <- model$threshold
  outcome   <- model$outcome

  X_test_clf <- model.matrix(fm_classifier, data = test)
  X_test_ext <- model.matrix(fm_extreme,    data = test)
  X_test_typ <- model.matrix(fm_typical,    data = test)

  post_logit   <- rstan::extract(model$fit_logit)
  post_extreme <- rstan::extract(model$fit_extreme)
  post_normal  <- rstan::extract(model$fit_normal)

  n_post <- min(length(post_logit$alpha), length(post_extreme$alpha), length(post_normal$alpha))
  set.seed(seed)
  draw_idx <- sample(seq_len(n_post), size = n_draws, replace = n_draws > n_post)

  n_test  <- nrow(test)
  sampled <- matrix(NA_real_, nrow = n_test, ncol = n_draws)
  is_ext_mat <- matrix(NA_integer_, nrow = n_test, ncol = n_draws)

  # Rejection-sampling truncated Gamma
  rgamma_truncated <- function(n, shape, rate, lower, max_attempts = 10000) {
    out <- numeric(n)
    needed <- rep(TRUE, n)
    attempts <- 0

    while (any(needed) && attempts < max_attempts) {
      candidates <- rgamma(sum(needed), shape = shape, rate = rate)
      accepted <- candidates > lower
      out[which(needed)[accepted]] <- candidates[accepted]
      needed[which(needed)[accepted]] <- FALSE
      attempts <- attempts + 1
    }

    if (any(needed)) {
      warning(sprintf(
        "%d of %d truncated-Gamma draws did not accept within %d attempts (shape/rate imply almost no mass above threshold=%.2f). Falling back to `lower` for these.",
        sum(needed), n, max_attempts, lower
      ))
      out[needed] <- lower
    }

    out
  }

  n_nonfinite_warned <- FALSE

  for (d in seq_len(n_draws)) {
    di <- draw_idx[d]

    p_ext_d <- as.numeric(plogis(post_logit$alpha[di] + X_test_clf %*% post_logit$beta[di, ]))
    is_ext  <- rbinom(n_test, size = 1, prob = p_ext_d)

    mu_ext_d   <- as.numeric(exp(post_extreme$alpha[di] + X_test_ext %*% post_extreme$beta[di, ]))
    phi_ext_d  <- post_extreme$phi[di]
    rate_ext_d <- phi_ext_d / mu_ext_d

    mu_typ_d   <- as.numeric(exp(post_normal$alpha[di] + X_test_typ %*% post_normal$beta[di, ]))
    phi_typ_d  <- post_normal$phi[di]
    rate_typ_d <- phi_typ_d / mu_typ_d

    # Guard: catch a genuinely broken upstream mu/rate immediately and
    # loudly, rather than let a non-finite value propagate silently into
    # rgamma()/rgamma_truncated() and produce an exploded result
    if (!n_nonfinite_warned && (any(!is.finite(mu_ext_d)) || any(!is.finite(rate_ext_d)) ||
                                 any(!is.finite(mu_typ_d)) || any(!is.finite(rate_typ_d)))) {
      warning(sprintf(
        "[%s] Non-finite mu/rate encountered at draw %d (posterior index %d). Check X_test_ext/X_test_typ column alignment against this model's fitted beta.",
        model$method, d, di
      ))
      n_nonfinite_warned <- TRUE
    }

    ext_draw <- if (truncate_extreme) {
      rgamma_truncated(n_test, shape = phi_ext_d, rate = rate_ext_d, lower = threshold)
    } else {
      rgamma(n_test, shape = phi_ext_d, rate = rate_ext_d)
    }
    typ_draw <- rgamma(n_test, shape = phi_typ_d, rate = rate_typ_d)

    sampled[, d]    <- ifelse(is_ext == 1, ext_draw, typ_draw)
    is_ext_mat[, d] <- is_ext
  }

  list(
    sampled_draws   = sampled,
    is_extreme_draw = is_ext_mat,
    observed        = test[[outcome]],
    n_draws         = n_draws,
    truncated       = truncate_extreme,
    model_label     = model$method
  )
}

# =============================================================================
# 2. METRICS: coverage, PIT, accuracy from the sampled mean
# =============================================================================
compute_sampled_metrics <- function(sample_result, n_boot = 1000, seed = 123) {
  sampled  <- sample_result$sampled_draws
  observed <- sample_result$observed
  n_test   <- nrow(sampled)

  lower <- apply(sampled, 1, quantile, probs = 0.025, na.rm = TRUE)
  upper <- apply(sampled, 1, quantile, probs = 0.975, na.rm = TRUE)
  covered <- (observed >= lower) & (observed <= upper)
  coverage_95 <- mean(covered, na.rm = TRUE)

  pit_values <- sapply(seq_len(n_test), function(i) {
    mean(sampled[i, ] <= observed[i], na.rm = TRUE)
  })

  pred_mean <- rowMeans(sampled, na.rm = TRUE)

  if (exists("compute_burned_metrics")) {
    accuracy <- compute_burned_metrics(observed, pred_mean, n_boot = n_boot, seed = seed)
  } else {
    accuracy <- list(
      nrmse = sqrt(mean((pred_mean - observed)^2, na.rm = TRUE)) / mean(observed, na.rm = TRUE),
      cor   = cor(pred_mean, observed, use = "complete.obs"),
      r2    = 1 - sum((observed - pred_mean)^2, na.rm = TRUE) /
                   sum((observed - mean(observed, na.rm = TRUE))^2, na.rm = TRUE)
    )
  }

  cat(sprintf("[%s] 95%% interval coverage: %.1f%% (target: 95%%)\n",
              sample_result$model_label, 100 * coverage_95))
  cat(sprintf("[%s] Mean PIT value: %.3f (target: 0.500 if well-calibrated)\n",
              sample_result$model_label, mean(pit_values, na.rm = TRUE)))

  list(
    coverage_95 = coverage_95,
    pit_values  = pit_values,
    lower_95    = lower,
    upper_95    = upper,
    accuracy_from_sampled_mean = accuracy
  )
}

# =============================================================================
# 3. HISTOGRAM: observed vs. sampled predicted distribution
# =============================================================================
plot_sampled_distribution <- function(sample_result, var_label, title = NULL,
                                       sscore_bins = 15, log_x = FALSE,
                                       remove_zeros = TRUE,
                                       low_width = 10, low_max = 100, high_width = 100) {

  n_draws <- sample_result$n_draws
  pred_vals_raw <- as.vector(sample_result$sampled_draws)
  obs_vals_raw  <- sample_result$observed

  pct_pred_zero <- mean(pred_vals_raw == 0, na.rm = TRUE) * 100
  pct_obs_zero  <- mean(obs_vals_raw == 0, na.rm = TRUE) * 100

  if (remove_zeros) {
    pred_vals <- pred_vals_raw[pred_vals_raw != 0]
    obs_vals  <- obs_vals_raw[obs_vals_raw != 0]
  } else {
    pred_vals <- pred_vals_raw
    obs_vals  <- obs_vals_raw
  }

  max_val <- max(c(pred_vals, obs_vals), na.rm = TRUE)

  # Custom piecewise bin scheme: fine (low_width) bins from 0 to low_max,
  # coarser (high_width) bins above that, up through the actual max value
  low_breaks <- seq(0, low_max, by = low_width)
  if (max_val > low_max) {
    high_breaks <- seq(low_max + high_width, ceiling(max_val / high_width) * high_width, by = high_width)
    breaks <- c(low_breaks, high_breaks)
  } else {
    breaks <- low_breaks
  }

  pred_binned <- data.table(bin = cut(pred_vals, breaks = breaks, include.lowest = TRUE))
  obs_binned  <- data.table(bin = cut(obs_vals,  breaks = breaks, include.lowest = TRUE))

  bin_levels <- levels(pred_binned$bin)

  pred_counts <- pred_binned[, .(count = .N / n_draws), by = bin]
  obs_counts  <- obs_binned[, .(count = .N), by = bin]

  pred_counts[, source := "Predicted (sampled)"]
  obs_counts[, source  := "Observed"]

  combined <- rbind(pred_counts, obs_counts)
  combined[, bin := factor(bin, levels = bin_levels)]

  bin_lookup <- data.table(
    bin = factor(bin_levels, levels = bin_levels),
    bin_lower = breaks[-length(breaks)],
    bin_upper = breaks[-1]
  )
  combined <- merge(combined, bin_lookup, by = "bin")

  # Shrink the TOTAL bin width slightly (inset_frac) so there's a visible
  # gap BETWEEN adjacent bins, not just between the two source-halves
  # within one bin -- this is what fixes the "fused into one weird-width
  # bar" look from before.
  inset_frac <- 0.10   # 10% of bin width reserved as inter-bin gap
  gap_frac   <- 0.05   # 5% of the remaining width as the intra-bin gap

  combined[, `:=`(
    bin_span   = bin_upper - bin_lower,
    bin_inset  = (bin_upper - bin_lower) * inset_frac / 2
  )]
  combined[, `:=`(
    inset_lower = bin_lower + bin_inset,
    inset_upper = bin_upper - bin_inset
  )]
  combined[, bin_mid_split := (inset_lower + inset_upper) / 2]

  combined[source == "Observed", `:=`(
    xmin = inset_lower,
    xmax = bin_mid_split - (inset_upper - inset_lower) * gap_frac / 2
  )]
  combined[source == "Predicted (sampled)", `:=`(
    xmin = bin_mid_split + (inset_upper - inset_lower) * gap_frac / 2,
    xmax = inset_upper
  )]

  summary_stats <- rbind(
    data.table(source = "Predicted (sampled)", stat = "Max", value = max(pred_vals, na.rm = TRUE)),
    data.table(source = "Observed",            stat = "Max", value = max(obs_vals,  na.rm = TRUE))
  )

  metric_label <- ""
  if (exists("compute_sscore")) {
    pred_dt <- data.table(
      draw_id = rep(seq_len(n_draws), each = nrow(sample_result$sampled_draws)),
      value   = as.vector(sample_result$sampled_draws)
    )
    if (remove_zeros) {
      pred_dt   <- pred_dt[value != 0]
      obs_nonzero <- obs_vals_raw[obs_vals_raw != 0]
    } else {
      obs_nonzero <- obs_vals_raw
    }
    sscore_draws <- compute_sscore(pred_dt, obs_nonzero, sscore_bins, bin_method = "quantile_union")
    sscore_mean  <- mean(sscore_draws)
    sscore_ci    <- quantile(sscore_draws, c(0.025, 0.975))
    metric_label <- sprintf("S-score: %.2f [%.2f, %.2f]", sscore_mean, sscore_ci[1], sscore_ci[2])
  }

  zero_label <- sprintf("Zeros removed:\nPredicted: %.1f%%\nObserved: %.1f%%\n%s\n%s",
                         if (remove_zeros) pct_pred_zero else 0,
                         if (remove_zeros) pct_obs_zero  else 0,
                         metric_label,
                         ifelse(sample_result$truncated, "(truncated extreme sampling)", "(untruncated)"))

  p <- ggplot() +
    geom_rect(data = combined, aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = count, fill = source)) +
    geom_vline(data = summary_stats,
               aes(xintercept = value, color = source, linetype = stat),
               linewidth = 0.7) +
    scale_fill_manual(values = c("Observed" = "firebrick", "Predicted (sampled)" = "steelblue"), name = NULL) +
    scale_color_manual(values = c("Observed" = "firebrick", "Predicted (sampled)" = "steelblue"), guide = "none") +
    scale_linetype_manual(values = c("Max" = "dotted"), name = NULL) +
    guides(linetype = guide_legend(override.aes = list(color = "black"))) +
    annotate("text", x = Inf, y = Inf, label = zero_label,
             hjust = 1.05, vjust = 1.1, size = 3.0, color = "black") +
    labs(title = title, x = var_label, y = "Count", fill = NULL) +
    theme_minimal()

  if (log_x) p <- p + scale_x_log10()
  p
}
```

```{r}
# =============================================================================
# Sampling + metrics + histogram for every MIXTURE-family model (m1_mixture
# and m3-m19, the three-component classifier+extreme+typical models).
# The single-component models (m1_bayes_g, m1_bayes_ln, m2_bayes_g,
# m2_bayes_ln) are NOT mixture models -- they have no $fit_logit/$fit_extreme/
# $fit_normal, so sample_mixture_predictions() doesn't apply to them and
# they're deliberately excluded here.
# =============================================================================

mixture_model_names <- c(
"m20_mixture_interactions"
)

# Sanity check: confirm every name actually exists as an object before
# looping, so a typo/missing model fails loudly and immediately rather
# than partway through an 18-model loop
missing_models <- mixture_model_names[!sapply(mixture_model_names, exists)]
if (length(missing_models) > 0) {
  stop("These model objects were not found in the environment: ",
       paste(missing_models, collapse = ", "))
}

samples_list <- list()
metrics_list <- list()
plots_list   <- list()

for (nm in mixture_model_names) {
  cat(sprintf("=== %s ===\n", nm))
  model_obj <- get(nm)

  samples_list[[nm]] <- sample_mixture_predictions(model_obj, test, n_draws = 500)
  metrics_list[[nm]] <- compute_sampled_metrics(samples_list[[nm]])
  plots_list[[nm]]   <- plot_sampled_distribution(
    samples_list[[nm]], "Burned area (km\u00b2)",
    title = nm, log_x = FALSE,
    low_width = 25, low_max = 100, high_width = 25
  )

  cat("\n")
}

# =============================================================================
# Quick-access summary table: coverage + PIT mean + accuracy, one row per
# model, so you can scan all 18 at once before opening any individual plot
# =============================================================================
sampled_summary <- do.call(rbind, lapply(names(metrics_list), function(nm) {
  m <- metrics_list[[nm]]
  data.frame(
    model        = nm,
    coverage_95  = m$coverage_95,
    mean_pit     = mean(m$pit_values, na.rm = TRUE),
    nrmse        = m$accuracy_from_sampled_mean$nrmse,
    cor          = m$accuracy_from_sampled_mean$cor,
    r2           = if (!is.null(m$accuracy_from_sampled_mean$r2)) m$accuracy_from_sampled_mean$r2 else NA
  )
}))

print(sampled_summary)
```


```{r}
plots_list[["m20_mixture_interactions"]]
plots_list[["m20_mixture_interactions"]] + xlim(0, 200)
```

- Overall accuracygood, AUC=0.887 and the specific pattern, year-conditional bias looks good. The classifier does make individual mistakes, a few true-typical fires get elevated P(extreme), a few true-extreme fires get a lowish one, but this is ordinary classifier noise at this AUC level, not a systematic failure. The mean bynyear group table: 0.031, 0.039, 0.035. No meaningful difference across year groups. If the classifier's imperfections were what's driving the regime-difference problem, expect 2007/2010 specifically to show elevated misclassification relative to the other groups, and it doesn't. So: real, but non-year-specific classifier noise, not the mechanism behind the regime failure. 

Next, test the two Gamma components: predicted burned areas for fires classified as extreme, versus fires classified as typical. This might reveal whether the two gamma models are too similar to carry over the classification into predictions of butned area. .
```{r}
library(dplyr)
library(data.table)
library(ggplot2)

# =============================================================================
# For each posterior draw, split predicted burned area by which component
# (extreme vs. typical) actually generated it that draw -- using
# is_extreme_draw, the per-draw classification outcome already stored in
# sample_mixture_predictions()'s output. This directly compares the two
# Gamma components' OUTPUT distributions, to check whether classification
# is actually carrying through to meaningfully different predictions.
# =============================================================================
compare_component_outputs <- function(sample_result, threshold) {
  sampled  <- sample_result$sampled_draws       # n_test x n_draws
  is_ext   <- sample_result$is_extreme_draw       # n_test x n_draws, same shape

  # Flatten both matrices in parallel, tagging each (fire, draw) cell by
  # which component generated it
  dt <- data.table(
    predicted_value = as.vector(sampled),
    component       = as.vector(is_ext)
  )
  dt[, component := fifelse(component == 1, "Classified extreme", "Classified typical")]

  dt
}

component_outputs_m20 <- compare_component_outputs(samples_list[["m20_mixture_interactions"]], threshold = 254.2795)

# ---- Summary stats per component ----
cat("Predicted burned area, by which component generated it (all draws pooled):\n")
print(component_outputs_m20[, .(
  median = median(predicted_value),
  mean   = mean(predicted_value),
  p75    = quantile(predicted_value, 0.75),
  p95    = quantile(predicted_value, 0.95),
  max    = max(predicted_value),
  n      = .N
), by = component])

# ---- Overlapping density plot: are the two components' OUTPUT
# distributions actually well-separated, or largely overlapping? ----
ggplot(component_outputs_m20[predicted_value > 0], aes(x = predicted_value, fill = component)) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +
  scale_fill_manual(values = c("Classified extreme" = "#D55E00", "Classified typical" = "#0072B2")) +
  labs(title = "m20: predicted burned area distribution, by classified component",
       subtitle = "If the two densities overlap heavily, the extreme/typical Gamma\ncomponents are not producing meaningfully different predictions",
       x = "Predicted burned area (km\u00b2, log scale)", y = "Density", fill = NULL) +
  theme_minimal()

#get a closer look at the overlap in burned area between extreme typical fires by removing log scale and limiting burned areas to the threshold (95% 254.2795):
ggplot(component_outputs_m20[predicted_value > 0], aes(x = predicted_value, fill = component)) +
    geom_density(alpha = 0.5) +
    xlim(100, 5000) +
    scale_fill_manual(values = c("Classified extreme" = "#D55E00", "Classified typical" = "#0072B2")) +
    theme_minimal()

# ---- Direct overlap check: what fraction of "classified typical" draws
# actually exceed the extreme threshold anyway, and vice versa? ----
pct_typical_exceeding_threshold <- component_outputs_m20[component == "Classified typical",
                                                            mean(predicted_value >= threshold)] * 100
pct_extreme_below_threshold <- component_outputs_m20[component == "Classified extreme",
                                                        mean(predicted_value < threshold)] * 100

cat(sprintf("\n%.1f%% of draws classified TYPICAL still predicted burned area >= threshold (%.1f km\u00b2)\n",
            pct_typical_exceeding_threshold, threshold))
cat(sprintf("%.1f%% of draws classified EXTREME still predicted burned area < threshold\n",
            pct_extreme_below_threshold))
```

Truncated m20 extreme predictions at the 95% threshold (254.2795), meaning fires classified as extremem cannot be less thatn 254.2795 km2. 
```{r}
samples_list[["m20_mixture_interactions_truncated"]] <- sample_mixture_predictions(
  m20_mixture_interactions, test, n_draws = 500, truncate_extreme = TRUE
)
samples_list[["m20_mixture_interactions_truncated"]] <- attach_identifiers(
  samples_list[["m20_mixture_interactions_truncated"]], test
)

component_outputs_m20_trunc <- compare_component_outputs(
  samples_list[["m20_mixture_interactions_truncated"]], threshold = 254.2795
)

pct_typical_m20_trunc <- component_outputs_m20_trunc[component == "Classified typical", mean(predicted_value >= 254.2795)] * 100
pct_extreme_m20_trunc <- component_outputs_m20_trunc[component == "Classified extreme", mean(predicted_value < 254.2795)] * 100

cat(sprintf("m20 (truncated): %.1f%% of TYPICAL draws >= threshold | %.1f%% of EXTREME draws < threshold\n",
            pct_typical_m20_trunc, pct_extreme_m20_trunc))

#-------------
# CLASSIFIER
#-------------
build_classifier_diagnostic <- function(sample_result, threshold, test) {
  p_ext_mean <- rowMeans(sample_result$is_extreme_draw, na.rm = TRUE)

  dt <- data.table(
    lat        = sample_result$lat,
    lon        = sample_result$lon,
    year       = sample_result$year,
    observed   = sample_result$observed,
    p_extreme  = p_ext_mean,
    true_extreme = sample_result$observed >= threshold
  )
  dt
}

clf_m <- build_classifier_diagnostic(samples_list[["m20_mixture_interactions_truncated"]], threshold = 254.2795, test = test)

# ---- 1. ROC / AUC: rebuilt from clf_m, NOT reused from any earlier model ----
roc_obj <- pROC::roc(clf_m$true_extreme, clf_m$p_extreme, quiet = TRUE)
cat(sprintf("Classifier AUC: %.3f\n", pROC::auc(roc_obj)))
plot(roc_obj, main = sprintf("Classifier ROC (AUC = %.3f)", pROC::auc(roc_obj)))

# ---- 2. Distribution of P(extreme) split by TRUE label ----
ggplot(clf_m, aes(x = true_extreme, y = p_extreme)) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.3, color = "steelblue") +
  geom_boxplot(width = 0.3, alpha = 0.3, outlier.shape = NA) +
  scale_x_discrete(labels = c("FALSE" = "True: typical (< threshold)", "TRUE" = "True: extreme (\u2265 threshold)")) +
  labs(title = "Classifier-predicted P(extreme) by true fire class",
       x = NULL, y = "Mean P(extreme) across 500 draws") +
  theme_minimal()

# ---- 3. Year-group check ----
clf_m[, year_group := fifelse(year %in% c(2007, 2010), "2007/2010 (many small)",
                          fifelse(year %in% c(2004, 2005), "2004/2005 (severe)", "Other years"))]

true_typical_only <- clf_m[true_extreme == FALSE]

ggplot(true_typical_only, aes(x = year_group, y = p_extreme)) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.3, aes(color = year_group)) +
  geom_boxplot(width = 0.3, alpha = 0.3, outlier.shape = NA) +
  scale_color_manual(values = c("2007/2010 (many small)" = "#0072B2",
                                  "2004/2005 (severe)"     = "#D55E00",
                                  "Other years"             = "grey60"), guide = "none") +
  labs(title = "P(extreme) among TRULY TYPICAL (small) fires only, by year group",
       subtitle = "If 2007/2010 shows elevated P(extreme) here, the model is spuriously\ninflating some genuinely small fires toward the extreme component",
       x = NULL, y = "Mean P(extreme) across 500 draws") +
  theme_minimal()

cat("\nMean P(extreme) among true-typical fires, by year group:\n")
print(true_typical_only[, .(mean_p_extreme = mean(p_extreme), n = .N), by = year_group])

#-------------
# GAMMA MODELS
#-------------
compare_component_outputs <- function(sample_result, threshold) {
  sampled  <- sample_result$sampled_draws       # n_test x n_draws
  is_ext   <- sample_result$is_extreme_draw       # n_test x n_draws, same shape

  # Flatten both matrices in parallel, tagging each (fire, draw) cell by
  # which component generated it
  dt <- data.table(
    predicted_value = as.vector(sampled),
    component       = as.vector(is_ext)
  )
  dt[, component := fifelse(component == 1, "Classified extreme", "Classified typical")]

  dt
}

component_outputs <- compare_component_outputs(samples_list[["m20_mixture_interactions_truncated"]], threshold = 254.2795)

# ---- Summary stats per component ----
cat("Predicted burned area, by which component generated it (all draws pooled):\n")
print(component_outputs[, .(
  median = median(predicted_value),
  mean   = mean(predicted_value),
  p75    = quantile(predicted_value, 0.75),
  p95    = quantile(predicted_value, 0.95),
  max    = max(predicted_value),
  n      = .N
), by = component])

# ---- Overlapping density plot: are the two components' OUTPUT
# distributions actually well-separated, or largely overlapping? ----
ggplot(component_outputs[predicted_value > 0], aes(x = predicted_value, fill = component)) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +
  scale_fill_manual(values = c("Classified extreme" = "#D55E00", "Classified typical" = "#0072B2")) +
  labs(title = "Predicted burned area distribution, by classified component",
       subtitle = "If the two densities overlap heavily, the extreme/typical Gamma\ncomponents are not producing meaningfully different predictions",
       x = "Predicted burned area (km\u00b2, log scale)", y = "Density", fill = NULL) +
  theme_minimal()

#get a closer look at the overlap in burned area between extreme typical fires by removing log scale and limiting burned areas to the threshold (95% 254.2795):
ggplot(component_outputs[predicted_value > 0], aes(x = predicted_value, fill = component)) +
    geom_density(alpha = 0.5) +
    xlim(200, 5000) +
    scale_fill_manual(values = c("Classified extreme" = "#D55E00", "Classified typical" = "#0072B2")) +
    theme_minimal()

hist(component_outputs[component == "Classified extreme"]$predicted_value, xlim=c(200, 5000))

ggplot(component_outputs[predicted_value > 0], aes(x = predicted_value, fill = component)) +
    geom_density(alpha = 0.5) +
    xlim(1, 5000) +
    scale_fill_manual(values = c("Classified extreme" = "#D55E00", "Classified typical" = "#0072B2")) +
    theme_minimal()

hist(component_outputs[predicted_value > 0]$predicted_value) #frequency is higher because of 500 draws
hist(component_outputs[predicted_value > 0]$predicted_value, xlim=c(0, max(test$burned_area_km2_total)))
hist(test$burned_area_km2_total)

# ---- Direct overlap check: what fraction of "classified typical" draws
# actually exceed the extreme threshold anyway, and vice versa? ----
pct_typical_exceeding_threshold <- component_outputs[component == "Classified typical",
                                                            mean(predicted_value >= threshold)] * 100
pct_extreme_below_threshold <- component_outputs[component == "Classified extreme",
                                                        mean(predicted_value < threshold)] * 100

cat(sprintf("\n%.1f%% of draws classified TYPICAL still predicted burned area >= threshold (%.1f km\u00b2)\n",
            pct_typical_exceeding_threshold, threshold))
cat(sprintf("%.1f%% of draws classified EXTREME still predicted burned area < threshold\n",
            pct_extreme_below_threshold))
```


Sensitivity analysis: feed both gamma models with the full test set (exclude the classification model) to anser: "how different are the simulated distributions between the typical veruss extreme gamma models?"

```{r}
library(dplyr)
library(data.table)
library(ggplot2)

# =============================================================================
# SENSITIVITY TEST: feed the FULL test set through BOTH Gamma components
# independently (ignoring classification entirely), to compare how
# differently the extreme and typical sub-models behave in isolation.
# =============================================================================
sensitivity_test_gamma_components <- function(model, test, n_draws = 500, seed = 123,
                                                truncate_extreme = FALSE) {

  if (!is.null(model$formula_extreme)) {
    fm_extreme <- update(model$formula_extreme, ~ . - 1)
    fm_typical <- update(model$formula_typical, ~ . - 1)
  } else {
    preds_extreme <- if (!is.null(model$preds_extreme)) model$preds_extreme else model$predictors
    preds_typical <- if (!is.null(model$preds_typical)) model$preds_typical else model$predictors
    fm_extreme <- as.formula(paste("~", paste(preds_extreme, collapse = " + "), "- 1"))
    fm_typical <- as.formula(paste("~", paste(preds_typical, collapse = " + "), "- 1"))
  }

  threshold <- model$threshold

  X_test_ext <- model.matrix(fm_extreme, data = test)
  X_test_typ <- model.matrix(fm_typical, data = test)

  post_extreme <- rstan::extract(model$fit_extreme)
  post_normal  <- rstan::extract(model$fit_normal)

  n_post <- min(length(post_extreme$alpha), length(post_normal$alpha))
  set.seed(seed)
  draw_idx <- sample(seq_len(n_post), size = n_draws, replace = n_draws > n_post)

  n_test <- nrow(test)
  extreme_draws <- matrix(NA_real_, nrow = n_test, ncol = n_draws)
  typical_draws <- matrix(NA_real_, nrow = n_test, ncol = n_draws)

  rgamma_truncated <- function(n, shape, rate, lower, max_attempts = 10000) {
    out <- numeric(n)
    needed <- rep(TRUE, n)
    attempts <- 0
    while (any(needed) && attempts < max_attempts) {
      candidates <- rgamma(sum(needed), shape = shape, rate = rate)
      accepted <- candidates > lower
      out[which(needed)[accepted]] <- candidates[accepted]
      needed[which(needed)[accepted]] <- FALSE
      attempts <- attempts + 1
    }
    if (any(needed)) out[needed] <- lower
    out
  }

  for (d in seq_len(n_draws)) {
    di <- draw_idx[d]

    mu_ext_d   <- as.numeric(exp(post_extreme$alpha[di] + X_test_ext %*% post_extreme$beta[di, ]))
    phi_ext_d  <- post_extreme$phi[di]
    rate_ext_d <- phi_ext_d / mu_ext_d

    mu_typ_d   <- as.numeric(exp(post_normal$alpha[di] + X_test_typ %*% post_normal$beta[di, ]))
    phi_typ_d  <- post_normal$phi[di]
    rate_typ_d <- phi_typ_d / mu_typ_d

    # EVERY test row goes through BOTH components, regardless of classification
    extreme_draws[, d] <- if (truncate_extreme) {
      rgamma_truncated(n_test, shape = phi_ext_d, rate = rate_ext_d, lower = threshold)
    } else {
      rgamma(n_test, shape = phi_ext_d, rate = rate_ext_d)
    }
    typical_draws[, d] <- rgamma(n_test, shape = phi_typ_d, rate = rate_typ_d)
  }

  list(extreme_draws = extreme_draws, typical_draws = typical_draws,
       truncated = truncate_extreme, threshold = threshold)
}

# =============================================================================
# PLOT + SUMMARY: compare the two components' full-test-set distributions
# =============================================================================
plot_sensitivity_test <- function(sens_result, title = NULL) {
  dt <- data.table(
    predicted_value = c(as.vector(sens_result$extreme_draws), as.vector(sens_result$typical_draws)),
    model = rep(c("Extreme model (all fires)", "Typical model (all fires)"),
                times = c(length(sens_result$extreme_draws), length(sens_result$typical_draws)))
  )

  cat("Summary, full test set through EACH component independently:\n")
  print(dt[predicted_value > 0, .(
    median = median(predicted_value), mean = mean(predicted_value),
    p75 = quantile(predicted_value, 0.75), p95 = quantile(predicted_value, 0.95),
    max = max(predicted_value), n = .N
  ), by = model])

  # Approximate KS test on the pooled distributions -- note: draws from the
  # same posterior sample are reused across many test rows, so this is a
  # descriptive/approximate comparison of distributional SHAPE, not a
  # formally independent-sample hypothesis test
  ks_result <- suppressWarnings(ks.test(sens_result$extreme_draws[sens_result$extreme_draws > 0],
                                          sens_result$typical_draws[sens_result$typical_draws > 0]))
  cat(sprintf("\nApproximate KS statistic (distributional overlap): D = %.3f, p %s 0.001\n",
              ks_result$statistic, ifelse(ks_result$p.value < 0.001, "<", paste("=", round(ks_result$p.value, 4)))))

  p <- ggplot(dt[predicted_value > 0], aes(x = predicted_value, fill = model)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = c("Extreme model (all fires)" = "#D55E00", "Typical model (all fires)" = "#0072B2")) +
    geom_vline(xintercept = sens_result$threshold, linetype = "dashed", color = "grey30") +
    labs(title = title,
         subtitle = sprintf("Dashed line: extreme threshold (%.1f km\u00b2) %s",
                             sens_result$threshold,
                             ifelse(sens_result$truncated, "| extreme model TRUNCATED", "| extreme model UNTRUNCATED")),
         x = "Predicted burned area (km\u00b2, log scale)", y = "Density", fill = NULL) +
    theme_minimal()

  p
}

# =============================================================================
# RUN: m20, both truncated and untruncated, compared side by side
# =============================================================================
sens_m20_untrunc <- sensitivity_test_gamma_components(m20_mixture_interactions, test, n_draws = 500,
                                                        truncate_extreme = FALSE)
sens_m20_trunc   <- sensitivity_test_gamma_components(m20_mixture_interactions, test, n_draws = 500,
                                                        truncate_extreme = TRUE)

hist(sens_m20_trunc$extreme_draws, xlim=c(threshold, max(sens_m20_trunc$extreme_draws)))
hist(sens_m20_trunc$typical_draws)

hist(sens_m20_trunc$extreme_draws, xlim = c(threshold, max(sens_m20_trunc$extreme_draws)),
     freq = FALSE, main = "Extreme model (truncated), full test set",
     xlab = "Predicted burned area (km\u00b2)")

hist(sens_m20_trunc$typical_draws, freq = FALSE,
     main = "Typical model, full test set",
     xlab = "Predicted burned area (km\u00b2)")

p_untrunc <- plot_sensitivity_test(sens_m20_untrunc, title = "m20: full test set through both components (UNTRUNCATED)")
p_trunc   <- plot_sensitivity_test(sens_m20_trunc,   title = "m20: full test set through both components (TRUNCATED)")

p_untrunc + scale_x_log10() 
p_trunc + scale_x_log10() 

p_untrunc + xlim(5, max(sens_m20_untrunc$extreme_draws))
p_trunc + xlim(5, max(sens_m20_trunc$extreme_draws))
```



Reliability diagram for m20's extreme classifier:
Recomputes p_ext_d DIRECTLY from the posterior (not from the sampled binary is_extreme_draw outcomes), since a reliability diagram needs the continuous predicted probability, not a Monte Carlo-sampled 0/1 draw.
```{r}
compute_burned_area_reliability_by_draw <- function(model, test, n_bins = 10, n_draws = 500, seed = 123) {

  if (!is.null(model$formula_classifier)) {
    fm_classifier <- update(model$formula_classifier, ~ . - 1)
  } else {
    preds_classifier <- if (!is.null(model$preds_classifier)) model$preds_classifier else model$predictors
    fm_classifier <- as.formula(paste("~", paste(preds_classifier, collapse = " + "), "- 1"))
  }

  X_test_clf <- model.matrix(fm_classifier, data = test)
  threshold  <- model$threshold
  true_extreme <- as.integer(test[[model$outcome]] >= threshold)

  post_logit <- rstan::extract(model$fit_logit)
  n_post <- length(post_logit$alpha)
  set.seed(seed)
  draw_idx <- sample(seq_len(n_post), size = n_draws, replace = n_draws > n_post)

  # p_ext_d for every draw, pooled, to set FIXED bin edges shared across draws
  p_ext_all_draws <- sapply(draw_idx, function(di) {
    as.numeric(plogis(post_logit$alpha[di] + X_test_clf %*% post_logit$beta[di, ]))
  })

  bin_probs <- seq(0, 1, length.out = n_bins + 1)
  breaks <- unique(quantile(as.vector(p_ext_all_draws), probs = bin_probs, na.rm = TRUE))
  bin_levels <- levels(cut(as.vector(p_ext_all_draws), breaks = breaks, include.lowest = TRUE))

  rel_by_draw <- rbindlist(lapply(seq_len(n_draws), function(d) {
    p_ext_d <- p_ext_all_draws[, d]
    bin <- cut(p_ext_d, breaks = breaks, include.lowest = TRUE)

    dt <- data.table(bin = factor(bin, levels = bin_levels), p_ext_d = p_ext_d, true_extreme = true_extreme)
    dt[, .(
      mean_predicted = mean(p_ext_d),
      observed_rate  = mean(true_extreme),
      n_fires        = .N
    ), by = bin][, draw_id := d]
  }))

  rel_by_draw
}

plot_burned_area_reliability <- function(model, test, n_bins = 10, n_draws = 500, seed = 123,
                                          title = NULL) {

  rel_by_draw <- compute_burned_area_reliability_by_draw(model, test, n_bins, n_draws, seed)

  rel_summary <- rel_by_draw[, .(
    mean_predicted_avg   = mean(mean_predicted),
    mean_predicted_lower = quantile(mean_predicted, 0.025),
    mean_predicted_upper = quantile(mean_predicted, 0.975),
    observed_rate_avg    = mean(observed_rate),
    observed_rate_lower  = quantile(observed_rate, 0.025),
    observed_rate_upper  = quantile(observed_rate, 0.975),
    n_fires_avg          = mean(n_fires)
  ), by = bin][order(mean_predicted_avg)]

  p <- ggplot(rel_summary, aes(x = mean_predicted_avg, y = observed_rate_avg)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.5) +
    geom_errorbar(aes(ymin = observed_rate_lower, ymax = observed_rate_upper),
                  width = 0, alpha = 0.6, color = "firebrick", linewidth = 0.6) +
    geom_errorbarh(aes(xmin = mean_predicted_lower, xmax = mean_predicted_upper),
                   height = 0, alpha = 0.6, color = "firebrick", linewidth = 0.5) +
    geom_line(color = "black", linewidth = 0.6) +
    geom_point(aes(size = n_fires_avg), color = "black", alpha = 0.75) +
    scale_size_continuous(name = "Mean fires\nin bin", range = c(2, 10)) +
    labs(title = title,
         x = "Mean predicted P(extreme) (bin)",
         y = "Observed extreme-fire rate (fires \u2265 threshold / fires in bin)") +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(size = 12),
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10)
    )

  p
}

p_reliability_m20 <- plot_burned_area_reliability(
  m20_mixture_interactions, test, n_bins = 10, n_draws = 500,
  title = "m20: extreme-fire classifier reliability"
)
p_reliability_m20
```

