---
title: "Rate_Calibration"
output: html_document
---

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(patchwork)
library(ggplot2)
library(gridExtra)
library(terra)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(glmnet)
library(pROC)
library(PRROC)
```
Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]


performance_root <- file.path(performance_root, "ignition_model_redo")
rds_root <- file.path(rds_root, "Ignition_Efficiency/redo")

```


Read in drivers
```{r}
df <- read.csv(file.path(drivers_root, "lightning_ignition_jja_2002_2011_new.csv"))
df <- df[,2:26]

df <- df[,c("lat","lon","year","month","rh","tair","precip","wind","precip_1m","precip_2m","precip_3m","precip_5y","tair_1m","tair_2m","tair_3m","tair_5y","B", "C", "U", "strike_count","fire_count")]

#save gridcell-month-years where strike count > fire count to check against burned area training data
null_rows <- df  %>% filter(strike_count > 0 & fire_count > strike_count) %>% dplyr::select(year, lat, lon, month, strike_count, fire_count)
write.csv(null_rows, file.path(drivers_root, "null_ignition_rows_training_data_2002-2011.csv"))

#remove rows where strike count > fire count (impossible, plus the model won't be abke to handle this)
df <- df  %>% filter(!(strike_count > 0 & fire_count > strike_count))

head(df)
```

Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]
```

```{r}
# ── Save scaling parameters ───────────────────────────────────────────────────
predictors_ignition <- c("rh", "tair", "precip", "wind",
                          "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                          "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B", "C", "U")

mu_rh        <- mean(train$rh);        sigma_rh        <- sd(train$rh)
mu_tair      <- mean(train$tair);      sigma_tair      <- sd(train$tair)
mu_precip    <- mean(train$precip);    sigma_precip    <- sd(train$precip)
mu_wind      <- mean(train$wind);      sigma_wind      <- sd(train$wind)
mu_precip_1m <- mean(train$precip_1m); sigma_precip_1m <- sd(train$precip_1m)
mu_precip_2m <- mean(train$precip_2m); sigma_precip_2m <- sd(train$precip_2m)
mu_precip_3m <- mean(train$precip_3m); sigma_precip_3m <- sd(train$precip_3m)
mu_precip_5y <- mean(train$precip_5y); sigma_precip_5y <- sd(train$precip_5y)
mu_tair_1m   <- mean(train$tair_1m);   sigma_tair_1m   <- sd(train$tair_1m)
mu_tair_2m   <- mean(train$tair_2m);   sigma_tair_2m   <- sd(train$tair_2m)
mu_tair_3m   <- mean(train$tair_3m);   sigma_tair_3m   <- sd(train$tair_3m)
mu_tair_5y   <- mean(train$tair_5y);   sigma_tair_5y   <- sd(train$tair_5y)
mu_B         <- mean(train$B);         sigma_B         <- sd(train$B)
mu_C         <- mean(train$C);         sigma_C         <- sd(train$C)
mu_U         <- mean(train$U);         sigma_U         <- sd(train$U)

# Store and save
mean_sd_ignition <- data.frame(
  var   = predictors_ignition,
  mu    = c(mu_rh, mu_tair, mu_precip, mu_wind,
            mu_precip_1m, mu_precip_2m, mu_precip_3m, mu_precip_5y,
            mu_tair_1m, mu_tair_2m, mu_tair_3m, mu_tair_5y, mu_B, mu_C, mu_U),
  sigma = c(sigma_rh, sigma_tair, sigma_precip, sigma_wind,
            sigma_precip_1m, sigma_precip_2m, sigma_precip_3m, sigma_precip_5y,
            sigma_tair_1m, sigma_tair_2m, sigma_tair_3m, sigma_tair_5y, sigma_B, sigma_C, sigma_U)
)

write.csv(mean_sd_ignition,
          file.path(mu_sigma_root, "mu_sigma_ignition_predictors_redo.csv"),
          row.names = FALSE)

# ── Standardize train ─────────────────────────────────────────────────────────
train <- train %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B,
    C         = (C         - mu_C)         / sigma_C,
    U         = (U         - mu_U)         / sigma_U
  )

train <- train[, c("lat", "lon", "year", "month",
                   "strike_count", "fire_count",
                   "rh", "tair", "precip", "wind",
                   "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                   "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B", "C", "U")]

# ── Standardize test ──────────────────────────────────────────────────────────
# Use training means and sds 
test <- test %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B,
    C         = (C         - mu_C)         / sigma_C,
    U         = (U         - mu_U)         / sigma_U
  )

test <- test[, c("lat", "lon", "year", "month",
                 "strike_count", "fire_count",
                 "rh", "tair", "precip", "wind",
                 "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                 "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B", "C", "U")]

head(train)

```

##########################################
##### IGNITION EFFICINCY (p_ignite) #####
##########################################

Y ~ Binomial(n, p_ignite) --> technically, this is applied in the next step, to estimate fire count.

logit(p_ignite) = alpha + X * beta

alpha ~ normal(mu_alpha, sigma_alpha)
beta  ~ normal(0, sigma_beta)

p_ignite_corrected = plogis(qlogis(pie) + cal_shift)

p_ignite: probability of 1 strike resulting in an ignition in a given gridcell and month. 
X: average precipitation rate, temperature, relative humidity, cumulative precipitation for the previous 1, 2, 3 months, and 5 years, mean temperature for the previous 1, 2, 3 months, and 5 years, and % broadleaf cover. 


```{r}

#Bayes recalibrated
#Load calibrated Bayesian model
ignition_bayes_obj <- readRDS(file.path(rds_root, 
                                         "m9_bayes_calibrated.rds"))

ignition_bayes_fit <- ignition_bayes_obj$fit        # stanfit object
cal_shift_bayes    <- ignition_bayes_obj$cal_shift  # intercept shift
coef_means         <- ignition_bayes_obj$coef_means # posterior mean alpha and beta

cat("Bayesian calibration shift:", round(cal_shift_bayes, 3), "\n")

# ── Build predictor matrix from standardised data
predictors_m9 <- c("rh", "tair", "precip", "precip_1m", "precip_2m", "precip_3m", "tair_2m", "tair_3m", "tair_5y", "C", "U")

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m9, collapse = " + "), "- 1")),
  data = train
)

#generate predictions using posterior mean
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

train <- train |>
  mutate(
    #inverse_logit of linear predictor, linpred
    p_ignite_bayes     = plogis(linpred),
    #add calibration shift 
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,     #pmin and pmax make sure arguments in plogis from reaching infinity
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

#verify
cat("Raw Bayes p_ignite summary:\n")
print(summary(train$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(train$p_ignite_bayes_cal))

```

```{r}
# Does real interannual signal exist in the logit BEFORE the flat shift is
# applied? If yes, the shift's placement in the sigmoid tail is compressing
# it. If the signal is already weak here, the problem is upstream in m9's
# predictors, not something recalibration can fix.
train %>%
  group_by(year) %>%
  summarise(mean_logit_precal = mean(qlogis(pmin(pmax(p_ignite_bayes_cal, 1e-8), 1 - 1e-8)))) %>%
  ggplot(aes(year, mean_logit_precal)) + geom_point() + geom_line()

train <- train |>
  filter(strike_count > 0) |>   #the lightning model takes care of 0 stirke counts, so dont need the P_ignite model to worry about zeros.
  mutate(
    obs_rate    = fire_count / strike_count,
    eps         = 1e-8,
    logit_p_cal = qlogis(pmin(pmax(p_ignite_bayes_cal, eps), 1 - eps))
  )
#intercept only model:
rate_cal_model <- glm(
  obs_rate ~ offset(logit_p_cal),
  data    = train,
  family  = binomial(),
  weights = strike_count
)

# Upper bound on what's achievable: how much does a genuinely year-varying
# calibration improve fit, versus the flat single-shift version?
rate_cal_model_yearvarying <- glm(
  obs_rate ~ offset(logit_p_cal) + factor(year),
  data = train, family = binomial(), weights = strike_count
)


# Instead of a flat intercept, let the shift vary with an annual climate
# anomaly that generalizes out of sample (example: annual mean temperature
# anomaly, already standardized the same way as your other predictors)
rate_cal_model_climate <- glm(
  obs_rate ~ offset(logit_p_cal) + tair_5y,
  data = train, family = binomial(), weights = strike_count
)

anova(rate_cal_model, rate_cal_model_yearvarying, test = "Chisq")
anova(rate_cal_model, rate_cal_model_climate, test = "Chisq")
anova(rate_cal_model_yearvarying, rate_cal_model_climate, test = "Chisq")

AIC(rate_cal_model, rate_cal_model_yearvarying, rate_cal_model_climate)
```

There is a clear interannual signal (year explains a highly significant chunk of deviance, and AIC drops by ~185 points), but tair_5y isn't the variable carrying it (essentially zero improvement, and AIC gets worse once you account for the extra parameter). So the fix isn't "no interannual signal exists," it's "you haven't found the right proxy for it yet."

Find a variable that can act as a proxy for year variation:
```{r}
library(dplyr)
library(purrr)
library(tibble)

# 1. Get the per-year "required shift" directly -- fit the intercept-only
# calibration model SEPARATELY within each year, rather than pulling apart
# factor(year) reference-level algebra
year_shifts <- train %>%
  group_by(year) %>%
  group_modify(~ {
    m <- glm(obs_rate ~ offset(logit_p_cal), data = .x, family = binomial(), weights = strike_count)
    tibble(required_shift = coef(m)[1])
  }) %>%
  ungroup()

# 2. Annual (strike-count-weighted) mean of every candidate predictor,
# matching the weighting used in the calibration models themselves
candidate_predictors <- c("rh", "tair", "precip", "wind",
                           "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                           "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B", "C", "U")

annual_predictor_means <- train %>%
  group_by(year) %>%
  summarise(across(all_of(candidate_predictors),
                    ~ weighted.mean(.x, w = strike_count)),
            .groups = "drop")

# 3. Merge and screen correlations against the per-year required shift
screen_df <- year_shifts %>% left_join(annual_predictor_means, by = "year")

cor_results <- map_dfr(candidate_predictors, function(p) {
  tibble(predictor = p,
         correlation = cor(screen_df$required_shift, screen_df[[p]]))
}) %>%
  arrange(desc(abs(correlation)))

print(cor_results, n = Inf)
```
B is a strange result - may be driven by strike count varaition, not interannual. Plus, in paloe context, it won't be a good predictor (if part of just one or two predictors) when landcover data are so sparse. 
```{r}
train %>%
  group_by(year) %>%
  summarise(annual_mean_B = weighted.mean(B, w = strike_count),
            n_gridcells = n_distinct(interaction(lat, lon))) %>%
  summarise(range_B = diff(range(annual_mean_B)),
            within_gridcell_sd_B = train %>% distinct(lat, lon, B) %>% pull(B) %>% sd())
```


On B: the range across years (0.216) is roughly a fifth of the within-gridcell spatial spread (1.05), so most of B's variation is genuinely spatial, not temporal, consistent with what you'd expect from landcover updating only occasionally in the training data rather than every year.

Check for colinearity within top candidate predictors
```{r}
train %>%
  group_by(year) %>%
  summarise(across(c(rh, precip, precip_1m, precip_2m, precip_3m, precip_5y),
                    ~ weighted.mean(.x, w = strike_count))) %>%
  select(-year) %>%
  cor() %>%
  round(2)
```

```{r}
candidates <- list(
  c("precip_2m"),
  c("precip_2m", "precip_5y"),
  c("precip_3m", "precip_5y"),
  c("rh", "precip_5y")
)

results <- purrr::map_dfr(candidates, function(vars) {
  f <- as.formula(paste("obs_rate ~ offset(logit_p_cal) +", paste(vars, collapse = " + ")))
  m <- glm(f, data = train, family = binomial(), weights = strike_count)
  tibble::tibble(predictors = paste(vars, collapse = " + "), AIC = AIC(m))
})

results
AIC(rate_cal_model, rate_cal_model_yearvarying)
```

##### RATE CALIBRATION SHIFT FOR P_IGNITE ##### 

Ignition efficiency model not capture the observed fire/strike rate. 


This is due to a combination of model type and data structure.

What the model was trained to predict
The binomial logistic regression was trained on gridcell-month outcomes — cbind(fire_count, strike_count - fire_count) — where the outcome is whether each strike results in a fire. The model maximises the binomial likelihood across all gridcell-months simultaneously.
However, the vast majority of gridcell-months have fire_count = 0 even when strike_count > 0. The model sees thousands of zero-fire months and adjusts its intercept downward to fit them. But "zero fires in a gridcell-month with 50 strikes" doesn't mean each individual strike had zero probability of igniting — it means the probability was low enough that none of 50 independent trials succeeded. The model conflates these two things.

Rare event bias in logistic regression
When events are rare (ie. lightning-caused fires are extremely rare at the gridcell-month level (prevalence ~0.004)) maximum likelihood estimation of the intercept in logistic regression is biased downward. The intercept gets pulled toward negative infinity to fit the overwhelming proportion of zeros.

The binary calibration only partially fixes it
The binary calibration corrects the model to match the observed gridcell-month fire rate — the fraction of gridcell-months with at least one fire. But this is still not the same as the per-strike ignition rate.
Consider a gridcell-month with 100 strikes and 1 fire:
* Binary outcome: 1 (fire occurred) — contributes to binary calibration target
* Per-strike rate: 1/100 = 0.01 — this is what you need for p_fire = 1-(1-p)^n
The binary calibration makes the model predict the right fraction of fire-months, but p_ignite ends up being the gridcell-month fire probability, not the per-strike probability. When you then raise it to the power of hundreds of strikes, you get enormous fire probabilities.

The rate calibration fixes the scale
The rate calibration specifically targets sum(fire_count) / sum(strike_count) — the true per-strike ignition rate — and estimates the additional shift needed to align p_ignite with this quantity. This is the number that belongs in 1 - (1-p)^n_strikes.

In summary, as a flow:

Model training:
  Binomial GLM learns P(fire | gridcell-month climate)
  → Rare event bias pulls intercept too low
  → Binary calibration corrects to P(fire-month | climate)
  → But P(fire-month) ≠ P(fire | single strike)
  → Rate calibration corrects to P(fire | single strike)
  → Now correctly used in 1-(1-p)^n_strikes
The need for two calibration steps reflects the fact that the model was specified at the gridcell-month level for practical reasons (that's the resolution of your climate data) but needs to produce predictions at the per-strike level for the fire probability formula. These are fundamentally different quantities that happen to be related through the binomial formula, and the two calibration steps bridge that gap.

so the initial binomial logistic regression model is trained to predict the probability of a fire given x strikes. but not the probability of a fire for each strike. calibration fixes this. 


```{r}
#Estimate rate calibration shift from train data
#wses gridcell-months with strikes > 0
#weights by strike_count so high-strike months have more influence

train <- train |>
  filter(strike_count > 0) |>   #the lightning model takes care of 0 stirke counts, so dont need the P_ignite model to worry about zeros.
  mutate(
    obs_rate    = fire_count / strike_count,
    eps         = 1e-8,
    logit_p_cal = qlogis(pmin(pmax(p_ignite_bayes_cal, eps), 1 - eps))
  )

rate_cal_model <- glm(
  obs_rate ~ offset(logit_p_cal),
  data    = train,
  family  = binomial(),
  weights = strike_count
)

rate_cal_shift <- coef(rate_cal_model)[1]
cat("Rate calibration shift:", round(rate_cal_shift, 3), "\n")
# Expect a negative value — shifts predictions down toward true per-strike rate

#apply combined shift to all predictions
eps <- 1e-8

train <- train |>
  mutate(
    p_ignite_rate_cal = plogis(
      qlogis(pmin(pmax(p_ignite_bayes_cal, eps), 1 - eps)) + rate_cal_shift
    )
  )

# verify annual rates now match observed 
train |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate           = sum(fire_count) / sum(strike_count),
    pred_binary_cal    = mean(p_ignite_bayes_cal),
    pred_rate_cal      = mean(p_ignite_rate_cal),
    .groups            = "drop"
  )

#check expected fire count
cat("Expected fires (binary cal):  ",
    sum(test$strike_count * test$p_ignite_bayes_cal, na.rm = TRUE), "\n")
cat("Expected fires (rate cal):    ",
    sum(test$strike_count * test$p_ignite_rate_cal,  na.rm = TRUE), "\n")
cat("Observed fires:               ",
    sum(test$lightning_ignition), "\n")

#save shift for paleo application
saveRDS(list(
  cal_shift_bayes = cal_shift_bayes,   # binary calibration shift
  rate_cal_shift  = rate_cal_shift,    # rate calibration shift
  combined_shift  = cal_shift_bayes + rate_cal_shift
), file.path(rds_root, "calibration_shifts.rds"))

```


### Apply full ignition model to test set


```{r}

############################
# Intercept calibration
############################

X_new <- model.matrix(
  as.formula(paste("~", paste(predictors_m9, collapse = " + "), "- 1")),
  data = test
)

#generate predictions using posterior mean
alpha_mean <- coef_means$alpha
beta_means <- coef_means$beta

eps <- 1e-8

linpred <- as.numeric(alpha_mean + X_new %*% beta_means)

test <- test |>
  mutate(
    #inverse_logit of linear predictor, linpred
    p_ignite_bayes     = plogis(linpred),
    #add calibration shift 
    p_ignite_bayes_cal = plogis(pmin(pmax(linpred + cal_shift_bayes,     #pmin and pmax make sure arguments in plogis from reaching infinity
                                           log(eps / (1 - eps))),
                                      log((1 - eps) / eps)))
  )

# Note: adding cal_shift directly to the linear predictor (before plogis)
# is equivalent to plogis(qlogis(p) + cal_shift) but avoids the
# double transformation and is numerically more stable

#verify
cat("Raw Bayes p_ignite summary:\n")
print(summary(test$p_ignite_bayes))

cat("Calibrated Bayes p_ignite summary:\n")
print(summary(test$p_ignite_bayes_cal))

############################
# Rate Calibration Shift
############################


test <- test |>
  mutate(
    p_ignite_rate_cal = plogis(
      qlogis(pmin(pmax(p_ignite_bayes_cal, eps), 1 - eps)) + rate_cal_shift
    )
  )

# verify annual rates now match observed 
test |>
  filter(strike_count > 0) |>
  group_by(year) |>
  summarise(
    obs_rate           = sum(fire_count) / sum(strike_count),
    pred_binary_cal    = mean(p_ignite_bayes_cal),
    pred_rate_cal      = mean(p_ignite_rate_cal),
    .groups            = "drop"
  )





```

Try beta binomial instead of binomial+rate shift
```{r}
library(glmmTMB)
library(dplyr)

# ---- 1. Fit Beta-Binomial ignition model on the same predictors as m9 ----
# cbind(successes, failures) = cbind(fire_count, strike_count - fire_count)
# Beta-Binomial adds one dispersion parameter (phi) that absorbs
# extra-binomial variance -- the same fix NegBinomial provided over Poisson
# for strike counts, applied here to the ignition-efficiency stage.

predictors_beta_bin <- c("rh", "tair", "precip", "precip_1m", "tair_3m", "tair_5y", "C", "U")

bb_formula_1 <- as.formula(
  paste("cbind(fire_count, strike_count - fire_count) ~", paste(predictors_m9, collapse = " + "))
)

bb_formula_2 <- as.formula(
  paste("cbind(fire_count, strike_count - fire_count) ~", paste(predictors_beta_bin, collapse = " + "))
)

bb_model_1 <- glmmTMB(
  bb_formula_1,
  data   = train,   # already filtered to strike_count > 0 and standardized
  family = betabinomial(link = "logit")
)

bb_model_2 <- glmmTMB(
  bb_formula_2,
  data   = train,   # already filtered to strike_count > 0 and standardized
  family = betabinomial(link = "logit")
)

summary(bb_model_1); summary(bb_model_2)

# ---- 2. Predict on the test set (already standardized with training mu/sigma) ----
test <- test %>%
  mutate(p_ignite_betabinom_1 = predict(bb_model_1, newdata = test, type = "response"),
         p_ignite_betabinom_2 = predict(bb_model_2, newdata = test, type = "response"))

# ---- 3. Compare against the existing rate-calibrated intercept-only model ----
# test$p_ignite_rate_cal should already exist from your earlier pipeline run

comparison <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate              = sum(fire_count) / sum(strike_count),
    pred_rate_cal_shift    = mean(p_ignite_rate_cal),
    pred_rate_betabinom_1    = mean(p_ignite_betabinom_1),
    pred_rate_betabinom_2    = mean(p_ignite_betabinom_2),
    .groups = "drop"
  )

print(comparison, n = Inf)

# ---- 4. Summary metrics for both approaches, same structure as before ----
comparison %>%
  summarise(
    spearman_cal_shift  = cor(obs_rate, pred_rate_cal_shift, method = "spearman"),
    spearman_betabinom_1  = cor(obs_rate, pred_rate_betabinom_1, method = "spearman"),
    spearman_betabinom_2  = cor(obs_rate, pred_rate_betabinom_2, method = "spearman"),
    pearson_cal_shift   = cor(obs_rate, pred_rate_cal_shift),
    pearson_betabinom_1   = cor(obs_rate, pred_rate_betabinom_1),
    pearson_betabinom_2   = cor(obs_rate, pred_rate_betabinom_2),
    ratio_cal_shift     = sum(pred_rate_cal_shift) / sum(obs_rate),
    ratio_betabinom_1     = sum(pred_rate_betabinom_1) / sum(obs_rate),
    ratio_betabinom_2     = sum(pred_rate_betabinom_2) / sum(obs_rate)
  )

# ---- 5. Expected vs. observed fire count check (same test you already ran) ----
cat("Observed fires:                      ", sum(test$fire_count), "\n")
cat("Expected fires (rate-cal shift):     ", sum(test$strike_count * test$p_ignite_rate_cal, na.rm = TRUE), "\n")
cat("Expected fires (beta-binomial 1):      ", sum(test$strike_count * test$p_ignite_betabinom_2, na.rm = TRUE), "\n")
cat("Expected fires (beta-binomial 2):      ", sum(test$strike_count * test$p_ignite_betabinom_2, na.rm = TRUE), "\n")

# ---- 6. Visual comparison across years ----
library(tidyr)
library(ggplot2)

comparison_long <- comparison %>%
  pivot_longer(cols = c(obs_rate, pred_rate_cal_shift, pred_rate_betabinom_1, pred_rate_betabinom_2),
               names_to = "source", values_to = "rate")

ggplot(comparison_long, aes(x = year, y = rate, color = source)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(
    values = c(obs_rate = "firebrick", pred_rate_cal_shift = "steelblue", pred_rate_betabinom_1 = "darkgreen", pred_rate_betabinom_2 = "orange"),
    labels = c("Observed", "Predicted (rate-cal shift)", "Predicted (beta-binomial 1)", "Predicted (beta-binomial 2)")
  ) +
  labs(x = "Year", y = "Per-strike ignition rate", color = NULL) +
  theme_minimal()
```

Apply a rate shift to the beta binomial predictions:

```{r}
library(dplyr)

# Fit a rate-calibration shift on TOP of the beta-binomial's predictions,
# using the same intercept-only approach as before, just against a
# better-discriminating base model (the beta binimial)
train <- train %>%
  mutate(
    p_ignite_bb_2     = predict(bb_model_2, newdata = ., type = "response"),
    logit_p_bb_2      = qlogis(pmin(pmax(p_ignite_bb_2, 1e-8), 1 - 1e-8))
  )

rate_cal_model_bb <- glm(
  obs_rate ~ offset(logit_p_bb_2),
  data    = train,
  family  = binomial(),
  weights = strike_count
)
rate_cal_shift_bb <- coef(rate_cal_model_bb)[1]

test <- test %>%
  mutate(
    logit_p_bb_2        = qlogis(pmin(pmax(p_ignite_betabinom_2, 1e-8), 1 - 1e-8)),
    p_ignite_bb_ratecal  = plogis(logit_p_bb_2 + rate_cal_shift_bb)
  )

# Re-check both metrics on this combined approach
comparison_bb_cal <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate                = sum(fire_count) / sum(strike_count),
    pred_rate_bb_ratecal     = mean(p_ignite_bb_ratecal),
    .groups = "drop"
  )

comparison_bb_cal %>%
  summarise(
    spearman = cor(obs_rate, pred_rate_bb_ratecal, method = "spearman"),
    pearson  = cor(obs_rate, pred_rate_bb_ratecal),
    ratio    = sum(pred_rate_bb_ratecal) / sum(obs_rate)
  )

cat("Expected fires (beta-binomial + rate-cal shift): ",
    sum(test$strike_count * test$p_ignite_bb_ratecal, na.rm = TRUE), "\n")
```

Spatial comparison test:
```{r}
spatial_comparison <- test %>%
  filter(strike_count > 0) %>%
  group_by(lat, lon) %>%
  summarise(
    obs_rate             = sum(fire_count) / sum(strike_count),
    pred_rate_cal_shift   = mean(p_ignite_rate_cal),
    pred_rate_betabinom_1 = mean(p_ignite_betabinom_1),
    pred_rate_betabinom_2 = mean(p_ignite_betabinom_2),
    .groups = "drop"
  )

spatial_comparison %>%
  summarise(
    spearman_cal_shift     = cor(obs_rate, pred_rate_cal_shift, method = "spearman"),
    spearman_betabinom_1   = cor(obs_rate, pred_rate_betabinom_1, method = "spearman"),
    spearman_betabinom_2   = cor(obs_rate, pred_rate_betabinom_2, method = "spearman"),
    pearson_cal_shift      = cor(obs_rate, pred_rate_cal_shift),
    pearson_betabinom_1    = cor(obs_rate, pred_rate_betabinom_1),
    pearson_betabinom_2    = cor(obs_rate, pred_rate_betabinom_2)
  )
```
Spatial performance:
The switch to beta-binomial barely moved the needle here, and actually made Pearson marginally worse. Compare that to the temporal result, where Spearman more than doubled (0.273 → 0.576). This is a real, meaningful contrast: whatever extra-binomial variance the beta-binomial's dispersion parameter is absorbing appears to be predominantly temporal (year-to-year), not spatial (gridcell-to-gridcell). The model's improvement in discriminating high/low ignition-efficiency years didn't transfer into better discriminating high/low ignition-efficiency places — these are separable sources of variation, and your fix only addressed one of them.

A plausible explanation: the predictor set (rh, tair, precip-family, tair_5y, C, U) already carries a fair amount of spatial information through C/U (landcover) and the fact that climate variables themselves vary systematically by location, so the mean structure of both models was already doing whatever spatial work these predictors can do, and switching the variance structure from Binomial to Beta-Binomial doesn't add anything new on that axis. The dispersion parameter is capturing something else entirely, likely genuine year-to-year process variability (drought years, fire-weather anomalies) that isn't well represented by any static spatial covariate.

Performance over time:
Spearman stayed exactly at 0.576, which is expected; adding a constant in logit space is a strictly monotonic transformation, so it can never change rank order. The rate-cal-shift only adjusted the overall level, leaving the beta-binomial's improved year-to-year discrimination intact.

Aggregate bias is now essentially fixed: 264.0 expected fires against 271 observed is a ~2.6% underprediction. This is an improvement over the beta-binomial alone's 41% overprediction, and now in the same ballpark of accuracy as the original approach's near-perfect 271.3.

Discrimination improvement (Spearman more than doubled) without paying beta-binomial's calibration cost. 

Note the ratio column in printed summary (1.22) and the "expected fires" total (264.0 vs 271, ratio ≈ 0.97) look like they're telling different stories, but they're measuring different things. The ratio in summarise() is sum(pred_rate_bb_ratecal) / sum(obs_rate) — an unweighted sum of 10 annual mean rates. "Expected fires" is sum(strike_count * p_ignite) — a strike-count-weighted total across all ~48,000 gridcell-months. Since rate_cal_shift_bb was fit with weights = strike_count, it's specifically optimized to get the strike-weighted total right (which it now does),not the simple unweighted average of 10 year-level ratios, which is a somewhat different and more idiosyncratic quantity. The expected-fires number is the more meaningful one for your actual downstream use (since p_ignite ultimately gets multiplied by strike count to produce fire counts in the pipeline), so I'd treat 264.0 vs. 271 as the trustworthy bias check.

Pearson barely moving (0.284 → 0.285) is expected too, and worth stating plainly rather than glossing over: this confirms the magnitude-compression problem is still there — the model ranks years correctly (and better than the binomial model), but still doesn't reproduce how large the swings between high- and low-ignition years really are. That's a real, remaining limitation to be upfront about, not something this fix resolved.

The beta binomial on its own better discriminates ignition, but does not capture the 10-year expected fire count well. However, adding the calibration shift adjusts this, maintaining good discrimination, though it does no better a job of tracking anomaly magnitude (ie. gets the direction, but not magnitude). 

Plots:
```{r}

comparison <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate              = sum(fire_count) / sum(strike_count),
    pred_rate_cal_shift    = mean(p_ignite_rate_cal),
    pred_rate_betabinom_1    = mean(p_ignite_betabinom_1),
    pred_rate_betabinom_2    = mean(p_ignite_betabinom_2),
    pred_rate_betabinom_2_cal_shift    = mean(p_ignite_bb_ratecal),
    .groups = "drop"
  )

comparison_long <- comparison %>%
  pivot_longer(cols = c(obs_rate, pred_rate_cal_shift, pred_rate_betabinom_1, pred_rate_betabinom_2, pred_rate_betabinom_2_cal_shift),
               names_to = "source", values_to = "rate")

ggplot(comparison_long, aes(x = year, y = rate, color = source)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(
    values = c(obs_rate = "firebrick", pred_rate_cal_shift = "steelblue", pred_rate_betabinom_1 = "darkgreen", pred_rate_betabinom_2 = "orange", pred_rate_betabinom_2_cal_shift = "purple"),
    labels = c("Observed", "Predicted (binomial rate-cal shift)", "Predicted (beta-binomial 1)", "Predicted (beta-binomial 2 rate-cal shift)")
  ) +
  labs(x = "Year", y = "Per-strike ignition rate", color = NULL) +
  theme_minimal()

comparison <- test %>%
    filter(strike_count > 0) %>%
  mutate(
    obs_rate = case_when(
      strike_count == 0          ~ NA_real_,
      fire_count > strike_count  ~ 0,
      TRUE                        ~ fire_count / strike_count
    )
  )

ggplot() + geom_point(data=comparison, aes(x=p_ignite_bb_ratecal, y=obs_rate))
```



Apply beta binomial as a bayesian model:
```{r}

library(rstan)
library(dplyr)

# ---- 1. Prepare data for Stan, using the reduced predictor set (bb_model_2) ----
predictors_beta_bin <- c("rh", "tair", "precip", "precip_1m", "tair_3m", "tair_5y", "C", "U")

stan_data <- list(
  N            = nrow(train),
  K            = length(predictors_beta_bin),
  X            = as.matrix(train[, predictors_beta_bin]),
  fire_count   = train$fire_count,
  strike_count = train$strike_count
)

# ---- 2. Fit the Bayesian beta-binomial in Stan ----
# NOTE: N is large (~48,000 rows) -- this will take considerably longer to
# sample than the glmmTMB version. Consider running this from the terminal
# (same pattern as your unified pipeline) if its slow interactively.
bb_bayes_fit <- stan(
  file    = "beta_binomial_ignition.stan",   # save the Stan code above to this path
  data    = stan_data,
  chains  = 4,
  iter    = 2000,
  warmup  = 1000,
  cores   = 4,
  seed    = 123
)

# Basic convergence check before trusting anything downstream
print(bb_bayes_fit, pars = c("alpha", "beta", "phi"))
# Rhat should be ~1.00-1.01 for every parameter; check for divergent
# transitions in the sampler output above -- if present, consider
# tightening priors or reparameterizing before proceeding.

# ---- 3. Extract posterior draws and posterior-mean coefficients ----
bb_bayes_posterior <- rstan::extract(bb_bayes_fit, pars = c("alpha", "beta", "phi"))

coef_means_bb <- list(
  alpha = mean(bb_bayes_posterior$alpha),
  beta  = colMeans(bb_bayes_posterior$beta),
  phi   = mean(bb_bayes_posterior$phi)
)

# ---- 4. Posterior-mean predictions on train, for fitting the rate-cal shift ----
X_train <- as.matrix(train[, predictors_beta_bin])
linpred_train <- as.numeric(coef_means_bb$alpha + X_train %*% coef_means_bb$beta)

train <- train %>%
  mutate(
    p_ignite_bb_bayes = plogis(linpred_train),
    logit_p_bb_bayes  = qlogis(pmin(pmax(p_ignite_bb_bayes, 1e-8), 1 - 1e-8))
  )

# ---- 5. Rate-calibration shift, unchanged approach from before ----
rate_cal_model_bb_bayes <- glm(
  obs_rate ~ offset(logit_p_bb_bayes),
  data    = train,
  family  = binomial(),
  weights = strike_count
)
rate_cal_shift_bb_bayes <- coef(rate_cal_model_bb_bayes)[1]
cat("Rate calibration shift (Bayesian beta-binomial):", round(rate_cal_shift_bb_bayes, 3), "\n")

# ---- 6. Apply to test set and check the same metrics as before ----
X_test <- as.matrix(test[, predictors_beta_bin])
linpred_test <- as.numeric(coef_means_bb$alpha + X_test %*% coef_means_bb$beta)

test <- test %>%
  mutate(
    p_ignite_bb_bayes         = plogis(linpred_test),
    logit_p_bb_bayes          = qlogis(pmin(pmax(p_ignite_bb_bayes, 1e-8), 1 - 1e-8)),
    p_ignite_bb_bayes_ratecal = plogis(logit_p_bb_bayes + rate_cal_shift_bb_bayes)
  )

comparison_bb_bayes <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate                   = sum(fire_count) / sum(strike_count),
    pred_rate_bb_bayes_ratecal = mean(p_ignite_bb_bayes_ratecal),
    .groups = "drop"
  )

comparison_bb_bayes %>%
  summarise(
    spearman = cor(obs_rate, pred_rate_bb_bayes_ratecal, method = "spearman"),
    pearson  = cor(obs_rate, pred_rate_bb_bayes_ratecal),
    ratio    = sum(pred_rate_bb_bayes_ratecal) / sum(obs_rate)
  )

cat("Expected fires (Bayesian beta-binomial + rate-cal shift): ",
    sum(test$strike_count * test$p_ignite_bb_bayes_ratecal, na.rm = TRUE), "\n")

# ---- 7. Save in the same structure your unified pipeline already expects ----
# This mirrors m9_bayes_calibrated.rds's format ($fit, $coef_means, plus the
# separate calibration_shifts.rds pattern) so it drops directly into
# predict_ignition() with minimal changes.
saveRDS(list(
  fit        = bb_bayes_fit,
  coef_means = coef_means_bb,
  predictors = predictors_beta_bin
), file.path(rds_root, "beta_binomial_ignition_bayes.rds"))

saveRDS(list(
  rate_cal_shift = rate_cal_shift_bb_bayes
), file.path(rds_root, "beta_binomial_calibration_shift.rds"))
```



Save betal binomial and calibration shift
```{r}

```



