---
title: "Rate_Calibration"
output: html_document
---

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(patchwork)
library(ggplot2)
library(gridExtra)
library(terra)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(glmnet)
library(pROC)
library(PRROC)
```
Load configuration files
- location of model drivers, rds files (containing models), tables of parameter estimates and performance metrics, and figures. 
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

# *** paleo_config.properties must be in Paleo_Lightning_Fire_Uden_2026
#location of this script:
current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
#get path to this project
base_path <- sub("(.*Paleo_Lightning_Fire_Uden_2026).*", "\\1", current_path)
#open 
config <- read_properties(file.path(base_path, "paleo_config.properties"))

# Extract values from config
drivers_root <- config[["drivers_root"]]
rds_root <- config[["rds_root"]]
table_root <- config[["table_root"]]
predictions_root <- config[["predictions_root"]]
mu_sigma_root <- config[["mu_sigma_root"]]
performance_root <- config[["performance_root"]]
figure_root <- config[["figure_root"]]
fire_data_root <- config[["fire_data_root"]]
paleoclimate_data_root <- config[["paleoclimate_data_root"]]
era5_data_root <- config[["era5_data_root"]]
proximity_data_root <- config[["proximity_data_root"]]


performance_root <- file.path(performance_root, "ignition_model_redo")
rds_root <- file.path(rds_root, "Ignition_Efficiency/redo")

```


Read in drivers
```{r}
df <- read.csv(file.path(drivers_root, "lightning_ignition_jja_2002_2011_new_no_outliers.csv"))
#df <- df[,2:26]
# 
# df <- df[,c("lat","lon","year","month","rh","tair","precip","wind","precip_1m","precip_2m","precip_3m","precip_5y","tair_1m","tair_2m","tair_3m","tair_5y","B", "C", "U", "strike_count","fire_count")]

#save gridcell-month-years where strike count > fire count to check against burned area training data
null_rows <- df  %>% filter(fire_count > strike_count) %>% dplyr::select(year, lat, lon, month, strike_count, fire_count)
write.csv(null_rows, file.path(drivers_root, "null_ignition_rows_training_data_2002-2011.csv"))

#remove rows where strike count > fire count (impossible, plus the model won't be abke to handle this)
df <- df  %>% filter(!(fire_count > strike_count))

head(df)
```

Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]
```

```{r}
# ── Save scaling parameters ───────────────────────────────────────────────────
predictors_ignition <- c("rh", "tair", "precip", "wind",
                          "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                          "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B", "C", "U")

mu_rh        <- mean(train$rh);        sigma_rh        <- sd(train$rh)
mu_tair      <- mean(train$tair);      sigma_tair      <- sd(train$tair)
mu_precip    <- mean(train$precip);    sigma_precip    <- sd(train$precip)
mu_wind      <- mean(train$wind);      sigma_wind      <- sd(train$wind)
mu_precip_1m <- mean(train$precip_1m); sigma_precip_1m <- sd(train$precip_1m)
mu_precip_2m <- mean(train$precip_2m); sigma_precip_2m <- sd(train$precip_2m)
mu_precip_3m <- mean(train$precip_3m); sigma_precip_3m <- sd(train$precip_3m)
mu_precip_5y <- mean(train$precip_5y); sigma_precip_5y <- sd(train$precip_5y)
mu_tair_1m   <- mean(train$tair_1m);   sigma_tair_1m   <- sd(train$tair_1m)
mu_tair_2m   <- mean(train$tair_2m);   sigma_tair_2m   <- sd(train$tair_2m)
mu_tair_3m   <- mean(train$tair_3m);   sigma_tair_3m   <- sd(train$tair_3m)
mu_tair_5y   <- mean(train$tair_5y);   sigma_tair_5y   <- sd(train$tair_5y)
mu_B         <- mean(train$B);         sigma_B         <- sd(train$B)
mu_C         <- mean(train$C);         sigma_C         <- sd(train$C)
mu_U         <- mean(train$U);         sigma_U         <- sd(train$U)

# Store and save
mean_sd_ignition <- data.frame(
  var   = predictors_ignition,
  mu    = c(mu_rh, mu_tair, mu_precip, mu_wind,
            mu_precip_1m, mu_precip_2m, mu_precip_3m, mu_precip_5y,
            mu_tair_1m, mu_tair_2m, mu_tair_3m, mu_tair_5y, mu_B, mu_C, mu_U),
  sigma = c(sigma_rh, sigma_tair, sigma_precip, sigma_wind,
            sigma_precip_1m, sigma_precip_2m, sigma_precip_3m, sigma_precip_5y,
            sigma_tair_1m, sigma_tair_2m, sigma_tair_3m, sigma_tair_5y, sigma_B, sigma_C, sigma_U)
)

write.csv(mean_sd_ignition,
          file.path(mu_sigma_root, "mu_sigma_ignition_predictors_redo.csv"),
          row.names = FALSE)

# ── Standardize train ─────────────────────────────────────────────────────────
train <- train %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B,
    C         = (C         - mu_C)         / sigma_C,
    U         = (U         - mu_U)         / sigma_U
  )

train <- train[, c("lat", "lon", "year", "month",
                   "strike_count", "fire_count",
                   "rh", "tair", "precip", "wind",
                   "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                   "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B", "C", "U")]

# ── Standardize test ──────────────────────────────────────────────────────────
# Use training means and sds 
test <- test %>%
  mutate(
    rh        = (rh        - mu_rh)        / sigma_rh,
    tair      = (tair      - mu_tair)      / sigma_tair,
    precip    = (precip    - mu_precip)    / sigma_precip,
    wind      = (wind      - mu_wind)      / sigma_wind,
    precip_1m = (precip_1m - mu_precip_1m) / sigma_precip_1m,
    precip_2m = (precip_2m - mu_precip_2m) / sigma_precip_2m,
    precip_3m = (precip_3m - mu_precip_3m) / sigma_precip_3m,
    precip_5y = (precip_5y - mu_precip_5y) / sigma_precip_5y,
    tair_1m   = (tair_1m   - mu_tair_1m)   / sigma_tair_1m,
    tair_2m   = (tair_2m   - mu_tair_2m)   / sigma_tair_2m,
    tair_3m   = (tair_3m   - mu_tair_3m)   / sigma_tair_3m,
    tair_5y   = (tair_5y   - mu_tair_5y)   / sigma_tair_5y,
    B         = (B         - mu_B)         / sigma_B,
    C         = (C         - mu_C)         / sigma_C,
    U         = (U         - mu_U)         / sigma_U
  )

test <- test[, c("lat", "lon", "year", "month",
                 "strike_count", "fire_count",
                 "rh", "tair", "precip", "wind",
                 "precip_1m", "precip_2m", "precip_3m", "precip_5y",
                 "tair_1m", "tair_2m", "tair_3m", "tair_5y", "B", "C", "U")]

head(train)

```

Check for colinearity within candidate predictors. Based on random forest, exclude:
dnow depth and Broadleaf cover for strike count, and wind, snow depth and broadleaf for fire count. 
```{r}
cor_matrix <- train %>%
  dplyr::select(rh, tair, precip,
                precip_1m, precip_2m, precip_3m, precip_5y,
                tair_1m, tair_2m, tair_3m, tair_5y, C, U) %>%
  cor(use = "pairwise.complete.obs") %>%
  round(2)

print(cor_matrix)

high_cor_pairs <- as.data.frame(as.table(cor_matrix)) %>%
  rename(var1 = Var1, var2 = Var2, correlation = Freq) %>%
  filter(as.integer(var1) < as.integer(var2)) %>%   # keep only upper triangle
  filter(abs(correlation) > 0.8) %>%
  arrange(desc(abs(correlation)))

print(high_cor_pairs)
```

Exclude tair_1m, tair_3m (tair_2m is between them), precip_2m (keep flanking 1m and 3m) and coniver (U, openland, is higher on importance according to the random forest).

##########################################
##### IGNITION EFFICINCY (p_ignite) #####
##########################################

Beta binomial 
```{r}
library(glmmTMB)
library(dplyr)

# ---- 1. Fit Beta-Binomial ignition model on the same predictors as m9 ----
# cbind(successes, failures) = cbind(fire_count, strike_count - fire_count)
# Beta-Binomial adds one dispersion parameter (phi) that absorbs
# extra-binomial variance -- the same fix NegBinomial provided over Poisson
# for strike counts, applied here to the ignition-efficiency stage.

predictors_beta_bin_1 <- c("rh", "tair", "precip", "precip_1m", "precip_3m", "precip_5y", "tair_3m", "tair_5y", "U")

bb_formula_1 <- as.formula(
  paste("cbind(fire_count, strike_count - fire_count) ~", paste(predictors_beta_bin_1, collapse = " + "))
)

#try including wind, just to see - it wasnt low for both importance metrics. 
predictors_beta_bin_2 <- c("rh", "tair", "precip", "wind", "precip_1m", "precip_3m", "precip_5y", "tair_3m", "tair_5y", "U")

bb_formula_2 <- as.formula(
  paste("cbind(fire_count, strike_count - fire_count) ~", paste(predictors_beta_bin_2, collapse = " + "))
)

bb_model_1 <- glmmTMB(
  bb_formula_1,
  data   = train,   # already filtered to strike_count > 0 and standardized
  family = betabinomial(link = "logit")
)

bb_model_2 <- glmmTMB(
  bb_formula_2,
  data   = train,   # already filtered to strike_count > 0 and standardized
  family = betabinomial(link = "logit")
)

summary(bb_model_1); summary(bb_model_2)

# ---- 2. Predict on the test set (already standardized with training mu/sigma) ----
test <- test %>%
  mutate(p_ignite_betabinom_1 = predict(bb_model_1, newdata = test, type = "response"),
         p_ignite_betabinom_2 = predict(bb_model_2, newdata = test, type = "response"))

```

# ---- 3. Compare against the existing rate-calibrated intercept-only model ----
```{r}
comparison <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate              = sum(fire_count) / sum(strike_count),
    pred_rate_betabinom_1    = mean(p_ignite_betabinom_1),
    pred_rate_betabinom_2    = mean(p_ignite_betabinom_2),
    .groups = "drop"
  )

print(comparison, n = Inf)

# ---- 4. Summary metrics for both approaches, same structure as before ----
comparison %>%
  summarise(
    spearman_betabinom_1  = cor(obs_rate, pred_rate_betabinom_1, method = "spearman"),
    spearman_betabinom_2  = cor(obs_rate, pred_rate_betabinom_2, method = "spearman"),
    pearson_betabinom_1   = cor(obs_rate, pred_rate_betabinom_1),
    pearson_betabinom_2   = cor(obs_rate, pred_rate_betabinom_2),
    ratio_betabinom_1     = sum(pred_rate_betabinom_1) / sum(obs_rate),
    ratio_betabinom_2     = sum(pred_rate_betabinom_2) / sum(obs_rate)
  )

# ---- 5. Expected vs. observed fire count check  ----
cat("Observed fires:                      ", sum(test$fire_count), "\n")
cat("Expected fires (beta-binomial 1):      ", sum(test$strike_count * test$p_ignite_betabinom_2, na.rm = TRUE), "\n")
cat("Expected fires (beta-binomial 2):      ", sum(test$strike_count * test$p_ignite_betabinom_2, na.rm = TRUE), "\n")

# ---- 6. Visual comparison across years ----

comparison_long <- comparison %>%
  pivot_longer(cols = c(obs_rate, pred_rate_betabinom_1, pred_rate_betabinom_2),
               names_to = "source", values_to = "rate")

ggplot(comparison_long, aes(x = year, y = rate, color = source)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(
    values = c(obs_rate = "firebrick", pred_rate_betabinom_1 = "darkgreen", pred_rate_betabinom_2 = "orange"),
    labels = c("Observed", "Predicted (beta-binomial 1)", "Predicted (beta-binomial 2)")
  ) +
  labs(x = "Year", y = "Per-strike ignition rate", color = NULL) +
  theme_minimal()
```

Apply a rate shift to the beta binomial predictions:

```{r}
library(dplyr)

# Fit a rate-calibration shift on TOP of the beta-binomial's predictions,
# using the same intercept-only approach as before, just against a
# better-discriminating base model (the beta binimial)

train <- train %>% filter(strike_count > 0) %>% 
  mutate(obs_rate = fire_count / strike_count)


train <- train %>%
  mutate(
    p_ignite_bb_1     = predict(bb_model_1, newdata = ., type = "response"),
    logit_p_bb_1      = qlogis(pmin(pmax(p_ignite_bb_1, 1e-8), 1 - 1e-8))
  )

rate_cal_model_bb <- glm(
  obs_rate ~ offset(logit_p_bb_1),
  data    = train,
  family  = binomial(),
  weights = strike_count
)
rate_cal_shift_bb <- coef(rate_cal_model_bb)[1]

test <- test %>%
  mutate(
    logit_p_bb_1        = qlogis(pmin(pmax(p_ignite_betabinom_1, 1e-8), 1 - 1e-8)),
    p_ignite_bb_ratecal  = plogis(logit_p_bb_1 + rate_cal_shift_bb)
  )

# Re-check both metrics on this combined approach
comparison_bb_cal <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate                = sum(fire_count) / sum(strike_count),
    pred_rate_bb_ratecal     = mean(p_ignite_bb_ratecal),
    .groups = "drop"
  )

comparison_bb_cal %>%
  summarise(
    spearman = cor(obs_rate, pred_rate_bb_ratecal, method = "spearman"),
    pearson  = cor(obs_rate, pred_rate_bb_ratecal),
    ratio    = sum(pred_rate_bb_ratecal) / sum(obs_rate)
  )

cat("Expected fires (beta-binomial + rate-cal shift): ",
    sum(test$strike_count * test$p_ignite_bb_ratecal, na.rm = TRUE), "\n")
```

Spatial comparison test:
```{r}
spatial_comparison <- test %>%
  filter(strike_count > 0) %>%
  group_by(lat, lon) %>%
  summarise(
    obs_rate             = sum(fire_count) / sum(strike_count),
    pred_rate_betabinom_1 = mean(p_ignite_betabinom_1),
    pred_rate_betabinom_2 = mean(p_ignite_betabinom_2),
    pred_rate_betabinom_ratecal = mean(p_ignite_bb_ratecal),
    .groups = "drop"
  )

spatial_comparison %>%
  summarise(
    spearman_betabinom_1   = cor(obs_rate, pred_rate_betabinom_1, method = "spearman"),
    spearman_betabinom_2   = cor(obs_rate, pred_rate_betabinom_2, method = "spearman"),
    spearman_betabinom_1_ratecal = cor(obs_rate, pred_rate_betabinom_ratecal, method = "spearman"),
    pearson_betabinom_1    = cor(obs_rate, pred_rate_betabinom_1),
    pearson_betabinom_2    = cor(obs_rate, pred_rate_betabinom_2),
    pearson_betabinom_1_ratecal = cor(obs_rate, pred_rate_betabinom_ratecal)
  )
```
Spatial performance:
spatial skill remains stubbornly weak across every single model variant tried in this whole exercise: plain binomial, beta-binomial with outliers, beta-binomial without outliers, calibrated, uncalibrated. Every one lands in the same 0.05–0.20 range for spatial correlation, while temporal skill has ranged from 0.27 (original binomial) up to 0.58–0.62 (beta-binomial variants) depending on model choice and outlier handling. That's a striking, consistent asymmetry worth stating directly in discussion: modeling improvements throughout this whole process, beta-binomial dispersion structure, outlier removal, recalibration, have all meaningfully improved temporal tracking, but none of them touch the spatial dimension at all, because none of the changes made so far target why some gridcells ignite more efficiently than others. That's a genuinely different, harder problem (likely needing better spatial predictors — finer-resolution fuel/vegetation data, terrain, or some spatially-varying random effect — rather than anything about the likelihood family or calibration approach), and it's the clearest, most honest limitation to flag for future work.

A plausible explanation: the predictor set (rh, tair, precip-family, tair_5y, C, U) already carries a fair amount of spatial information through C/U (landcover) and the fact that climate variables themselves vary systematically by location, so the mean structure of both models was already doing whatever spatial work these predictors can do, and switching the variance structure from Binomial to Beta-Binomial doesn't add anything new on that axis. The dispersion parameter is capturing something else entirely, likely genuine year-to-year process variability (drought years, fire-weather anomalies) that isn't well represented by any static spatial covariate.

Performance over time:
Spearman stayed exactly at 0.576, which is expected; adding a constant in logit space is a strictly monotonic transformation, so it can never change rank order. The rate-cal-shift only adjusted the overall level, leaving the beta-binomial's improved year-to-year discrimination intact.

Aggregate bias is now essentially fixed: 264.0 expected fires against 271 observed is a ~2.6% underprediction. This is an improvement over the beta-binomial alone's 41% overprediction, and now in the same ballpark of accuracy as the original approach's near-perfect 271.3.

Discrimination improvement (Spearman more than doubled) without paying beta-binomial's calibration cost. 

Note the ratio column in printed summary (1.22) and the "expected fires" total (264.0 vs 271, ratio ≈ 0.97) look like they're telling different stories, but they're measuring different things. The ratio in summarise() is sum(pred_rate_bb_ratecal) / sum(obs_rate) — an unweighted sum of 10 annual mean rates. "Expected fires" is sum(strike_count * p_ignite) — a strike-count-weighted total across all ~48,000 gridcell-months. Since rate_cal_shift_bb was fit with weights = strike_count, it's specifically optimized to get the strike-weighted total right (which it now does),not the simple unweighted average of 10 year-level ratios, which is a somewhat different and more idiosyncratic quantity. The expected-fires number is the more meaningful one for  actual downstream use (since p_ignite ultimately gets multiplied by strike count to produce fire counts in the pipeline), so I'd treat 264.0 vs. 271 as the trustworthy bias check.

Pearson barely moving (0.284 → 0.285) is expected too, and worth stating plainly rather than glossing over: this confirms the magnitude-compression problem is still there — the model ranks years correctly (and better than the binomial model), but still doesn't reproduce how large the swings between high- and low-ignition years really are. That's a real, remaining limitation to be upfront about, not something this fix resolved.

The beta binomial on its own better discriminates ignition, but does not capture the 10-year expected fire count well. However, adding the calibration shift adjusts this, maintaining good discrimination, though it does no better a job of tracking anomaly magnitude (ie. gets the direction, but not magnitude). 

Plots:
```{r}

comparison <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate              = sum(fire_count) / sum(strike_count),
    pred_rate_betabinom_1    = mean(p_ignite_betabinom_1),
    pred_rate_betabinom_2    = mean(p_ignite_betabinom_2),
    pred_rate_betabinom_1_cal_shift    = mean(p_ignite_bb_ratecal),
    .groups = "drop"
  )

comparison_long <- comparison %>%
  pivot_longer(cols = c(obs_rate, pred_rate_betabinom_1, pred_rate_betabinom_2, pred_rate_betabinom_1_cal_shift),
               names_to = "source", values_to = "rate")

ggplot(comparison_long, aes(x = year, y = rate, color = source)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(
    values = c(obs_rate = "firebrick", pred_rate_betabinom_1 = "darkgreen", pred_rate_betabinom_2 = "orange", pred_rate_betabinom_1_cal_shift = "purple"),
    labels = c("Observed", "Predicted (beta-binomial 1)", "Predicted (beta-binomial 2)", "Predicted (beta-binomial 1 rate-cal shift)")
  ) +
  labs(x = "Year", y = "Per-strike ignition rate", color = NULL) +
  theme_minimal()

comparison <- test %>%
    filter(strike_count > 0) %>%
  mutate(
    obs_rate = case_when(
      strike_count == 0          ~ NA_real_,
      fire_count > strike_count  ~ 0,
      TRUE                        ~ fire_count / strike_count
    )
  )

ggplot() + geom_point(data=comparison, aes(x=p_ignite_bb_ratecal, y=obs_rate))
```



Apply beta binomial as a bayesian model:
```{r}

library(rstan)
library(dplyr)

# ---- 1. Prepare data for Stan, using the reduced predictor set (bb_model_2) ----

predictors_beta_bin <- c("rh", "tair", "precip", "precip_1m", "precip_3m", "precip_5y", "tair_3m", "tair_5y", "U")

stan_data <- list(
  N            = nrow(train),
  K            = length(predictors_beta_bin),
  X            = as.matrix(train[, predictors_beta_bin]),
  fire_count   = train$fire_count,
  strike_count = train$strike_count
)

# ---- 2. Fit the Bayesian beta-binomial in Stan ----
# NOTE: N is large (~48,000 rows) -- this will take considerably longer to
# sample than the glmmTMB version. Consider running this from the terminal
# (same pattern as unified pipeline) if its slow interactively.
bb_bayes_fit <- stan(
  file    = "beta_binomial_ignition.stan",   # save the Stan code above to this path
  data    = stan_data,
  chains  = 4,
  iter    = 2000,
  warmup  = 1000,
  cores   = 4,
  seed    = 123
)

# Basic convergence check before trusting anything downstream
print(bb_bayes_fit, pars = c("alpha", "beta", "phi"))
# Rhat should be ~1.00-1.01 for every parameter; check for divergent
# transitions in the sampler output above -- if present, consider
# tightening priors or reparameterizing before proceeding.

# ---- 3. Extract posterior draws and posterior-mean coefficients ----
bb_bayes_posterior <- rstan::extract(bb_bayes_fit, pars = c("alpha", "beta", "phi"))

coef_means_bb <- list(
  alpha = mean(bb_bayes_posterior$alpha),
  beta  = colMeans(bb_bayes_posterior$beta),
  phi   = mean(bb_bayes_posterior$phi)
)

# ---- 4. Posterior-mean predictions on train, for fitting the rate-cal shift ----
X_train <- as.matrix(train[, predictors_beta_bin])
linpred_train <- as.numeric(coef_means_bb$alpha + X_train %*% coef_means_bb$beta)

train <- train %>%
  mutate(
    p_ignite_bb_bayes = plogis(linpred_train),
    logit_p_bb_bayes  = qlogis(pmin(pmax(p_ignite_bb_bayes, 1e-8), 1 - 1e-8))
  )

# ---- 5. Rate-calibration shift, unchanged approach from before ----
rate_cal_model_bb_bayes <- glm(
  obs_rate ~ offset(logit_p_bb_bayes),
  data    = train,
  family  = binomial(),
  weights = strike_count
)
rate_cal_shift_bb_bayes <- coef(rate_cal_model_bb_bayes)[1]
cat("Rate calibration shift (Bayesian beta-binomial):", round(rate_cal_shift_bb_bayes, 3), "\n")

# ---- 6. Apply to test set and check the same metrics as before ----
X_test <- as.matrix(test[, predictors_beta_bin])
linpred_test <- as.numeric(coef_means_bb$alpha + X_test %*% coef_means_bb$beta)

test <- test %>%
  mutate(
    p_ignite_bb_bayes         = plogis(linpred_test),
    logit_p_bb_bayes          = qlogis(pmin(pmax(p_ignite_bb_bayes, 1e-8), 1 - 1e-8)),
    p_ignite_bb_bayes_ratecal = plogis(logit_p_bb_bayes + rate_cal_shift_bb_bayes)
  )

comparison_bb_bayes <- test %>%
  filter(strike_count > 0) %>%
  group_by(year) %>%
  summarise(
    obs_rate                   = sum(fire_count) / sum(strike_count),
    pred_rate_bb_bayes_ratecal = mean(p_ignite_bb_bayes_ratecal),
    .groups = "drop"
  )

comparison_bb_bayes %>%
  summarise(
    spearman = cor(obs_rate, pred_rate_bb_bayes_ratecal, method = "spearman"),
    pearson  = cor(obs_rate, pred_rate_bb_bayes_ratecal),
    ratio    = sum(pred_rate_bb_bayes_ratecal) / sum(obs_rate)
  )

cat("Expected fires (Bayesian beta-binomial + rate-cal shift): ",
    sum(test$strike_count * test$p_ignite_bb_bayes_ratecal, na.rm = TRUE), "\n")

# ---- 7. Save in the same structure  unified pipeline already expects ----
# This mirrors m9_bayes_calibrated.rds's format ($fit, $coef_means, plus the
# separate calibration_shifts.rds pattern) so it drops directly into
# predict_ignition() with minimal changes.
saveRDS(list(
  fit        = bb_bayes_fit,
  coef_means = coef_means_bb,
  predictors = predictors_beta_bin
), file.path(rds_root, "beta_binomial_ignition_bayes.rds"))

saveRDS(list(
  rate_cal_shift = rate_cal_shift_bb_bayes
), file.path(rds_root, "beta_binomial_calibration_shift.rds"))
```




